---
title: "LoL Early-Game Win Predictor"
author: "OZNAL — Capstone"
format:
  dashboard:
    orientation: columns
    nav-buttons: [github]
    theme: cosmo
server: shiny
---

```{r}
#| context: setup
#| message: false
#| warning: false

library(shiny)
library(tidyverse)
library(caret)

models_dir <- "../shiny/models"
lr_model    <- readRDS(file.path(models_dir, "lr_model.rds"))
rf_model    <- readRDS(file.path(models_dir, "rf_model.rds"))
nb_model    <- readRDS(file.path(models_dir, "nb_model.rds"))
knn_model   <- readRDS(file.path(models_dir, "knn_model.rds"))
cart_model  <- readRDS(file.path(models_dir, "cart_model.rds"))
preproc     <- readRDS(file.path(models_dir, "preproc.rds"))
X_train     <- readRDS(file.path(models_dir, "X_train.rds"))

feature_names <- names(X_train)
train_means   <- colMeans(X_train)

build_input_row <- function(input) {
    row <- as.data.frame(as.list(train_means))
    row$side          <- as.numeric(input$side)
    row$golddiffat15  <- input$golddiffat15
    row$xpdiffat15    <- input$xpdiffat15
    row$csdiffat15    <- input$csdiffat15
    row$firstblood    <- as.numeric(input$firstblood)
    row$firstdragon   <- as.numeric(input$firstdragon)
    row$firstherald   <- as.numeric(input$firstherald)
    row$firsttower    <- as.numeric(input$firsttower)
    row$grub_diff     <- input$grub_diff
    row$winrate_diff  <- input$wr_diff / 100
    row[, feature_names, drop = FALSE]
}

predict_prob <- function(model, newdata) {
    p <- tryCatch(
        predict(model, newdata = newdata, type = "prob"),
        error = function(e) NULL
    )
    if (is.null(p)) return(NA_real_)
    if ("Win" %in% colnames(p)) p[["Win"]] else p[, ncol(p)]
}

probs <- reactive({
    row <- build_input_row(input)
    row_scaled <- predict(preproc, row)
    list(
        LR   = predict_prob(lr_model,   row_scaled),
        RF   = predict_prob(rf_model,   row_scaled),
        NB   = predict_prob(nb_model,   row_scaled),
        KNN  = predict_prob(knn_model,  row_scaled),
        CART = predict_prob(cart_model, row_scaled)
    )
})

fmt_pct <- function(x) if (is.na(x)) "—" else sprintf("%.1f%%", x * 100)
```

# Match Predictor

## Column {.sidebar width="320px"}

**Map side & objectives**

```{r}
radioButtons("side", "Map side",
             choices = c("Blue" = 1, "Red" = 0),
             selected = 1, inline = TRUE)

checkboxInput("firstblood",  "First blood",  TRUE)
checkboxInput("firstdragon", "First dragon", TRUE)
checkboxInput("firstherald", "First herald", FALSE)
checkboxInput("firsttower",  "First tower",  FALSE)

sliderInput("grub_diff", "Void grub advantage",
            min = -6, max = 6, value = 2, step = 1)

sliderInput("wr_diff", "Pre-game win rate advantage (%)",
            min = -50, max = 50, value = 10, step = 1)
```

**Economy at 15 minutes**

```{r}
sliderInput("golddiffat15", "Gold difference",
            min = -10000, max = 10000, value = 1500, step = 100)

sliderInput("xpdiffat15", "Experience difference",
            min = -6000, max = 6000, value = 800, step = 100)

sliderInput("csdiffat15", "CS difference",
            min = -80, max = 80, value = 12, step = 1)
```

## Column

### Row {height="22%"}

```{r}
#| content: valuebox
#| title: "Logistic Regression"
list(
    color = "primary",
    value = textOutput("vb_lr")
)
```

```{r}
#| content: valuebox
#| title: "Random Forest"
list(
    color = "primary",
    value = textOutput("vb_rf")
)
```

```{r}
#| content: valuebox
#| title: "Naive Bayes"
list(
    color = "primary",
    value = textOutput("vb_nb")
)
```

```{r}
#| content: valuebox
#| title: "K-Nearest Neighbors"
list(
    color = "primary",
    value = textOutput("vb_knn")
)
```

```{r}
#| content: valuebox
#| title: "Decision Tree (CART)"
list(
    color = "primary",
    value = textOutput("vb_cart")
)
```

```{r}
#| content: valuebox
#| title: "Average"
list(
    color = "success",
    value = textOutput("vb_avg")
)
```

```{r}
#| context: server

output$vb_lr   <- renderText(fmt_pct(probs()$LR))
output$vb_rf   <- renderText(fmt_pct(probs()$RF))
output$vb_nb   <- renderText(fmt_pct(probs()$NB))
output$vb_knn  <- renderText(fmt_pct(probs()$KNN))
output$vb_cart <- renderText(fmt_pct(probs()$CART))
output$vb_avg  <- renderText({
    p <- unlist(probs())
    fmt_pct(mean(p, na.rm = TRUE))
})

output$prob_plot <- renderPlot({
    p <- probs()
    df <- tibble(
        model = c("LR", "RF", "NB", "KNN", "CART"),
        prob  = c(p$LR, p$RF, p$NB, p$KNN, p$CART)
    )
    ggplot(df, aes(x = reorder(model, prob), y = prob)) +
        geom_col(fill = "#5b8def", width = 0.6) +
        geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
        geom_text(aes(label = sprintf("%.0f%%", prob * 100)),
                  hjust = -0.1, size = 4) +
        coord_flip() +
        scale_y_continuous(labels = scales::percent, limits = c(0, 1.05)) +
        labs(x = NULL, y = "Predicted win probability") +
        theme_minimal(base_size = 13) +
        theme(panel.grid.major.y = element_blank())
}, res = 96)

output$prob_table <- renderTable({
    p <- probs()
    data.frame(
        Model = c("Logistic Regression", "Random Forest", "Naive Bayes",
                  "K-Nearest Neighbors", "Decision Tree (CART)", "Average"),
        `Win probability` = c(
            fmt_pct(p$LR), fmt_pct(p$RF), fmt_pct(p$NB),
            fmt_pct(p$KNN), fmt_pct(p$CART),
            fmt_pct(mean(unlist(p), na.rm = TRUE))
        ),
        check.names = FALSE
    )
}, striped = TRUE, hover = TRUE, width = "100%", align = "lr")
```

### Row {height="78%"}

#### Column {width="60%"}

```{r}
#| title: Win probability by model
plotOutput("prob_plot", height = "100%")
```

#### Column {width="40%"}

```{r}
#| title: Summary table
tableOutput("prob_table")
```

# About

::: {.card title="What is this?"}

Five classification models trained on the first 15 minutes of professional
**League of Legends** matches. Adjust the early-game state on the **Match Predictor**
page to see each model's predicted win probability live.

The full app additionally covers hyperparameter tuning, cross-validated model
comparison, feature selection, and dataset exploration. This dashboard is a
stripped-down preview to evaluate the visual style.

:::

::: {.card title="Models"}

| Model | Hyperparameter | Description |
|------|----------------|-------------|
| Logistic Regression | none | Standard GLM, all features standardized. |
| Random Forest | `mtry` | 500 trees, `mtry` controls features per split. |
| Naive Bayes | `usekernel`, `adjust` | Probabilistic classifier with kernel density tuning. |
| K-Nearest Neighbors | `k` | Majority vote among `k` nearest scaled neighbors. |
| Decision Tree (CART) | `cp` | Recursive binary splitting, `cp` controls pruning. |

:::
