---
title: "League of Legends: Early-Game Win Prediction"
subtitle: "Five classification models on the first 15 minutes of professional matches"
author: "OZNAL — Capstone Project"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    code-fold: true
    fig-align: center
server: shiny
---

## Overview

Can the first 15 minutes of a professional **League of Legends** match predict the final winner?
We trained five classification models — Logistic Regression, Random Forest, Naive Bayes,
K-Nearest Neighbors, and a Decision Tree (CART) — on competitive match data and compare
their predictions side by side.

The interactive panel below lets you set the early-game state of a hypothetical match and
see each model's predicted win probability.

::: {.callout-note}
This is a **UI preview**: only the Match Predictor is wired up. The full app contains
additional tabs for hyperparameter tuning, model comparison, feature selection, and
data exploration.
:::

## Match Predictor

```{r}
#| context: server
#| message: false
#| warning: false

library(shiny)
library(tidyverse)
library(caret)

models_dir <- "../shiny/models"
lr_model    <- readRDS(file.path(models_dir, "lr_model.rds"))
rf_model    <- readRDS(file.path(models_dir, "rf_model.rds"))
nb_model    <- readRDS(file.path(models_dir, "nb_model.rds"))
knn_model   <- readRDS(file.path(models_dir, "knn_model.rds"))
cart_model  <- readRDS(file.path(models_dir, "cart_model.rds"))
preproc     <- readRDS(file.path(models_dir, "preproc.rds"))
X_train     <- readRDS(file.path(models_dir, "X_train.rds"))

feature_names <- names(X_train)
train_means   <- colMeans(X_train)

build_input_row <- function(input) {
    row <- as.data.frame(as.list(train_means))
    row$side          <- as.numeric(input$side)
    row$golddiffat15  <- input$golddiffat15
    row$xpdiffat15    <- input$xpdiffat15
    row$csdiffat15    <- input$csdiffat15
    row$firstblood    <- as.numeric(input$firstblood)
    row$firstdragon   <- as.numeric(input$firstdragon)
    row$firstherald   <- as.numeric(input$firstherald)
    row$firsttower    <- as.numeric(input$firsttower)
    row$grub_diff     <- input$grub_diff
    row$winrate_diff  <- input$wr_diff / 100
    row[, feature_names, drop = FALSE]
}

predict_prob <- function(model, newdata) {
    p <- tryCatch(
        predict(model, newdata = newdata, type = "prob"),
        error = function(e) NULL
    )
    if (is.null(p)) return(NA_real_)
    if ("Win" %in% colnames(p)) p[["Win"]] else p[, ncol(p)]
}

probs <- reactive({
    row <- build_input_row(input)
    row_scaled <- predict(preproc, row)
    list(
        LR   = predict_prob(lr_model,   row_scaled),
        RF   = predict_prob(rf_model,   row_scaled),
        NB   = predict_prob(nb_model,   row_scaled),
        KNN  = predict_prob(knn_model,  row_scaled),
        CART = predict_prob(cart_model, row_scaled)
    )
})

fmt_pct <- function(x) if (is.na(x)) "—" else sprintf("%.1f%%", x * 100)

output$prob_lr   <- renderText(fmt_pct(probs()$LR))
output$prob_rf   <- renderText(fmt_pct(probs()$RF))
output$prob_nb   <- renderText(fmt_pct(probs()$NB))
output$prob_knn  <- renderText(fmt_pct(probs()$KNN))
output$prob_cart <- renderText(fmt_pct(probs()$CART))
output$prob_avg  <- renderText({
    p <- unlist(probs())
    fmt_pct(mean(p, na.rm = TRUE))
})

output$prob_table <- renderTable({
    p <- probs()
    data.frame(
        Model = c("Logistic Regression", "Random Forest", "Naive Bayes",
                  "K-Nearest Neighbors", "Decision Tree (CART)", "Average"),
        `Win probability` = c(
            fmt_pct(p$LR), fmt_pct(p$RF), fmt_pct(p$NB),
            fmt_pct(p$KNN), fmt_pct(p$CART),
            fmt_pct(mean(unlist(p), na.rm = TRUE))
        ),
        check.names = FALSE
    )
}, striped = TRUE, hover = TRUE, width = "100%", align = "lr")

output$prob_plot <- renderPlot({
    p <- probs()
    df <- tibble(
        model = c("LR", "RF", "NB", "KNN", "CART"),
        prob  = c(p$LR, p$RF, p$NB, p$KNN, p$CART)
    )
    ggplot(df, aes(x = reorder(model, prob), y = prob)) +
        geom_col(fill = "#5b8def", width = 0.6) +
        geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
        geom_text(aes(label = sprintf("%.0f%%", prob * 100)),
                  hjust = -0.1, size = 4) +
        coord_flip() +
        scale_y_continuous(labels = scales::percent, limits = c(0, 1.05)) +
        labs(x = NULL, y = "Predicted win probability") +
        theme_minimal(base_size = 13) +
        theme(panel.grid.major.y = element_blank())
}, res = 96)
```

### Match conditions

::: {.grid}

::: {.g-col-12 .g-col-md-5}

**Map side & objectives**

```{r}
radioButtons("side", "Map side",
             choices = c("Blue" = 1, "Red" = 0),
             selected = 1, inline = TRUE)

checkboxInput("firstblood",  "First blood",  TRUE)
checkboxInput("firstdragon", "First dragon", TRUE)
checkboxInput("firstherald", "First herald", FALSE)
checkboxInput("firsttower",  "First tower",  FALSE)

sliderInput("grub_diff", "Void grub advantage",
            min = -6, max = 6, value = 2, step = 1)

sliderInput("wr_diff", "Pre-game win rate advantage (%)",
            min = -50, max = 50, value = 10, step = 1)
```

:::

::: {.g-col-12 .g-col-md-7}

**Economic state at 15 minutes**

```{r}
sliderInput("golddiffat15", "Gold difference",
            min = -10000, max = 10000, value = 1500, step = 100)

sliderInput("xpdiffat15", "Experience difference",
            min = -6000, max = 6000, value = 800, step = 100)

sliderInput("csdiffat15", "CS (creep score) difference",
            min = -80, max = 80, value = 12, step = 1)
```

:::

:::

### Predicted win probability

```{r}
plotOutput("prob_plot", height = "260px")
```

```{r}
tableOutput("prob_table")
```

::: {.callout-tip}
A probability above 50% means that model predicts the team to win, given the
early-game state above. Each model is trained on the same features but uses a
different decision boundary, which is why their predictions can disagree —
especially in close games.
:::

## Notes on the models

| Model | Hyperparameter | Description |
|------|----------------|-------------|
| Logistic Regression | none | Standard GLM, all features standardized. |
| Random Forest | `mtry` | 500 trees, `mtry` controls features per split. |
| Naive Bayes | `usekernel`, `adjust` | Probabilistic classifier, kernel density tuning. |
| K-Nearest Neighbors | `k` | Majority vote among `k` nearest scaled neighbors. |
| Decision Tree (CART) | `cp` | Recursive binary splitting, `cp` controls pruning. |

The full app reports cross-validated performance across all five models, including
ROC curves, confusion matrices, and a feature-importance comparison.
