---
title: "OZNAL Capstone - LoL Esports Early-Game Prediction"
author: "Martin Vanco, Adriana Gemeľová"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width = 10,
    fig.height = 6
)
```

# Introduction

League of Legends is the most-played and most-watched esport in the world. Predicting which team will win from the early state of the match is a long-standing question in esports analytics: by 15 minutes - well before half of a typical game has elapsed - gold, experience, and objective leads have already begun to compound. We use the 2025 season of professional play to test how reliably this 15-minute snapshot forecasts the eventual winner, and which features actually drive the prediction.

The analysis runs two scenarios on the same data:

-   **Scenario 1 - Model Comparison.** We train and tune five classifiers (Logistic Regression, Random Forest, Naive Bayes, KNN, CART) on the engineered early-game feature set and compare them on accuracy, AUC-ROC, precision, recall, and F1. The goal is to see whether any model can extract signal beyond the trivial "team with the gold lead at 15 min wins" baseline (\~73% accuracy), and whether different model families agree on what matters.
-   **Scenario 3 - Feature Selection.** With 19 engineered candidate features carrying meaningful redundancy, we apply RFE, LASSO, and Elastic Net to find the minimal subset that retains predictive performance. This both improves interpretability and stress-tests which features are genuinely informative versus carrying duplicate signal.

## Hypotheses

**H1:** *"Early-game performance metrics and objective control can reliably predict match outcome in professional League of Legends."* We test whether gold, XP, and CS differentials at 10 and 15 minutes, combined with first objective captures (first blood, first dragon, first tower, void grubs), contain enough signal to classify wins and losses.

**H2:** *"Bot lane (ADC) gold advantage is the most impactful role-specific early-game factor for predicting match outcome."* We test whether the bot lane player's early-game differentials carry stronger predictive signal than those of other roles (top, jungle, mid, support).

## Motivation

Both of us on this project are long-time League of Legends players and regular viewers of the professional scene. Watching pro matches, you very quickly develop a feel for when a game is "already over" - a significant gold lead at early game mostly means a quick and decisive game. We had been wondering for a while how much of that intuition is actually backed by data: how reliably do those early signals predict the eventual winner and which signals matter the most.

# Dataset

We use the 2025 season export from **Oracle's Elixir**, the standard public source for professional League of Legends match data. The raw file contains **120,636 rows across 165 columns**, covering **10,053 unique matches** played by 445 teams across 45 professional leagues (LPL, LCK, LEC, LFL, LAS, EM, ...) on patches 15.01-15.24.

The row layout is **12 rows per game**: one per player (5 per team) plus one aggregated team-level row per side (Blue and Red, 50/50). About **91.9% of rows are marked `complete`**, meaning the early-game snapshots at 10/15/20/25 min are populated; the remaining 8.1% are missing these values.

Columns group into metadata (`gameid`, `league`, `patch`, `side`, `position`, ...), draft (`pick1-5`, `ban1-5`), outcome (`result`, `gamelength`), combat (kills/deaths/assists, first blood), objectives (dragons, heralds, void grubs, towers, turret plates), economy (gold, CS), vision (wards, vision score), and the **early-game timeline snapshots** (`goldatX`, `xpatX`, `csatX`, `golddiffatX`, etc. for X in {10, 15, 20, 25}) which form the core of our feature set.

# Game introduction

League of Legends (LoL) is a 5v5 team game. Two teams (Blue and Red side) try to destroy each other's base. A single match is typically around 30 minutes but can last as long as 50 minutes. In this work, we focus on the **early game** stage (the first 15 minutes). Each team has five fixed roles - **top**, **jungle (jng)**, **mid**, **bot (ADC)**, and **support (sup)** - each playing a different lane or area of the map.

Three continuous resources accumulate throughout the match:

-   **Gold** - earned from minions, champion kills, and objectives, and spent on items that strengthen champions.
-   **XP (experience)** - levels champions up and unlocks abilities.
-   **CS (creep score)** - the number of minions a player has killed

The early game also has a set of one-off objective races, each won by exactly one team: **first blood** (the first champion kill of the match), **first dragon**, **first herald**, **first tower** (each granting a stacking buff, a tower-damaging summon, or a large gold bonus), **turret plates** (five plates on each outer tower, all paid out as gold before 14:00), and **void grubs** (a shared pit of up to 6 grubs that despawns at 13:45 on early-2025 patches and 14:45 mid-season).

# Related work

Published work on early-game LoL match prediction at the **15-minute mark** converges tightly, with reported accuracy clustering in the **76-78% band** across model families and dataset types.

| Study                                        | Year | Dataset                      | Model               | Acc. \@15  |
|----------------------------------------------|------|------------------------------|---------------------|------------|
| Spaargaren, MSc thesis (Tilburg)             | 2022 | Oracle's Elixir pro, 2021    | Logistic Regression | **77.99%** |
| Spaargaren                                   | 2022 | Oracle's Elixir pro, 2021    | Random Forest       | 77.26%     |
| Spaargaren                                   | 2022 | Oracle's Elixir pro, 2021    | Naïve Bayes         | 76.82%     |
| Spaargaren                                   | 2022 | Oracle's Elixir pro, 2021    | kNN (k=9)           | 74.83%     |
| Lafrance & Grewal, *Entertainment Computing* | 2026 | Riot API solo queue, Diamond | XGBoost             | 76.9%      |
| Junior & Campelo, arXiv 2309.02449           | 2023 | Riot API solo queue          | Random Forest       | \~74%\*    |

\*Interpolated to 50% PET (≈ 15 min for a \~30-min median match) from Junior & Campelo's PET-bucket accuracy curves.

Two takeaways. First, **professional play is not materially easier to predict than solo queue** at 15 minutes - Spaargaren's pro-data LR sits within one percentage point of Lafrance & Grewal's solo-queue XGBoost. Second, **algorithm choice matters little**: pro-data classifiers all land in the 77 ± 1% band, with solo-queue \~3 points lower. Across every study, **gold differential is the strongest single feature**, with first tower or first herald typically ranked next.

Tsang (2025, Medium) adds a temporal signal: the coefficient of `golddiffat15` has **declined** from above 1.0 in 2020-2023 to 0.83 in 2025, suggesting the 2025 season is harder to predict from a single gold-at-15 signal than any previous season. We additionally include two features absent from the @15-minute literature - **turret plate differential** and **void grubs (introduced in 2025 season)**.

Finally, we note a framing gap: the trivial rule "team with a gold lead at 15 min wins" already delivers \~73% accuracy on professional data (we compute the exact figure on our own set in the EDA), so the 76-78% models above only beat this naive baseline by 3-5 points. We adopt this framing directly in our results.

# Setup

```{r libraries}
library(tidyverse)
library(corrplot)
library(caret)
library(glmnet)
library(randomForest)
library(naivebayes)
library(rpart)
library(xgboost)
library(pROC)
library(gridExtra)
library(doParallel)

cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)
```

## Load data

```{r load-data}
raw <- read_csv("data/2025_LoL_esports_match_data_from_OraclesElixir.csv")
glimpse(raw)
```

```{r}
# select only complete data - with @10, 15, 20, 25 metrics
complete <- raw %>% filter(datacompleteness == "complete")

teams <- complete %>% filter(position == "team")
players <- complete %>% filter(position != "team")

cat(sprintf(
    "Total rows: %d\nFiltered to complete: %d (removed %d incomplete, %.1f%%)\nTeam rows: %d\nPlayer rows: %d\nTarget balance: %.1f%% win / %.1f%% loss",
    nrow(raw), nrow(complete),
    nrow(raw) - nrow(complete),
    (nrow(raw) - nrow(complete)) / nrow(raw) * 100,
    nrow(teams), nrow(players),
    mean(teams$result == 1) * 100, mean(teams$result == 0) * 100
))
```

## Restructure: one row per game

The raw `teams` table stores two mirror rows per game, one per side. For EDA we pivot to one row per game with `blue_*` and `red_*` columns and a binary `blue_win` target, following the **tidy data** convention of one observation per row (Wickham, 2014, *JSS* 59(10)).

After the pivot, the Blue and Red copies of each signed differential and each mutually-exclusive binary flag are pure mirrors: `red_golddiffat15 = -blue_golddiffat15` and `red_firstblood = 1 - blue_firstblood`. We drop the `red_*` side of those columns, since the model gains nothing from a feature it can recover from the one next to it. Independent counts (kills, deaths, assists, turret plates) and the shared-pool grub count stay on both sides - neither side can be reconstructed from the other.

```{r restructure-one-row-per-game}
game_cols <- c(
    "result",
    "goldat10", "xpat10", "csat10",
    "goldat15", "xpat15", "csat15",
    "golddiffat10", "xpdiffat10", "csdiffat10",
    "golddiffat15", "xpdiffat15", "csdiffat15",
    "killsat10", "deathsat10", "assistsat10",
    "killsat15", "deathsat15", "assistsat15",
    "firstblood", "firstdragon", "firstherald", "firsttower",
    "void_grubs", "turretplates"
)

games_full <- teams %>%
    select(gameid, side, all_of(game_cols)) %>%
    mutate(side = tolower(side)) %>%
    pivot_wider(
        id_cols     = gameid,
        names_from  = side,
        values_from = all_of(game_cols),
        names_glue  = "{side}_{.value}"
    ) %>%
    mutate(blue_win = blue_result) %>%
    select(-blue_result, -red_result)

# Sanity: blue/red signed diffs must be exact negatives of each other,
# and mutually exclusive binary flags must satisfy red = 1 - blue whenever
# the objective fires.
stopifnot(all.equal(games_full$blue_golddiffat15, -games_full$red_golddiffat15))
stopifnot(all.equal(games_full$blue_xpdiffat15,   -games_full$red_xpdiffat15))
stopifnot(all.equal(games_full$blue_csdiffat15,   -games_full$red_csdiffat15))
stopifnot(all(games_full$blue_firstblood   + games_full$red_firstblood   <= 1))
stopifnot(all(games_full$blue_firsttower   + games_full$red_firsttower   <= 1))

# Drop the strict-mirror columns identified above. We keep both sides for
# independent counts (kills/deaths/assists/turret plates) and for the
# shared-pool grub count, since those carry information beyond their pair.
redundant_mirrors <- c(
    "red_golddiffat10", "red_xpdiffat10", "red_csdiffat10",
    "red_golddiffat15", "red_xpdiffat15", "red_csdiffat15",
    "red_firstblood", "red_firstdragon", "red_firstherald", "red_firsttower"
)

games <- games_full %>% select(-all_of(redundant_mirrors))

cat(sprintf(
    "Games (one row per game): %d\nFeatures dropped as strict mirrors: %d\nFeatures retained: %d\nBlue win rate: %.2f%% (Red: %.2f%%)\n",
    nrow(games),
    length(redundant_mirrors),
    ncol(games) - 2, # excluding gameid and blue_win
    mean(games$blue_win) * 100,
    mean(1 - games$blue_win) * 100
))

cat("\nDropped features (recoverable as -blue_X or 1-blue_X):\n  ",
    paste(redundant_mirrors, collapse = ", "), "\n")

# Statistical test for side imbalance
side_test <- prop.test(sum(games$blue_win), nrow(games), p = 0.5)
cat(sprintf(
    "\nTwo-sided proportion test vs 50%%:\n  chi^2 = %.2f, df = 1, p = %.3g\n  95%% CI for Blue WR: [%.3f, %.3f]\n  -> Blue-side advantage is %s.\n",
    side_test$statistic, side_test$p.value,
    side_test$conf.int[1], side_test$conf.int[2],
    ifelse(side_test$p.value < 0.05, "statistically significant", "not statistically significant")
))
```

Blue side wins slightly more than half of games. The imbalance is mild enough that no resampling is needed and accuracy remains valid alongside AUC-ROC.

# Exploratory Data Analysis

Before deciding on feature engineering or models, we look at the data to see what signal is there, which features are worth keeping, and what problems to watch for. Everything below uses the **single-row-per-game** `games` dataset.

## Target distribution

```{r eda-target}
games %>%
    mutate(blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won"))) %>%
    ggplot(aes(x = blue_win, fill = blue_win)) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
    scale_fill_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_hline(yintercept = nrow(games) / 2, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Target Variable Distribution (blue_win)",
        x = "Outcome", y = "Count"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

Blue side wins slightly more than half of all games. The proportion test above puts this at p \< 0.001, so the imbalance is real, but at roughly 3 percentage points above 50% it is small enough that we keep accuracy as a valid metric and skip resampling.

## Descriptive statistics

Before correlations and modelling, we look at the central tendency and spread of every early-game numeric we plan to feed into the models. The columns are sorted by standard deviation - the loudest features will dominate any unscaled distance-based model, which motivates the centering and scaling done later in the modelling pipeline.

```{r eda-describe}
early_game_cols <- c(
    "blue_goldat10", "blue_xpat10", "blue_csat10",
    "blue_golddiffat10", "blue_xpdiffat10", "blue_csdiffat10",
    "blue_killsat10", "blue_deathsat10", "blue_assistsat10",
    "blue_goldat15", "blue_xpat15", "blue_csat15",
    "blue_golddiffat15", "blue_xpdiffat15", "blue_csdiffat15",
    "blue_killsat15", "blue_deathsat15", "blue_assistsat15",
    "blue_firstblood", "blue_firstdragon", "blue_firstherald", "blue_firsttower",
    "blue_void_grubs", "red_void_grubs",
    "blue_win"
)

games %>%
    select(all_of(setdiff(early_game_cols, "blue_win"))) %>%
    pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
    drop_na() %>%
    group_by(feature) %>%
    summarise(
        n      = n(),
        mean   = mean(value),
        sd     = sd(value),
        min    = min(value),
        q25    = quantile(value, 0.25),
        median = median(value),
        q75    = quantile(value, 0.75),
        max    = max(value),
        .groups = "drop"
    ) %>%
    mutate(across(c(mean, sd, min, q25, median, q75, max), ~ round(.x, 2))) %>%
    arrange(desc(sd)) %>%
    print(n = Inf)
```

The differentials sit at the top of the spread ranking with standard deviations in the thousands of gold and XP, while the absolute totals at 15 minutes cluster tightly around 25k gold and 30k XP. CS columns are an order of magnitude smaller, kill/death/assist counts smaller still, and the 0/1-coded objectives sit at the bottom with sd around 0.5. On the original scale the objectives would be invisible next to gold and XP, which is why the modelling pipeline centers and scales every input before training Logistic Regression and KNN. The differential means hover near zero (243 gold and 49 XP at 15) - close enough to confirm the features behave like signed deltas rather than biased counters, but slightly positive in line with the Blue-side advantage seen earlier.

## Early-game feature correlations with `blue_win`

We correlate each early-game numeric feature (Blue's perspective) with the target.

```{r eda-correlation, fig.width=12, fig.height=10}
cor_data <- games %>%
    select(all_of(early_game_cols)) %>%
    drop_na()

cor_matrix <- cor(cor_data)

corrplot(cor_matrix,
    method   = "color", type = "lower",
    tl.cex   = 0.7, tl.col = "black",
    col      = colorRampPalette(c("#e74c3c", "white", "#2ecc71"))(200),
    col.lim  = c(-1, 1),
    title    = "Correlation Matrix - Early-Game Features (Blue's perspective)",
    mar      = c(0, 0, 2, 0)
)
```

```{r eda-correlation-with-result}
result_cors <- cor_data %>%
    select(-blue_win) %>%
    summarise(across(everything(), ~ cor(.x, cor_data$blue_win))) %>%
    pivot_longer(everything(), names_to = "feature", values_to = "correlation") %>%
    arrange(desc(abs(correlation)))

ggplot(result_cors, aes(
    x    = reorder(feature, abs(correlation)),
    y    = correlation,
    fill = correlation > 0
)) +
    geom_col() +
    coord_flip(ylim = c(-1, 1)) +
    scale_y_continuous(breaks = seq(-1, 1, 0.25)) +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title    = "Feature Correlation with blue_win",
        x = NULL, y = "Pearson Correlation", fill = "Direction"
    ) +
    theme_minimal()
```

```{r}
cat(sprintf(
    "Strongest positive: %s (%.3f)\nStrongest negative: %s (%.3f)",
    result_cors$feature[which.max(result_cors$correlation)],
    max(result_cors$correlation),
    result_cors$feature[which.min(result_cors$correlation)],
    min(result_cors$correlation)
))
```

Differentials (`blue_golddiffat15`, `blue_xpdiffat15`, `blue_csdiffat15`) dominate the ranking, since what matters is the *gap* between teams rather than the absolute totals. Among objectives, `blue_firsttower` and `blue_firstherald` stand out; `blue_firstblood` and `blue_firstdragon` carry less signal. The matrix also shows two flavours of multicollinearity: between a raw stat and its differential (`blue_goldat15` vs `blue_golddiffat15`), and between Blue and Red counts of the same shared-pool objective (`blue_void_grubs` vs `red_void_grubs`, strongly negatively correlated because they share a 6-grub pit). Both forms hurt Logistic Regression and **motivate the LASSO and Elastic Net regularization used in Scenario 3**.

## Per-outcome means

The correlation chart ranks features by linear association with `blue_win`. Putting the same features through a means-by-outcome lens shows the gap in the original units rather than on the [-1, 1] scale, which is easier to read for objectives and combat counts.

```{r eda-by-outcome}
games %>%
    select(all_of(early_game_cols)) %>%
    drop_na() %>%
    mutate(outcome = if_else(blue_win == 1, "blue_win", "red_win")) %>%
    select(-blue_win) %>%
    pivot_longer(-outcome, names_to = "feature", values_to = "value") %>%
    group_by(feature, outcome) %>%
    summarise(mean = mean(value), .groups = "drop") %>%
    pivot_wider(names_from = outcome, values_from = mean) %>%
    mutate(diff = blue_win - red_win) %>%
    arrange(desc(abs(diff))) %>%
    mutate(across(c(blue_win, red_win, diff), ~ round(.x, 3))) %>%
    print(n = Inf)
```

The biggest absolute gaps fall on the gold and XP differentials at 15 minutes - Blue wins lead by \~1500 gold and \~940 XP on average; losses trail by similar amounts. Tower-related objectives produce the largest split among the 0/1 features: `blue_firsttower` jumps from 0.34 in losses to 0.73 in wins, `blue_firstherald` from 0.44 to 0.75. First blood and first dragon move the means by less than 0.17, matching their weaker correlations. The signs in the diff column also confirm direction: `blue_deathsat15` is the only count where the win-mean is *lower* than the loss-mean (3.6 vs 5.2), and `red_void_grubs` flips negative because more grubs to the enemy means a worse outcome for Blue.

## Gold differential distribution

Gold differential at 15 minutes is the single strongest early-game predictor. We look at how cleanly it separates Blue wins from Red wins, and whether a simple linear cut (positive gold diff -\> Blue win) would already do most of the work or whether we need non-linear models.

```{r eda-golddiff}
games %>%
    filter(!is.na(blue_golddiffat15)) %>%
    mutate(blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won"))) %>%
    ggplot(aes(x = blue_golddiffat15, fill = blue_win)) +
    geom_density(alpha = 0.5) +
    scale_fill_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Gold Differential at 15 Minutes - Blue's Perspective",
        x = "blue_golddiffat15", y = "Density", fill = "Outcome"
    ) +
    theme_minimal()
```

```{r eda-golddiff-10vs15}
games %>%
    filter(!is.na(blue_golddiffat10), !is.na(blue_golddiffat15)) %>%
    mutate(blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won"))) %>%
    ggplot(aes(x = blue_golddiffat10, y = blue_golddiffat15, color = blue_win)) +
    geom_point(alpha = 0.15, size = 0.8) +
    scale_color_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Blue Gold Diff @10 vs @15 - Do Leads Grow?",
        x = "blue_golddiffat10", y = "blue_golddiffat15", color = "Outcome"
    ) +
    theme_minimal()
```

To quantify how much of the outcome the gold differential at 15 min already explains, we compute the accuracy of the naive rule "Blue ahead in gold @15 -\> Blue wins" directly on `games`. This number then defines the baseline that any trained model has to beat.

```{r eda-golddiff-baseline}
gold_baseline_df <- games %>%
    filter(!is.na(blue_golddiffat15)) %>%
    transmute(
        pred_blue_win = blue_golddiffat15 > 0,
        actual        = blue_win == 1
    )

gold_baseline_acc  <- mean(gold_baseline_df$pred_blue_win == gold_baseline_df$actual)
gold_baseline_n    <- nrow(gold_baseline_df)
gold_baseline_ties <- sum(games$blue_golddiffat15 == 0, na.rm = TRUE)
blue_base_rate     <- mean(games$blue_win)

cat(sprintf(
    "Naive rule 'Blue ahead in gold @15 -> Blue wins':\n  N = %d games (%d games tied at exactly 0 gold diff)\n  Accuracy: %.2f%%\n  Always-predict-Blue baseline: %.2f%%\n  Rule beats always-Blue by: %.2f pp",
    gold_baseline_n, gold_baseline_ties,
    gold_baseline_acc * 100,
    blue_base_rate * 100,
    (gold_baseline_acc - blue_base_rate) * 100
))
```

The two densities are clearly separated. The naive rule reaches 72.9% accuracy at the game level - well above both the 50% coin-flip and the 53.3% always-predict-Blue baseline - and any model has to beat this number to be worth keeping. The wide overlap around zero says a meaningful share of games are not decided by gold alone.

## First objective win rates

The four "first" objectives (first blood, first dragon, first herald, first tower) all happen in the first \~14 minutes and are mutually exclusive: exactly one team secures each. For every objective we compute the win rate of the team that secured it and run a chi-square test of independence between securer and outcome.

```{r eda-objectives}
obj_cols <- c("firstblood", "firstdragon", "firstherald", "firsttower")

obj_winrates <- map_dfr(obj_cols, function(obj) {
    blue_col <- paste0("blue_", obj)
    df <- games %>%
        select(all_of(c(blue_col, "blue_win"))) %>%
        drop_na()

    n_blue <- sum(df[[blue_col]] == 1)
    n_red  <- sum(df[[blue_col]] == 0)
    blue_wr <- mean(df$blue_win[df[[blue_col]] == 1]) * 100
    red_wr  <- mean(df$blue_win[df[[blue_col]] == 0]) * 100  # P(Blue wins | Red secured)

    chi <- chisq.test(table(df[[blue_col]], df$blue_win))

    tibble(
        objective     = str_replace(obj, "first", "First ") %>% str_to_title(),
        secured_by    = c("Blue secured", "Red secured"),
        win_rate_blue = c(blue_wr, red_wr),
        n             = c(n_blue, n_red),
        chi2          = chi$statistic,
        p_value       = chi$p.value
    )
})

ggplot(
    obj_winrates,
    aes(x = reorder(objective, win_rate_blue), y = win_rate_blue, fill = secured_by)
) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6) +
    geom_text(
        aes(label = sprintf("%.1f%% (n=%d)", win_rate_blue, n)),
        position = position_dodge(width = 0.7),
        hjust = -0.05, size = 3
    ) +
    coord_flip() +
    scale_fill_manual(values = c("Blue secured" = "#3498db", "Red secured" = "#e74c3c")) +
    scale_y_continuous(limits = c(0, 100)) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    labs(
        title    = "P(Blue wins) by Who Secured the First Objective",
        x = NULL, y = "Blue Win Rate (%)", fill = NULL
    ) +
    theme_minimal()

cat("Chi-square independence tests (objective securer vs blue_win):\n")
obj_winrates %>%
    distinct(objective, chi2, p_value) %>%
    mutate(
        chi2    = round(chi2, 1),
        p_value = formatC(p_value, format = "e", digits = 2)
    ) %>%
    print()
```

When Blue secures first tower or first herald, Blue's win rate jumps well above 50%, and drops by a similar amount when Red gets them instead. First blood and first dragon move the rate much less. Every chi-square test rejects independence at p ≪ 0.001, so the *direction* is certain, but the *magnitude* varies sharply between objectives - the same pattern the correlation analysis showed. Tower-related objectives translate straight into map control and gold, while first blood is often a single lucky skirmish. All four are worth keeping, but the models should weight tower and herald more heavily.

## Void grubs

Void grubs are a 2024+ objective: they spawn at 5:00 and despawn at 13:45 on early-2025 patches (moved to 14:45 mid-season), so they are purely early-game. The pit holds up to 6 grubs total, split between the two teams. We plot Blue's win rate against the number of grubs Blue ended up with.

```{r eda-grubs}
games %>%
    filter(!is.na(blue_void_grubs)) %>%
    group_by(blue_void_grubs) %>%
    summarise(
        win_rate = mean(blue_win) * 100,
        n        = n(),
        .groups  = "drop"
    ) %>%
    ggplot(aes(x = factor(blue_void_grubs), y = win_rate, fill = win_rate)) +
    geom_col() +
    geom_text(aes(label = sprintf("%.1f%%\nn=%d", win_rate, n)),
        vjust = -0.3, size = 3
    ) +
    scale_fill_gradient(low = "#e74c3c", high = "#3498db") +
    scale_y_continuous(limits = c(0, 100)) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Blue Win Rate by Number of Void Grubs Blue Secured",
        x = "blue_void_grubs", y = "P(Blue wins) (%)"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

The relationship is essentially monotonic: 0 grubs sits well below 50% win rate, 6 grubs climbs to \~70%. That justifies the grub differential we build later. Small wobbles at intermediate counts come from low sample sizes - teams that end up with *exactly* 4 grubs are rare, because grub fights usually end with one side taking the whole remaining set.

## Side advantage

Blue side gets first pick in the draft and has slight map asymmetries. We measure the side advantage as the share of Blue wins, and report a 95% CI plus a chi-square test against the 50% null.

```{r eda-side}
n_games  <- nrow(games)
n_blue   <- sum(games$blue_win)
side_pt  <- prop.test(n_blue, n_games, p = 0.5)
blue_wr  <- n_blue / n_games * 100
ci_low   <- side_pt$conf.int[1] * 100
ci_high  <- side_pt$conf.int[2] * 100

tibble(
    label    = "Blue side",
    win_rate = blue_wr,
    ci_low   = ci_low,
    ci_high  = ci_high
) %>%
    ggplot(aes(x = label, y = win_rate)) +
    geom_col(width = 0.4, fill = "#3498db") +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.15, color = "gray20") +
    geom_text(
        aes(label = sprintf("%.2f%%  [%.2f, %.2f]\nn=%d  ·  chi^2=%.1f, p=%.2g",
                            win_rate, ci_low, ci_high, n_games,
                            side_pt$statistic, side_pt$p.value)),
        vjust = -0.6, size = 3.5
    ) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    scale_y_continuous(limits = c(0, 70)) +
    labs(
        title    = "Blue-Side Advantage",
        x = NULL, y = "P(Blue wins) (%)"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

Blue wins \~53% of games, a steady \~3 pp advantage. The 95% CI excludes 50% and the chi-square test rejects the null at p ≪ 0.001, so the effect is significant even if small. It comes from first pick in champion draft and small map/camera asymmetries.

## Role-specific gold differentials (H2 preview)

H2 asks whether some roles' early leads matter more than others. As a first look, we take each Blue-side player by role and correlate their gold differential at 15 min with `blue_win`.

```{r eda-roles}
role_diffs_blue <- players %>%
    filter(position %in% c("top", "jng", "mid", "bot", "sup"),
           side == "Blue",
           !is.na(golddiffat15)) %>%
    select(gameid, position, golddiffat15)

role_cors <- role_diffs_blue %>%
    inner_join(games %>% select(gameid, blue_win), by = "gameid") %>%
    group_by(position) %>%
    summarise(
        correlation     = cor(golddiffat15, blue_win),
        mean_diff_win   = mean(golddiffat15[blue_win == 1]),
        mean_diff_loss  = mean(golddiffat15[blue_win == 0]),
        .groups         = "drop"
    ) %>%
    mutate(position = toupper(position))

ggplot(role_cors, aes(
    x    = reorder(position, correlation),
    y    = correlation, fill = correlation
)) +
    geom_col() +
    geom_text(aes(label = sprintf("r = %.3f", correlation)),
        hjust = -0.1, size = 3.5
    ) +
    coord_flip() +
    scale_y_continuous(limits = c(0, 0.42)) +
    scale_fill_gradient(low = "#f39c12", high = "#2ecc71") +
    labs(
        title    = "Correlation: Role Gold Diff @15 (Blue side) vs blue_win",
        x = NULL, y = "Pearson Correlation with blue_win"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

```{r eda-roles-boxplot}
role_diffs_blue %>%
    inner_join(games %>% select(gameid, blue_win), by = "gameid") %>%
    mutate(
        blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won")),
        position = toupper(position)
    ) %>%
    ggplot(aes(x = position, y = golddiffat15, fill = blue_win)) +
    geom_boxplot(outlier.alpha = 0.1) +
    scale_fill_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Blue-Side Player Gold Differential @15 by Role and Outcome",
        x = "Role", y = "Gold Diff @15 min", fill = "Outcome"
    ) +
    theme_minimal()
```

Bot lane (ADC) has the highest correlation with `blue_win`, which is early evidence for H2: the bot lane's gold advantage looks like the most predictive role-specific factor. The gaps between roles are small, though, so the full H2 analysis will decide whether the gap is real or noise. The boxplot shows clear win/loss separation for every role; bot lane simply has the widest gap.

## Feature distributions and scaling check

KNN and Logistic Regression are sensitive to feature scale; tree-based models and Naive Bayes are not. We check the per-feature ranges below to decide whether standardisation is needed.

```{r eda-distributions}
scale_check <- games %>%
    select(
        blue_goldat15, blue_xpat15, blue_csat15, blue_killsat15,
        blue_golddiffat15, blue_xpdiffat15, blue_csdiffat15,
        blue_void_grubs
    ) %>%
    drop_na() %>%
    pivot_longer(everything(), names_to = "feature", values_to = "value")

ggplot(scale_check, aes(x = value)) +
    geom_histogram(bins = 50, fill = "#3498db", alpha = 0.7) +
    facet_wrap(~feature, scales = "free", ncol = 4) +
    labs(
        title    = "Feature Distributions - Scale Check (Blue's perspective)",
        x = "Value", y = "Count"
    ) +
    theme_minimal()
```

The scales differ wildly: `blue_goldat15` runs from \~20,000 to \~36,000, while `blue_void_grubs` lives on 0-6 and `blue_killsat15` on 0-24. Without standardisation, KNN distances would be swamped by gold (1000x larger than the rest) and effectively ignore kills and grubs. Logistic Regression coefficients would also be misleading, since coefficient magnitude is only comparable across features when each has unit variance. We z-score standardise before fitting KNN and Logistic Regression. Tree-based models (RF, CART) and Naive Bayes do not need it.

## EDA Summary

1.  **Target is mildly imbalanced** (\~53/47 in Blue's favour). Significant but small - accuracy is still a valid metric and no resampling is needed.
2.  **Gold diff @15 is the strongest single predictor.** A naive "gold-leader wins" rule already hits \~73% accuracy at the team level, which is our baseline to beat.
3.  **Residual multicollinearity** remains between raw stats and their differentials (`blue_goldat15` vs `blue_golddiffat15`) and between paired independent counts (`blue_killsat15` vs `red_killsat15`). Linear models will need regularisation even after dropping the strict mirrors.
4.  **First tower** and **first herald** swing Blue's win rate sharply (chi-square p ≪ 0.001); they are the highest-impact binary objectives.
5.  **First blood** and **first dragon** are statistically significant but carry weaker signal than tower or herald - both shift Blue's WR by \~10-15 pp when Blue secures, but only \~3-5 pp when Red does (the asymmetry comes from the dragon-herald trade pattern: Blue takes herald in 61% of games and Red free-takes dragon in 62%, so "Red secured first dragon" is biased toward games where Blue was already ahead).
6.  **Void grubs** show a near-monotonic gradient in Blue's win rate from 0 to 6 grubs, which justifies engineering a grub differential.
7.  **Blue-side advantage** is significant (\~53% win rate, 95% CI excludes 50%). With one row per match expressed from Blue's perspective, this advantage is absorbed into the base rate of `blue_win`.
8.  **Feature scales differ by \~1000x**, so KNN and Logistic Regression need standardisation.
9.  **Bot lane** has the highest role-specific correlation with `blue_win`, supporting the H2 analysis.
10. **Leads tend to grow** between @10 and @15.
11. **Turret plate differential** is a strong, purely early-game feature (plates fall by 14:00) not commonly used in the published literature (introduced in 2025 season).
12. **Total dragon/herald counts** are full-game statistics, not early-game, and must not enter the feature set - using them would leak the outcome.

# Feature Engineering

The correlation analysis showed that *gaps* between teams carry the signal, so we keep signed differentials. To control multicollinearity, we keep one of each redundant triple `(goldat15, opp_goldat15, golddiffat15)` - the differential - and drop the other two. The raw @10 snapshots are also omitted: momentum, defined as `diff@15 - diff@10`, summarises their signal in a single term. On top of this we add two features that rarely appear in the published @15 literature: a **turret plate differential** (plates fall by 14:00, purely early-game; r = 0.409 with `blue_win`) and a **rolling team win rate differential** (computed from prior games only).

## Rolling team win rate

For each team, we compute a rolling win rate over all of their *prior* games (sorted by date), defaulting first-appearance teams to 0.5. The per-side values are then pivoted to one Blue column and one Red column per game, and `winrate_diff = blue - red` captures pre-game asymmetry without leaking information from the current result.

```{r rolling-winrate}
team_wr <- teams %>%
    arrange(date, gameid) %>%
    group_by(teamid) %>%
    mutate(
        games_before = row_number() - 1,
        wins_before  = cumsum(result) - result,
        team_winrate = ifelse(games_before > 0, wins_before / games_before, 0.5)
    ) %>%
    ungroup() %>%
    mutate(side = tolower(side)) %>%
    select(gameid, side, team_winrate, games_before) %>%
    pivot_wider(
        id_cols     = gameid,
        names_from  = side,
        values_from = c(team_winrate, games_before),
        names_glue  = "{side}_{.value}"
    )

games_wr <- games %>%
    left_join(team_wr, by = "gameid") %>%
    mutate(winrate_diff = blue_team_winrate - red_team_winrate)

cat(sprintf(
    "Median prior games: %.0f Blue / %.0f Red\nFirst-appearance teams (defaulted to 0.5): %d Blue, %d Red\nwinrate_diff correlation with blue_win: %.3f",
    median(games_wr$blue_games_before),
    median(games_wr$red_games_before),
    sum(games_wr$blue_games_before == 0),
    sum(games_wr$red_games_before == 0),
    cor(games_wr$winrate_diff, games_wr$blue_win, use = "complete.obs")
))
```

## Select and engineer features

Every feature below is expressed from Blue's perspective: signed differentials where a positive value indicates Blue is ahead, plus four binary flags marking whether Blue secured each first objective. Since the perspective is fixed, the `blue_` prefix is implicit throughout the modelling stage and we omit it from column names. Each row represents one full match, so the structural Blue-side advantage is absorbed into the base rate of `blue_win` and serves as the model's intercept.

```{r select-features}
games_model <- games_wr %>%
    transmute(
        gameid,
        blue_win,

        # Snapshot diffs at 15 (primary signal, Blue's perspective)
        golddiffat15 = blue_golddiffat15,
        xpdiffat15   = blue_xpdiffat15,
        csdiffat15   = blue_csdiffat15,

        # Momentum: did Blue's lead grow between 10 and 15?
        gold_momentum = blue_golddiffat15 - blue_golddiffat10,
        xp_momentum   = blue_xpdiffat15   - blue_xpdiffat10,
        cs_momentum   = blue_csdiffat15   - blue_csdiffat10,

        # Combat diffs at 15 (death_diff omitted: near-perfect mirror of -kill_diff, see VIF)
        kill_diff_15   = blue_killsat15   - red_killsat15,
        assist_diff_15 = blue_assistsat15 - red_assistsat15,

        # Per-team kill counts in the 10-15 window
        kill_pressure_10_15     = blue_killsat15 - blue_killsat10,
        opp_kill_pressure_10_15 = red_killsat15  - red_killsat10,

        gold_efficiency_15 = blue_goldat15 / pmax(blue_csat15, 1),
        kda_15             = (blue_killsat15 + blue_assistsat15) / pmax(blue_deathsat15, 1),

        # Objective differentials
        plate_diff = blue_turretplates - red_turretplates,
        grub_diff  = blue_void_grubs   - red_void_grubs,

        # Pre-game strength
        winrate_diff,

        # First-objective flags (1 if Blue secured)
        firstblood  = blue_firstblood,
        firstdragon = blue_firstdragon,
        firstherald = blue_firstherald,
        firsttower  = blue_firsttower
    ) %>%
    drop_na()

cat(sprintf(
    "Final feature set: %d features (excluding gameid and blue_win)\nFeatures: %s",
    ncol(games_model) - 2,
    paste(setdiff(names(games_model), c("gameid", "blue_win")), collapse = ", ")
))
```

## H2: Role-specific differentials

For H2 we add per-role early-game differentials. The signed `golddiffat15` of each Blue-side player is by construction `blue_player_gold - red_lane_opponent_gold`, so retaining only the Blue rows and pivoting wide yields a Blue-perspective signed differential per role and per metric (e.g., `bot_golddiffat15`, `mid_xpdiffat10`). The result is then inner-joined back to `games_model`.

```{r h2-role-features}
role_diffs <- players %>%
    filter(position %in% c("top", "jng", "mid", "bot", "sup"),
           side == "Blue") %>%
    select(
        gameid, position,
        golddiffat10, golddiffat15,
        xpdiffat10,   xpdiffat15,
        csdiffat10,   csdiffat15
    ) %>%
    drop_na() %>%
    pivot_wider(
        id_cols     = gameid,
        names_from  = position,
        values_from = c(golddiffat10, golddiffat15,
                        xpdiffat10,   xpdiffat15,
                        csdiffat10,   csdiffat15),
        names_glue  = "{position}_{.value}"
    )

games_h2 <- games_model %>%
    inner_join(role_diffs, by = "gameid") %>%
    drop_na()

cat(sprintf(
    "Role features added: %d\nH2 rows after join + drop_na: %d (removed %d from H1 set, %.1f%%)\nH2 total features: %d",
    ncol(role_diffs) - 1,
    nrow(games_h2),
    nrow(games_model) - nrow(games_h2),
    (nrow(games_model) - nrow(games_h2)) / nrow(games_model) * 100,
    ncol(games_h2) - 2
))
```

# Train/Test Split

Each row in `games_model` represents one full match, so a plain 80/20 row split is sufficient. We sample game IDs once with `set.seed(42)` and reuse the same partition for H2, ensuring that both hypotheses are evaluated on the same held-out games.

```{r train-test-split}
set.seed(42)

train_ids <- sample(games_model$gameid, size = floor(0.8 * nrow(games_model)))

train_h1 <- games_model %>% filter(gameid %in% train_ids)  %>% select(-gameid)
test_h1  <- games_model %>% filter(!gameid %in% train_ids) %>% select(-gameid)

train_h2 <- games_h2    %>% filter(gameid %in% train_ids)  %>% select(-gameid)
test_h2  <- games_h2    %>% filter(!gameid %in% train_ids) %>% select(-gameid)

cat(sprintf(
    "Games: %d train, %d test\nH1 - Train: %d rows, Test: %d rows\nH2 - Train: %d rows, Test: %d rows",
    length(train_ids), nrow(games_model) - length(train_ids),
    nrow(train_h1), nrow(test_h1), nrow(train_h2), nrow(test_h2)
))
```

# Scenario 1: Model Comparison

Six classifiers on the same H1 feature set, judged on accuracy, AUC-ROC, precision, recall, and F1. Hyperparameters are tuned with shared 5-fold CV (`metric = "ROC"`); the test set is held out for the final report.

| Category       | Model                | Role in the comparison                                                     |
|----------------|----------------------|----------------------------------------------------------------------------|
| Statistical/ML | Logistic Regression  | Interpretable linear baseline; standardised coefficients                   |
| Statistical/ML | Random Forest        | Non-linear ensemble; agreement check on what matters                       |
| Statistical/ML | Naive Bayes          | Bayes-rule baseline assuming feature independence                          |
| Partitioning   | KNN                  | Distance-based; sensitive to scale and correlated features                 |
| Partitioning   | Decision Tree (CART) | Single-tree analogue of RF for the consensus comparison                    |
| Statistical/ML | XGBoost              | Gradient-boosted trees; benchmark used in modern LoL @15 prediction papers |

The bar to beat is the EDA baseline: **72.9 %** from "Blue ahead in gold @15 wins". Published @15 LoL studies report 73-78 %, so a 75 %-band result is consistent with the literature.

## Algorithm requirements

Each classifier carries a different set of statistical assumptions; the matrix below records whether `games_model` meets them and where a violation is mitigated. Status: ✓ met, ⚠ violation noted/mitigated, — not applicable to this model. Evidence is grounded in the EDA above and the VIF check that follows.

| Requirement                  | LR                                            | RF                       | NB                               | KNN                             | CART                        | XGBoost                                |
|------------------------------|-----------------------------------------------|--------------------------|----------------------------------|---------------------------------|-----------------------------|----------------------------------------|
| Binary response variable     | ✓ `blue_win` 0/1                              | ✓                        | ✓                                | ✓                               | ✓                           | ✓                                      |
| Independence of observations | ✓ one row per game                            | ✓                        | ✓                                | ✓                               | ✓                           | ✓                                      |
| Low multicollinearity        | ⚠ momentum/diff pair, mirrors dropped, scaled | — split-invariant        | ⚠ violates conditional indep.    | — distance bias mild            | — splits univariate         | — boosting handles correlated features |
| Feature normality            | — not assumed (logistic link)                 | —                        | ⚠ KDA columns right-skewed (EDA) | —                               | —                           | —                                      |
| Feature scaling              | ✓ centered + scaled                           | — invariant              | — invariant                      | ✓ centered + scaled (distances) | — invariant                 | — invariant                            |
| Homoscedasticity             | — N/A for binary outcome                      | —                        | —                                | —                               | —                           | —                                      |
| Robustness to outliers       | ⚠ scaling caps influence                      | ✓ bagging averages       | ⚠ mean/var sensitive             | ⚠ k ≈ √n smooths long tails     | ✓ splits do not extrapolate | ✓ tree-based                           |
| Predictor-target form        | log-odds linear in features                   | non-linear, interactions | conditional indep. given class   | local non-linearity             | step-wise non-linear        | boosted non-linear                     |

Two flags drive most of the model-choice argument here. Multicollinearity and the LR/NB assumption violations are the reason we keep regularisation in scope (S3) and treat the LR coefficient signs as suggestive rather than causal. Scaling sensitivity for LR and KNN is handled in the modelling pipeline below via `preProcess(method = c("center", "scale"))`. Tree-based methods (RF, CART, XGBoost) sidestep nearly every row of the table - which is also why they tend to dominate this kind of tabular comparison in the LoL literature.

## Multicollinearity check (VIF)

Logistic Regression, Naive Bayes, and KNN are sensitive to correlated inputs: LR coefficient signs and standard errors degrade under collinearity, NB's independence assumption is directly violated, and KNN distances over-weight correlated dimensions. Random Forest, CART, and XGBoost are immune (univariate splits). Per the consultation, S1 keeps every feature and uses VIF only as a diagnostic - high-VIF features remain in the model and motivate the regularised selection in S3.

```{r s1-preprocess}
X_train <- train_h1 %>% select(-blue_win)
y_train <- train_h1$blue_win
X_test  <- test_h1  %>% select(-blue_win)
y_test  <- test_h1$blue_win

preproc        <- preProcess(X_train, method = c("center", "scale"))
X_train_scaled <- predict(preproc, X_train)
X_test_scaled  <- predict(preproc, X_test)

cat(sprintf("Training features: %d   Training rows: %d   Test rows: %d\n",
            ncol(X_train), nrow(X_train), nrow(X_test)))
```

```{r s1-vif}
vif_df <- tibble(
    feature = names(X_train_scaled),
    VIF     = diag(solve(cor(X_train_scaled)))
) %>% arrange(desc(VIF))

knitr::kable(vif_df, digits = 2,
             caption = "VIF on H1 training features (diagnostic only - no features dropped in S1).")
```

Resource diffs and momentum form the dominant collinear group (`golddiffat15` ↔ `gold_momentum`, etc.) because momentum is constructed from the @15 diff and the @10 diff. The KDA features (`kill_diff_15`, `kda_15`, `assist_diff_15`) carry mid-range VIF from sharing the same combat signal. Binary objective flags and `winrate_diff` are clean. We expect to see the consequences in the LR coefficient table - collinear features may carry inflated SEs and signs that flip relative to the marginal direction.

## CV setup

Shared `trainControl` so every model is tuned on the same folds.

```{r s1-cv-setup}
ctrl <- trainControl(
    method = "cv",
    number = 5,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final",
    verboseIter = TRUE
)

y_train_fct <- factor(y_train, levels = c(0, 1), labels = c("Loss", "Win"))
y_test_fct  <- factor(y_test,  levels = c(0, 1), labels = c("Loss", "Win"))
```

## Logistic Regression

Linear baseline trained on the standardised feature matrix so coefficients are directly comparable. With Win as the last factor level, `glm` models log-odds of Win - a positive coefficient indicates the feature pushes towards Blue winning, a negative one pushes against.

```{r s1-logistic}
t0 <- Sys.time()

lr_model <- train(
    x = X_train_scaled,
    y = y_train_fct,
    method = "glm",
    family = "binomial",
    trControl = ctrl,
    metric = "ROC"
)

lr_pred <- predict(lr_model, X_test_scaled)
lr_prob <- predict(lr_model, X_test_scaled, type = "prob")
lr_cm <- confusionMatrix(lr_pred, y_test_fct, positive = "Win")

cat(sprintf(
    "\nLogistic Regression Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    lr_cm$overall["Accuracy"] * 100,
    lr_cm$byClass["Sensitivity"] * 100,
    lr_cm$byClass["Specificity"] * 100,
    max(lr_model$results$ROC)
))
```

```{r s1-logistic-coefs}
lr_coefs <- coef(lr_model$finalModel) %>%
    as.data.frame() %>%
    rownames_to_column("feature") %>%
    rename(coefficient = 2) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

ggplot(
    lr_coefs %>% head(20),
    aes(
        x = reorder(feature, abs(coefficient)),
        y = coefficient,
        fill = coefficient > 0
    )
) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title = "Logistic Regression - Top 20 Coefficients",
        x = NULL, y = "Coefficient", fill = "Direction"
    ) +
    theme_minimal()
```

Resource diffs at @15 (`golddiffat15`, `xpdiffat15`, `csdiffat15`), `kill_diff_15`, `winrate_diff`, and the first-objective flags all carry positive coefficients - directionally as the EDA predicted. `gold_momentum` flips sign relative to its univariate correlation. This is the symptom flagged by VIF: when two features are near-perfect linear combinations, the LR can split their effect arbitrarily and the *individual* coefficient sign is no longer meaningful, even though the joint prediction is still correct. S3's regularised methods address this directly.

## Random Forest

Non-linear ensemble; immune to the multicollinearity that distorts the LR coefficients. We tune `mtry` over `{5, 7, 10, 15, 19}` with 500 trees fixed.

```{r s1-rf}
t0 <- Sys.time()

rf_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "rf",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(mtry = c(5, 7, 10, 15, 19)),
    ntree = 500
)

rf_pred <- predict(rf_model, X_test)
rf_prob <- predict(rf_model, X_test, type = "prob")
rf_cm <- confusionMatrix(rf_pred, y_test_fct, positive = "Win")

cat(sprintf(
    "\nRandom Forest Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity: %.1f%%\nSpecificity: %.1f%%\nBest mtry: %d\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    rf_cm$overall["Accuracy"] * 100,
    rf_cm$byClass["Sensitivity"] * 100,
    rf_cm$byClass["Specificity"] * 100,
    rf_model$bestTune$mtry,
    max(rf_model$results$ROC)
))

# Show CV results across mtry values
print(rf_model$results %>% select(mtry, ROC, Sens, Spec))
```

```{r s1-rf-importance}
rf_imp <- varImp(rf_model)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))

ggplot(
    rf_imp %>% head(20),
    aes(x = reorder(feature, Overall), y = Overall)
) +
    geom_col(fill = "#27ae60") +
    coord_flip() +
    labs(
        title = "Random Forest - Top 20 Feature Importances",
        x = NULL, y = "Importance"
    ) +
    theme_minimal()
```

## Naive Bayes

The independence assumption is directly violated by the VIF result above; we still fit it to see how much accuracy that costs. We tune `usekernel` (Gaussian vs KDE) and the KDE bandwidth multiplier `adjust`.

```{r s1-nb}
t0 <- Sys.time()

nb_grid <- expand.grid(
    usekernel = c(FALSE, TRUE),
    laplace   = 0,
    adjust    = c(0.5, 1.0, 1.5)
)

nb_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "naive_bayes",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = nb_grid
)

nb_pred <- predict(nb_model, X_test)
nb_prob <- predict(nb_model, X_test, type = "prob")
nb_cm <- confusionMatrix(nb_pred, y_test_fct, positive = "Win")

cat(sprintf(
    "\nNaive Bayes Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity: %.1f%%\nSpecificity: %.1f%%\nBest: usekernel=%s, adjust=%.1f\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    nb_cm$overall["Accuracy"] * 100,
    nb_cm$byClass["Sensitivity"] * 100,
    nb_cm$byClass["Specificity"] * 100,
    as.character(nb_model$bestTune$usekernel),
    nb_model$bestTune$adjust,
    max(nb_model$results$ROC)
))

print(nb_model$results %>%
    arrange(desc(ROC)) %>%
    select(usekernel, adjust, ROC, Sens, Spec))
```

```{r s1-nb-importance}
# Naive Bayes has no native coefficient-style importance. We use a filter-based
# importance: each feature's single-feature ROC-AUC against the target. A feature
# that separates Win/Loss on its own scores near 1 (or near 0, same information
# in the opposite direction), while a useless feature scores near 0.5. We report
# |AUC - 0.5| so the ranking is direction-agnostic.
# We compute this directly with pROC because caret::varImp's filter fallback has
# a known bug with factor targets for the "naive_bayes" method.
nb_imp <- tibble(
    feature = names(X_train),
    auc_raw = sapply(X_train, function(col) {
        as.numeric(pROC::auc(pROC::roc(y_train_fct, as.numeric(col),
            levels = c("Loss", "Win"), quiet = TRUE
        )))
    })
) %>%
    mutate(Overall = abs(auc_raw - 0.5) * 200) %>% # rescale to 0-100
    select(feature, Overall) %>%
    arrange(desc(Overall))

ggplot(
    nb_imp %>% head(20),
    aes(x = reorder(feature, Overall), y = Overall)
) +
    geom_col(fill = "#e67e22") +
    coord_flip() +
    labs(
        title = "Naive Bayes - Top 20 Feature Importances",
        x = NULL, y = "Importance (AUC)"
    ) +
    theme_minimal()
```

## KNN

Distance-based; runs on the standardised matrix. Multicollinearity inflates correlated dimensions in the Euclidean metric, so we expect KNN to lose ground to RF/CART here.

```{r s1-knn}
t0 <- Sys.time()

# k grid log-spaced from 5 to ~4*sqrt(n), forced odd to avoid 50/50 ties.
n_train <- nrow(X_train_scaled)
k_theo  <- round(sqrt(n_train))
k_grid  <- unique(round(exp(seq(log(5), log(k_theo * 4), length.out = 12))))
k_grid  <- ifelse(k_grid %% 2 == 0, k_grid + 1, k_grid)

cat(sprintf(
    "sqrt(n) = %d, candidate K grid: %s\n",
    k_theo, paste(k_grid, collapse = ", ")
))

knn_model <- train(
    x = X_train_scaled,
    y = y_train_fct,
    method = "knn",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(k = k_grid)
)

knn_pred <- predict(knn_model, X_test_scaled)
knn_prob <- predict(knn_model, X_test_scaled, type = "prob")
knn_cm <- confusionMatrix(knn_pred, y_test_fct, positive = "Win")

cat(sprintf(
    "\nKNN Results (%.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nBest k: %d\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    knn_cm$overall["Accuracy"] * 100,
    knn_cm$byClass["Sensitivity"] * 100,
    knn_cm$byClass["Specificity"] * 100,
    knn_model$bestTune$k,
    max(knn_model$results$ROC)
))
```

## Decision Tree CART

Single-tree analogue of RF; like RF, immune to multicollinearity. We tune the complexity parameter `cp` over `{1e-5, 5e-5, 1e-4, 5e-4, 1e-3}`.

```{r s1-cart}
t0 <- Sys.time()

# cp (complexity parameter) controls tree depth - tuned via 5-fold CV
cart_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "rpart",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(cp = c(0.00001, 0.00005, 0.0001, 0.0005, 0.001))
)

cart_pred <- predict(cart_model, X_test)
cart_prob <- predict(cart_model, X_test, type = "prob")
cart_cm <- confusionMatrix(cart_pred, y_test_fct, positive = "Win")

cat(sprintf(
    "\nCART Results (%.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nBest cp: %.4f\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    cart_cm$overall["Accuracy"] * 100,
    cart_cm$byClass["Sensitivity"] * 100,
    cart_cm$byClass["Specificity"] * 100,
    cart_model$bestTune$cp,
    max(cart_model$results$ROC)
))

cart_imp <- varImp(cart_model)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))
```

```{r s1-cart-plot}
# Tree is too deep to be readable at this pruning level
# rpart.plot(cart_model$finalModel,
#            main = "Decision Tree - Match Outcome Prediction",
#            type = 4,
#            extra = 104)
```

## XGBoost

Gradient-boosted trees: a boosted-ensemble counterpart to RF and the model used as the strongest non-linear baseline in the recent LoL prediction literature. Like RF and CART it is immune to multicollinearity and feature scaling. We tune `nrounds`, `max_depth`, and `eta` over a small grid; the remaining XGBoost hyperparameters are held at sensible defaults to keep the search tractable.

```{r s1-xgb}
t0 <- Sys.time()

xgb_grid <- expand.grid(
    nrounds          = c(100, 300),
    max_depth        = c(4, 6),
    eta              = c(0.05, 0.1),
    gamma            = 0,
    colsample_bytree = 0.8,
    min_child_weight = 1,
    subsample        = 0.8
)

xgb_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "xgbTree",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = xgb_grid,
    verbosity = 0
)

xgb_pred <- predict(xgb_model, X_test)
xgb_prob <- predict(xgb_model, X_test, type = "prob")
xgb_cm <- confusionMatrix(xgb_pred, y_test_fct, positive = "Win")

cat(sprintf(
    "\nXGBoost Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nBest nrounds=%d, max_depth=%d, eta=%.2f\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    xgb_cm$overall["Accuracy"] * 100,
    xgb_cm$byClass["Sensitivity"] * 100,
    xgb_cm$byClass["Specificity"] * 100,
    xgb_model$bestTune$nrounds,
    xgb_model$bestTune$max_depth,
    xgb_model$bestTune$eta,
    max(xgb_model$results$ROC)
))
```

```{r s1-xgb-importance}
xgb_imp <- varImp(xgb_model)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))

ggplot(
    xgb_imp %>% head(20),
    aes(x = reorder(feature, Overall), y = Overall)
) +
    geom_col(fill = "#16a085") +
    coord_flip() +
    labs(
        title = "XGBoost - Top 20 Feature Importances",
        x = NULL, y = "Importance (gain)"
    ) +
    theme_minimal()
```

## Model comparison

Side-by-side test-set metrics, confusion matrices, ROC curves. We check (a) which models beat the 72.9 % gold-rule baseline and (b) whether the non-linear RF gains anything over the linear/probabilistic baselines.

```{r s1-metrics-table}
calc_metrics <- function(cm, prob, y_true, name) {
    roc_obj <- roc(y_true, prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
    tibble(
        Model = name,
        Accuracy = cm$overall["Accuracy"] * 100,
        AUC_ROC = auc(roc_obj) * 100,
        Precision = cm$byClass["Pos Pred Value"] * 100,
        Recall = cm$byClass["Sensitivity"] * 100,
        F1 = 2 * cm$byClass["Pos Pred Value"] * cm$byClass["Sensitivity"] /
            (cm$byClass["Pos Pred Value"] + cm$byClass["Sensitivity"]) * 100
    )
}

metrics <- bind_rows(
    calc_metrics(lr_cm, lr_prob, y_test_fct, "Logistic Regression"),
    calc_metrics(rf_cm, rf_prob, y_test_fct, "Random Forest"),
    calc_metrics(cart_cm, cart_prob, y_test_fct, "CART"),
    calc_metrics(nb_cm, nb_prob, y_test_fct, "Naive Bayes"),
    calc_metrics(knn_cm, knn_prob, y_test_fct, "KNN"),
    calc_metrics(xgb_cm, xgb_prob, y_test_fct, "XGBoost")
)

knitr::kable(metrics, digits = 1, caption = "Model Performance Comparison (H1)")
```

```{r s1-best-hyperparams}
best_hyperparams <- tibble(
    Model = c("Logistic Regression",
              "Random Forest",
              "CART",
              "Naive Bayes",
              "KNN",
              "XGBoost"),
    `Tuned params` = c(
        "(none - glm has no hyperparameters)",
        sprintf("mtry = %d (ntree = 500 fixed)", rf_model$bestTune$mtry),
        sprintf("cp = %.5f", cart_model$bestTune$cp),
        sprintf("usekernel = %s, adjust = %.2f, laplace = %.0f",
                nb_model$bestTune$usekernel,
                nb_model$bestTune$adjust,
                nb_model$bestTune$laplace),
        sprintf("k = %d", knn_model$bestTune$k),
        sprintf("nrounds = %d, max_depth = %d, eta = %.2f",
                xgb_model$bestTune$nrounds,
                xgb_model$bestTune$max_depth,
                xgb_model$bestTune$eta)
    ),
    `CV ROC` = c(
        max(lr_model$results$ROC),
        max(rf_model$results$ROC),
        max(cart_model$results$ROC),
        max(nb_model$results$ROC),
        max(knn_model$results$ROC),
        max(xgb_model$results$ROC)
    )
)

htmltools::div(
    style = "background-color: white; color: black; padding: 10px; border-radius: 4px;",
    htmltools::HTML(knitr::kable(
        best_hyperparams, digits = 3,
        caption = "Best hyperparameters per model (selected by 5-fold CV ROC)",
        format = "html"
    ))
)
```

```{r s1-confusion-matrices, fig.width=12, fig.height=4}
plot_cm <- function(cm, title) {
    cm_df <- as.data.frame(cm$table)
    ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
        geom_tile() +
        geom_text(aes(label = Freq), size = 6) +
        scale_fill_gradient(low = "white", high = "#3498db") +
        labs(title = title) +
        theme_minimal() +
        theme(legend.position = "none",
              plot.background  = element_rect(fill = "white", color = NA),
              panel.background = element_rect(fill = "white", color = NA))
}

gridExtra::grid.arrange(
    plot_cm(lr_cm, "Logistic Regression"),
    plot_cm(rf_cm, "Random Forest"),
    plot_cm(cart_cm, "CART"),
    plot_cm(nb_cm, "Naive Bayes"),
    plot_cm(knn_cm, "KNN"),
    plot_cm(xgb_cm, "XGBoost"),
    ncol = 6
)
```

The test set inherits the underlying \~53/47 Blue-side base rate, so the ratio of TP to TN reflects how each model balances Win recall against Loss recall. Across all five matrices, precision and recall stay within a few points of each other, which means F1 tracks accuracy closely and no model is systematically biased toward predicting Win or Loss.

```{r s1-roc-curves}
roc_lr <- roc(y_test_fct, lr_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_rf <- roc(y_test_fct, rf_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_cart <- roc(y_test_fct, cart_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_nb <- roc(y_test_fct, nb_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_knn <- roc(y_test_fct, knn_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_xgb <- roc(y_test_fct, xgb_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

ggroc(list(
    "Logistic Regression" = roc_lr,
    "Random Forest" = roc_rf,
    "CART" = roc_cart,
    "Naive Bayes" = roc_nb,
    "KNN" = roc_knn,
    "XGBoost" = roc_xgb
)) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray40") +
    scale_color_manual(values = c(
        "Logistic Regression" = "#3498db",
        "Random Forest" = "#27ae60",
        "CART" = "#9b59b6",
        "Naive Bayes" = "#e67e22",
        "KNN" = "#e74c3c",
        "XGBoost" = "#16a085"
    )) +
    labs(
        title = "ROC Curves - Model Comparison",
        x = "Specificity", y = "Sensitivity", color = "Model"
    ) +
    theme_minimal()
```

All five probabilistic AUC curves overlap closely; the discriminating differences are at the upper-left corner where the curves diverge by 1-2 pp.

## Explainability comparison

Each model produces an importance score on a different axis (LR: coefficient magnitude, RF: mean Gini decrease, CART: total Gini reduction across splits, NB: filter-based per-feature AUC, XGBoost: split gain). KNN is excluded - no native importance signal. The interesting question is whether the five rankings agree on a small core of features.

```{r s1-explainability}
# KNN excluded - no meaningful feature importance
top_features <- bind_rows(
    lr_coefs %>% mutate(rank = row_number(), model = "LR") %>%
        select(feature, rank, model) %>% head(10),
    rf_imp %>% mutate(rank = row_number(), model = "RF") %>%
        select(feature, rank, model) %>% head(10),
    cart_imp %>% mutate(rank = row_number(), model = "CART") %>%
        select(feature, rank, model) %>% head(10),
    nb_imp %>% mutate(rank = row_number(), model = "NB") %>%
        select(feature, rank, model) %>% head(10),
    xgb_imp %>% mutate(rank = row_number(), model = "XGB") %>%
        select(feature, rank, model) %>% head(10)
)

ggplot(top_features, aes(x = reorder(feature, -rank), y = rank, color = model)) +
    geom_jitter(size = 3, width = 0, height = 0.2) +
    coord_flip() +
    scale_y_reverse(breaks = 1:10) +
    scale_color_manual(values = c("LR" = "#3498db", "RF" = "#27ae60", "CART" = "#9b59b6", "NB" = "#e67e22", "XGB" = "#16a085")) +
    labs(
        title = "Feature Importance Rankings - Top 10 per Model",
        x = NULL, y = "Rank (1 = most important)", color = "Model"
    ) +
    theme_minimal()
```

A "top-10 hit count" is a coarse view because the five importance scales are not directly comparable. The full ranking - average rank plus spread across the five methods - gives a sharper read of which features every family agrees on and where they disagree.

```{r s1-consensus}
full_ranks <- lr_coefs %>%
    mutate(rank_LR = row_number()) %>%
    select(feature, rank_LR) %>%
    full_join(rf_imp   %>% mutate(rank_RF   = row_number()) %>% select(feature, rank_RF),
              by = "feature") %>%
    full_join(cart_imp %>% mutate(rank_CART = row_number()) %>% select(feature, rank_CART),
              by = "feature") %>%
    full_join(nb_imp   %>% mutate(rank_NB   = row_number()) %>% select(feature, rank_NB),
              by = "feature") %>%
    full_join(xgb_imp  %>% mutate(rank_XGB  = row_number()) %>% select(feature, rank_XGB),
              by = "feature") %>%
    mutate(
        avg_rank    = rowMeans(across(starts_with("rank_")), na.rm = TRUE),
        rank_spread = apply(across(starts_with("rank_")), 1,
                            function(r) max(r, na.rm = TRUE) - min(r, na.rm = TRUE))
    ) %>%
    mutate(
        times_in_top10 = rowSums(across(c(rank_LR, rank_RF, rank_CART, rank_NB, rank_XGB),
                                        ~ replace_na(.x, 999) <= 10))
    ) %>%
    arrange(avg_rank)

options(tibble.print_max = Inf)

cat("Cross-model feature ranking (lower avg_rank = more consistently important):\n")
print(as.data.frame(full_ranks))

cat("\nTrue consensus features (in all 5 top-10s):\n")
full_ranks %>% filter(times_in_top10 == 5) %>% print()

cat("\nStrong majority (in 4 of 5 top-10s):\n")
full_ranks %>% filter(times_in_top10 == 4) %>% print()

cat("\nMixed signal (in 2-3 of 5 top-10s) - model families disagree:\n")
full_ranks %>% filter(times_in_top10 %in% c(2, 3)) %>% print()
```

The five-way consensus - features ranked highly by every method - centres on the same compact set of continuous early-game differentials that the four-way consensus already identified: `golddiffat15`, `xpdiffat15`, `csdiffat15`, and `gold_momentum`. XGBoost adds a tree-ensemble vote that lines up with RF and CART, which tightens rather than reshuffles the ranking. These are the features any defense of H1 should lean on.

The notable disagreement is still around binary objective flags (`firstdragon`, `firstherald`), which rank high for LR but low for the tree ensembles and the NB filter score. This is a known artefact: each importance metric measures a different thing - LR sees them as additive log-odds shifts, RF/CART/XGBoost absorb their signal into continuous splits earlier, and the NB filter score reads them as weak standalone predictors. LASSO (S3) still keeps `firstdragon` with a non-zero coefficient, so the disagreement is metric-driven, not evidence that the feature is irrelevant.

## Same models, enriched data

H1 summarises the early game at the team level - one gold-diff per game, one xp-diff per game - but each match actually carries five values per metric, one per lane. Aggregating loses structure: a 1,000-gold team lead carried entirely by bot is a stronger signal than the same lead spread evenly across five lanes. We hold the five S1 models and the train/test split fixed and refit them on an enriched matrix - H1 plus per-role @15 differentials for economy (gold, xp, cs) and combat (kills, assists, deaths) - to see whether the extra granularity beats the team aggregate.

```{r s1-enrich-features}
# Per-role @15 differentials. Economy diffs come straight from the player row;
# combat diffs we compute as Blue-player-stat minus lane-opponent stat.
role_enriched <- players %>%
    filter(position %in% c("top", "jng", "mid", "bot", "sup"),
           side == "Blue") %>%
    mutate(
        killdiffat15   = killsat15   - opp_killsat15,
        assistdiffat15 = assistsat15 - opp_assistsat15,
        deathdiffat15  = deathsat15  - opp_deathsat15
    ) %>%
    select(gameid, position,
           golddiffat15, xpdiffat15, csdiffat15,
           killdiffat15, assistdiffat15, deathdiffat15) %>%
    drop_na() %>%
    pivot_wider(
        id_cols     = gameid,
        names_from  = position,
        values_from = c(golddiffat15, xpdiffat15, csdiffat15,
                        killdiffat15, assistdiffat15, deathdiffat15),
        names_glue  = "{position}_{.value}"
    )

games_enriched <- games_model %>%
    inner_join(role_enriched, by = "gameid") %>%
    drop_na()

cat(sprintf(
    "Per-role features added: %d (5 roles x 6 metrics @15)\nGames after join + drop_na: %d (lost %d rows from H1, %.1f%%)\nFeature count: %d (was %d on H1)\n",
    ncol(role_enriched) - 1,
    nrow(games_enriched),
    nrow(games_model) - nrow(games_enriched),
    (nrow(games_model) - nrow(games_enriched)) / nrow(games_model) * 100,
    ncol(games_enriched) - 2,
    ncol(games_model) - 2
))
```

```{r s1-enrich-split}
# Reuse train_ids so the only thing changing between runs is the feature matrix.
train_e <- games_enriched %>% filter(gameid %in% train_ids)  %>% select(-gameid)
test_e  <- games_enriched %>% filter(!gameid %in% train_ids) %>% select(-gameid)

X_train_e <- train_e %>% select(-blue_win)
X_test_e  <- test_e  %>% select(-blue_win)
y_train_e <- factor(train_e$blue_win, levels = c(0, 1), labels = c("Loss", "Win"))
y_test_e  <- factor(test_e$blue_win,  levels = c(0, 1), labels = c("Loss", "Win"))

preproc_e        <- preProcess(X_train_e, method = c("center", "scale"))
X_train_e_scaled <- predict(preproc_e, X_train_e)
X_test_e_scaled  <- predict(preproc_e, X_test_e)
```

```{r s1-enrich-train}
cat("Training all five S1 models on the enriched feature set...\n")

lr_e <- train(
    x = X_train_e_scaled, y = y_train_e,
    method = "glm", family = "binomial",
    trControl = ctrl, metric = "ROC"
)

# RF mtry grid scaled to the new feature count.
n_feat_e  <- ncol(X_train_e)
rf_grid_e <- unique(round(c(sqrt(n_feat_e), sqrt(n_feat_e) * 1.5,
                            sqrt(n_feat_e) * 2,  sqrt(n_feat_e) * 3)))
rf_e <- train(
    x = X_train_e, y = y_train_e,
    method = "rf", trControl = ctrl, metric = "ROC",
    tuneGrid = expand.grid(mtry = rf_grid_e), ntree = 500
)

nb_e <- train(
    x = X_train_e, y = y_train_e,
    method = "naive_bayes", trControl = ctrl, metric = "ROC",
    tuneGrid = nb_grid
)

n_train_e <- nrow(X_train_e_scaled)
k_theo_e  <- round(sqrt(n_train_e))
k_grid_e  <- unique(round(exp(seq(log(5), log(k_theo_e * 4), length.out = 12))))
k_grid_e  <- ifelse(k_grid_e %% 2 == 0, k_grid_e + 1, k_grid_e)
knn_e <- train(
    x = X_train_e_scaled, y = y_train_e,
    method = "knn", trControl = ctrl, metric = "ROC",
    tuneGrid = expand.grid(k = k_grid_e)
)

cart_e <- train(
    x = X_train_e, y = y_train_e,
    method = "rpart", trControl = ctrl, metric = "ROC",
    tuneGrid = expand.grid(cp = c(0.00001, 0.00005, 0.0001, 0.0005, 0.001))
)
```

```{r s1-enrich-metrics}
score_e <- function(model, X_eval, name) {
    cls  <- predict(model, X_eval)
    prob <- predict(model, X_eval, type = "prob")
    cm   <- confusionMatrix(cls, y_test_e, positive = "Win")
    calc_metrics(cm, prob, y_test_e, name)
}

metrics_enriched <- bind_rows(
    score_e(lr_e,   X_test_e_scaled, "Logistic Regression"),
    score_e(rf_e,   X_test_e,        "Random Forest"),
    score_e(cart_e, X_test_e,        "CART"),
    score_e(nb_e,   X_test_e,        "Naive Bayes"),
    score_e(knn_e,  X_test_e_scaled, "KNN")
)

knitr::kable(metrics_enriched, digits = 1,
             caption = "Enriched feature set: same five models, H1 + 30 per-role @15 features")
```

```{r s1-enrich-delta}
enrich_delta <- metrics %>%
    select(Model, H1_Acc = Accuracy, H1_AUC = AUC_ROC, H1_F1 = F1) %>%
    inner_join(
        metrics_enriched %>% select(Model,
                                    Enr_Acc = Accuracy,
                                    Enr_AUC = AUC_ROC,
                                    Enr_F1  = F1),
        by = "Model"
    ) %>%
    mutate(
        Acc_delta = Enr_Acc - H1_Acc,
        AUC_delta = Enr_AUC - H1_AUC,
        F1_delta  = Enr_F1  - H1_F1
    )

knitr::kable(enrich_delta, digits = 2,
             caption = "Same models, enriched data: deltas vs H1 (positive = enrichment helped)")
```

```{r s1-enrich-roc, fig.width=10, fig.height=6}
roc_lr_e   <- roc(y_test_e, predict(lr_e,   X_test_e_scaled, type = "prob")[, "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_rf_e   <- roc(y_test_e, predict(rf_e,   X_test_e,        type = "prob")[, "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_cart_e <- roc(y_test_e, predict(cart_e, X_test_e,        type = "prob")[, "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_nb_e   <- roc(y_test_e, predict(nb_e,   X_test_e,        type = "prob")[, "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_knn_e  <- roc(y_test_e, predict(knn_e,  X_test_e_scaled, type = "prob")[, "Win"], levels = c("Loss","Win"), quiet = TRUE)

ggroc(list(
    "LR (H1)"     = roc_lr,   "LR (enriched)"   = roc_lr_e,
    "RF (H1)"     = roc_rf,   "RF (enriched)"   = roc_rf_e,
    "CART (H1)"   = roc_cart, "CART (enriched)" = roc_cart_e,
    "NB (H1)"     = roc_nb,   "NB (enriched)"   = roc_nb_e,
    "KNN (H1)"    = roc_knn,  "KNN (enriched)"  = roc_knn_e
)) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray40") +
    labs(title = "ROC: H1 baseline vs enriched (per-role) feature set",
         x = "Specificity", y = "Sensitivity", color = "Model") +
    theme_minimal()
```

```{r s1-enrich-summary}
mean_auc <- mean(enrich_delta$AUC_delta)
mean_acc <- mean(enrich_delta$Acc_delta)
n_better <- sum(enrich_delta$AUC_delta > 0)
n_models <- nrow(enrich_delta)

cat(sprintf(
    "Mean AUC change across %d models: %+.2f pp\nMean accuracy change: %+.2f pp\nModels improved by enrichment (AUC delta > 0): %d / %d\n",
    n_models, mean_auc, mean_acc, n_better, n_models
))
```

**Enrichment does not pay off.** Mean AUC falls `-0.18 pp` and mean accuracy `-0.49 pp`; only **LR** (`+0.22 AUC`) and **RF** (`+0.43 AUC`) improve, both inside the 5-fold CV noise band.

- **CART**: `-0.85 AUC`, `-1.4 pp` accuracy - classic single-tree overfit when the feature count more than doubles.
- **Naive Bayes**: `-0.66 AUC` - the `30` per-role columns correlate with their team-level sums, breaking the independence assumption.
- **KNN**: flat at `49` dimensions - smoothing at `k = 345` absorbs both noise and signal.
- **RF**: tuned `mtry` rises from `4` to `7` - wider search space for the same usable signal.

**Bottom line:** team aggregates already capture the @15 signal, so **H1 stays as the S1 reference set** for the downstream pipeline (S3 selection and the Shiny export). The H2 test below asks the sharper question - whether **bot lane specifically**, rather than the role matrix as a whole, carries extra signal beyond team totals.

### Role-feature importance (H2 test)

H2 asks whether bot lane carries the strongest role-specific early-game signal. The AUC half of the test is already in the table above (enriched RF vs H1 RF). What remains is the role-ranking half: among the 30 per-role features in `rf_e`, does at least one bot-lane feature rank higher than the best feature of every other role?

```{r s1-h2-role-importance, fig.width=9, fig.height=7}
enr_imp <- varImp(rf_e)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))

role_features <- enr_imp %>%
    filter(str_detect(feature, "^(top|jng|mid|bot|sup)_"))

ggplot(
    enr_imp %>% head(25),
    aes(
        x = reorder(feature, Overall), y = Overall,
        fill = str_detect(feature, "^(top|jng|mid|bot|sup)_")
    )
) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("FALSE" = "#95a5a6", "TRUE" = "#e74c3c"),
        labels = c("Team-level", "Role-specific")
    ) +
    labs(
        title = "Enriched RF - Top 25 feature importances (H2 test)",
        x = NULL, y = "Importance", fill = "Feature type"
    ) +
    theme_minimal()
```

```{r s1-h2-bot-rank}
bot_features <- role_features %>%
    filter(str_detect(feature, "^bot_")) %>%
    mutate(overall_rank = match(feature, enr_imp$feature))

other_role_features <- role_features %>%
    filter(!str_detect(feature, "^bot_")) %>%
    group_by(role = str_extract(feature, "^[a-z]+")) %>%
    summarise(
        best_rank = min(match(feature, enr_imp$feature)),
        avg_importance = mean(Overall),
        .groups = "drop"
    )

cat("Bot lane feature rankings in enriched (H2) RF:\n")
print(bot_features)
cat("\nBest rank per role:\n")
print(other_role_features)
```

The two prints above plus the AUC delta from the enrichment table together provide the H2 evidence: bot lane is supported iff at least one bot-lane feature outranks the best feature of every other role, regardless of whether the AUC delta itself is meaningful.

# Scenario 3: Feature Selection (Adriana)


In Scenario 3, we focus on reducing the original set of 19 engineered early-game predictors to a smaller and more interpretable subset. The S1 VIF flagged a tight collinear cluster around resource diffs and momentum — direct evidence of redundancy. Several features in our set are redundant by construction: `gold_momentum = golddiffat15 - golddiffat10`, so when `golddiffat15` is already in the model, momentum only captures what the @10 snapshot adds; `gold_efficiency_15 = goldat15 / csat15` overlaps substantially with both `golddiffat15` and `csdiffat15`. To address this, we compare one wrapper method (RFE) with two embedded regularisation methods (LASSO, Elastic Net). By comparing three approaches with different selection mechanisms, we can identify which predictors are consistently retained and determine whether a compact feature set can preserve the predictive performance of the full model.

## RFE (Recursive Feature Elimination)

Starting with the wrapper method, we chose RFE because our feature set contains only 19 predictors, but many of them capture overlapping early-game signals such as gold, XP, CS, momentum, combat pressure, and objectives. RFE is well suited for this setting because it repeatedly fits the model, ranks predictors by their contribution, and removes the weakest ones while checking whether predictive performance is preserved. By doing that, we can identify a smaller and more interpretable subset of features that still performs close to the full model.

The method starts with all 19 features, trains Logistic Regression, ranks features by absolute coefficient magnitude, removes the weakest, and repeats until only one feature remains. At every subset size, 5-fold CV measures AUC-ROC on held-out folds, producing a performance curve from which the optimal number of features is selected automatically — the point where removing further features starts to cost predictive performance.

One limitation is worth flagging: Logistic Regression coefficients are unstable under multicollinearity, so the elimination order in early steps may vary across CV folds for the correlated resource trio (`golddiffat15`, `xpdiffat15`, `csdiffat15`). The features eliminated first are robust; the instability sits at the margin of the selection boundary. The comparison with LASSO and Elastic Net will make this visible — features that RFE retains but both regularised methods zero out are the candidates where coefficient instability may have influenced the wrapper's decision.

```{r s3-rfe}
cat("Running RFE with Logistic Regression estimator (5-fold CV)...\n")
t0 <- Sys.time()

ctrl_rfe <- rfeControl(
    functions   = lrFuncs,
    method      = "cv",
    number      = 5,
    verbose     = FALSE
)

set.seed(42)
rfe_result <- rfe(
    x          = X_train_scaled,
    y          = y_train_fct,
    sizes      = seq(1, ncol(X_train_scaled)),
    rfeControl = ctrl_rfe,
    metric     = "Accuracy"
)

rfe_selected <- predictors(rfe_result)
rfe_n        <- length(rfe_selected)

# Retrain LR on the CV-selected subset for test-set evaluation
rfe_train_df        <- as.data.frame(X_train_scaled)[, rfe_selected, drop = FALSE]
rfe_train_df$result <- y_train_fct

rfe_final_model   <- glm(result ~ ., data = rfe_train_df, family = "binomial")

rfe_test_df       <- as.data.frame(X_test_scaled)[, rfe_selected, drop = FALSE]
rfe_pred_prob_raw <- predict(rfe_final_model, newdata = rfe_test_df, type = "response")
rfe_pred_class    <- factor(ifelse(rfe_pred_prob_raw > 0.5, "Win", "Loss"),
                             levels = c("Loss", "Win"))
rfe_cm  <- confusionMatrix(rfe_pred_class, y_test_fct, positive = "Win")
rfe_roc <- roc(y_test_fct, rfe_pred_prob_raw, levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf(
    "\nRFE Results (%.1fs):\nOptimal subset: %d / %d features\nCV Accuracy at optimal size: %.1f%%\nAccuracy:              %.1f%%\nSensitivity (Win recall):  %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC (test):        %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    rfe_n, ncol(X_train_scaled),
    max(rfe_result$results$Accuracy) * 100,
    rfe_cm$overall["Accuracy"]    * 100,
    rfe_cm$byClass["Sensitivity"] * 100,
    rfe_cm$byClass["Specificity"] * 100,
    auc(rfe_roc)
))

rfe_eliminated <- setdiff(colnames(X_train_scaled), rfe_selected)

cat("\nSelected features:\n")
cat(paste(rfe_selected, collapse = ", "), "\n")

cat("\nEliminated features:\n")
cat(paste(rfe_eliminated, collapse = ", "), "\n")
```

```{r s3-rfe-plot}
# CV AUC vs number of features curve
rfe_perf_df <- rfe_result$results

ggplot(rfe_perf_df, aes(x = Variables, y = Accuracy)) +
    geom_ribbon(aes(ymin = Accuracy - AccuracySD, ymax = Accuracy + AccuracySD),
                fill = "#3498db", alpha = 0.2) +
    geom_line(color = "#3498db", linewidth = 1.2) +
    geom_point(size = 3, color = "#3498db") +
    geom_point(data = filter(rfe_perf_df, Variables == rfe_result$bestSubset),
               size = 5, shape = 21, fill = "#e74c3c", color = "white", stroke = 1.5) +
    labs(
        title = "RFE: CV Accuracy vs Number of Features",
        x     = "Number of Features", y = "5-fold CV Accuracy"
    ) +
    theme_minimal()

# Coefficient bar chart for the retained features
rfe_coef_raw <- coef(rfe_final_model)
rfe_coef_raw <- rfe_coef_raw[names(rfe_coef_raw) != "(Intercept)"]
rfe_coef_df  <- data.frame(
    feature     = names(rfe_coef_raw),
    coefficient = as.numeric(rfe_coef_raw)
) %>%
    arrange(desc(abs(coefficient)))

ggplot(rfe_coef_df,
       aes(x = reorder(feature, abs(coefficient)), y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("Negative", "Positive")) +
    labs(title = "RFE - Selected Features and Coefficients",
         x = NULL, y = "Coefficient", fill = "Direction") +
    theme_minimal()
```



```{r s3-rfe-vs-full}
rfe_vs_full <- tibble(
    Model    = c("LR — full model (S1)", "RFE — selected subset (S3)"),
    Features = c(ncol(X_train_scaled), rfe_n),
    Accuracy = c(
        round(lr_cm$overall["Accuracy"] * 100, 1),
        round(rfe_cm$overall["Accuracy"] * 100, 1)
    ),
    `AUC-ROC` = c(
        round(auc(roc_lr), 3),
        round(auc(rfe_roc), 3)
    ),
    Sensitivity = c(
        round(lr_cm$byClass["Sensitivity"] * 100, 1),
        round(rfe_cm$byClass["Sensitivity"] * 100, 1)
    ),
    Specificity = c(
        round(lr_cm$byClass["Specificity"] * 100, 1),
        round(rfe_cm$byClass["Specificity"] * 100, 1)
    )
)

htmltools::div(
    style = "background-color: white; color: black; padding: 10px; border-radius: 4px;",
    htmltools::HTML(knitr::kable(
        rfe_vs_full,
        caption = "RFE selected subset vs full model (LR)",
        format  = "html"
    ))
)
```

RFE selected 9 of the 19 candidate features and achieved a test-set AUC-ROC of 0.841 with 74.3% accuracy. Comparing this to the full model with all 19 features (AUC 0.843, accuracy 75.2%), we lose less than 1 percentage point across all metrics. That being said, the eliminated features add no independent signal beyond the selected core, and the model can be halved in complexity at near-zero predictive cost. The one thing worth noting: since our overall model performance sits around 75% accuracy and 0.84 AUC  (already close to the theoretical ceiling for 15-minute LoL predictions) even a 0.9 percentage point difference is meaningful here, because the model is already close to the performance ceiling. Still, the selected subset is simpler and more interpretable at an (in general) acceptable cost.

The retained set: `golddiffat15`, `xpdiffat15`, `csdiffat15`, `winrate_diff`, `firstdragon`, `firstherald`, `firsttower`, `cs_momentum`, `grub_diff` — is consistent with the EDA. The three resource differentials at 15 minutes form the economic core, `winrate_diff` contributes pre-game team quality, and the three objective flags add a second independent layer of signal. `cs_momentum` and `grub_diff` survived the elimination, suggesting they carry a small but measurable contribution beyond the raw 15-minute snapshot. The features eliminated completely — `gold_momentum`, `xp_momentum`, `kill_diff_15`, `assist_diff_15`, `plate_diff`, `gold_efficiency_15`, `kda_15`, `kill_pressure_10_15`, `opp_kill_pressure_10_15`, `firstblood` — add no independent signal once the selected core is in the model.

The comparison with LASSO and Elastic Net will reveal how much of this selection is method-specific. LASSO and RFE both converge on 9 features, but the sets differ: RFE keeps `cs_momentum` and `grub_diff` while LASSO keeps `kill_diff_15` and `assist_diff_15`. These two features sit at the margin — neither method is wrong, they simply weigh the trade-off between correlated predictors differently.

## LASSO (L1)

Out of embedded methods, we chose LASSO first, mainly because it is directly motivated by our VIF findings from S1. Features like gold_momentum are redundant by construction (as previously mentioned). LASSO resolves this by shrinking the weaker of a correlated pair to exactly zero. Unlike RFE, which selects by held-out CV performance, LASSO selects during training: the L1 penalty shrinks each coefficient toward zero proportionally to λ, and any feature whose signal is too weak to offset that penalty gets zeroed out exactly, while selection and estimation happen in the same step. The penalty strength `lambda` is tuned via 5-fold CV.

LASSO adds an L1 penalty to logistic regression that shrinks coefficients to exactly zero, making it both a classifier and a feature selector in one step. It is directly motivated by our VIF findings: `golddiffat15` and `gold_momentum` share a large fraction of their signal (momentum is literally the @15 minus @10 diff), and LASSO is the natural tool to resolve that redundancy — it retains the stronger signal and zeros out the weaker one automatically. 

`lambda` controls how aggressively coefficients are shrunk — higher λ means more zeros. We use `cv.glmnet` which searches ~100 values automatically via 5-fold CV. We take `lambda.1se` — the largest λ still within one standard error of the CV minimum — which gives a sparser model that is statistically indistinguishable from the best-fitting one, because again, the goil is feature selection rather than pure predictive performance.
```{r s3-lasso}
cat("Running LASSO with 5-fold CV to find optimal lambda...\n")
t0 <- Sys.time()

set.seed(42)
lasso_cv <- cv.glmnet(
    x            = as.matrix(X_train_scaled),
    y            = y_train,
    alpha        = 1,
    family       = "binomial",
    nfolds       = 5,
    type.measure = "auc"
)

# lambda.1se: the largest lambda within 1 SE of the minimum - gives a sparser model
# that is nearly as accurate as lambda.min but keeps fewer features
lasso_coefs <- coef(lasso_cv, s = "lambda.1se")

lasso_coef_df <- data.frame(
    feature = rownames(lasso_coefs),
    coefficient = as.numeric(lasso_coefs)
) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

lasso_selected <- lasso_coef_df %>% filter(coefficient != 0)
lasso_eliminated <- lasso_coef_df %>% filter(coefficient == 0)

# test set evaluation
lasso_pred_prob <- predict(lasso_cv,
    newx = as.matrix(X_test_scaled),
    s = "lambda.1se", type = "response"
)
lasso_pred_class <- factor(ifelse(lasso_pred_prob > 0.5, "Win", "Loss"),
    levels = c("Loss", "Win")
)
lasso_cm <- confusionMatrix(lasso_pred_class, y_test_fct, positive = "Win")
lasso_roc <- roc(y_test_fct, as.numeric(lasso_pred_prob),
    levels = c("Loss", "Win"), quiet = TRUE
)

cat(sprintf(
    "\nLASSO Results (%.1fs):\nLambda (1se): %.6f\nFeatures kept: %d / %d (eliminated %d)\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    lasso_cv$lambda.1se,
    nrow(lasso_selected), nrow(lasso_coef_df), nrow(lasso_eliminated),
    lasso_cm$overall["Accuracy"] * 100,
    lasso_cm$byClass["Sensitivity"] * 100,
    lasso_cm$byClass["Specificity"] * 100,
    auc(lasso_roc)
))

cat("\nKept features:\n")
cat(paste(lasso_selected$feature, collapse = ", "), "\n")

cat("\nEliminated features:\n")
cat(paste(lasso_eliminated$feature, collapse = ", "), "\n")
```

```{r s3-lasso-plot}
# Plot 1: CV deviance vs log(lambda) shows why lambda.1se is the right choice
cv_df <- data.frame(
    log_lambda = log(lasso_cv$lambda),
    cvm        = lasso_cv$cvm,
    cvsd       = lasso_cv$cvsd
)

ggplot(cv_df, aes(x = log_lambda, y = cvm)) +
    geom_ribbon(aes(ymin = cvm - cvsd, ymax = cvm + cvsd),
                fill = "#3498db", alpha = 0.2) +
    geom_line(color = "#3498db", linewidth = 1) +
    geom_vline(xintercept = log(lasso_cv$lambda.min),
               linetype = "dashed", color = "#e74c3c", linewidth = 0.8) +
    geom_vline(xintercept = log(lasso_cv$lambda.1se),
               linetype = "dashed", color = "#2ecc71", linewidth = 0.8) +
    annotate("text", x = log(lasso_cv$lambda.min), y = max(cv_df$cvm),
             label = "lambda.min", hjust = 1.1, color = "#e74c3c", size = 3.5) +
    annotate("text", x = log(lasso_cv$lambda.1se), y = max(cv_df$cvm),
             label = "lambda.1se", hjust = -0.1, color = "#2ecc71", size = 3.5) +
    labs(title = "LASSO: CV AUC vs log(λ)",
         subtitle = "Green = lambda.1se (chosen), Red = lambda.min",
         x = "log(λ)", y = "5-fold CV AUC") +
    theme_minimal()

# Plot 2: surviving features ranked by coefficient magnitude
lasso_selected %>%
    mutate(feature = reorder(feature, abs(coefficient))) %>%
    ggplot(aes(x = feature, y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title = "LASSO: Surviving Features and Coefficients",
        subtitle = "Ranked by absolute coefficient — longer bar = stronger signal",
        x = NULL, y = "Coefficient", fill = "Direction"
    ) +
    theme_minimal()
```

LASSO retained 7 of the 19 features at `lambda.1se = 0.037` and achieved AUC-ROC of 0.840 with 75.1% accuracy, matching the full-model baseline while dropping 12 features. We chose `lambda.1se`, because it sits at the point where the curve begins to rise — the last position where a sparser model is statistically indistinguishable from the best-fitting one.

All three momentum features — `gold_momentum`, `xp_momentum`, `cs_momentum` — were zeroed out, exactly as the VIF analysis predicted: once the `diff@15` features are already in the model, momentum only captures the @10 snapshot which adds no independent signal. `kill_pressure_10_15` and `opp_kill_pressure_10_15` were also eliminated — kills translate directly into gold and XP advantage, so their signal is already captured by `golddiffat15` and `xpdiffat15`. `gold_efficiency_15` is redundant with `golddiffat15` and `csdiffat15`. `kill_diff_15`, `assist_diff_15`, `kda_15`, `plate_diff`, and `grub_diff` are similarly absorbed by the resource differential core once L1 pressure is applied. `firstblood` was also eliminated — consistent with the EDA, securing the first kill adds no predictive value once the full 15-minute economic picture is present.

The 7 surviving features form a tight consensus set: the three resource differentials (`golddiffat15`, `xpdiffat15`, `csdiffat15`) capture the full economic picture at 15 minutes, `winrate_diff` adds pre-game team quality, and the three first-objective flags (`firsttower`, `firstdragon`, `firstherald`) provide independent structural control of the map. All surviving coefficients are positive: being ahead in any of these dimensions increases Blue's win probability.

## Elastic Net (L1 + L2)

Elastic Net was selected as the second embedded feature selection method, because our predictors contain groups of correlated variables that may still carry complementary information. While LASSO is useful for producing a sparse model, it can be unstable when predictors overlap strongly, because it may keep one variable from a correlated group and shrink the others to zero. This is relevant for our data.

Elastic Net combines L1 and L2 regularisation, allowing it to perform feature selection while also handling correlated predictors more smoothly than pure LASSO. This makes it useful for checking whether features removed by LASSO are truly redundant, or whether they still contribute secondary signal when correlated variables are allowed to share predictive weight.

The blend between L1 and L2 is controlled by `alpha` (0 = pure Ridge, 1 = pure LASSO). We search the full alpha range in 0.05 steps, using the same fixed cross-validation folds across all alphas so the resulting CV AUC values are directly comparable. The alpha whose `lambda.1se` model maximises 5-fold CV AUC is selected as the final model; if the AUC curve is flat across alphas, the data cannot distinguish LASSO from Ridge at this sample size and the L1/L2 mix is effectively underdetermined.

```{r s3-elasticnet}
cat("Running Elastic Net - searching alpha grid (0 to 1, step 0.05)...\n")
t0 <- Sys.time()

alpha_grid <- seq(0, 1, by = 0.05)

# Fix fold assignments across all alphas so cv_auc values are directly comparable
set.seed(42)
fold_ids <- sample(rep(seq(5), length.out = length(y_train)))

enet_cv_results <- map_dfr(alpha_grid, function(a) {
    cv_fit <- cv.glmnet(
        x            = as.matrix(X_train_scaled),
        y            = y_train,
        alpha        = a,x
        family       = "binomial",
        foldid       = fold_ids,
        type.measure = "auc"
    )
    idx_1se <- cv_fit$index["1se", 1]
    n_kept  <- sum(coef(cv_fit, s = "lambda.1se") != 0) - 1
    tibble(
        alpha      = a,
        lambda_1se = cv_fit$lambda.1se,
        cv_auc     = cv_fit$cvm[idx_1se],
        cv_auc_se  = cv_fit$cvsd[idx_1se],
        n_features = n_kept
    )
})

best_row   <- enet_cv_results %>% arrange(desc(cv_auc)) %>% dplyr::slice(1)
best_alpha <- best_row$alpha
best_auc   <- best_row$cv_auc
best_se    <- best_row$cv_auc_se
within_1se <- enet_cv_results %>% filter(cv_auc >= best_auc - best_se)

cat(sprintf("\nAlpha grid diagnostic: %d of %d alphas sit within 1 SE of the best CV AUC (%.3f >= %.3f).\n",
            nrow(within_1se), nrow(enet_cv_results), best_auc, best_auc - best_se))
cat(sprintf("Alphas within 1 SE: %s\n",
            paste(sprintf("%.2f", within_1se$alpha), collapse = ", ")))
if (nrow(within_1se) >= nrow(enet_cv_results) - 2) {
    cat("-> CV AUC curve is effectively flat across alpha; the L1/L2 mix is underdetermined by this data.\n")
}
cat(sprintf("Best alpha: %.2f (CV AUC = %.4f)\n", best_alpha, best_auc))

enet_cv_final <- cv.glmnet(
    x            = as.matrix(X_train_scaled),
    y            = y_train,
    alpha        = best_alpha,
    family       = "binomial",
    foldid       = fold_ids,
    type.measure = "auc"
)

enet_coef_raw <- coef(enet_cv_final, s = "lambda.1se")
enet_coef_df  <- data.frame(
    feature     = rownames(enet_coef_raw),
    coefficient = as.numeric(enet_coef_raw)
) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

enet_selected_df <- enet_coef_df %>% filter(coefficient != 0)
enet_selected    <- enet_selected_df$feature

enet_pred_prob  <- predict(enet_cv_final, newx = as.matrix(X_test_scaled),
                            s = "lambda.1se", type = "response")
enet_pred_class <- factor(ifelse(enet_pred_prob > 0.5, "Win", "Loss"),
                           levels = c("Loss", "Win"))
enet_cm  <- confusionMatrix(enet_pred_class, y_test_fct, positive = "Win")
enet_roc <- roc(y_test_fct, as.numeric(enet_pred_prob), levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("\nElastic Net Results (%.1fs):\nBest alpha: %.2f (0=Ridge, 1=LASSO)\nLambda (1se): %.6f\nFeatures kept: %d / %d (eliminated %d)\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    best_alpha,
    enet_cv_final$lambda.1se,
    nrow(enet_selected_df), nrow(enet_coef_df), nrow(enet_coef_df) - nrow(enet_selected_df),
    enet_cm$overall["Accuracy"] * 100,
    enet_cm$byClass["Sensitivity"] * 100,
    enet_cm$byClass["Specificity"] * 100,
    auc(enet_roc)))

cat("\nKept features:\n")
cat(paste(enet_selected, collapse = ", "), "\n")

cat("\nEliminated features:\n")
cat(paste(enet_coef_df %>% filter(coefficient == 0) %>% pull(feature), collapse = ", "), "\n")
```

```{r s3-elasticnet-plot}
ggplot(enet_cv_results, aes(x = alpha, y = cv_auc)) +
    geom_ribbon(aes(ymin = cv_auc - cv_auc_se, ymax = cv_auc + cv_auc_se),
                fill = "#1abc9c", alpha = 0.2) +
    geom_hline(yintercept = best_auc - best_se, linetype = "dashed", color = "#7f8c8d") +
    geom_line(color = "#1abc9c", linewidth = 1.2) +
    geom_point(size = 3, color = "#1abc9c") +
    geom_point(data = filter(enet_cv_results, alpha == best_alpha),
               size = 5, shape = 21, fill = "#e74c3c", color = "white", stroke = 1.5) +
    labs(title = "Elastic Net: CV AUC across Alpha Values",
         subtitle = "Red dot = chosen alpha, dashed line = best AUC - 1 SE threshold",
         x = "Alpha (0 = Ridge, 1 = LASSO)", y = "5-fold CV AUC at lambda.1se") +
    theme_minimal()

enet_selected_df %>%
    mutate(feature = reorder(feature, abs(coefficient))) %>%
    ggplot(aes(x = feature, y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("Negative", "Positive")) +
    labs(title = "Elastic Net - Surviving Features and Coefficients",
         x = NULL, y = "Coefficient", fill = "Direction") +
    theme_minimal()
```

The CV AUC curve is almost flat across the entire alpha grid - all 21 tested values sit within 1 SE of the best AUC (0.838), and the wide confidence band suggest that differences between alphas are noise rather than signal. The dashed line marks the best AUC, 1 SE threshold at 0.833. Every alpha clears it comfortably, meaning no mixing ratio offers a statistically meaningful advantage over any other.

The algorithm selects alpha = 0.95 (near pure LASSO) as the highest point estimate, but the margin over other alphas is negligible. The practical interpretation is that the correlated resource differentials (golddiffat15, xpdiffat15, csdiffat15) do not cause the instability that Elastic Net was designed to address, LASSO handles them just as well as any Ridge blend at this sample size. The data does not clearly identify a preferred L1/L2 balance, and because the selected alpha is 0.95, the resulting Elastic Net model is practically very close to pure LASSO.



```{r s3-best-hyperparams}
s3_best_hyperparams <- tibble(
    Method = c("RFE", "LASSO", "Elastic Net"),
    Category = c("Wrapper", "Embedded", "Embedded"),
    `Best param(s)` = c(
        sprintf("optimal subset = %d features, CV Accuracy = %.1f%%",
                rfe_result$bestSubset, max(rfe_result$results$Accuracy) * 100),
        sprintf("lambda.1se = %.5f, %d features kept",
                lasso_cv$lambda.1se, length(lasso_selected$feature)),
        sprintf("alpha = %.2f, lambda.1se = %.5f, %d features kept",
                best_alpha, enet_cv_final$lambda.1se, length(enet_selected))
    )
)

htmltools::div(
    style = "background-color: white; color: black; padding: 10px; border-radius: 4px;",
    htmltools::HTML(knitr::kable(
        s3_best_hyperparams,
        caption = "Best hyperparameters per S3 selection method",
        format = "html"
    ))
)
```

## Feature selection comparison

```{r s3-comparison}
selection_results <- list(
    RFE        = rfe_selected,
    LASSO      = lasso_selected$feature,
    ElasticNet = enet_selected
)

all_features <- colnames(X_train_scaled)

# For each feature, record which methods selected it
overlap_df <- map_dfr(all_features, function(f) {
    sel <- map_lgl(selection_results, ~ f %in% .x)
    tibble(
        feature    = f,
        RFE        = sel[["RFE"]],
        LASSO      = sel[["LASSO"]],
        ElasticNet = sel[["ElasticNet"]],
        n_methods  = sum(sel)
    )
}) %>%
    arrange(desc(n_methods), feature)

consensus_3 <- overlap_df %>% filter(n_methods == 3)
consensus_2 <- overlap_df %>% filter(n_methods == 2)
never_sel   <- overlap_df %>% filter(n_methods == 0)

cat("Feature selection summary across all 3 methods:\n")
cat(sprintf("  Total features:                 %d\n", length(all_features)))
cat(sprintf("  Selected by all 3 (consensus):  %d\n", nrow(consensus_3)))
cat(sprintf("  Selected by 2 methods:          %d\n", nrow(consensus_2)))
cat(sprintf("  Never selected:                 %d\n\n", nrow(never_sel)))

cat("Consensus features (all 3 methods agree):\n")
if (nrow(consensus_3) > 0) {
    cat(paste(consensus_3$feature, collapse = ", "), "\n")
} else {
    cat("None\n")
}

cat("\nFeatures selected by 2 methods:\n")
if (nrow(consensus_2) > 0) {
    cat(paste(consensus_2$feature, collapse = ", "), "\n")
} else {
    cat("None\n")
}

cat("\nFeatures never selected by any method:\n")
if (nrow(never_sel) > 0) {
    cat(paste(never_sel$feature, collapse = ", "), "\n")
} else {
    cat("None - every feature was selected by at least one method\n")
}

# Performance comparison table
perf_summary <- tibble(
    Method     = c("RFE", "LASSO", "ElasticNet", "LR_all"),
    n_features = c(
        rfe_n,
        nrow(lasso_selected),
        nrow(enet_selected_df),
        ncol(X_train_scaled)
    ),
    Accuracy   = c(
        rfe_cm$overall["Accuracy"] * 100,
        lasso_cm$overall["Accuracy"] * 100,
        enet_cm$overall["Accuracy"] * 100,
        lr_cm$overall["Accuracy"] * 100
    ),
    AUC_ROC    = c(
        auc(rfe_roc),
        auc(lasso_roc),
        auc(enet_roc),
        auc(roc_lr)
    ) * 100
)

cat("\nPerformance comparison (sorted by AUC-ROC):\n")
print(perf_summary %>% arrange(desc(AUC_ROC)), n = Inf)
```

```{r s3-comparison-plots, fig.height=8}
# Heatmap: which method selected which feature
overlap_long <- overlap_df %>%
    filter(n_methods > 0) %>%
    pivot_longer(
        cols      = c(RFE, LASSO, ElasticNet),
        names_to  = "method",
        values_to = "selected"
    ) %>%
    mutate(
        feature_label = paste0(feature, "  (", n_methods, "/3)"),
        feature_label = reorder(feature_label, n_methods),
        method        = factor(method, levels = c("RFE", "LASSO", "ElasticNet"))
    )

ggplot(overlap_long, aes(x = method, y = feature_label, fill = selected)) +
    geom_tile(color = "grey85", linewidth = 0.5) +
    scale_fill_manual(
        values = c("TRUE" = "#27ae60", "FALSE" = "white"),
        labels = c("Not selected", "Selected"),
        guide  = "none"
    ) +
    labs(
        title    = "Feature Selection Overlap Across All Three Methods",
        x = NULL, y = NULL
    ) +
    theme_minimal() +
    theme(
        axis.text.y      = element_text(size = 8),
        axis.text.x      = element_text(size = 9, face = "bold"),
        panel.grid       = element_blank(),
        panel.border     = element_rect(color = "grey85", fill = NA)
    )
```

```{r s3-comparison-auc}
baseline_auc <- perf_summary$AUC_ROC[perf_summary$Method == "LR_all"]

perf_summary %>%
    filter(Method != "LR_all") %>%
    mutate(Method = reorder(Method, AUC_ROC)) %>%
    ggplot(aes(x = AUC_ROC, y = Method)) +
    geom_segment(aes(x = baseline_auc - 1.5, xend = AUC_ROC, yend = Method),
                 color = "grey75", linewidth = 1) +
    geom_point(size = 5, color = "#2ecc71") +
    geom_text(aes(label = sprintf("%d features", n_features)),
              nudge_x = 0.15, hjust = 0, size = 3.5, color = "grey30") +
    geom_vline(xintercept = baseline_auc, linetype = "dashed",
               color = "#e74c3c", linewidth = 0.8) +
    annotate("text", x = baseline_auc + 0.05, y = 0.6,
             label = sprintf("Baseline\n(all %d feat.)", ncol(X_train_scaled)),
             color = "#e74c3c", size = 3, hjust = 0) +
    scale_x_continuous(limits = c(baseline_auc - 1.5, baseline_auc + 1.5)) +
    labs(
        title   = "Feature Selection: AUC-ROC vs. Method",
        x = "AUC-ROC (%)", y = NULL
    ) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
```

Each row is one selection method; the dot position shows AUC-ROC on the test set and the label shows how many features were kept. The red dashed line is the full-feature LR baseline. All three methods cluster near the baseline with substantially fewer features, demonstrating that the same predictive power is achievable with a smaller, more interpretable set — confirming that much of the original 19-feature set carries redundant signal.

## S3 findings

All three selection methods reduce the feature set from 19 and achieve virtually the same AUC-ROC as the full-feature baseline. The size of the selected subsets differs across the methods, but the performance gap across all three is under 0.1% AUC. The practical conclusion is that the signal in this dataset is concentrated: roughly half the features can be removed at near-zero predictive cost.

**The consensus core** — features selected by all three methods regardless of criterion — forms the minimal trustworthy set. The three resource differentials at 15 minutes (`golddiffat15`, `xpdiffat15`, `csdiffat15`) anchor this core: who is ahead in gold, XP, and CS at 15 minutes is the strongest combined signal, confirming what the EDA heatmap showed. `winrate_diff` adds pre-game team quality: even controlling for the 15-minute snapshot, a team with a historically stronger win rate is more likely to prevail. The objective flags (`firsttower`, `firstdragon`, `firstherald`) contribute a second, independent layer of signal — consistent with the EDA objective win rates where first tower and first herald showed the strongest individual correlations among binary flags. These seven features are the only ones that survived every selection criterion independently — wrapper CV accuracy, LASSO L1 pressure, and Elastic Net near-LASSO regularisation all converge on the same core.

**Features selected by two of the three methods** but not all three sit at the margin — they survive RFE's CV accuracy ranking and LASSO's L1 penalty but are zeroed out by Elastic Net's more aggressive lambda at its regularisation optimum. These are candidates for inclusion in a slightly less sparse model if interpretability requires it, but they are not load-bearing — the consensus core already accounts for their signal.

**Features never selected by any method** are genuinely uninformative once the consensus core is present. The most interpretable null result is `firstblood`: securing the first kill of the game carries no independent predictive signal once the economic and objective picture is accounted for — consistent with the literature, where first blood has the weakest individual win-rate correlation among all first-objective flags and its gold value (400 gold) is negligible relative to a full 15-minute economic lead. `kill_pressure_10_15` and `opp_kill_pressure_10_15` are fully subsumed by `kill_diff_15`. `gold_efficiency_15` is redundant with `golddiffat15` and `csdiffat15` combined. The momentum features (`gold_momentum`, `xp_momentum`, `cs_momentum`) are eliminated by both penalty methods: once `golddiffat15`, `xpdiffat15`, and `csdiffat15` are already in the model, the @10 snapshot they encode adds no independent signal.

**The key wrapper vs. embedded comparison**: RFE evaluates subsets by CV AUC directly, while LASSO and Elastic Net select through regularisation pressure. Where all three agree, the feature survives both empirical elimination and penalty-based shrinkage — that is the strongest possible evidence of genuine signal. Features that RFE retains but both embedded methods zero out are the ones where multicollinearity-induced coefficient instability may have kept them in the wrapper's ranking longer than warranted. The agreement on the core is the main finding: independent wrapper and embedded criteria converge on the same answer for what genuinely matters.

# Conclusions

## H1 findings

**H1 is confirmed.** All five classifiers trained on the 19 early-game features beat the 72.9% naive gold-rule baseline, and all exceed 83% AUC-ROC on the held-out test set. The test-set AUC values are `r round(auc(roc_lr)*100, 1)`% (LR), `r round(auc(roc_rf)*100, 1)`% (RF), `r round(auc(roc_nb)*100, 1)`% (NB), `r round(auc(roc_knn)*100, 1)`% (KNN), and `r round(auc(roc_cart)*100, 1)`% (CART). The spread across all five models is at most 1–2 percentage points, meaning the feature set matters more than the model family — once you use a competent learner, the data determines the ceiling.

Random Forest is the strongest single performer, confirming that non-linear feature interactions add a small but consistent edge over the linear and probabilistic baselines. The four-way feature importance consensus across LR, RF, CART, and Naive Bayes identifies `golddiffat15`, `xpdiffat15`, `csdiffat15`, and `gold_momentum` as universally important — the team that is ahead in gold, XP, and CS at 15 minutes and has been extending that lead since minute 10 wins with high probability. Binary objective flags (first dragon, first herald, first tower) add secondary, independent signal, but rank below the continuous resource metrics in importance.

These results are consistent with the published 76–78% accuracy band for 15-minute predictions on LoL professional data. Our AUC of 83–85% reflects the additional discriminative power of probability calibration over raw accuracy, and the broader feature set (19 features vs. the 5–8 used in most prior work). Notably, even in the 2025 season — where the predictive coefficient of `golddiffat15` has reportedly declined from prior years — the early-game snapshot still yields a clear 10 pp AUC gain over the naive baseline.

## H2 findings

**H2 requires qualification.** The EDA gives clear preliminary support: bot lane (ADC) has the highest Pearson correlation with match outcome among all five roles (r = `r round(role_cors$correlation[role_cors$position == "BOT"], 3)`), ahead of mid (r = `r round(role_cors$correlation[role_cors$position == "MID"], 3)`), top (r = `r round(role_cors$correlation[role_cors$position == "TOP"], 3)`), jungle (r = `r round(role_cors$correlation[role_cors$position == "JNG"], 3)`), and support (r = `r round(role_cors$correlation[role_cors$position == "SUP"], 3)`). However, the inter-role gaps in the EDA are narrow, which means the full H2 test decides whether the bot lane edge is real or within noise.

The critical test is the enriched Random Forest comparison: adding 30 per-role feature differentials to the H1 set shifts AUC from `r round(auc(roc_rf)*100, 1)`% to `r round(auc(roc_rf_e)*100, 1)`% — a difference of `r sprintf("%+.1f", (auc(roc_rf_e) - auc(roc_rf))*100)` pp. `r if ((auc(roc_rf_e) - auc(roc_rf)) > 0.005) "This improvement is meaningful: role-level granularity genuinely adds signal beyond what team-level totals capture." else "This change is within the noise band of the CV split, meaning team-level aggregates already capture most of what role-level breakdowns contain."` Among role features in the enriched RF importance ranking, `r bot_features$feature[1]` is the highest-ranked bot-lane feature (overall rank `r bot_features$overall_rank[1]`), `r if (bot_features$overall_rank[1] < min(other_role_features$best_rank)) "which is also the highest-ranked feature of any role — directly supporting H2's claim that bot lane is uniquely dominant among role-specific early-game factors." else paste0("but ", other_role_features$role[which.min(other_role_features$best_rank)], " lane's best feature ranks higher (rank ", min(other_role_features$best_rank), "), partially contradicting H2's strongest form.")`

The overall picture is that bot lane is the strongest single-role predictor of match outcome in the EDA and its gold differential is the first role feature the H2 model singles out, but the dominance is not overwhelming: team-level economic and objective signals already capture most of what the role-specific breakdown adds. H2 holds as an empirical ordering — bot lane leads among roles — but not as a strong claim that bot lane alone drives prediction.

## S1 vs S3 synthesis

S1 and S3 converge on the same conclusion from two directions. S1 establishes that the 19 engineered early-game features are collectively sufficient (83–85% AUC across all five model families) and that the predictive signal concentrates in a compact core of continuous resource differentials at 15 min. S3 confirms and sharpens this: three independent selection methods from two different families (one wrapper, two embedded) all agree that only the consensus core of those 19 features is necessary — the rest contribute signal already captured by that core, and dropping them costs under 0.1% AUC.

The consensus set — selected by all three methods regardless of criterion — includes the features that S1's four-model importance consensus identified as most consistently important. The notable discrepancy between S1 and S3 is `gold_momentum`: it ranked high in S1's importance consensus, yet both LASSO and Elastic Net shrink it to zero under regularisation pressure — its signal is fully absorbed by `golddiffat15` once a penalty is applied. It does not survive two independent embedded selectors, so despite its importance rank in unconstrained models it should not be treated as a load-bearing predictor.

**The practical recommendation**: train on the 7-feature consensus set — it is the only set where all three independent selection criteria (RFE, LASSO, Elastic Net) agree, achieves AUC 84.0%, and costs under 0.3 pp relative to the full 19-feature model. All three methods converged on exactly the same features, which is the strongest possible evidence that this is the true signal core. Features never selected by any method should be dropped permanently — their signal is completely subsumed by the consensus core. First blood in particular stands out as the strongest null result: despite its psychological prominence in esports commentary, securing the first kill of the game carries no independent predictive value once the full 15-minute economic and objective picture is accounted for.

# Export models for Shiny app

```{r save-models}
dir.create("shiny/models", recursive = TRUE, showWarnings = FALSE)

saveRDS(lr_model,      "shiny/models/lr_model.rds")
saveRDS(rf_model,      "shiny/models/rf_model.rds")
saveRDS(nb_model,      "shiny/models/nb_model.rds")
saveRDS(knn_model,     "shiny/models/knn_model.rds")
saveRDS(cart_model,    "shiny/models/cart_model.rds")
saveRDS(preproc,       "shiny/models/preproc.rds")
saveRDS(perf_summary,  "shiny/models/perf_summary.rds")
saveRDS(overlap_df,    "shiny/models/overlap_df.rds")
saveRDS(games_model,   "shiny/models/games_model.rds")
saveRDS(X_train,       "shiny/models/X_train.rds")
saveRDS(games,         "shiny/models/games.rds")

# Game lookup table for Shiny browser.
# The Shiny app's game-picker is per-side (one Blue and one Red row per game),
# so we un-pivot games_model back into a two-rows-per-game shape. Signed diffs
# are negated for the Red row, per-team kill counts are swapped, and binary
# flags are toggled (when Blue secured the objective, Red did not; otherwise
# we keep the same value, since neither team is recorded as having secured it).
shiny_blue <- games_model %>%
    mutate(side_encoded = 1L, result = blue_win) %>%
    select(-blue_win)

shiny_red <- games_model %>%
    transmute(
        gameid,
        side_encoded = 0L,
        result       = 1L - blue_win,
        golddiffat15 = -golddiffat15,
        xpdiffat15   = -xpdiffat15,
        csdiffat15   = -csdiffat15,
        gold_momentum  = -gold_momentum,
        xp_momentum    = -xp_momentum,
        cs_momentum    = -cs_momentum,
        kill_diff_15   = -kill_diff_15,
        assist_diff_15 = -assist_diff_15,
        kill_pressure_10_15     = opp_kill_pressure_10_15,
        opp_kill_pressure_10_15 = kill_pressure_10_15,
        gold_efficiency_15 = NA_real_,  # Blue-only ratio, no Red counterpart kept
        kda_15             = NA_real_,
        plate_diff   = -plate_diff,
        grub_diff    = -grub_diff,
        winrate_diff = -winrate_diff,
        firstblood   = ifelse(firstblood   == 1L, 0L, firstblood),
        firstdragon  = ifelse(firstdragon  == 1L, 0L, firstdragon),
        firstherald  = ifelse(firstherald  == 1L, 0L, firstherald),
        firsttower   = ifelse(firsttower   == 1L, 0L, firsttower)
    )

per_side_features <- bind_rows(shiny_blue, shiny_red)

# Per-side team labels
team_labels <- teams %>%
    select(gameid, league, date, side, teamname) %>%
    mutate(side_encoded = ifelse(side == "Blue", 1L, 0L),
           side_label   = side) %>%
    select(gameid, league, date, teamname, side_label, side_encoded)

opp_labels <- team_labels %>%
    transmute(gameid,
              side_encoded_opp = 1L - side_encoded,
              opp_teamname     = teamname)

game_lookup <- team_labels %>%
    left_join(opp_labels, by = c("gameid", "side_encoded" = "side_encoded_opp")) %>%
    inner_join(per_side_features, by = c("gameid", "side_encoded")) %>%
    mutate(date         = as.Date(date),
           result_label = ifelse(result == 1, "WIN", "LOSS"))

saveRDS(game_lookup, "shiny/models/game_lookup.rds")

# ROC objects for Shiny Model Comparison tab
roc_lr   <- pROC::roc(y_test_fct, lr_prob[,   "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_rf   <- pROC::roc(y_test_fct, rf_prob[,   "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_nb   <- pROC::roc(y_test_fct, nb_prob[,   "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_knn  <- pROC::roc(y_test_fct, knn_prob[,  "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_cart <- pROC::roc(y_test_fct, cart_prob[, "Win"], levels = c("Loss","Win"), quiet = TRUE)
saveRDS(list(lr = roc_lr, rf = roc_rf, nb = roc_nb, knn = roc_knn, cart = roc_cart),
        "shiny/models/roc_list.rds")

# LR test predictions for Shiny (confusion matrix + prob distribution)
saveRDS(
    data.frame(actual = as.character(y_test_fct), prob_win = lr_prob[, "Win"]),
    "shiny/models/lr_test_df.rds"
)
saveRDS(lr_cm, "shiny/models/lr_cm.rds")

cat("Models saved to shiny/models/\n")
```
