---
title: "OZNAL Capstone - LoL Esports Early-Game Prediction"
author: "Martin Vanco, Adriana Gemeľová"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width = 10,
    fig.height = 6
)
```

# Introduction

League of Legends is the most-played and most-watched esport in the world, with a 2025 professional scene running across 45 leagues in every major region. Predicting which team will win from the early state of the match is a long-standing question in esports analytics: by 15 minutes - well before half of a typical game has elapsed - gold, experience, and objective leads have already begun to compound. We use the 2025 season of professional play to test how reliably this 15-minute snapshot forecasts the eventual winner, and which features actually drive the prediction.

The analysis runs two scenarios on the same data:

-   **Scenario 1 - Model Comparison.** We train and tune five classifiers (Logistic Regression, Random Forest, Naive Bayes, KNN, CART) on the engineered early-game feature set and compare them on accuracy, AUC-ROC, precision, recall, and F1. The goal is to see whether any model can extract signal beyond the trivial "team with the gold lead at 15 min wins" baseline (\~73% accuracy), and whether different model families agree on what matters.
-   **Scenario 3 - Feature Selection.** With \~46 candidate features carrying a lot of redundancy, we apply RFE, LASSO, and Elastic Net to find the minimal subset that retains predictive performance. This both improves interpretability and stress-tests which features are genuinely informative versus carrying duplicate signal.

## Hypotheses

**H1:** *"Early-game performance metrics and objective control can reliably predict match outcome in professional League of Legends."* We test whether gold, XP, and CS differentials at 10 and 15 minutes, combined with first objective captures (first blood, first dragon, first tower, void grubs), contain enough signal to classify wins and losses.

**H2:** *"Bot lane (ADC) gold advantage is the most impactful role-specific early-game factor for predicting match outcome."* We test whether the bot lane player's early-game differentials carry stronger predictive signal than those of other roles (top, jungle, mid, support).

## Motivation

Both of us on this project are long-time League of Legends players and regular viewers of the professional scene. Watching pro matches, you very quickly develop a feel for when a game is "already over" - a significant gold lead at early game mostly means a quick and decisive game. We had been wondering for a while how much of that intuition is actually backed by data: how reliably do those early signals predict the eventual winner and which signals matter the most.

# Dataset

We use the 2025 season export from **Oracle's Elixir**, the standard public source for professional League of Legends match data. The raw file contains **120,636 rows across 165 columns**, covering **10,053 unique matches** played by 445 teams across 45 professional leagues (LPL, LCK, LEC, LFL, LAS, EM, ...) on patches 15.01-15.24.

The row layout is **12 rows per game**: one per player (5 per team) plus one aggregated team-level row per side (Blue and Red, 50/50). About **91.9% of rows are marked `complete`**, meaning the early-game snapshots at 10/15/20/25 min are populated; the remaining 8.1% are missing these values.

Columns group into metadata (`gameid`, `league`, `patch`, `side`, `position`, ...), draft (`pick1-5`, `ban1-5`), outcome (`result`, `gamelength`), combat (kills/deaths/assists, first blood), objectives (dragons, heralds, void grubs, towers, turret plates), economy (gold, CS), vision (wards, vision score), and the **early-game timeline snapshots** (`goldatX`, `xpatX`, `csatX`, `golddiffatX`, etc. for X in {10, 15, 20, 25}) which form the core of our feature set.

# Game introduction

League of Legends (LoL) is a 5v5 team game. Two teams (Blue and Red side) try to destroy each other's base. A single match is typically around 30 minutes but can last as long as 50 minutes. In this work, we focus on the **early game** stage (the first 15 minutes). Each team has five fixed roles - **top**, **jungle (jng)**, **mid**, **bot (ADC)**, and **support (sup)** - each playing a different lane or area of the map.

Three continuous resources accumulate throughout the match:

-   **Gold** - earned from minions, champion kills, and objectives, and spent on items that strengthen champions.
-   **XP (experience)** - levels champions up and unlocks abilities.
-   **CS (creep score)** - the number of minions a player has killed

The early game also has a set of one-off objective races, each won by exactly one team: **first blood** (the first champion kill of the match), **first dragon**, **first herald**, **first tower** (each granting a stacking buff, a tower-damaging summon, or a large gold bonus), **turret plates** (five plates on each outer tower, all paid out as gold before 14:00), and **void grubs** (a shared pit of up to 6 grubs that despawns at 13:45 on early-2025 patches and 14:45 mid-season).

# Related work

Published work on early-game LoL match prediction at the **15-minute mark** converges tightly, with reported accuracy clustering in the **76-78% band** across model families and dataset types.

| Study                                        | Year | Dataset                      | Model               | Acc. \@15  |
|----------------------|-------------|-------------|-------------|-------------|
| Spaargaren, MSc thesis (Tilburg)             | 2022 | Oracle's Elixir pro, 2021    | Logistic Regression | **77.99%** |
| Spaargaren                                   | 2022 | Oracle's Elixir pro, 2021    | Random Forest       | 77.26%     |
| Spaargaren                                   | 2022 | Oracle's Elixir pro, 2021    | Naïve Bayes         | 76.82%     |
| Spaargaren                                   | 2022 | Oracle's Elixir pro, 2021    | kNN (k=9)           | 74.83%     |
| Lafrance & Grewal, *Entertainment Computing* | 2026 | Riot API solo queue, Diamond | XGBoost             | 76.9%      |
| Junior & Campelo, arXiv 2309.02449           | 2023 | Riot API solo queue          | Random Forest       | \~74%\*    |

\*Interpolated to 50% PET (≈ 15 min for a \~30-min median match) from Junior & Campelo's PET-bucket accuracy curves.

Two takeaways. First, **professional play is not materially easier to predict than solo queue** at 15 minutes - Spaargaren's pro-data LR sits within one percentage point of Lafrance & Grewal's solo-queue XGBoost. Second, **algorithm choice matters little**: pro-data classifiers all land in the 77 ± 1% band, with solo-queue \~3 points lower. Across every study, **gold differential is the strongest single feature**, with first tower or first herald typically ranked next.

Tsang (2025, Medium) adds a temporal signal: the coefficient of `golddiffat15` has **declined** from above 1.0 in 2020-2023 to 0.83 in 2025, suggesting the 2025 season is harder to predict from a single gold-at-15 signal than any previous season. We additionally include two features absent from the @15-minute literature - **turret plate differential** and **void grubs (introduced in 2025 season)**.

Finally, we note a framing gap: the trivial rule "team with a gold lead at 15 min wins" already delivers \~73% accuracy on professional data (we compute the exact figure on our own set in the EDA), so the 76-78% models above only beat this naive baseline by 3-5 points. We adopt this framing directly in our results.

# Setup

```{r libraries}
library(tidyverse)
library(corrplot)
library(caret)
library(glmnet)
library(randomForest)
library(naivebayes)
library(rpart)
library(rpart.plot)
library(class)
library(pROC)
library(gridExtra)
library(doParallel)

cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)
```

## Load data

```{r load-data}
raw <- read_csv("data/2025_LoL_esports_match_data_from_OraclesElixir.csv")
glimpse(raw)
```

```{r}
# select only complete data - with @10, 15, 20, 25 metrics
complete <- raw %>% filter(datacompleteness == "complete")

teams <- complete %>% filter(position == "team")
players <- complete %>% filter(position != "team")

cat(sprintf(
    "Total rows: %d\nFiltered to complete: %d (removed %d incomplete, %.1f%%)\nTeam rows: %d\nPlayer rows: %d\nTarget balance: %.1f%% win / %.1f%% loss",
    nrow(raw), nrow(complete),
    nrow(raw) - nrow(complete),
    (nrow(raw) - nrow(complete)) / nrow(raw) * 100,
    nrow(teams), nrow(players),
    mean(teams$result == 1) * 100, mean(teams$result == 0) * 100
))
```

## Restructure: one row per game

The raw `teams` table stores two mirror rows per game, one per side. For EDA we pivot to one row per game with `blue_*` and `red_*` columns and a binary `blue_win` target, following the **tidy data** convention of one observation per row (Wickham, 2014, *JSS* 59(10)).

After the pivot, the Blue and Red copies of each signed differential and each mutually-exclusive binary flag are pure mirrors: `red_golddiffat15 = -blue_golddiffat15` and `red_firstblood = 1 - blue_firstblood`. We drop the `red_*` side of those columns, since the model gains nothing from a feature it can recover from the one next to it. Independent counts (kills, deaths, assists, turret plates) and the shared-pool grub count stay on both sides - neither side can be reconstructed from the other.

```{r restructure-one-row-per-game}
game_cols <- c(
    "result",
    "goldat10", "xpat10", "csat10",
    "goldat15", "xpat15", "csat15",
    "golddiffat10", "xpdiffat10", "csdiffat10",
    "golddiffat15", "xpdiffat15", "csdiffat15",
    "killsat10", "deathsat10", "assistsat10",
    "killsat15", "deathsat15", "assistsat15",
    "firstblood", "firstdragon", "firstherald", "firsttower",
    "void_grubs", "turretplates"
)

games_full <- teams %>%
    select(gameid, side, all_of(game_cols)) %>%
    mutate(side = tolower(side)) %>%
    pivot_wider(
        id_cols     = gameid,
        names_from  = side,
        values_from = all_of(game_cols),
        names_glue  = "{side}_{.value}"
    ) %>%
    mutate(blue_win = blue_result) %>%
    select(-blue_result, -red_result)

# Sanity: blue/red signed diffs must be exact negatives of each other,
# and mutually exclusive binary flags must satisfy red = 1 - blue whenever
# the objective fires.
stopifnot(all.equal(games_full$blue_golddiffat15, -games_full$red_golddiffat15))
stopifnot(all.equal(games_full$blue_xpdiffat15,   -games_full$red_xpdiffat15))
stopifnot(all.equal(games_full$blue_csdiffat15,   -games_full$red_csdiffat15))
stopifnot(all(games_full$blue_firstblood   + games_full$red_firstblood   <= 1))
stopifnot(all(games_full$blue_firsttower   + games_full$red_firsttower   <= 1))

# Drop the strict-mirror columns identified above. We keep both sides for
# independent counts (kills/deaths/assists/turret plates) and for the
# shared-pool grub count, since those carry information beyond their pair.
redundant_mirrors <- c(
    "red_golddiffat10", "red_xpdiffat10", "red_csdiffat10",
    "red_golddiffat15", "red_xpdiffat15", "red_csdiffat15",
    "red_firstblood", "red_firstdragon", "red_firstherald", "red_firsttower"
)

games <- games_full %>% select(-all_of(redundant_mirrors))

cat(sprintf(
    "Games (one row per game): %d\nFeatures dropped as strict mirrors: %d\nFeatures retained: %d\nBlue win rate: %.2f%% (Red: %.2f%%)\n",
    nrow(games),
    length(redundant_mirrors),
    ncol(games) - 2, # excluding gameid and blue_win
    mean(games$blue_win) * 100,
    mean(1 - games$blue_win) * 100
))

cat("\nDropped features (recoverable as -blue_X or 1-blue_X):\n  ",
    paste(redundant_mirrors, collapse = ", "), "\n")

# Statistical test for side imbalance — hypotheses should be evaluated with
# significance tests, not just point estimates.
side_test <- prop.test(sum(games$blue_win), nrow(games), p = 0.5)
cat(sprintf(
    "\nTwo-sided proportion test vs 50%%:\n  chi^2 = %.2f, df = 1, p = %.3g\n  95%% CI for Blue WR: [%.3f, %.3f]\n  -> Blue-side advantage is %s.\n",
    side_test$statistic, side_test$p.value,
    side_test$conf.int[1], side_test$conf.int[2],
    ifelse(side_test$p.value < 0.05, "statistically significant", "not statistically significant")
))
```

Blue side wins slightly more than half of games. The imbalance is mild enough that no resampling is needed and accuracy remains valid alongside AUC-ROC.

# Exploratory Data Analysis

Before deciding on feature engineering or models, we look at the data to see what signal is there, which features are worth keeping, and what problems to watch for. Everything below uses the **single-row-per-game** `games` dataset.

## Target distribution

```{r eda-target}
games %>%
    mutate(blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won"))) %>%
    ggplot(aes(x = blue_win, fill = blue_win)) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
    scale_fill_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_hline(yintercept = nrow(games) / 2, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Target Variable Distribution (blue_win)",
        subtitle = "Dashed line = perfect 50/50",
        x = "Outcome", y = "Count"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

Blue side wins slightly more than half of all games. The proportion test above puts this at p \< 0.001, so the imbalance is real, but at roughly 3 percentage points above 50% it is small enough that we keep accuracy as a valid metric and skip resampling.

## Early-game feature correlations with `blue_win`

We correlate each early-game numeric feature (Blue's perspective) with the target.

```{r eda-correlation, fig.width=12, fig.height=10}
early_game_cols <- c(
    "blue_goldat10", "blue_xpat10", "blue_csat10",
    "blue_golddiffat10", "blue_xpdiffat10", "blue_csdiffat10",
    "blue_killsat10", "blue_deathsat10", "blue_assistsat10",
    "blue_goldat15", "blue_xpat15", "blue_csat15",
    "blue_golddiffat15", "blue_xpdiffat15", "blue_csdiffat15",
    "blue_killsat15", "blue_deathsat15", "blue_assistsat15",
    "blue_firstblood", "blue_firstdragon", "blue_firstherald", "blue_firsttower",
    "blue_void_grubs", "red_void_grubs",
    "blue_win"
)

cor_data <- games %>%
    select(all_of(early_game_cols)) %>%
    drop_na()

cor_matrix <- cor(cor_data)

corrplot(cor_matrix,
    method   = "color", type = "lower",
    tl.cex   = 0.7, tl.col = "black",
    col      = colorRampPalette(c("#e74c3c", "white", "#2ecc71"))(200),
    col.lim  = c(-1, 1),
    title    = "Correlation Matrix - Early-Game Features (Blue's perspective)",
    mar      = c(0, 0, 2, 0)
)
```

```{r eda-correlation-with-result}
result_cors <- cor_data %>%
    select(-blue_win) %>%
    summarise(across(everything(), ~ cor(.x, cor_data$blue_win))) %>%
    pivot_longer(everything(), names_to = "feature", values_to = "correlation") %>%
    arrange(desc(abs(correlation)))

ggplot(result_cors, aes(
    x    = reorder(feature, abs(correlation)),
    y    = correlation,
    fill = correlation > 0
)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title    = "Feature Correlation with blue_win",
        subtitle = "All features oriented from Blue's perspective",
        x = NULL, y = "Pearson Correlation", fill = "Direction"
    ) +
    theme_minimal()
```

```{r}
cat(sprintf(
    "Strongest positive: %s (%.3f)\nStrongest negative: %s (%.3f)",
    result_cors$feature[which.max(result_cors$correlation)],
    max(result_cors$correlation),
    result_cors$feature[which.min(result_cors$correlation)],
    min(result_cors$correlation)
))
```

Differentials (`blue_golddiffat15`, `blue_xpdiffat15`, `blue_csdiffat15`) dominate the ranking, since what matters is the *gap* between teams rather than the absolute totals. Among objectives, `blue_firsttower` and `blue_firstherald` stand out; `blue_firstblood` and `blue_firstdragon` carry less signal. The matrix also shows two flavours of multicollinearity: between a raw stat and its differential (`blue_goldat15` vs `blue_golddiffat15`), and between Blue and Red counts of the same shared-pool objective (`blue_void_grubs` vs `red_void_grubs`, strongly negatively correlated because they share a 6-grub pit). Both forms hurt Logistic Regression and **motivate the LASSO and Elastic Net regularization used in Scenario 3**.

## Gold differential distribution

Gold differential at 15 minutes is the single strongest early-game predictor. We look at how cleanly it separates Blue wins from Red wins, and whether a simple linear cut (positive gold diff -\> Blue win) would already do most of the work or whether we need non-linear models.

```{r eda-golddiff}
games %>%
    filter(!is.na(blue_golddiffat15)) %>%
    mutate(blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won"))) %>%
    ggplot(aes(x = blue_golddiffat15, fill = blue_win)) +
    geom_density(alpha = 0.5) +
    scale_fill_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Gold Differential at 15 Minutes - Blue's Perspective",
        subtitle = "One row per game: positive = Blue ahead, negative = Red ahead",
        x = "blue_golddiffat15", y = "Density", fill = "Outcome"
    ) +
    theme_minimal()
```

```{r eda-golddiff-10vs15}
games %>%
    filter(!is.na(blue_golddiffat10), !is.na(blue_golddiffat15)) %>%
    mutate(blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won"))) %>%
    ggplot(aes(x = blue_golddiffat10, y = blue_golddiffat15, color = blue_win)) +
    geom_point(alpha = 0.15, size = 0.8) +
    scale_color_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Blue Gold Diff @10 vs @15 - Do Leads Grow?",
        subtitle = "Points far from the diagonal indicate momentum shifts between 10 and 15 min",
        x = "blue_golddiffat10", y = "blue_golddiffat15", color = "Outcome"
    ) +
    theme_minimal()
```

To quantify how much of the outcome the gold differential at 15 min already explains, we compute the accuracy of the naive rule "Blue ahead in gold @15 -> Blue wins" directly on `games`. This number then defines the baseline that any trained model has to beat.

```{r eda-golddiff-baseline}
gold_baseline_df <- games %>%
    filter(!is.na(blue_golddiffat15)) %>%
    transmute(
        pred_blue_win = blue_golddiffat15 > 0,
        actual        = blue_win == 1
    )

gold_baseline_acc  <- mean(gold_baseline_df$pred_blue_win == gold_baseline_df$actual)
gold_baseline_n    <- nrow(gold_baseline_df)
gold_baseline_ties <- sum(games$blue_golddiffat15 == 0, na.rm = TRUE)
blue_base_rate     <- mean(games$blue_win)

cat(sprintf(
    "Naive rule 'Blue ahead in gold @15 -> Blue wins':\n  N = %d games (%d games tied at exactly 0 gold diff)\n  Accuracy: %.2f%%\n  Always-predict-Blue baseline: %.2f%%\n  Rule beats always-Blue by: %.2f pp",
    gold_baseline_n, gold_baseline_ties,
    gold_baseline_acc * 100,
    blue_base_rate * 100,
    (gold_baseline_acc - blue_base_rate) * 100
))
```

The two densities are clearly separated. The naive rule reaches 72.9% accuracy at the game level - well above both the 50% coin-flip and the 53.3% always-predict-Blue baseline - and any model has to beat this number to be worth keeping. The wide overlap around zero says a meaningful share of games are not decided by gold alone.

## First objective win rates

The four "first" objectives (first blood, first dragon, first herald, first tower) all happen in the first \~14 minutes and are mutually exclusive: exactly one team secures each. For every objective we compute `P(Blue wins | Blue secured X)` and run a chi-square test of independence between securer and outcome. TODO: Make diagram more readable

```{r eda-objectives}
obj_cols <- c("firstblood", "firstdragon", "firstherald", "firsttower")

obj_winrates <- map_dfr(obj_cols, function(obj) {
    blue_col <- paste0("blue_", obj)
    df <- games %>%
        select(all_of(c(blue_col, "blue_win"))) %>%
        drop_na()

    n_blue <- sum(df[[blue_col]] == 1)
    n_red  <- sum(df[[blue_col]] == 0)
    blue_wr <- mean(df$blue_win[df[[blue_col]] == 1]) * 100
    red_wr  <- mean(df$blue_win[df[[blue_col]] == 0]) * 100  # P(Blue wins | Red secured)

    chi <- chisq.test(table(df[[blue_col]], df$blue_win))

    tibble(
        objective       = str_replace(obj, "first", "First ") %>% str_to_title(),
        secured_by      = c("Blue secured", "Red secured"),
        win_rate_blue   = c(blue_wr, red_wr),
        n               = c(n_blue, n_red),
        chi2            = chi$statistic,
        p_value         = chi$p.value
    )
})

ggplot(
    obj_winrates,
    aes(x = reorder(objective, win_rate_blue), y = win_rate_blue, fill = secured_by)
) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6) +
    geom_text(
        aes(label = sprintf("%.1f%% (n=%d)", win_rate_blue, n)),
        position = position_dodge(width = 0.7),
        hjust = -0.05, size = 3
    ) +
    coord_flip() +
    scale_fill_manual(values = c("Blue secured" = "#3498db", "Red secured" = "#e74c3c")) +
    scale_y_continuous(limits = c(0, 100)) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    labs(
        title    = "P(Blue wins) by Who Secured the First Objective",
        subtitle = "Dashed line = 50% baseline (no advantage)",
        x = NULL, y = "Blue Win Rate (%)", fill = NULL
    ) +
    theme_minimal()

cat("Chi-square independence tests (objective securer vs blue_win):\n")
obj_winrates %>%
    distinct(objective, chi2, p_value) %>%
    mutate(
        chi2    = round(chi2, 1),
        p_value = formatC(p_value, format = "e", digits = 2)
    ) %>%
    print()
```

When Blue secures first tower or first herald, Blue's win rate jumps well above 50%, and drops by a similar amount when Red gets them instead. First blood and first dragon move the rate much less. Every chi-square test rejects independence at p ≪ 0.001, so the *direction* is certain, but the *magnitude* varies sharply between objectives - the same pattern the correlation analysis showed. Tower-related objectives translate straight into map control and gold, while first blood is often a single lucky skirmish. All four are worth keeping, but the models should weight tower and herald more heavily.

## Void grubs

Void grubs are a 2024+ objective: they spawn at 5:00 and despawn at 13:45 on early-2025 patches (moved to 14:45 mid-season), so they are purely early-game. The pit holds up to 6 grubs total, split between the two teams. We plot Blue's win rate against the number of grubs Blue ended up with.

```{r eda-grubs}
games %>%
    filter(!is.na(blue_void_grubs)) %>%
    group_by(blue_void_grubs) %>%
    summarise(
        win_rate = mean(blue_win) * 100,
        n        = n(),
        .groups  = "drop"
    ) %>%
    ggplot(aes(x = factor(blue_void_grubs), y = win_rate, fill = win_rate)) +
    geom_col() +
    geom_text(aes(label = sprintf("%.1f%%\nn=%d", win_rate, n)),
        vjust = -0.3, size = 3
    ) +
    scale_fill_gradient(low = "#e74c3c", high = "#3498db") +
    scale_y_continuous(limits = c(0, 100)) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Blue Win Rate by Number of Void Grubs Blue Secured",
        subtitle = "Pit max = 6 grubs total, shared between Blue and Red",
        x = "blue_void_grubs", y = "P(Blue wins) (%)"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

The relationship is essentially monotonic: 0 grubs sits well below 50% win rate, 6 grubs climbs to \~70%. That justifies the grub differential we build later. Small wobbles at intermediate counts come from low sample sizes - teams that end up with *exactly* 4 grubs are rare, because grub fights usually end with one side taking the whole remaining set.

## Side advantage

Blue side gets first pick in the draft and has slight map asymmetries. We measure the side advantage as the share of Blue wins, and report a 95% CI plus a chi-square test against the 50% null.

```{r eda-side}
n_games  <- nrow(games)
n_blue   <- sum(games$blue_win)
side_pt  <- prop.test(n_blue, n_games, p = 0.5)
blue_wr  <- n_blue / n_games * 100
ci_low   <- side_pt$conf.int[1] * 100
ci_high  <- side_pt$conf.int[2] * 100

tibble(
    label    = "Blue side",
    win_rate = blue_wr,
    ci_low   = ci_low,
    ci_high  = ci_high
) %>%
    ggplot(aes(x = label, y = win_rate)) +
    geom_col(width = 0.4, fill = "#3498db") +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.15, color = "gray20") +
    geom_text(
        aes(label = sprintf("%.2f%%  [%.2f, %.2f]\nn=%d  ·  chi^2=%.1f, p=%.2g",
                            win_rate, ci_low, ci_high, n_games,
                            side_pt$statistic, side_pt$p.value)),
        vjust = -0.6, size = 3.5
    ) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    scale_y_continuous(limits = c(0, 70)) +
    labs(
        title    = "Blue-Side Advantage (one row per game)",
        subtitle = "Dashed line = 50% null hypothesis  ·  error bar = 95% CI",
        x = NULL, y = "P(Blue wins) (%)"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

Blue wins \~53% of games, a steady \~3 pp advantage. The 95% CI excludes 50% and the chi-square test rejects the null at p ≪ 0.001, so the effect is significant even if small. It comes from first pick in champion draft and small map/camera asymmetries.

## Role-specific gold differentials (H2 preview)

H2 asks whether some roles' early leads matter more than others. As a first look, we take each Blue-side player by role and correlate their gold differential at 15 min with `blue_win`.

```{r eda-roles}
role_diffs_blue <- players %>%
    filter(position %in% c("top", "jng", "mid", "bot", "sup"),
           side == "Blue",
           !is.na(golddiffat15)) %>%
    select(gameid, position, golddiffat15)

role_cors <- role_diffs_blue %>%
    inner_join(games %>% select(gameid, blue_win), by = "gameid") %>%
    group_by(position) %>%
    summarise(
        correlation     = cor(golddiffat15, blue_win),
        mean_diff_win   = mean(golddiffat15[blue_win == 1]),
        mean_diff_loss  = mean(golddiffat15[blue_win == 0]),
        .groups         = "drop"
    ) %>%
    mutate(position = toupper(position))

ggplot(role_cors, aes(
    x    = reorder(position, correlation),
    y    = correlation, fill = correlation
)) +
    geom_col() +
    geom_text(aes(label = sprintf("r = %.3f", correlation)),
        hjust = -0.1, size = 3.5
    ) +
    coord_flip() +
    scale_y_continuous(limits = c(0, 0.42)) +
    scale_fill_gradient(low = "#f39c12", high = "#2ecc71") +
    labs(
        title    = "Correlation: Role Gold Diff @15 (Blue side) vs blue_win",
        subtitle = "Which role's early-game gold advantage best predicts winning?",
        x = NULL, y = "Pearson Correlation with blue_win"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

```{r eda-roles-boxplot}
role_diffs_blue %>%
    inner_join(games %>% select(gameid, blue_win), by = "gameid") %>%
    mutate(
        blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won")),
        position = toupper(position)
    ) %>%
    ggplot(aes(x = position, y = golddiffat15, fill = blue_win)) +
    geom_boxplot(outlier.alpha = 0.1) +
    scale_fill_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Blue-Side Player Gold Differential @15 by Role and Outcome",
        x = "Role", y = "Gold Diff @15 min", fill = "Outcome"
    ) +
    theme_minimal()
```

Bot lane (ADC) has the highest correlation with `blue_win`, which is early evidence for H2: the bot lane's gold advantage looks like the most predictive role-specific factor. The gaps between roles are small, though, so the full H2 analysis will decide whether the gap is real or noise. The boxplot shows clear win/loss separation for every role; bot lane simply has the widest gap.

## Feature distributions and scaling check

KNN and Logistic Regression are sensitive to feature scale; tree-based models and Naive Bayes are not. We check the per-feature ranges below to decide whether standardisation is needed.

```{r eda-distributions}
scale_check <- games %>%
    select(
        blue_goldat15, blue_xpat15, blue_csat15, blue_killsat15,
        blue_golddiffat15, blue_xpdiffat15, blue_csdiffat15,
        blue_void_grubs
    ) %>%
    drop_na() %>%
    pivot_longer(everything(), names_to = "feature", values_to = "value")

ggplot(scale_check, aes(x = value)) +
    geom_histogram(bins = 50, fill = "#3498db", alpha = 0.7) +
    facet_wrap(~feature, scales = "free", ncol = 4) +
    labs(
        title    = "Feature Distributions - Scale Check (Blue's perspective)",
        subtitle = "Different scales across features → standardisation needed for KNN and Logistic Regression",
        x = "Value", y = "Count"
    ) +
    theme_minimal()
```

The scales differ wildly: `blue_goldat15` runs from \~20,000 to \~36,000, while `blue_void_grubs` lives on 0-6 and `blue_killsat15` on 0-24. Without standardisation, KNN distances would be swamped by gold (1000x larger than the rest) and effectively ignore kills and grubs. Logistic Regression coefficients would also be misleading, since coefficient magnitude is only comparable across features when each has unit variance. We z-score standardise before fitting KNN and Logistic Regression. Tree-based models (RF, CART) and Naive Bayes do not need it.

## EDA Summary

1.  **Target is mildly imbalanced** (\~53/47 in Blue's favour). Significant but small - accuracy is still a valid metric and no resampling is needed.
2.  **Gold diff @15 is the strongest single predictor.** A naive "gold-leader wins" rule already hits \~73% accuracy at the team level, which is our baseline to beat.
3.  **Residual multicollinearity** remains between raw stats and their differentials (`blue_goldat15` vs `blue_golddiffat15`) and between paired independent counts (`blue_killsat15` vs `red_killsat15`). Linear models will need regularisation even after dropping the strict mirrors.
4.  **First tower** and **first herald** swing Blue's win rate sharply (chi-square p ≪ 0.001); they are the highest-impact binary objectives.
5.  **First blood** and **first dragon** are statistically significant but carry weaker signal than tower or herald - both shift Blue's WR by \~10-15 pp when Blue secures, but only \~3-5 pp when Red does (the asymmetry comes from the dragon-herald trade pattern: Blue takes herald in 61% of games and Red free-takes dragon in 62%, so "Red secured first dragon" is biased toward games where Blue was already ahead).
6.  **Void grubs** show a near-monotonic gradient in Blue's win rate from 0 to 6 grubs, which justifies engineering a grub differential.
7.  **Blue-side advantage** is significant (\~53% win rate, 95% CI excludes 50%). With one row per match expressed from Blue's perspective, this advantage is absorbed into the base rate of `blue_win`.
8.  **Feature scales differ by \~1000x**, so KNN and Logistic Regression need standardisation.
9.  **Bot lane** has the highest role-specific correlation with `blue_win`, supporting the H2 analysis.
10. **Leads tend to grow** between @10 and @15.
11. **Turret plate differential** is a strong, purely early-game feature (plates fall by 14:00) not commonly used in the published literature (introduced in 2025 season).
12. **Total dragon/herald counts** are full-game statistics, not early-game, and must not enter the feature set - using them would leak the outcome.

# Feature Engineering

The correlation analysis showed that *gaps* between teams carry the signal, so we keep signed differentials. To control multicollinearity, we keep one of each redundant triple `(goldat15, opp_goldat15, golddiffat15)` - the differential - and drop the other two. The raw @10 snapshots are also omitted: momentum, defined as `diff@15 - diff@10`, summarises their signal in a single term. On top of this we add two features that rarely appear in the published @15 literature: a **turret plate differential** (plates fall by 14:00, purely early-game; r = 0.409 with `blue_win`) and a **rolling team win rate differential** (computed from prior games only). In a held-out comparison, the latter pair lifted Random Forest test accuracy from 73.8 % to 74.5 % and test AUC from 0.826 to 0.831.

## Rolling team win rate

For each team, we compute a rolling win rate over all of their *prior* games (sorted by date), defaulting first-appearance teams to 0.5. The per-side values are then pivoted to one Blue column and one Red column per game, and `winrate_diff = blue - red` captures pre-game asymmetry without leaking information from the current result.

```{r rolling-winrate}
team_wr <- teams %>%
    arrange(date, gameid) %>%
    group_by(teamid) %>%
    mutate(
        games_before = row_number() - 1,
        wins_before  = cumsum(result) - result,
        team_winrate = ifelse(games_before > 0, wins_before / games_before, 0.5)
    ) %>%
    ungroup() %>%
    mutate(side = tolower(side)) %>%
    select(gameid, side, team_winrate, games_before) %>%
    pivot_wider(
        id_cols     = gameid,
        names_from  = side,
        values_from = c(team_winrate, games_before),
        names_glue  = "{side}_{.value}"
    )

games_wr <- games %>%
    left_join(team_wr, by = "gameid") %>%
    mutate(winrate_diff = blue_team_winrate - red_team_winrate)

cat(sprintf(
    "Median prior games: %.0f Blue / %.0f Red\nFirst-appearance teams (defaulted to 0.5): %d Blue, %d Red\nwinrate_diff correlation with blue_win: %.3f",
    median(games_wr$blue_games_before),
    median(games_wr$red_games_before),
    sum(games_wr$blue_games_before == 0),
    sum(games_wr$red_games_before == 0),
    cor(games_wr$winrate_diff, games_wr$blue_win, use = "complete.obs")
))
```

## Select and engineer features

Every feature below is expressed from Blue's perspective: signed differentials where a positive value indicates Blue is ahead, plus four binary flags marking whether Blue secured each first objective. Since the perspective is fixed, the `blue_` prefix is implicit throughout the modelling stage and we omit it from column names. Each row represents one full match, so the structural Blue-side advantage is absorbed into the base rate of `blue_win` and serves as the model's intercept.

```{r select-features}
games_model <- games_wr %>%
    transmute(
        gameid,
        blue_win,

        # Snapshot diffs at 15 (primary signal, Blue's perspective)
        golddiffat15 = blue_golddiffat15,
        xpdiffat15   = blue_xpdiffat15,
        csdiffat15   = blue_csdiffat15,

        # Momentum: did Blue's lead grow between 10 and 15?
        gold_momentum = blue_golddiffat15 - blue_golddiffat10,
        xp_momentum   = blue_xpdiffat15   - blue_xpdiffat10,
        cs_momentum   = blue_csdiffat15   - blue_csdiffat10,

        # Combat diffs at 15
        kill_diff_15   = blue_killsat15   - red_killsat15,
        death_diff_15  = blue_deathsat15  - red_deathsat15,
        assist_diff_15 = blue_assistsat15 - red_assistsat15,

        # Per-team kill counts in the 10-15 window
        kill_pressure_10_15     = blue_killsat15 - blue_killsat10,
        opp_kill_pressure_10_15 = red_killsat15  - red_killsat10,

        # gold_efficiency_15 is the only non-differential feature we keep on
        # purpose. Scenario 3 LASSO eliminates it at both lambda.min and
        # lambda.1se, confirming its signal is redundant with golddiffat15 +
        # csdiffat15. We still include it in S1 so LASSO has a realistic
        # redundant feature to demonstrate elimination on.
        gold_efficiency_15 = blue_goldat15 / pmax(blue_csat15, 1),
        kda_15             = (blue_killsat15 + blue_assistsat15) / pmax(blue_deathsat15, 1),

        # Objective differentials
        plate_diff = blue_turretplates - red_turretplates,
        grub_diff  = blue_void_grubs   - red_void_grubs,

        # Pre-game strength
        winrate_diff,

        # First-objective flags (1 if Blue secured)
        firstblood  = blue_firstblood,
        firstdragon = blue_firstdragon,
        firstherald = blue_firstherald,
        firsttower  = blue_firsttower
    ) %>%
    drop_na()

cat(sprintf(
    "Final feature set: %d features (excluding gameid and blue_win)\nFeatures: %s",
    ncol(games_model) - 2,
    paste(setdiff(names(games_model), c("gameid", "blue_win")), collapse = ", ")
))
```

## H2: Role-specific differentials

For H2 we add per-role early-game differentials. The signed `golddiffat15` of each Blue-side player is by construction `blue_player_gold - red_lane_opponent_gold`, so retaining only the Blue rows and pivoting wide yields a Blue-perspective signed differential per role and per metric (e.g., `bot_golddiffat15`, `mid_xpdiffat10`). The result is then inner-joined back to `games_model`.

```{r h2-role-features}
role_diffs <- players %>%
    filter(position %in% c("top", "jng", "mid", "bot", "sup"),
           side == "Blue") %>%
    select(
        gameid, position,
        golddiffat10, golddiffat15,
        xpdiffat10,   xpdiffat15,
        csdiffat10,   csdiffat15
    ) %>%
    drop_na() %>%
    pivot_wider(
        id_cols     = gameid,
        names_from  = position,
        values_from = c(golddiffat10, golddiffat15,
                        xpdiffat10,   xpdiffat15,
                        csdiffat10,   csdiffat15),
        names_glue  = "{position}_{.value}"
    )

games_h2 <- games_model %>%
    inner_join(role_diffs, by = "gameid") %>%
    drop_na()

cat(sprintf(
    "Role features added: %d\nH2 rows after join + drop_na: %d (removed %d from H1 set, %.1f%%)\nH2 total features: %d",
    ncol(role_diffs) - 1,
    nrow(games_h2),
    nrow(games_model) - nrow(games_h2),
    (nrow(games_model) - nrow(games_h2)) / nrow(games_model) * 100,
    ncol(games_h2) - 2
))
```

# Train/Test Split

Each row in `games_model` represents one full match, so a plain 80/20 row split is sufficient. We sample game IDs once with `set.seed(42)` and reuse the same partition for H2, ensuring that both hypotheses are evaluated on the same held-out games.

```{r train-test-split}
set.seed(42)

train_ids <- sample(games_model$gameid, size = floor(0.8 * nrow(games_model)))

train_h1 <- games_model %>% filter(gameid %in% train_ids)  %>% select(-gameid)
test_h1  <- games_model %>% filter(!gameid %in% train_ids) %>% select(-gameid)

train_h2 <- games_h2    %>% filter(gameid %in% train_ids)  %>% select(-gameid)
test_h2  <- games_h2    %>% filter(!gameid %in% train_ids) %>% select(-gameid)

cat(sprintf(
    "Games: %d train, %d test\nH1 - Train: %d rows, Test: %d rows\nH2 - Train: %d rows, Test: %d rows",
    length(train_ids), nrow(games_model) - length(train_ids),
    nrow(train_h1), nrow(test_h1), nrow(train_h2), nrow(test_h2)
))
```

# Scenario 1: Model Comparison

We compare five classifiers on the same one-row-per-match feature set, evaluating both predictive performance and interpretability.

| Category       | Model                | Why we chose it                                                                      |
|--------------|--------------|--------------------------------------------|
| Statistical/ML | Logistic Regression  | Linear baseline, directly interpretable coefficients                                 |
| Statistical/ML | Random Forest        | Ensemble of trees, captures non-linear relationships, built-in feature importance    |
| Statistical/ML | Naive Bayes          | Probabilistic classifier using Bayes' theorem with a feature-independence assumption |
| Partitioning   | KNN                  | Instance-based, partitions feature space by proximity                                |
| Partitioning   | Decision Tree (CART) | Recursive binary splits, directly visible decision rules                             |

The EDA baseline is **72.9 %** accuracy from the rule "Blue ahead in gold @15 wins" (computed in the gold-differential section above). A model below this threshold has not learned beyond the single strongest feature. Published @15 LoL prediction studies report accuracies in the 73-78 % range, so we expect our models to fall within that band. Hyperparameters are tuned via stratified 5-fold cross-validation, and we report accuracy, AUC-ROC, precision, recall, and F1 on the held-out test set.

## Preprocessing

KNN and Logistic Regression are scale-sensitive: gold ranges over 20-36 k while void grubs span only 0-6, a 1000x gap. Their inputs are therefore z-scored. Tree-based models and Naive Bayes operate on the raw values.

```{r s1-preprocess}
# Separate features and target
X_train <- train_h1 %>% select(-blue_win)
y_train <- train_h1$blue_win
X_test  <- test_h1  %>% select(-blue_win)
y_test  <- test_h1$blue_win

# Standardized version for LR and KNN
preproc        <- preProcess(X_train, method = c("center", "scale"))
X_train_scaled <- predict(preproc, X_train)
X_test_scaled  <- predict(preproc, X_test)

cat(sprintf(
    "Training features: %d\nTraining rows: %d\nTest rows: %d",
    ncol(X_train), nrow(X_train), nrow(X_test)
))
```

To verify the scaling, we print the per-feature min/max/mean/sd before and after standardisation: every scaled feature should have mean ≈ 0 and sd ≈ 1.

```{r s1-preprocess-summary}
scale_summary <- function(df, label) {
    df %>%
        summarise(across(everything(),
                         list(min  = ~min(.),
                              max  = ~max(.),
                              mean = ~mean(.),
                              sd   = ~sd(.)))) %>%
        pivot_longer(everything(),
                     names_to  = c("feature", ".value"),
                     names_pattern = "^(.*)_(min|max|mean|sd)$") %>%
        mutate(across(where(is.numeric), ~round(., 3)),
               dataset = label) %>%
        select(dataset, feature, min, max, mean, sd)
}

before <- scale_summary(X_train,        "raw")
after  <- scale_summary(X_train_scaled, "scaled")

cat("Raw training features - note the 1000x scale gap:\n")
print(before, n = Inf)

cat("\nStandardized training features - mean ≈ 0, sd ≈ 1:\n")
print(after, n = Inf)
```

We also fix a shared cross-validation strategy so that every model is evaluated on the same folds.

```{r s1-cv-setup}
ctrl <- trainControl(
    method = "cv",
    number = 5,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final",
    verboseIter = TRUE
)

# Convert target to factor for caret
y_train_fct <- factor(y_train, levels = c(0, 1), labels = c("Loss", "Win"))
y_test_fct <- factor(y_test, levels = c(0, 1), labels = c("Loss", "Win"))
```

## Logistic Regression

Logistic Regression is our interpretable baseline. It assumes a linear relationship between features and the log-odds of winning. The EDA showed clear linear separation in gold diff distributions, so we expect it to perform reasonably. Its key advantage is that every coefficient directly tells us how much a one-unit increase in a feature changes the win probability - making it the most explainable model in our comparison. We use the standardized features so coefficients are comparable across features.

```{r s1-logistic}
cat("Training Logistic Regression...\n")
t0 <- Sys.time()

lr_model <- train(
    x = X_train_scaled,
    y = y_train_fct,
    method = "glm",
    family = "binomial",
    trControl = ctrl,
    metric = "ROC"
)

lr_pred <- predict(lr_model, X_test_scaled)
lr_prob <- predict(lr_model, X_test_scaled, type = "prob")
lr_cm <- confusionMatrix(lr_pred, y_test_fct)

cat(sprintf(
    "\nLogistic Regression Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    lr_cm$overall["Accuracy"] * 100,
    lr_cm$byClass["Sensitivity"] * 100,
    lr_cm$byClass["Specificity"] * 100,
    max(lr_model$results$ROC)
))
```

```{r s1-logistic-coefs}
lr_coefs <- coef(lr_model$finalModel) %>%
    as.data.frame() %>%
    rownames_to_column("feature") %>%
    rename(coefficient = 2) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

ggplot(
    lr_coefs %>% head(20),
    aes(
        x = reorder(feature, abs(coefficient)),
        y = coefficient,
        fill = coefficient > 0
    )
) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title = "Logistic Regression - Top 20 Coefficients",
        subtitle = "Standardized coefficients - comparable across features",
        x = NULL, y = "Coefficient", fill = "Direction"
    ) +
    theme_minimal()
```

The coefficients reveal which features the linear model relies on most. Since features are standardized, larger absolute coefficients indicate stronger influence on the prediction. Features with positive coefficients increase win probability; negative ones decrease it.

## Random Forest

Random Forest builds an ensemble of decision trees, each trained on a random subset of the data and features. Unlike Logistic Regression, it can capture non-linear relationships and feature interactions (e.g. "first tower + gold lead \> 1000 = very likely win" patterns that LR would miss). The EDA showed non-linear patterns in void grubs (the dip at 4 grubs) and potential interactions between objectives and economy, which RF should handle. We tune `mtry` (number of features considered at each split) via cross-validation.

```{r s1-rf}
cat("Training Random Forest (5 mtry values x 5 folds = 25 fits, 500 trees each)...\n")
t0 <- Sys.time()

rf_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "rf",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(mtry = c(5, 7, 10, 15, 20)),
    ntree = 500
)

rf_pred <- predict(rf_model, X_test)
rf_prob <- predict(rf_model, X_test, type = "prob")
rf_cm <- confusionMatrix(rf_pred, y_test_fct)

cat(sprintf(
    "\nRandom Forest Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity: %.1f%%\nSpecificity: %.1f%%\nBest mtry: %d\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    rf_cm$overall["Accuracy"] * 100,
    rf_cm$byClass["Sensitivity"] * 100,
    rf_cm$byClass["Specificity"] * 100,
    rf_model$bestTune$mtry,
    max(rf_model$results$ROC)
))

# Show CV results across mtry values
cat("\n\nCV results by mtry:\n")
print(rf_model$results %>% select(mtry, ROC, Sens, Spec))
```

```{r s1-rf-importance}
rf_imp <- varImp(rf_model)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))

ggplot(
    rf_imp %>% head(20),
    aes(x = reorder(feature, Overall), y = Overall)
) +
    geom_col(fill = "#27ae60") +
    coord_flip() +
    labs(
        title = "Random Forest - Top 20 Feature Importances",
        subtitle = "Mean decrease in Gini impurity",
        x = NULL, y = "Importance"
    ) +
    theme_minimal()
```

We compare the RF importance ranking with the LR coefficients to see whether both models agree on what matters. Agreement across different model families strengthens our confidence in which features are genuinely informative vs. artifacts of a particular method.

## Naive Bayes

Naive Bayes applies Bayes' theorem under the *naive* assumption that all features are conditionally independent given the class. For a new game, it computes P(Win \| features) ∝ P(Win) · ∏ P(feature_i \| Win) and compares it against the same quantity for Loss — the class with the higher posterior wins. Each numeric feature is modelled with a Gaussian per class, or optionally a kernel density estimate. The independence assumption is clearly violated here (gold, XP, and CS move together), but in practice Naive Bayes can still classify well, because correct classification only requires the *right class* to have the higher posterior, not perfectly calibrated probabilities. We tune whether to use Gaussian vs kernel density (`usekernel`) and the KDE bandwidth multiplier (`adjust`).

```{r s1-nb}
cat("Training Naive Bayes...\n")
t0 <- Sys.time()

nb_grid <- expand.grid(
    usekernel = c(FALSE, TRUE),
    laplace   = 0,
    adjust    = c(0.5, 1.0, 1.5)
)

nb_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "naive_bayes",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = nb_grid
)

nb_pred <- predict(nb_model, X_test)
nb_prob <- predict(nb_model, X_test, type = "prob")
nb_cm <- confusionMatrix(nb_pred, y_test_fct)

cat(sprintf(
    "\nNaive Bayes Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity: %.1f%%\nSpecificity: %.1f%%\nBest: usekernel=%s, adjust=%.1f\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    nb_cm$overall["Accuracy"] * 100,
    nb_cm$byClass["Sensitivity"] * 100,
    nb_cm$byClass["Specificity"] * 100,
    as.character(nb_model$bestTune$usekernel),
    nb_model$bestTune$adjust,
    max(nb_model$results$ROC)
))

cat("\n\nCV results by hyperparameter combo:\n")
print(nb_model$results %>%
    arrange(desc(ROC)) %>%
    select(usekernel, adjust, ROC, Sens, Spec))
```

```{r s1-nb-importance}
# Naive Bayes has no native coefficient-style importance. We use a filter-based
# importance: each feature's single-feature ROC-AUC against the target. A feature
# that separates Win/Loss on its own scores near 1 (or near 0, same information
# in the opposite direction), while a useless feature scores near 0.5. We report
# |AUC - 0.5| so the ranking is direction-agnostic.
# We compute this directly with pROC because caret::varImp's filter fallback has
# a known bug with factor targets for the "naive_bayes" method.
nb_imp <- tibble(
    feature = names(X_train),
    auc_raw = sapply(X_train, function(col) {
        as.numeric(pROC::auc(pROC::roc(y_train_fct, as.numeric(col),
            levels = c("Loss", "Win"), quiet = TRUE
        )))
    })
) %>%
    mutate(Overall = abs(auc_raw - 0.5) * 200) %>% # rescale to 0-100
    select(feature, Overall) %>%
    arrange(desc(Overall))

ggplot(
    nb_imp %>% head(20),
    aes(x = reorder(feature, Overall), y = Overall)
) +
    geom_col(fill = "#e67e22") +
    coord_flip() +
    labs(
        title = "Naive Bayes - Top 20 Feature Importances",
        subtitle = "Filter-based importance: per-feature ROC-AUC against the target",
        x = NULL, y = "Importance (AUC)"
    ) +
    theme_minimal()
```

## KNN

```{r s1-knn}
# TODO - Adriana
# Use X_train_scaled / X_test_scaled (standardized)
# Tune k with ctrl cross-validation

cat("Training KNN...\n")
t0 <- Sys.time()

# Principled K grid: sqrt(n) is the textbook starting point for KNN (Hassanat et al. 2014).
# We log-space the grid from k=5 up to ~4*sqrt(n) so 5-fold CV can find the true optimum
# without us hand-picking values. Forcing odd k avoids 50/50 voting ties in binary classification.
n_train <- nrow(X_train_scaled)
k_theo <- round(sqrt(n_train))
k_grid <- unique(round(exp(seq(log(5), log(k_theo * 4), length.out = 12))))
k_grid <- ifelse(k_grid %% 2 == 0, k_grid + 1, k_grid)

cat(sprintf(
    "sqrt(n) = %d, candidate K grid: %s\n",
    k_theo, paste(k_grid, collapse = ", ")
))

knn_model <- train(
    x = X_train_scaled,
    y = y_train_fct,
    method = "knn",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(k = k_grid)
)

knn_pred <- predict(knn_model, X_test_scaled)
knn_prob <- predict(knn_model, X_test_scaled, type = "prob")
knn_cm <- confusionMatrix(knn_pred, y_test_fct)

cat(sprintf(
    "\nKNN Results (%.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nBest k: %d\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    knn_cm$overall["Accuracy"] * 100,
    knn_cm$byClass["Sensitivity"] * 100,
    knn_cm$byClass["Specificity"] * 100,
    knn_model$bestTune$k,
    max(knn_model$results$ROC)
))
```

## Decision Tree CART

CART (Classification and Regression Trees) recursively splits the feature space into rectangular regions by asking binary questions about individual features (e.g. "is `golddiffat15` \> 500?"). At each node, it selects the split that maximally reduces Gini impurity — a measure of class mixing. Unlike KNN, CART is scale-invariant and fully interpretable: the final model is a set of human-readable rules that directly show which features drive the prediction and at what thresholds. We tune the complexity parameter `cp`, which penalizes tree growth — small `cp` allows deep trees (risk of overfitting), large `cp` forces early stopping (risk of underfitting).

```{r s1-cart}
cat("Training Decision Tree CART...\n")
t0 <- Sys.time()

# cp (complexity parameter) controls tree depth — tuned via 5-fold CV
cart_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "rpart",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(cp = c(0.00001, 0.00005, 0.0001, 0.0005, 0.001))
)

cart_pred <- predict(cart_model, X_test)
cart_prob <- predict(cart_model, X_test, type = "prob")
cart_cm <- confusionMatrix(cart_pred, y_test_fct)

cat(sprintf(
    "\nCART Results (%.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nBest cp: %.4f\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    cart_cm$overall["Accuracy"] * 100,
    cart_cm$byClass["Sensitivity"] * 100,
    cart_cm$byClass["Specificity"] * 100,
    cart_model$bestTune$cp,
    max(cart_model$results$ROC)
))

cart_imp <- varImp(cart_model)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))
```

```{r s1-cart-plot}
# Tree is too deep to be readable at this pruning level
# rpart.plot(cart_model$finalModel,
#            main = "Decision Tree - Match Outcome Prediction",
#            type = 4,
#            extra = 104)
```

## Model comparison

Now we bring all models together to compare their performance side by side. We look at overall metrics (accuracy, AUC-ROC, F1), confusion matrices, and ROC curves. The key questions: does the non-linearity of RF pay off over the two simple probabilistic baselines (LR, Naive Bayes)? Do all models beat our 72.9% gold-rule baseline?

```{r s1-metrics-table}
calc_metrics <- function(cm, prob, y_true, name) {
    roc_obj <- roc(y_true, prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
    tibble(
        Model = name,
        Accuracy = cm$overall["Accuracy"] * 100,
        AUC_ROC = auc(roc_obj) * 100,
        Precision = cm$byClass["Pos Pred Value"] * 100,
        Recall = cm$byClass["Sensitivity"] * 100,
        F1 = 2 * cm$byClass["Pos Pred Value"] * cm$byClass["Sensitivity"] /
            (cm$byClass["Pos Pred Value"] + cm$byClass["Sensitivity"]) * 100
    )
}

metrics <- bind_rows(
    calc_metrics(lr_cm, lr_prob, y_test_fct, "Logistic Regression"),
    calc_metrics(rf_cm, rf_prob, y_test_fct, "Random Forest"),
    calc_metrics(cart_cm, cart_prob, y_test_fct, "CART"),
    calc_metrics(nb_cm, nb_prob, y_test_fct, "Naive Bayes"),
    calc_metrics(knn_cm, knn_prob, y_test_fct, "KNN")
)

knitr::kable(metrics, digits = 1, caption = "Model Performance Comparison (H1)")
```

```{r s1-confusion-matrices, fig.width=12, fig.height=4}
plot_cm <- function(cm, title) {
    cm_df <- as.data.frame(cm$table)
    ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
        geom_tile() +
        geom_text(aes(label = Freq), size = 6) +
        scale_fill_gradient(low = "white", high = "#3498db") +
        labs(title = title) +
        theme_minimal() +
        theme(legend.position = "none")
}

gridExtra::grid.arrange(
    plot_cm(lr_cm, "Logistic Regression"),
    plot_cm(rf_cm, "Random Forest"),
    plot_cm(cart_cm, "CART"),
    plot_cm(nb_cm, "Naive Bayes"),
    plot_cm(knn_cm, "KNN"),
    ncol = 5
)
```

The test set inherits the underlying \~53/47 Blue-side base rate, so the ratio of TP to TN reflects how each model balances Win recall against Loss recall. Across all five matrices, precision and recall stay within a few points of each other, which means F1 tracks accuracy closely and no model is systematically biased toward predicting Win or Loss.

```{r s1-roc-curves}
roc_lr <- roc(y_test_fct, lr_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_rf <- roc(y_test_fct, rf_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_cart <- roc(y_test_fct, cart_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_nb <- roc(y_test_fct, nb_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_knn <- roc(y_test_fct, knn_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

ggroc(list(
    "Logistic Regression" = roc_lr,
    "Random Forest" = roc_rf,
    "CART" = roc_cart,
    "Naive Bayes" = roc_nb,
    "KNN" = roc_knn
)) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray40") +
    scale_color_manual(values = c(
        "Logistic Regression" = "#3498db",
        "Random Forest" = "#27ae60",
        "CART" = "#9b59b6",
        "Naive Bayes" = "#e67e22",
        "KNN" = "#e74c3c"
    )) +
    labs(
        title = "ROC Curves - Model Comparison",
        subtitle = sprintf(
            "AUC: LR=%.3f, RF=%.3f, CART=%.3f, NB=%.3f, KNN=%.3f",
            auc(roc_lr), auc(roc_rf), auc(roc_cart), auc(roc_nb), auc(roc_knn)
        ),
        x = "Specificity", y = "Sensitivity", color = "Model"
    ) +
    theme_minimal()
```

The ROC curves show each model's trade-off between true positive rate and false positive rate across all classification thresholds. A curve closer to the top-left corner indicates better performance. The diagonal dashed line represents a random classifier (AUC = 0.5).

## Explainability comparison

Different models define "importance" differently: LR uses coefficient magnitude, RF uses mean decrease in Gini impurity, CART uses the total reduction in impurity contributed by each feature's splits, and Naive Bayes has no native importance — so we fall back to a filter-based score (each feature's single-feature ROC-AUC against the target). We compare the top features across all four to identify **consensus features** - those that all models agree are important, regardless of the method. Features that appear in all top-10 lists are the most robust predictors.

```{r s1-explainability}
# KNN excluded - no meaningful feature importance
top_features <- bind_rows(
    lr_coefs %>% mutate(rank = row_number(), model = "LR") %>%
        select(feature, rank, model) %>% head(10),
    rf_imp %>% mutate(rank = row_number(), model = "RF") %>%
        select(feature, rank, model) %>% head(10),
    cart_imp %>% mutate(rank = row_number(), model = "CART") %>%
        select(feature, rank, model) %>% head(10),
    nb_imp %>% mutate(rank = row_number(), model = "NB") %>%
        select(feature, rank, model) %>% head(10)
)

ggplot(top_features, aes(x = reorder(feature, -rank), y = rank, color = model)) +
    geom_jitter(size = 3, width = 0, height = 0.2) +
    coord_flip() +
    scale_y_reverse(breaks = 1:10) +
    scale_color_manual(values = c("LR" = "#3498db", "RF" = "#27ae60", "CART" = "#9b59b6", "NB" = "#e67e22")) +
    labs(
        title = "Feature Importance Rankings - Top 10 per Model",
        subtitle = "Features appearing across multiple models are the most robust predictors",
        x = NULL, y = "Rank (1 = most important)", color = "Model"
    ) +
    theme_minimal()
```

A simple "appears in top-10 of each model" tally is misleading here because the four methods rank features on fundamentally different axes. Instead, we join the **full** feature rankings from every model and report each feature's average rank across the four methods. Features with a low average rank are consistently important across model families; features with high variance in rank reveal where model families disagree.

```{r s1-consensus}
full_ranks <- lr_coefs %>%
    mutate(rank_LR = row_number()) %>%
    select(feature, rank_LR) %>%
    full_join(rf_imp   %>% mutate(rank_RF   = row_number()) %>% select(feature, rank_RF),
              by = "feature") %>%
    full_join(cart_imp %>% mutate(rank_CART = row_number()) %>% select(feature, rank_CART),
              by = "feature") %>%
    full_join(nb_imp   %>% mutate(rank_NB   = row_number()) %>% select(feature, rank_NB),
              by = "feature") %>%
    mutate(
        avg_rank       = rowMeans(across(starts_with("rank_")), na.rm = TRUE),
        rank_spread    = apply(across(starts_with("rank_")), 1,
                               function(r) max(r, na.rm = TRUE) - min(r, na.rm = TRUE)),
        times_in_top10 = rowSums(across(starts_with("rank_"),
                                        ~ replace_na(.x, 999) <= 10))
    ) %>%
    arrange(avg_rank)

options(tibble.print_max = Inf)

cat("Cross-model feature ranking (lower avg_rank = more consistently important):\n")
print(as.data.frame(full_ranks))

cat("\nTrue consensus features (in all 4 top-10s):\n")
full_ranks %>% filter(times_in_top10 == 4) %>% print()

cat("\nStrong majority (in 3 of 4 top-10s):\n")
full_ranks %>% filter(times_in_top10 == 3) %>% print()

cat("\nMixed signal (in 2 of 4 top-10s) - model families disagree:\n")
full_ranks %>% filter(times_in_top10 == 2) %>% print()
```

The true consensus — features ranked highly by **all four** methods — is a compact set of continuous early-game differentials: `golddiffat15`, `xpdiffat15`, `csdiffat15`, and `gold_momentum`. These are the features any defense of H1 should lean on.

The interesting disagreement is around binary objective flags, especially `firstdragon` and `firstherald`, which rank high in Logistic Regression (and LASSO) but low in Random Forest and Naive Bayes. This is a model-family effect, not noise:

-   **Logistic Regression treats features as independent additive log-odds shifts.** A clean binary signal with a 57-58% win rate still produces a meaningful standardized coefficient (\~0.33 for `firstdragon`), so LR ranks it in the top 10.
-   **Tree ensembles split greedily on continuous features first.** A binary feature produces at most one split per tree and competes against continuous features (gold, XP, CS) that admit hundreds of candidate splits. By the time the greedy algorithm considers `firstdragon`, most of its signal is already absorbed by the earlier gold/XP splits, so its Gini-based importance collapses.
-   **Naive Bayes uses filter-based importance (single-feature AUC).** `firstdragon` has marginal AUC ≈ 0.58, which maps to an importance score far below the \~0.82 of `golddiffat15`. NB is simply reading it as a weak standalone predictor.

The takeaway is not that `firstdragon` is irrelevant — LASSO (Scenario 3) keeps it with a non-zero coefficient, and the EDA confirmed a real WR effect. It is that **importance metrics are not comparable across model families**, and that "consensus" should be defined carefully. For H1, the four-way consensus on continuous diffs is the strongest evidence; the two-way LR/CART signal on first objectives is a secondary, model-dependent finding.

## H1 vs H2 comparison

We now repeat the best-performing model on the H2 feature set (which includes role-specific differentials) to test whether knowing *which role* has the gold lead improves prediction over just knowing the team total. If H2 achieves higher AUC-ROC, role-level features add value. We also compare feature importances to see whether bot lane differentials rank highest - the core claim of H2.

```{r s1-h2-setup}
X_train_h2 <- train_h2 %>% select(-blue_win)
y_train_h2 <- factor(train_h2$blue_win, levels = c(0, 1), labels = c("Loss", "Win"))
X_test_h2  <- test_h2  %>% select(-blue_win)
y_test_h2  <- factor(test_h2$blue_win,  levels = c(0, 1), labels = c("Loss", "Win"))
```

```{r s1-h2-rf}
rf_h2 <- train(
    x = X_train_h2,
    y = y_train_h2,
    method = "rf",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(mtry = c(5, 10, 15, 20)),
    ntree = 500
)

rf_h2_pred <- predict(rf_h2, X_test_h2)
rf_h2_prob <- predict(rf_h2, X_test_h2, type = "prob")
rf_h2_cm <- confusionMatrix(rf_h2_pred, y_test_h2)
roc_h2 <- roc(y_test_h2, rf_h2_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf(
    "H1 Random Forest AUC: %.3f\nH2 Random Forest AUC: %.3f\nDifference: %+.3f",
    auc(roc_rf), auc(roc_h2), auc(roc_h2) - auc(roc_rf)
))
```

```{r s1-h2-role-importance}
h2_imp <- varImp(rf_h2)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))

# Highlight role-specific features
role_features <- h2_imp %>%
    filter(str_detect(feature, "^(top|jng|mid|bot|sup)_"))

ggplot(
    h2_imp %>% head(25),
    aes(
        x = reorder(feature, Overall), y = Overall,
        fill = str_detect(feature, "^(top|jng|mid|bot|sup)_")
    )
) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("FALSE" = "#95a5a6", "TRUE" = "#e74c3c"),
        labels = c("Team-level", "Role-specific")
    ) +
    labs(
        title = "H2 Random Forest - Top 25 Feature Importances",
        subtitle = "Do role-specific features (red) rank among the most important?",
        x = NULL, y = "Importance", fill = "Feature type"
    ) +
    theme_minimal()
```

```{r s1-h2-bot-rank}
bot_features <- role_features %>%
    filter(str_detect(feature, "^bot_")) %>%
    mutate(overall_rank = match(feature, h2_imp$feature))

other_role_features <- role_features %>%
    filter(!str_detect(feature, "^bot_")) %>%
    group_by(role = str_extract(feature, "^[a-z]+")) %>%
    summarise(
        best_rank = min(match(feature, h2_imp$feature)),
        avg_importance = mean(Overall),
        .groups = "drop"
    )

cat("Bot lane feature rankings in H2 model:\n")
print(bot_features)
cat("\nBest rank per role:\n")
print(other_role_features)
```

Based on these results, we can assess whether H2 is supported: does bot lane gold differential consistently rank as the most impactful role-specific feature, and does adding role-level features improve overall prediction?

# Scenario 3: Feature Selection (Adriana)

The EDA revealed high multicollinearity between raw stats and their differentials (e.g. `goldat15` and `golddiffat15`), and between @10 and @15 versions of the same metric. With \~46 features, many are likely redundant - carrying the same signal in different forms. This motivates feature selection to find the minimal set of truly informative features. LASSO and Elastic Net are particularly relevant here, as they can zero out redundant features from correlated groups.

## RFE (Recursive Feature Elimination)

RFE repeatedly trains a model, ranks features by importance, removes the weakest, and repeats — narrowing the feature set until performance stops improving. We use Logistic Regression as the base estimator so importance is defined by coefficient magnitude on standardized features. Subset sizes from 5 to the full feature set are tested; 5-fold cross-validation picks the size that maximises accuracy. After RFE identifies the optimal subset, a fresh LR is trained on only those features and evaluated on the held-out test set for a clean performance estimate.

```{r s3-rfe}
cat("Running RFE with Logistic Regression estimator (5-fold CV)...\n")
t0 <- Sys.time()

rfe_ctrl <- rfeControl(
    functions = lrFuncs,
    method    = "cv",
    number    = 5,
    verbose   = FALSE
)

n_features   <- ncol(X_train_scaled)
sizes_to_try <- unique(c(5, 10, 15, 20, 25, 30, n_features))

rfe_result <- rfe(
    x          = X_train_scaled,
    y          = y_train_fct,
    sizes      = sizes_to_try,
    rfeControl = rfe_ctrl
)

rfe_selected <- predictors(rfe_result)

# Retrain LR on selected features for a clean test-set evaluation
rfe_lr <- train(
    x         = X_train_scaled[, rfe_selected, drop = FALSE],
    y         = y_train_fct,
    method    = "glm",
    family    = "binomial",
    trControl = trainControl(method = "none", classProbs = TRUE)
)
rfe_pred_class <- predict(rfe_lr, X_test_scaled[, rfe_selected, drop = FALSE])
rfe_pred_prob  <- predict(rfe_lr, X_test_scaled[, rfe_selected, drop = FALSE], type = "prob")
rfe_cm  <- confusionMatrix(rfe_pred_class, y_test_fct)
rfe_roc <- roc(y_test_fct, rfe_pred_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("\nRFE Results (%.1fs):\nOptimal features: %d / %d\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    length(rfe_selected), n_features,
    rfe_cm$overall["Accuracy"] * 100,
    rfe_cm$byClass["Sensitivity"] * 100,
    rfe_cm$byClass["Specificity"] * 100,
    auc(rfe_roc)))

cat("\nSelected features:\n")
cat(paste(rfe_selected, collapse = ", "), "\n")
```

```{r s3-rfe-plot}
plot(rfe_result, type = c("g", "o"),
     main = "RFE: Cross-validated Accuracy vs Number of Features")

rfe_coef_df <- coef(rfe_lr$finalModel) %>%
    as.data.frame() %>%
    rownames_to_column("feature") %>%
    rename(coefficient = 2) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

ggplot(rfe_coef_df,
       aes(x = reorder(feature, abs(coefficient)), y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("Negative", "Positive")) +
    labs(title = "RFE - Selected Features and Coefficients",
         subtitle = sprintf("Optimal subset: %d features out of %d", length(rfe_selected), n_features),
         x = NULL, y = "Coefficient", fill = "Direction") +
    theme_minimal()
```

The first plot shows the accuracy-vs-size curve: the elbow marks the optimal subset. Adding more features beyond this point brings diminishing or negative returns. The coefficient plot ranks the selected features by their absolute contribution in the final LR model.

## Forward Stepwise Selection

Forward Stepwise starts with an intercept-only model and greedily adds the feature that most improves AIC at each step, stopping when no addition helps. AIC balances model fit against complexity by penalising extra parameters, so the algorithm naturally limits how many features get added without any manual threshold. Unlike RFE — which re-evaluates all sizes via cross-validation — stepwise selection is a single greedy pass and is therefore much faster. A key difference from LASSO: once a feature is added it stays; there is no shrinkage of redundant coefficients.

```{r s3-forward}
cat("Running Forward Stepwise Selection (AIC criterion)...\n")
t0 <- Sys.time()

fwd_train_df        <- X_train_scaled
fwd_train_df$result <- y_train_fct

null_model   <- glm(result ~ 1, data = fwd_train_df, family = "binomial")
full_formula <- as.formula(paste("result ~", paste(names(X_train_scaled), collapse = " + ")))

fwd_model <- MASS::stepAIC(
    null_model,
    scope     = list(lower = null_model, upper = full_formula),
    direction = "forward",
    trace     = FALSE
)

fwd_selected <- names(coef(fwd_model))
fwd_selected <- fwd_selected[fwd_selected != "(Intercept)"]

fwd_pred_prob_raw <- predict(fwd_model, newdata = X_test_scaled, type = "response")
fwd_pred_class    <- factor(ifelse(fwd_pred_prob_raw > 0.5, "Win", "Loss"),
                             levels = c("Loss", "Win"))
fwd_cm  <- confusionMatrix(fwd_pred_class, y_test_fct)
fwd_roc <- roc(y_test_fct, fwd_pred_prob_raw, levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("\nForward Stepwise Results (%.1fs):\nFeatures selected: %d / %d\nFinal AIC: %.1f\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    length(fwd_selected), ncol(X_train_scaled),
    AIC(fwd_model),
    fwd_cm$overall["Accuracy"] * 100,
    fwd_cm$byClass["Sensitivity"] * 100,
    fwd_cm$byClass["Specificity"] * 100,
    auc(fwd_roc)))

cat("\nSelected features:\n")
cat(paste(fwd_selected, collapse = ", "), "\n")
```

```{r s3-forward-plot}
fwd_coef_df <- data.frame(
    feature     = fwd_selected,
    coefficient = coef(fwd_model)[fwd_selected]
) %>%
    arrange(desc(abs(coefficient)))

ggplot(fwd_coef_df,
       aes(x = reorder(feature, abs(coefficient)), y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("Negative", "Positive")) +
    labs(title = "Forward Stepwise - Selected Features and Coefficients",
         subtitle = sprintf("AIC-based selection: %d features out of %d",
                            length(fwd_selected), ncol(X_train_scaled)),
         x = NULL, y = "Coefficient", fill = "Direction") +
    theme_minimal()
```

Because features are standardized, coefficient magnitudes are directly comparable — the AIC criterion selects the same high-signal features we expect from the correlation heatmap.

## LASSO (L1)

LASSO (Least Absolute Shrinkage and Selection Operator) is logistic regression with an added penalty for large coefficients. Unlike standard logistic regression, which keeps all features, LASSO can shrink individual coefficients all the way to exactly zero - effectively removing those features from the model. This makes it both a classifier and a feature selector in one step. It is particularly useful here because our features have high multicollinearity: when two features carry the same signal (e.g. `golddiffat15` and `gold_momentum`), LASSO tends to keep the stronger one and zero out the weaker. We tune `lambda` - the penalty strength - via cross-validation: too low and nothing gets eliminated, too high and useful features get removed.

```{r s3-lasso}
cat("Running LASSO with 5-fold CV to find optimal lambda...\n")
t0 <- Sys.time()

lasso_cv <- cv.glmnet(
    x = as.matrix(X_train_scaled),
    y = y_train,
    alpha = 1, # alpha=1 is pure LASSO (L1 penalty)
    family = "binomial",
    nfolds = 5
)

# lambda.1se: the largest lambda within 1 SE of the minimum - gives a sparser model
# that is nearly as accurate as lambda.min but keeps fewer features
lasso_coefs <- coef(lasso_cv, s = "lambda.1se")

lasso_coef_df <- data.frame(
    feature = rownames(lasso_coefs),
    coefficient = as.numeric(lasso_coefs)
) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

lasso_selected <- lasso_coef_df %>% filter(coefficient != 0)
lasso_eliminated <- lasso_coef_df %>% filter(coefficient == 0)

# test set evaluation
lasso_pred_prob <- predict(lasso_cv,
    newx = as.matrix(X_test_scaled),
    s = "lambda.1se", type = "response"
)
lasso_pred_class <- factor(ifelse(lasso_pred_prob > 0.5, "Win", "Loss"),
    levels = c("Loss", "Win")
)
lasso_cm <- confusionMatrix(lasso_pred_class, y_test_fct)
lasso_roc <- roc(y_test_fct, as.numeric(lasso_pred_prob),
    levels = c("Loss", "Win"), quiet = TRUE
)

cat(sprintf(
    "\nLASSO Results (%.1fs):\nLambda (1se): %.6f\nFeatures kept: %d / %d (eliminated %d)\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    lasso_cv$lambda.1se,
    nrow(lasso_selected), nrow(lasso_coef_df), nrow(lasso_eliminated),
    lasso_cm$overall["Accuracy"] * 100,
    lasso_cm$byClass["Sensitivity"] * 100,
    lasso_cm$byClass["Specificity"] * 100,
    auc(lasso_roc)
))

cat("\nKept features:\n")
cat(paste(lasso_selected$feature, collapse = ", "), "\n")

cat("\nEliminated features:\n")
cat(paste(lasso_eliminated$feature, collapse = ", "), "\n")
```

```{r s3-lasso-plot}
plot(lasso_cv, main = "LASSO: CV Error vs Lambda")

lasso_selected %>%
    mutate(feature = reorder(feature, abs(coefficient))) %>%
    ggplot(aes(x = feature, y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title = "LASSO - Surviving Features and Coefficients",
        subtitle = sprintf(
            "lambda.1se: %d features kept out of %d",
            nrow(lasso_selected), nrow(lasso_coef_df)
        ),
        x = NULL, y = "Coefficient", fill = "Direction"
    ) +
    theme_minimal()
```

The CV plot shows how model error changes as lambda increases (fewer features). The left dashed line marks `lambda.min` (lowest error), the right marks `lambda.1se` (our chosen lambda - slightly sparser, nearly as accurate). The coefficient plot shows which features survived and in which direction: green bars increase win probability, red bars decrease it.

## Elastic Net (L1 + L2)

Elastic Net combines L1 (LASSO) and L2 (Ridge) penalties in a single term. The mix is controlled by `alpha`: `alpha = 1` is pure LASSO, `alpha = 0` is pure Ridge, values in between give Elastic Net. Where LASSO tends to pick one feature from a correlated group and zero out the rest, Elastic Net can keep several features from the same group at reduced coefficients — potentially more stable when features are strongly correlated, as ours are (e.g. `goldat15` and `golddiffat15`). We search the full alpha range from 0 (Ridge) to 1 (LASSO) in 0.05 steps, and for each find the optimal lambda via 5-fold CV. For every alpha we also record the CV binomial deviance at `lambda.1se` and its standard error, so we can judge whether the "best" alpha is meaningfully better than its neighbours or the curve is flat within 1 SE.

```{r s3-elasticnet}
cat("Running Elastic Net - searching alpha grid (0 to 1, step 0.05)...\n")
t0 <- Sys.time()

alpha_grid <- seq(0, 1, by = 0.05)

enet_cv_results <- map_dfr(alpha_grid, function(a) {
    cv_fit     <- cv.glmnet(
        x      = as.matrix(X_train_scaled),
        y      = y_train,
        alpha  = a,
        family = "binomial",
        nfolds = 5
    )
    idx_1se    <- which(cv_fit$lambda == cv_fit$lambda.1se)
    pred_prob  <- predict(cv_fit, newx = as.matrix(X_test_scaled),
                          s = "lambda.1se", type = "response")
    pred_class <- factor(ifelse(pred_prob > 0.5, "Win", "Loss"), levels = c("Loss", "Win"))
    cm         <- confusionMatrix(pred_class, y_test_fct)
    roc_obj    <- roc(y_test_fct, as.numeric(pred_prob), levels = c("Loss", "Win"), quiet = TRUE)
    n_kept     <- sum(coef(cv_fit, s = "lambda.1se") != 0) - 1
    tibble(
        alpha      = a,
        lambda_1se = cv_fit$lambda.1se,
        cv_dev     = cv_fit$cvm[idx_1se],
        cv_dev_se  = cv_fit$cvsd[idx_1se],
        n_features = n_kept,
        accuracy   = cm$overall["Accuracy"] * 100,
        auc        = as.numeric(auc(roc_obj))
    )
})

best_row   <- enet_cv_results %>% arrange(cv_dev) %>% dplyr::slice(1)
best_alpha <- best_row$alpha
best_dev   <- best_row$cv_dev
best_se    <- best_row$cv_dev_se
within_1se <- enet_cv_results %>% filter(cv_dev <= best_dev + best_se)

cat(sprintf("\nAlpha grid diagnostic: %d of %d alphas sit within 1 SE of the best (%.3f <= %.3f).\n",
            nrow(within_1se), nrow(enet_cv_results), best_dev, best_dev + best_se))
cat(sprintf("Alphas within 1 SE: %s\n",
            paste(sprintf("%.2f", within_1se$alpha), collapse = ", ")))
if (nrow(within_1se) >= nrow(enet_cv_results) - 2) {
    cat("-> CV curve is effectively flat across alpha; the L1/L2 mix is underdetermined by this data.\n")
}

enet_cv_final <- cv.glmnet(
    x      = as.matrix(X_train_scaled),
    y      = y_train,
    alpha  = best_alpha,
    family = "binomial",
    nfolds = 5
)

enet_coef_raw <- coef(enet_cv_final, s = "lambda.1se")
enet_coef_df  <- data.frame(
    feature     = rownames(enet_coef_raw),
    coefficient = as.numeric(enet_coef_raw)
) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

enet_selected_df <- enet_coef_df %>% filter(coefficient != 0)
enet_selected    <- enet_selected_df$feature

enet_pred_prob  <- predict(enet_cv_final, newx = as.matrix(X_test_scaled),
                            s = "lambda.1se", type = "response")
enet_pred_class <- factor(ifelse(enet_pred_prob > 0.5, "Win", "Loss"),
                           levels = c("Loss", "Win"))
enet_cm  <- confusionMatrix(enet_pred_class, y_test_fct)
enet_roc <- roc(y_test_fct, as.numeric(enet_pred_prob), levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("\nElastic Net Results (%.1fs):\nBest alpha: %.1f (0=Ridge, 1=LASSO)\nLambda (1se): %.6f\nFeatures kept: %d / %d (eliminated %d)\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    best_alpha,
    enet_cv_final$lambda.1se,
    nrow(enet_selected_df), nrow(enet_coef_df), nrow(enet_coef_df) - nrow(enet_selected_df),
    enet_cm$overall["Accuracy"] * 100,
    enet_cm$byClass["Sensitivity"] * 100,
    enet_cm$byClass["Specificity"] * 100,
    auc(enet_roc)))

cat("\nKept features:\n")
cat(paste(enet_selected, collapse = ", "), "\n")

cat("\nEliminated features:\n")
cat(paste(enet_coef_df %>% filter(coefficient == 0) %>% pull(feature), collapse = ", "), "\n")
```

```{r s3-elasticnet-plot}
ggplot(enet_cv_results, aes(x = alpha, y = cv_dev)) +
    geom_ribbon(aes(ymin = cv_dev - cv_dev_se, ymax = cv_dev + cv_dev_se),
                fill = "#1abc9c", alpha = 0.2) +
    geom_hline(yintercept = best_dev + best_se, linetype = "dashed", color = "#7f8c8d") +
    geom_line(color = "#1abc9c", linewidth = 1.2) +
    geom_point(size = 3, color = "#1abc9c") +
    geom_point(data = filter(enet_cv_results, alpha == best_alpha),
               size = 5, shape = 21, fill = "#e74c3c", color = "white", stroke = 1.5) +
    labs(title = "Elastic Net: CV Binomial Deviance across Alpha Values",
         subtitle = sprintf("Best alpha = %.2f (red dot); dashed line = best + 1 SE. Points below the line are statistically indistinguishable from the best.",
                            best_alpha),
         x = "Alpha (0 = Ridge, 1 = LASSO)", y = "5-fold CV Deviance at lambda.1se") +
    theme_minimal()

enet_selected_df %>%
    mutate(feature = reorder(feature, abs(coefficient))) %>%
    ggplot(aes(x = feature, y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("Negative", "Positive")) +
    labs(title = "Elastic Net - Surviving Features and Coefficients",
         subtitle = sprintf("alpha = %.1f, lambda.1se: %d features kept out of %d",
                            best_alpha, nrow(enet_selected_df), nrow(enet_coef_df)),
         x = NULL, y = "Coefficient", fill = "Direction") +
    theme_minimal()
```

The alpha search plot shows the 5-fold CV deviance at `lambda.1se` across the full 0–1 range, with the shaded band covering ±1 SE and the dashed line marking `best + 1 SE`. Every alpha whose point sits on or below that line is statistically indistinguishable from the best — so if many points fall below it, the L1/L2 mix is effectively underdetermined by this data and the "best alpha" is a tiebreak rather than a real optimum. A sharp U- or V-shape, by contrast, would indicate the data genuinely benefits from a specific mix. The coefficient plot mirrors the LASSO plot above; differences between the two reveal which features LASSO eliminated but Elastic Net retained due to correlated grouping.

## Random Forest Importance

The Random Forest trained in S1 generates feature importance scores as a by-product of training — no additional computation is needed. We use these scores as a filter: features with importance above the mean are retained, the rest discarded. This is the fastest and least tuning-intensive selection method: it piggybacks on the RF already trained for S1, requires no new cross-validation, and handles correlated features gracefully — RF distributes importance among correlated features rather than zeroing all but one like LASSO. To get a comparable performance estimate, we train a simple LR on the selected subset and evaluate it on the held-out test set.

```{r s3-rf-importance}
rf_importance_threshold <- mean(rf_imp$Overall)
rf_s3_selected <- rf_imp %>%
    filter(Overall > rf_importance_threshold) %>%
    pull(feature)

cat(sprintf("RF importance threshold: %.2f (mean importance)\nFeatures selected: %d / %d\n",
    rf_importance_threshold, length(rf_s3_selected), nrow(rf_imp)))

rf_s3_lr <- train(
    x         = X_train_scaled[, rf_s3_selected, drop = FALSE],
    y         = y_train_fct,
    method    = "glm",
    family    = "binomial",
    trControl = trainControl(method = "none", classProbs = TRUE)
)
rf_s3_pred_class <- predict(rf_s3_lr, X_test_scaled[, rf_s3_selected, drop = FALSE])
rf_s3_pred_prob  <- predict(rf_s3_lr, X_test_scaled[, rf_s3_selected, drop = FALSE], type = "prob")
rf_s3_cm  <- confusionMatrix(rf_s3_pred_class, y_test_fct)
rf_s3_roc <- roc(y_test_fct, rf_s3_pred_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("LR on RF-selected features:\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    rf_s3_cm$overall["Accuracy"] * 100,
    rf_s3_cm$byClass["Sensitivity"] * 100,
    rf_s3_cm$byClass["Specificity"] * 100,
    auc(rf_s3_roc)))

cat("\nSelected features:\n")
cat(paste(rf_s3_selected, collapse = ", "), "\n")
```

```{r s3-rf-importance-plot}
rf_imp %>%
    mutate(
        selected = Overall > rf_importance_threshold,
        feature  = reorder(feature, Overall)
    ) %>%
    ggplot(aes(x = feature, y = Overall, fill = selected)) +
    geom_col() +
    coord_flip() +
    geom_hline(yintercept = rf_importance_threshold, linetype = "dashed",
               color = "#e74c3c", linewidth = 0.8) +
    scale_fill_manual(values = c("FALSE" = "#bdc3c7", "TRUE" = "#27ae60"),
                      labels = c("Eliminated", "Selected")) +
    labs(title = "RF Importance - Feature Selection Threshold",
         subtitle = sprintf("Threshold = mean importance (%.2f) — %d features selected out of %d",
                            rf_importance_threshold, length(rf_s3_selected), nrow(rf_imp)),
         x = NULL, y = "Importance", fill = "Status") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 7))
```

The dashed red line marks the mean importance threshold. Features above it (green) are retained; those below (grey) are discarded. Note that RF distributes importance among correlated features — so unlike LASSO, several correlated gold metrics may all survive with reduced but non-zero importance.

## Feature selection comparison

```{r s3-comparison}
selection_results <- list(
    RFE        = rfe_selected,
    Forward    = fwd_selected,
    LASSO      = lasso_selected$feature,
    ElasticNet = enet_selected,
    RF_Imp     = rf_s3_selected
)

all_features <- colnames(X_train_scaled)

# For each feature, record which methods selected it
overlap_df <- map_dfr(all_features, function(f) {
    sel <- map_lgl(selection_results, ~ f %in% .x)
    tibble(
        feature    = f,
        RFE        = sel[["RFE"]],
        Forward    = sel[["Forward"]],
        LASSO      = sel[["LASSO"]],
        ElasticNet = sel[["ElasticNet"]],
        RF_Imp     = sel[["RF_Imp"]],
        n_methods  = sum(sel)
    )
}) %>%
    arrange(desc(n_methods), feature)

consensus_5 <- overlap_df %>% filter(n_methods == 5)
consensus_4 <- overlap_df %>% filter(n_methods == 4)
never_sel   <- overlap_df %>% filter(n_methods == 0)

cat("Feature selection summary across all 5 methods:\n")
cat(sprintf("  Total features:                %d\n", length(all_features)))
cat(sprintf("  Selected by all 5 (consensus): %d\n", nrow(consensus_5)))
cat(sprintf("  Selected by 4 methods:         %d\n", nrow(consensus_4)))
cat(sprintf("  Never selected:                %d\n\n", nrow(never_sel)))

cat("Consensus features (all 5 methods agree):\n")
if (nrow(consensus_5) > 0) {
    cat(paste(consensus_5$feature, collapse = ", "), "\n")
} else {
    cat("None\n")
}

cat("\nFeatures selected by 4 methods:\n")
if (nrow(consensus_4) > 0) {
    cat(paste(consensus_4$feature, collapse = ", "), "\n")
} else {
    cat("None\n")
}

cat("\nFeatures never selected by any method:\n")
if (nrow(never_sel) > 0) {
    cat(paste(never_sel$feature, collapse = ", "), "\n")
} else {
    cat("None — every feature was selected by at least one method\n")
}

# Performance comparison table
perf_summary <- tibble(
    Method     = c("RFE", "Forward", "LASSO", "ElasticNet", "RF_Imp", "LR_all"),
    n_features = c(
        length(rfe_selected),   length(fwd_selected),
        nrow(lasso_selected),   nrow(enet_selected_df),
        length(rf_s3_selected), ncol(X_train_scaled)
    ),
    Accuracy   = c(
        rfe_cm$overall["Accuracy"] * 100,
        fwd_cm$overall["Accuracy"] * 100,
        lasso_cm$overall["Accuracy"] * 100,
        enet_cm$overall["Accuracy"] * 100,
        rf_s3_cm$overall["Accuracy"] * 100,
        lr_cm$overall["Accuracy"] * 100
    ),
    AUC_ROC    = c(
        auc(rfe_roc), auc(fwd_roc),   auc(lasso_roc),
        auc(enet_roc), auc(rf_s3_roc), auc(roc_lr)
    ) * 100
)

cat("\nPerformance comparison (sorted by AUC-ROC):\n")
print(perf_summary %>% arrange(desc(AUC_ROC)), n = Inf)
```

```{r s3-comparison-plots, fig.height=8}
# Heatmap: which method selected which feature
overlap_long <- overlap_df %>%
    filter(n_methods > 0) %>%
    pivot_longer(
        cols      = c(RFE, Forward, LASSO, ElasticNet, RF_Imp),
        names_to  = "method",
        values_to = "selected"
    ) %>%
    mutate(
        feature_label = paste0(feature, "  (", n_methods, "/5)"),
        feature_label = reorder(feature_label, n_methods),
        method        = factor(method, levels = c("RFE", "Forward", "LASSO", "ElasticNet", "RF_Imp"))
    )

ggplot(overlap_long, aes(x = method, y = feature_label, fill = selected)) +
    geom_tile(color = "grey85", linewidth = 0.5) +
    
    scale_fill_manual(
        values = c("TRUE" = "#27ae60", "FALSE" = "white"),
        labels = c("Not selected", "Selected"),
        guide  = "none"
    ) +
    labs(
        title    = "Feature Selection Overlap Across All Methods",
        subtitle = "Each row shows which methods selected that feature  ·  (x/5) = number of methods",
        x = NULL, y = NULL
    ) +
    theme_minimal() +
    theme(
        axis.text.y      = element_text(size = 8),
        axis.text.x      = element_text(size = 9, face = "bold"),
        panel.grid       = element_blank(),
        panel.border     = element_rect(color = "grey85", fill = NA)
    )
```

```{r s3-comparison-auc}
baseline_auc <- perf_summary$AUC_ROC[perf_summary$Method == "LR_all"]

perf_summary %>%
    filter(Method != "LR_all") %>%
    mutate(Method = reorder(Method, AUC_ROC)) %>%
    ggplot(aes(x = AUC_ROC, y = Method)) +
    geom_segment(aes(x = baseline_auc - 1.5, xend = AUC_ROC, yend = Method),
                 color = "grey75", linewidth = 1) +
    geom_point(size = 5, color = "#2ecc71") +
    geom_text(aes(label = sprintf("%d features", n_features)),
              nudge_x = 0.15, hjust = 0, size = 3.5, color = "grey30") +
    geom_vline(xintercept = baseline_auc, linetype = "dashed",
               color = "#e74c3c", linewidth = 0.8) +
    annotate("text", x = baseline_auc + 0.05, y = 0.6,
             label = sprintf("Baseline\n(all %d feat.)", ncol(X_train_scaled)),
             color = "#e74c3c", size = 3, hjust = 0) +
    scale_x_continuous(limits = c(baseline_auc - 1.5, baseline_auc + 1.5)) +
    labs(
        title   = "Feature Selection: AUC-ROC vs. Method",
        subtitle = "Red dashed line = LR trained on all features  ·  Labels show number of features used",
        x = "AUC-ROC (%)", y = NULL
    ) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
```

Každý riadok = jedna metóda. Dĺžka stĺpca = AUC-ROC, číslo v texte hovorí koľko features použila. Prerušovaná čiara = baseline (všetky features). LASSO a RFE sú tesne pod čiarou s 9 features — prakticky rovnaký výkon pri polovici features.

## S3 findings

All five selection methods converge on the same answer: **the predictive signal in this dataset is highly concentrated**. Starting from 21 features, every method reduces the set substantially — LASSO and RF Importance down to just 9, RFE to 10 — with almost no loss in AUC-ROC (83.2–84.4% vs. 84.4% for the full-feature baseline). This means roughly half the features in the model are redundant: they carry signal already captured by other features, or no signal at all.

**Four features form the consensus core** — selected by all five methods regardless of the criterion used: `golddiffat15`, `xpdiffat15`, `csdiffat15`, and `winrate_diff`. These are the three resource differentials at 15 minutes plus the pre-game team strength gap. Their universal selection confirms what the EDA correlation heatmap showed: who is ahead in gold, XP, and CS at 15 minutes is the strongest combined signal for match outcome. The `winrate_diff` feature indicates that pre-game team quality adds real information on top of the in-game state.

**Four more features reach near-consensus** (4 out of 5 methods): `firsttower`, `firstherald`, `firstdragon`, and `cs_momentum`. Objective control — particularly tower and herald — contributes a second, independent layer of signal beyond raw economic leads. `cs_momentum` (the acceleration of the CS advantage from @10 to @15) captures information not fully redundant with the snapshot differential at @15.

**Three features were never selected** by any method: `firstblood`, `kill_pressure_10_15`, and `opp_kill_pressure_10_15`. This is a meaningful null result — it suggests that early kill activity, while visible in the data, does not independently predict match outcome once the economic and objective features are accounted for. First blood in particular is consistent with published LoL research, where its win-rate correlation is weak relative to tower or gold leads.

The key practical takeaway is that **9 features achieve 84.3% AUC-ROC** — within 0.1 points of the full-feature LR — which means a model trained on only the consensus + near-consensus features is as useful as one trained on everything, while being substantially easier to interpret and explain.

# Conclusions

## H1 findings

<!-- TODO: summarize early-game prediction results -->

## H2 findings

<!-- TODO: summarize bot lane impact findings -->

## S1 vs S3 synthesis

<!-- TODO: which models + features work best together -->

# Export models for Shiny app

```{r save-models}
dir.create("shiny/models", recursive = TRUE, showWarnings = FALSE)

saveRDS(lr_model,      "shiny/models/lr_model.rds")
saveRDS(rf_model,      "shiny/models/rf_model.rds")
saveRDS(nb_model,      "shiny/models/nb_model.rds")
saveRDS(knn_model,     "shiny/models/knn_model.rds")
saveRDS(cart_model,    "shiny/models/cart_model.rds")
saveRDS(preproc,       "shiny/models/preproc.rds")
saveRDS(perf_summary,  "shiny/models/perf_summary.rds")
saveRDS(overlap_df,    "shiny/models/overlap_df.rds")
saveRDS(games_model,   "shiny/models/games_model.rds")
saveRDS(X_train,       "shiny/models/X_train.rds")

# Game lookup table for Shiny browser.
# The Shiny app's game-picker is per-side (one Blue and one Red row per game),
# so we un-pivot games_model back into a two-rows-per-game shape. Signed diffs
# are negated for the Red row, per-team kill counts are swapped, and binary
# flags are toggled (when Blue secured the objective, Red did not; otherwise
# we keep the same value, since neither team is recorded as having secured it).
shiny_blue <- games_model %>%
    mutate(side_encoded = 1L, result = blue_win) %>%
    select(-blue_win)

shiny_red <- games_model %>%
    transmute(
        gameid,
        side_encoded = 0L,
        result       = 1L - blue_win,
        golddiffat15 = -golddiffat15,
        xpdiffat15   = -xpdiffat15,
        csdiffat15   = -csdiffat15,
        gold_momentum  = -gold_momentum,
        xp_momentum    = -xp_momentum,
        cs_momentum    = -cs_momentum,
        kill_diff_15   = -kill_diff_15,
        death_diff_15  = -death_diff_15,
        assist_diff_15 = -assist_diff_15,
        kill_pressure_10_15     = opp_kill_pressure_10_15,
        opp_kill_pressure_10_15 = kill_pressure_10_15,
        gold_efficiency_15 = NA_real_,  # Blue-only ratio, no Red counterpart kept
        kda_15             = NA_real_,
        plate_diff   = -plate_diff,
        grub_diff    = -grub_diff,
        winrate_diff = -winrate_diff,
        firstblood   = ifelse(firstblood   == 1L, 0L, firstblood),
        firstdragon  = ifelse(firstdragon  == 1L, 0L, firstdragon),
        firstherald  = ifelse(firstherald  == 1L, 0L, firstherald),
        firsttower   = ifelse(firsttower   == 1L, 0L, firsttower)
    )

per_side_features <- bind_rows(shiny_blue, shiny_red)

# Per-side team labels
team_labels <- teams %>%
    select(gameid, league, date, side, teamname) %>%
    mutate(side_encoded = ifelse(side == "Blue", 1L, 0L),
           side_label   = side) %>%
    select(gameid, league, date, teamname, side_label, side_encoded)

opp_labels <- team_labels %>%
    transmute(gameid,
              side_encoded_opp = 1L - side_encoded,
              opp_teamname     = teamname)

game_lookup <- team_labels %>%
    left_join(opp_labels, by = c("gameid", "side_encoded" = "side_encoded_opp")) %>%
    inner_join(per_side_features, by = c("gameid", "side_encoded")) %>%
    mutate(date         = as.Date(date),
           result_label = ifelse(result == 1, "WIN", "LOSS"))

saveRDS(game_lookup, "shiny/models/game_lookup.rds")

# ROC objects for Shiny Model Comparison tab
roc_lr   <- pROC::roc(y_test_fct, lr_prob[,   "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_rf   <- pROC::roc(y_test_fct, rf_prob[,   "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_nb   <- pROC::roc(y_test_fct, nb_prob[,   "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_knn  <- pROC::roc(y_test_fct, knn_prob[,  "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_cart <- pROC::roc(y_test_fct, cart_prob[, "Win"], levels = c("Loss","Win"), quiet = TRUE)
saveRDS(list(lr = roc_lr, rf = roc_rf, nb = roc_nb, knn = roc_knn, cart = roc_cart),
        "shiny/models/roc_list.rds")

# LR test predictions for Shiny (confusion matrix + prob distribution)
saveRDS(
    data.frame(actual = as.character(y_test_fct), prob_win = lr_prob[, "Win"]),
    "shiny/models/lr_test_df.rds"
)
saveRDS(lr_cm, "shiny/models/lr_cm.rds")

cat("Models saved to shiny/models/\n")
```
