---
title: "OZNAL Capstone - LoL Esports Early-Game Prediction"
author: "Martin Vanco, Adriana Gemeľová"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width = 10,
    fig.height = 6
)
```

# Introduction

League of Legends is the most-played and most-watched esport in the world. Predicting which team will win from the early state of the match is a long-standing question in esports analytics: by 15 minutes - well before half of a typical game has elapsed - gold, experience, and objective leads have already begun to compound. We use the 2025 season of professional play to test how reliably this 15-minute snapshot forecasts the eventual winner, and which features actually drive the prediction.

The analysis runs two scenarios on the same data:

-   **Scenario 1 - Model Comparison.** We train and tune five classifiers (Logistic Regression, Random Forest, Naive Bayes, KNN, CART) on the engineered early-game feature set and compare them on accuracy, AUC-ROC, precision, recall, and F1. The goal is to see whether any model can extract signal beyond the trivial "team with the gold lead at 15 min wins" baseline (\~73% accuracy), and whether different model families agree on what matters.
-   **Scenario 3 - Feature Selection.** With 19 engineered candidate features carrying meaningful redundancy, we apply RFE, LASSO, and Elastic Net to find the minimal subset that retains predictive performance. This both improves interpretability and stress-tests which features are genuinely informative versus carrying duplicate signal.

## Hypotheses

**H1:** *"Early-game performance metrics and objective control can reliably predict match outcome in professional League of Legends."* We test whether gold, XP, and CS differentials at 10 and 15 minutes, combined with first objective captures (first blood, first dragon, first tower, void grubs), contain enough signal to classify wins and losses.

**H2:** *"Bot lane (ADC) gold advantage is the most impactful role-specific early-game factor for predicting match outcome."* We test whether the bot lane player's early-game differentials carry stronger predictive signal than those of other roles (top, jungle, mid, support).

## Motivation

Both of us on this project are long-time League of Legends players and regular viewers of the professional scene. Watching pro matches, you very quickly develop a feel for when a game is "already over" - a significant gold lead at early game mostly means a quick and decisive game. We had been wondering for a while how much of that intuition is actually backed by data: how reliably do those early signals predict the eventual winner and which signals matter the most.

# Dataset

We use the 2025 season export from **Oracle's Elixir**, the standard public source for professional League of Legends match data. The raw file contains **120,636 rows across 165 columns**, covering **10,053 unique matches** played by 445 teams across 45 professional leagues (LPL, LCK, LEC, LFL, LAS, EM, ...) on patches 15.01-15.24.

The row layout is **12 rows per game**: one per player (5 per team) plus one aggregated team-level row per side (Blue and Red, 50/50). About **91.9% of rows are marked `complete`**, meaning the early-game snapshots at 10/15/20/25 min are populated; the remaining 8.1% are missing these values.

Columns group into metadata (`gameid`, `league`, `patch`, `side`, `position`, ...), draft (`pick1-5`, `ban1-5`), outcome (`result`, `gamelength`), combat (kills/deaths/assists, first blood), objectives (dragons, heralds, void grubs, towers, turret plates), economy (gold, CS), vision (wards, vision score), and the **early-game timeline snapshots** (`goldatX`, `xpatX`, `csatX`, `golddiffatX`, etc. for X in {10, 15, 20, 25}) which form the core of our feature set.

# Game introduction

League of Legends (LoL) is a 5v5 team game. Two teams (Blue and Red side) try to destroy each other's base. A single match is typically around 30 minutes but can last as long as 50 minutes. In this work, we focus on the **early game** stage (the first 15 minutes). Each team has five fixed roles - **top**, **jungle (jng)**, **mid**, **bot (ADC)**, and **support (sup)** - each playing a different lane or area of the map.

Three continuous resources accumulate throughout the match:

-   **Gold** - earned from minions, champion kills, and objectives, and spent on items that strengthen champions.
-   **XP (experience)** - levels champions up and unlocks abilities.
-   **CS (creep score)** - the number of minions a player has killed

The early game also has a set of one-off objective races, each won by exactly one team: **first blood** (the first champion kill of the match), **first dragon**, **first herald**, **first tower** (each granting a stacking buff, a tower-damaging summon, or a large gold bonus), **turret plates** (five plates on each outer tower, all paid out as gold before 14:00), and **void grubs** (a shared pit of up to 6 grubs that despawns at 13:45 on early-2025 patches and 14:45 mid-season).

# Related work

Published work on early-game LoL match prediction at the **15-minute mark** converges tightly, with reported accuracy clustering in the **76-78% band** across model families and dataset types.

| Study                                        | Year | Dataset                      | Model               | Acc. \@15  |
|----------------------------------------------|------|------------------------------|---------------------|------------|
| Spaargaren, MSc thesis (Tilburg)             | 2022 | Oracle's Elixir pro, 2021    | Logistic Regression | **77.99%** |
| Spaargaren                                   | 2022 | Oracle's Elixir pro, 2021    | Random Forest       | 77.26%     |
| Spaargaren                                   | 2022 | Oracle's Elixir pro, 2021    | Naïve Bayes         | 76.82%     |
| Spaargaren                                   | 2022 | Oracle's Elixir pro, 2021    | kNN (k=9)           | 74.83%     |
| Lafrance & Grewal, *Entertainment Computing* | 2026 | Riot API solo queue, Diamond | XGBoost             | 76.9%      |
| Junior & Campelo, arXiv 2309.02449           | 2023 | Riot API solo queue          | Random Forest       | \~74%\*    |

\*Interpolated to 50% PET (≈ 15 min for a \~30-min median match) from Junior & Campelo's PET-bucket accuracy curves.

Two takeaways. First, **professional play is not materially easier to predict than solo queue** at 15 minutes - Spaargaren's pro-data LR sits within one percentage point of Lafrance & Grewal's solo-queue XGBoost. Second, **algorithm choice matters little**: pro-data classifiers all land in the 77 ± 1% band, with solo-queue \~3 points lower. Across every study, **gold differential is the strongest single feature**, with first tower or first herald typically ranked next.

Tsang (2025, Medium) adds a temporal signal: the coefficient of `golddiffat15` has **declined** from above 1.0 in 2020-2023 to 0.83 in 2025, suggesting the 2025 season is harder to predict from a single gold-at-15 signal than any previous season. We additionally include two features absent from the @15-minute literature - **turret plate differential** and **void grubs (introduced in 2025 season)**.

Finally, we note a framing gap: the trivial rule "team with a gold lead at 15 min wins" already delivers \~73% accuracy on professional data (we compute the exact figure on our own set in the EDA), so the 76-78% models above only beat this naive baseline by 3-5 points. We adopt this framing directly in our results.

# Setup

```{r libraries}
library(tidyverse)
library(corrplot)
library(caret)
library(glmnet)
library(randomForest)
library(naivebayes)
library(rpart)
library(pROC)
library(gridExtra)
library(doParallel)

cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)
```

## Load data

```{r load-data}
raw <- read_csv("data/2025_LoL_esports_match_data_from_OraclesElixir.csv")
glimpse(raw)
```

```{r}
# select only complete data - with @10, 15, 20, 25 metrics
complete <- raw %>% filter(datacompleteness == "complete")

teams <- complete %>% filter(position == "team")
players <- complete %>% filter(position != "team")

cat(sprintf(
    "Total rows: %d\nFiltered to complete: %d (removed %d incomplete, %.1f%%)\nTeam rows: %d\nPlayer rows: %d\nTarget balance: %.1f%% win / %.1f%% loss",
    nrow(raw), nrow(complete),
    nrow(raw) - nrow(complete),
    (nrow(raw) - nrow(complete)) / nrow(raw) * 100,
    nrow(teams), nrow(players),
    mean(teams$result == 1) * 100, mean(teams$result == 0) * 100
))
```

## Restructure: one row per game

The raw `teams` table stores two mirror rows per game, one per side. For EDA we pivot to one row per game with `blue_*` and `red_*` columns and a binary `blue_win` target, following the **tidy data** convention of one observation per row (Wickham, 2014, *JSS* 59(10)).

After the pivot, the Blue and Red copies of each signed differential and each mutually-exclusive binary flag are pure mirrors: `red_golddiffat15 = -blue_golddiffat15` and `red_firstblood = 1 - blue_firstblood`. We drop the `red_*` side of those columns, since the model gains nothing from a feature it can recover from the one next to it. Independent counts (kills, deaths, assists, turret plates) and the shared-pool grub count stay on both sides - neither side can be reconstructed from the other.

```{r restructure-one-row-per-game}
game_cols <- c(
    "result",
    "goldat10", "xpat10", "csat10",
    "goldat15", "xpat15", "csat15",
    "golddiffat10", "xpdiffat10", "csdiffat10",
    "golddiffat15", "xpdiffat15", "csdiffat15",
    "killsat10", "deathsat10", "assistsat10",
    "killsat15", "deathsat15", "assistsat15",
    "firstblood", "firstdragon", "firstherald", "firsttower",
    "void_grubs", "turretplates"
)

games_full <- teams %>%
    select(gameid, side, all_of(game_cols)) %>%
    mutate(side = tolower(side)) %>%
    pivot_wider(
        id_cols     = gameid,
        names_from  = side,
        values_from = all_of(game_cols),
        names_glue  = "{side}_{.value}"
    ) %>%
    mutate(blue_win = blue_result) %>%
    select(-blue_result, -red_result)

# Sanity: blue/red signed diffs must be exact negatives of each other,
# and mutually exclusive binary flags must satisfy red = 1 - blue whenever
# the objective fires.
stopifnot(all.equal(games_full$blue_golddiffat15, -games_full$red_golddiffat15))
stopifnot(all.equal(games_full$blue_xpdiffat15,   -games_full$red_xpdiffat15))
stopifnot(all.equal(games_full$blue_csdiffat15,   -games_full$red_csdiffat15))
stopifnot(all(games_full$blue_firstblood   + games_full$red_firstblood   <= 1))
stopifnot(all(games_full$blue_firsttower   + games_full$red_firsttower   <= 1))

# Drop the strict-mirror columns identified above. We keep both sides for
# independent counts (kills/deaths/assists/turret plates) and for the
# shared-pool grub count, since those carry information beyond their pair.
redundant_mirrors <- c(
    "red_golddiffat10", "red_xpdiffat10", "red_csdiffat10",
    "red_golddiffat15", "red_xpdiffat15", "red_csdiffat15",
    "red_firstblood", "red_firstdragon", "red_firstherald", "red_firsttower"
)

games <- games_full %>% select(-all_of(redundant_mirrors))

cat(sprintf(
    "Games (one row per game): %d\nFeatures dropped as strict mirrors: %d\nFeatures retained: %d\nBlue win rate: %.2f%% (Red: %.2f%%)\n",
    nrow(games),
    length(redundant_mirrors),
    ncol(games) - 2, # excluding gameid and blue_win
    mean(games$blue_win) * 100,
    mean(1 - games$blue_win) * 100
))

cat("\nDropped features (recoverable as -blue_X or 1-blue_X):\n  ",
    paste(redundant_mirrors, collapse = ", "), "\n")

# Statistical test for side imbalance
side_test <- prop.test(sum(games$blue_win), nrow(games), p = 0.5)
cat(sprintf(
    "\nTwo-sided proportion test vs 50%%:\n  chi^2 = %.2f, df = 1, p = %.3g\n  95%% CI for Blue WR: [%.3f, %.3f]\n  -> Blue-side advantage is %s.\n",
    side_test$statistic, side_test$p.value,
    side_test$conf.int[1], side_test$conf.int[2],
    ifelse(side_test$p.value < 0.05, "statistically significant", "not statistically significant")
))
```

Blue side wins slightly more than half of games. The imbalance is mild enough that no resampling is needed and accuracy remains valid alongside AUC-ROC.

# Exploratory Data Analysis

Before deciding on feature engineering or models, we look at the data to see what signal is there, which features are worth keeping, and what problems to watch for. Everything below uses the **single-row-per-game** `games` dataset.

## Target distribution

```{r eda-target}
games %>%
    mutate(blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won"))) %>%
    ggplot(aes(x = blue_win, fill = blue_win)) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
    scale_fill_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_hline(yintercept = nrow(games) / 2, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Target Variable Distribution (blue_win)",
        x = "Outcome", y = "Count"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

Blue side wins slightly more than half of all games. The proportion test above puts this at p \< 0.001, so the imbalance is real, but at roughly 3 percentage points above 50% it is small enough that we keep accuracy as a valid metric and skip resampling.

## Descriptive statistics

Before correlations and modelling, we look at the central tendency and spread of every early-game numeric we plan to feed into the models. The columns are sorted by standard deviation - the loudest features will dominate any unscaled distance-based model, which motivates the centering and scaling done later in the modelling pipeline.

```{r eda-describe}
early_game_cols <- c(
    "blue_goldat10", "blue_xpat10", "blue_csat10",
    "blue_golddiffat10", "blue_xpdiffat10", "blue_csdiffat10",
    "blue_killsat10", "blue_deathsat10", "blue_assistsat10",
    "blue_goldat15", "blue_xpat15", "blue_csat15",
    "blue_golddiffat15", "blue_xpdiffat15", "blue_csdiffat15",
    "blue_killsat15", "blue_deathsat15", "blue_assistsat15",
    "blue_firstblood", "blue_firstdragon", "blue_firstherald", "blue_firsttower",
    "blue_void_grubs", "red_void_grubs",
    "blue_win"
)

games %>%
    select(all_of(setdiff(early_game_cols, "blue_win"))) %>%
    pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
    drop_na() %>%
    group_by(feature) %>%
    summarise(
        n      = n(),
        mean   = mean(value),
        sd     = sd(value),
        min    = min(value),
        q25    = quantile(value, 0.25),
        median = median(value),
        q75    = quantile(value, 0.75),
        max    = max(value),
        .groups = "drop"
    ) %>%
    mutate(across(c(mean, sd, min, q25, median, q75, max), ~ round(.x, 2))) %>%
    arrange(desc(sd)) %>%
    print(n = Inf)
```

The differentials sit at the top of the spread ranking with standard deviations in the thousands of gold and XP, while the absolute totals at 15 minutes cluster tightly around 25k gold and 30k XP. CS columns are an order of magnitude smaller, kill/death/assist counts smaller still, and the 0/1-coded objectives sit at the bottom with sd around 0.5. On the original scale the objectives would be invisible next to gold and XP, which is why the modelling pipeline centers and scales every input before training Logistic Regression and KNN. The differential means hover near zero (243 gold and 49 XP at 15) - close enough to confirm the features behave like signed deltas rather than biased counters, but slightly positive in line with the Blue-side advantage seen earlier.

## Early-game feature correlations with `blue_win`

We correlate each early-game numeric feature (Blue's perspective) with the target.

```{r eda-correlation, fig.width=12, fig.height=10}
cor_data <- games %>%
    select(all_of(early_game_cols)) %>%
    drop_na()

cor_matrix <- cor(cor_data)

corrplot(cor_matrix,
    method   = "color", type = "lower",
    tl.cex   = 0.7, tl.col = "black",
    col      = colorRampPalette(c("#e74c3c", "white", "#2ecc71"))(200),
    col.lim  = c(-1, 1),
    title    = "Correlation Matrix - Early-Game Features (Blue's perspective)",
    mar      = c(0, 0, 2, 0)
)
```

```{r eda-correlation-with-result}
result_cors <- cor_data %>%
    select(-blue_win) %>%
    summarise(across(everything(), ~ cor(.x, cor_data$blue_win))) %>%
    pivot_longer(everything(), names_to = "feature", values_to = "correlation") %>%
    arrange(desc(abs(correlation)))

ggplot(result_cors, aes(
    x    = reorder(feature, abs(correlation)),
    y    = correlation,
    fill = correlation > 0
)) +
    geom_col() +
    coord_flip(ylim = c(-1, 1)) +
    scale_y_continuous(breaks = seq(-1, 1, 0.25)) +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title    = "Feature Correlation with blue_win",
        x = NULL, y = "Pearson Correlation", fill = "Direction"
    ) +
    theme_minimal()
```

```{r}
cat(sprintf(
    "Strongest positive: %s (%.3f)\nStrongest negative: %s (%.3f)",
    result_cors$feature[which.max(result_cors$correlation)],
    max(result_cors$correlation),
    result_cors$feature[which.min(result_cors$correlation)],
    min(result_cors$correlation)
))
```

Differentials (`blue_golddiffat15`, `blue_xpdiffat15`, `blue_csdiffat15`) dominate the ranking, since what matters is the *gap* between teams rather than the absolute totals. Among objectives, `blue_firsttower` and `blue_firstherald` stand out; `blue_firstblood` and `blue_firstdragon` carry less signal. The matrix also shows two flavours of multicollinearity: between a raw stat and its differential (`blue_goldat15` vs `blue_golddiffat15`), and between Blue and Red counts of the same shared-pool objective (`blue_void_grubs` vs `red_void_grubs`, strongly negatively correlated because they share a 6-grub pit). Both forms hurt Logistic Regression and **motivate the LASSO and Elastic Net regularization used in Scenario 3**.

## Per-outcome means

The correlation chart ranks features by linear association with `blue_win`. Putting the same features through a means-by-outcome lens shows the gap in the original units rather than on the [-1, 1] scale, which is easier to read for objectives and combat counts.

```{r eda-by-outcome}
games %>%
    select(all_of(early_game_cols)) %>%
    drop_na() %>%
    mutate(outcome = if_else(blue_win == 1, "blue_win", "red_win")) %>%
    select(-blue_win) %>%
    pivot_longer(-outcome, names_to = "feature", values_to = "value") %>%
    group_by(feature, outcome) %>%
    summarise(mean = mean(value), .groups = "drop") %>%
    pivot_wider(names_from = outcome, values_from = mean) %>%
    mutate(diff = blue_win - red_win) %>%
    arrange(desc(abs(diff))) %>%
    mutate(across(c(blue_win, red_win, diff), ~ round(.x, 3))) %>%
    print(n = Inf)
```

The biggest absolute gaps fall on the gold and XP differentials at 15 minutes - Blue wins lead by \~1500 gold and \~940 XP on average; losses trail by similar amounts. Tower-related objectives produce the largest split among the 0/1 features: `blue_firsttower` jumps from 0.34 in losses to 0.73 in wins, `blue_firstherald` from 0.44 to 0.75. First blood and first dragon move the means by less than 0.17, matching their weaker correlations. The signs in the diff column also confirm direction: `blue_deathsat15` is the only count where the win-mean is *lower* than the loss-mean (3.6 vs 5.2), and `red_void_grubs` flips negative because more grubs to the enemy means a worse outcome for Blue.

## Gold differential distribution

Gold differential at 15 minutes is the single strongest early-game predictor. We look at how cleanly it separates Blue wins from Red wins, and whether a simple linear cut (positive gold diff -\> Blue win) would already do most of the work or whether we need non-linear models.

```{r eda-golddiff}
games %>%
    filter(!is.na(blue_golddiffat15)) %>%
    mutate(blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won"))) %>%
    ggplot(aes(x = blue_golddiffat15, fill = blue_win)) +
    geom_density(alpha = 0.5) +
    scale_fill_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Gold Differential at 15 Minutes - Blue's Perspective",
        x = "blue_golddiffat15", y = "Density", fill = "Outcome"
    ) +
    theme_minimal()
```

```{r eda-golddiff-10vs15}
games %>%
    filter(!is.na(blue_golddiffat10), !is.na(blue_golddiffat15)) %>%
    mutate(blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won"))) %>%
    ggplot(aes(x = blue_golddiffat10, y = blue_golddiffat15, color = blue_win)) +
    geom_point(alpha = 0.15, size = 0.8) +
    scale_color_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Blue Gold Diff @10 vs @15 - Do Leads Grow?",
        x = "blue_golddiffat10", y = "blue_golddiffat15", color = "Outcome"
    ) +
    theme_minimal()
```

To quantify how much of the outcome the gold differential at 15 min already explains, we compute the accuracy of the naive rule "Blue ahead in gold @15 -\> Blue wins" directly on `games`. This number then defines the baseline that any trained model has to beat.

```{r eda-golddiff-baseline}
gold_baseline_df <- games %>%
    filter(!is.na(blue_golddiffat15)) %>%
    transmute(
        pred_blue_win = blue_golddiffat15 > 0,
        actual        = blue_win == 1
    )

gold_baseline_acc  <- mean(gold_baseline_df$pred_blue_win == gold_baseline_df$actual)
gold_baseline_n    <- nrow(gold_baseline_df)
gold_baseline_ties <- sum(games$blue_golddiffat15 == 0, na.rm = TRUE)
blue_base_rate     <- mean(games$blue_win)

cat(sprintf(
    "Naive rule 'Blue ahead in gold @15 -> Blue wins':\n  N = %d games (%d games tied at exactly 0 gold diff)\n  Accuracy: %.2f%%\n  Always-predict-Blue baseline: %.2f%%\n  Rule beats always-Blue by: %.2f pp",
    gold_baseline_n, gold_baseline_ties,
    gold_baseline_acc * 100,
    blue_base_rate * 100,
    (gold_baseline_acc - blue_base_rate) * 100
))
```

The two densities are clearly separated. The naive rule reaches 72.9% accuracy at the game level - well above both the 50% coin-flip and the 53.3% always-predict-Blue baseline - and any model has to beat this number to be worth keeping. The wide overlap around zero says a meaningful share of games are not decided by gold alone.

## First objective win rates

The four "first" objectives (first blood, first dragon, first herald, first tower) all happen in the first \~14 minutes and are mutually exclusive: exactly one team secures each. For every objective we compute the win rate of the team that secured it and run a chi-square test of independence between securer and outcome.

```{r eda-objectives}
obj_cols <- c("firstblood", "firstdragon", "firstherald", "firsttower")

obj_winrates <- map_dfr(obj_cols, function(obj) {
    blue_col <- paste0("blue_", obj)
    df <- games %>%
        select(all_of(c(blue_col, "blue_win"))) %>%
        drop_na()

    n_blue <- sum(df[[blue_col]] == 1)
    n_red  <- sum(df[[blue_col]] == 0)
    blue_wr <- mean(df$blue_win[df[[blue_col]] == 1]) * 100
    red_wr  <- mean(df$blue_win[df[[blue_col]] == 0]) * 100  # P(Blue wins | Red secured)

    chi <- chisq.test(table(df[[blue_col]], df$blue_win))

    tibble(
        objective     = str_replace(obj, "first", "First ") %>% str_to_title(),
        secured_by    = c("Blue secured", "Red secured"),
        win_rate_blue = c(blue_wr, red_wr),
        n             = c(n_blue, n_red),
        chi2          = chi$statistic,
        p_value       = chi$p.value
    )
})

ggplot(
    obj_winrates,
    aes(x = reorder(objective, win_rate_blue), y = win_rate_blue, fill = secured_by)
) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6) +
    geom_text(
        aes(label = sprintf("%.1f%% (n=%d)", win_rate_blue, n)),
        position = position_dodge(width = 0.7),
        hjust = -0.05, size = 3
    ) +
    coord_flip() +
    scale_fill_manual(values = c("Blue secured" = "#3498db", "Red secured" = "#e74c3c")) +
    scale_y_continuous(limits = c(0, 100)) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    labs(
        title    = "P(Blue wins) by Who Secured the First Objective",
        x = NULL, y = "Blue Win Rate (%)", fill = NULL
    ) +
    theme_minimal()

cat("Chi-square independence tests (objective securer vs blue_win):\n")
obj_winrates %>%
    distinct(objective, chi2, p_value) %>%
    mutate(
        chi2    = round(chi2, 1),
        p_value = formatC(p_value, format = "e", digits = 2)
    ) %>%
    print()
```

When Blue secures first tower or first herald, Blue's win rate jumps well above 50%, and drops by a similar amount when Red gets them instead. First blood and first dragon move the rate much less. Every chi-square test rejects independence at p ≪ 0.001, so the *direction* is certain, but the *magnitude* varies sharply between objectives - the same pattern the correlation analysis showed. Tower-related objectives translate straight into map control and gold, while first blood is often a single lucky skirmish. All four are worth keeping, but the models should weight tower and herald more heavily.

## Void grubs

Void grubs are a 2024+ objective: they spawn at 5:00 and despawn at 13:45 on early-2025 patches (moved to 14:45 mid-season), so they are purely early-game. The pit holds up to 6 grubs total, split between the two teams. We plot Blue's win rate against the number of grubs Blue ended up with.

```{r eda-grubs}
games %>%
    filter(!is.na(blue_void_grubs)) %>%
    group_by(blue_void_grubs) %>%
    summarise(
        win_rate = mean(blue_win) * 100,
        n        = n(),
        .groups  = "drop"
    ) %>%
    ggplot(aes(x = factor(blue_void_grubs), y = win_rate, fill = win_rate)) +
    geom_col() +
    geom_text(aes(label = sprintf("%.1f%%\nn=%d", win_rate, n)),
        vjust = -0.3, size = 3
    ) +
    scale_fill_gradient(low = "#e74c3c", high = "#3498db") +
    scale_y_continuous(limits = c(0, 100)) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Blue Win Rate by Number of Void Grubs Blue Secured",
        x = "blue_void_grubs", y = "P(Blue wins) (%)"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

The relationship is essentially monotonic: 0 grubs sits well below 50% win rate, 6 grubs climbs to \~70%. That justifies the grub differential we build later. Small wobbles at intermediate counts come from low sample sizes - teams that end up with *exactly* 4 grubs are rare, because grub fights usually end with one side taking the whole remaining set.

## Side advantage

Blue side gets first pick in the draft and has slight map asymmetries. We measure the side advantage as the share of Blue wins, and report a 95% CI plus a chi-square test against the 50% null.

```{r eda-side}
n_games  <- nrow(games)
n_blue   <- sum(games$blue_win)
side_pt  <- prop.test(n_blue, n_games, p = 0.5)
blue_wr  <- n_blue / n_games * 100
ci_low   <- side_pt$conf.int[1] * 100
ci_high  <- side_pt$conf.int[2] * 100

tibble(
    label    = "Blue side",
    win_rate = blue_wr,
    ci_low   = ci_low,
    ci_high  = ci_high
) %>%
    ggplot(aes(x = label, y = win_rate)) +
    geom_col(width = 0.4, fill = "#3498db") +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.15, color = "gray20") +
    geom_text(
        aes(label = sprintf("%.2f%%  [%.2f, %.2f]\nn=%d  ·  chi^2=%.1f, p=%.2g",
                            win_rate, ci_low, ci_high, n_games,
                            side_pt$statistic, side_pt$p.value)),
        vjust = -0.6, size = 3.5
    ) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    scale_y_continuous(limits = c(0, 70)) +
    labs(
        title    = "Blue-Side Advantage",
        x = NULL, y = "P(Blue wins) (%)"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

Blue wins \~53% of games, a steady \~3 pp advantage. The 95% CI excludes 50% and the chi-square test rejects the null at p ≪ 0.001, so the effect is significant even if small. It comes from first pick in champion draft and small map/camera asymmetries.

## Role-specific gold differentials (H2 preview)

H2 asks whether some roles' early leads matter more than others. As a first look, we take each Blue-side player by role and correlate their gold differential at 15 min with `blue_win`.

```{r eda-roles}
role_diffs_blue <- players %>%
    filter(position %in% c("top", "jng", "mid", "bot", "sup"),
           side == "Blue",
           !is.na(golddiffat15)) %>%
    select(gameid, position, golddiffat15)

role_cors <- role_diffs_blue %>%
    inner_join(games %>% select(gameid, blue_win), by = "gameid") %>%
    group_by(position) %>%
    summarise(
        correlation     = cor(golddiffat15, blue_win),
        mean_diff_win   = mean(golddiffat15[blue_win == 1]),
        mean_diff_loss  = mean(golddiffat15[blue_win == 0]),
        .groups         = "drop"
    ) %>%
    mutate(position = toupper(position))

ggplot(role_cors, aes(
    x    = reorder(position, correlation),
    y    = correlation, fill = correlation
)) +
    geom_col() +
    geom_text(aes(label = sprintf("r = %.3f", correlation)),
        hjust = -0.1, size = 3.5
    ) +
    coord_flip() +
    scale_y_continuous(limits = c(0, 0.42)) +
    scale_fill_gradient(low = "#f39c12", high = "#2ecc71") +
    labs(
        title    = "Correlation: Role Gold Diff @15 (Blue side) vs blue_win",
        x = NULL, y = "Pearson Correlation with blue_win"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

```{r eda-roles-boxplot}
role_diffs_blue %>%
    inner_join(games %>% select(gameid, blue_win), by = "gameid") %>%
    mutate(
        blue_win = factor(blue_win, levels = c(0, 1), labels = c("Red won", "Blue won")),
        position = toupper(position)
    ) %>%
    ggplot(aes(x = position, y = golddiffat15, fill = blue_win)) +
    geom_boxplot(outlier.alpha = 0.1) +
    scale_fill_manual(values = c("Red won" = "#e74c3c", "Blue won" = "#3498db")) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title    = "Blue-Side Player Gold Differential @15 by Role and Outcome",
        x = "Role", y = "Gold Diff @15 min", fill = "Outcome"
    ) +
    theme_minimal()
```

Bot lane (ADC) has the highest correlation with `blue_win`, which is early evidence for H2: the bot lane's gold advantage looks like the most predictive role-specific factor. The gaps between roles are small, though, so the full H2 analysis will decide whether the gap is real or noise. The boxplot shows clear win/loss separation for every role; bot lane simply has the widest gap.

## Feature distributions and scaling check

KNN and Logistic Regression are sensitive to feature scale; tree-based models and Naive Bayes are not. We check the per-feature ranges below to decide whether standardisation is needed.

```{r eda-distributions}
scale_check <- games %>%
    select(
        blue_goldat15, blue_xpat15, blue_csat15, blue_killsat15,
        blue_golddiffat15, blue_xpdiffat15, blue_csdiffat15,
        blue_void_grubs
    ) %>%
    drop_na() %>%
    pivot_longer(everything(), names_to = "feature", values_to = "value")

ggplot(scale_check, aes(x = value)) +
    geom_histogram(bins = 50, fill = "#3498db", alpha = 0.7) +
    facet_wrap(~feature, scales = "free", ncol = 4) +
    labs(
        title    = "Feature Distributions - Scale Check (Blue's perspective)",
        x = "Value", y = "Count"
    ) +
    theme_minimal()
```

The scales differ wildly: `blue_goldat15` runs from \~20,000 to \~36,000, while `blue_void_grubs` lives on 0-6 and `blue_killsat15` on 0-24. Without standardisation, KNN distances would be swamped by gold (1000x larger than the rest) and effectively ignore kills and grubs. Logistic Regression coefficients would also be misleading, since coefficient magnitude is only comparable across features when each has unit variance. We z-score standardise before fitting KNN and Logistic Regression. Tree-based models (RF, CART) and Naive Bayes do not need it.

## EDA Summary

1.  **Target is mildly imbalanced** (\~53/47 in Blue's favour). Significant but small - accuracy is still a valid metric and no resampling is needed.
2.  **Gold diff @15 is the strongest single predictor.** A naive "gold-leader wins" rule already hits \~73% accuracy at the team level, which is our baseline to beat.
3.  **Residual multicollinearity** remains between raw stats and their differentials (`blue_goldat15` vs `blue_golddiffat15`) and between paired independent counts (`blue_killsat15` vs `red_killsat15`). Linear models will need regularisation even after dropping the strict mirrors.
4.  **First tower** and **first herald** swing Blue's win rate sharply (chi-square p ≪ 0.001); they are the highest-impact binary objectives.
5.  **First blood** and **first dragon** are statistically significant but carry weaker signal than tower or herald - both shift Blue's WR by \~10-15 pp when Blue secures, but only \~3-5 pp when Red does (the asymmetry comes from the dragon-herald trade pattern: Blue takes herald in 61% of games and Red free-takes dragon in 62%, so "Red secured first dragon" is biased toward games where Blue was already ahead).
6.  **Void grubs** show a near-monotonic gradient in Blue's win rate from 0 to 6 grubs, which justifies engineering a grub differential.
7.  **Blue-side advantage** is significant (\~53% win rate, 95% CI excludes 50%). With one row per match expressed from Blue's perspective, this advantage is absorbed into the base rate of `blue_win`.
8.  **Feature scales differ by \~1000x**, so KNN and Logistic Regression need standardisation.
9.  **Bot lane** has the highest role-specific correlation with `blue_win`, supporting the H2 analysis.
10. **Leads tend to grow** between @10 and @15.
11. **Turret plate differential** is a strong, purely early-game feature (plates fall by 14:00) not commonly used in the published literature (introduced in 2025 season).
12. **Total dragon/herald counts** are full-game statistics, not early-game, and must not enter the feature set - using them would leak the outcome.

# Feature Engineering

The correlation analysis showed that *gaps* between teams carry the signal, so we keep signed differentials. To control multicollinearity, we keep one of each redundant triple `(goldat15, opp_goldat15, golddiffat15)` - the differential - and drop the other two. The raw @10 snapshots are also omitted: momentum, defined as `diff@15 - diff@10`, summarises their signal in a single term. On top of this we add two features that rarely appear in the published @15 literature: a **turret plate differential** (plates fall by 14:00, purely early-game; r = 0.409 with `blue_win`) and a **rolling team win rate differential** (computed from prior games only).

## Rolling team win rate

For each team, we compute a rolling win rate over all of their *prior* games (sorted by date), defaulting first-appearance teams to 0.5. The per-side values are then pivoted to one Blue column and one Red column per game, and `winrate_diff = blue - red` captures pre-game asymmetry without leaking information from the current result.

```{r rolling-winrate}
team_wr <- teams %>%
    arrange(date, gameid) %>%
    group_by(teamid) %>%
    mutate(
        games_before = row_number() - 1,
        wins_before  = cumsum(result) - result,
        team_winrate = ifelse(games_before > 0, wins_before / games_before, 0.5)
    ) %>%
    ungroup() %>%
    mutate(side = tolower(side)) %>%
    select(gameid, side, team_winrate, games_before) %>%
    pivot_wider(
        id_cols     = gameid,
        names_from  = side,
        values_from = c(team_winrate, games_before),
        names_glue  = "{side}_{.value}"
    )

games_wr <- games %>%
    left_join(team_wr, by = "gameid") %>%
    mutate(winrate_diff = blue_team_winrate - red_team_winrate)

cat(sprintf(
    "Median prior games: %.0f Blue / %.0f Red\nFirst-appearance teams (defaulted to 0.5): %d Blue, %d Red\nwinrate_diff correlation with blue_win: %.3f",
    median(games_wr$blue_games_before),
    median(games_wr$red_games_before),
    sum(games_wr$blue_games_before == 0),
    sum(games_wr$red_games_before == 0),
    cor(games_wr$winrate_diff, games_wr$blue_win, use = "complete.obs")
))
```

## Select and engineer features

Every feature below is expressed from Blue's perspective: signed differentials where a positive value indicates Blue is ahead, plus four binary flags marking whether Blue secured each first objective. Since the perspective is fixed, the `blue_` prefix is implicit throughout the modelling stage and we omit it from column names. Each row represents one full match, so the structural Blue-side advantage is absorbed into the base rate of `blue_win` and serves as the model's intercept.

```{r select-features}
games_model <- games_wr %>%
    transmute(
        gameid,
        blue_win,

        # Snapshot diffs at 15 (primary signal, Blue's perspective)
        golddiffat15 = blue_golddiffat15,
        xpdiffat15   = blue_xpdiffat15,
        csdiffat15   = blue_csdiffat15,

        # Momentum: did Blue's lead grow between 10 and 15?
        gold_momentum = blue_golddiffat15 - blue_golddiffat10,
        xp_momentum   = blue_xpdiffat15   - blue_xpdiffat10,
        cs_momentum   = blue_csdiffat15   - blue_csdiffat10,

        # Combat diffs at 15 (death_diff omitted: near-perfect mirror of -kill_diff, see VIF)
        kill_diff_15   = blue_killsat15   - red_killsat15,
        assist_diff_15 = blue_assistsat15 - red_assistsat15,

        # Per-team kill counts in the 10-15 window
        kill_pressure_10_15     = blue_killsat15 - blue_killsat10,
        opp_kill_pressure_10_15 = red_killsat15  - red_killsat10,

        gold_efficiency_15 = blue_goldat15 / pmax(blue_csat15, 1),
        kda_15             = (blue_killsat15 + blue_assistsat15) / pmax(blue_deathsat15, 1),

        # Objective differentials
        plate_diff = blue_turretplates - red_turretplates,
        grub_diff  = blue_void_grubs   - red_void_grubs,

        # Pre-game strength
        winrate_diff,

        # First-objective flags (1 if Blue secured)
        firstblood  = blue_firstblood,
        firstdragon = blue_firstdragon,
        firstherald = blue_firstherald,
        firsttower  = blue_firsttower
    ) %>%
    drop_na()

cat(sprintf(
    "Final feature set: %d features (excluding gameid and blue_win)\nFeatures: %s",
    ncol(games_model) - 2,
    paste(setdiff(names(games_model), c("gameid", "blue_win")), collapse = ", ")
))
```

## H2: Role-specific differentials

For H2 we add per-role early-game differentials. The signed `golddiffat15` of each Blue-side player is by construction `blue_player_gold - red_lane_opponent_gold`, so retaining only the Blue rows and pivoting wide yields a Blue-perspective signed differential per role and per metric (e.g., `bot_golddiffat15`, `mid_xpdiffat10`). The result is then inner-joined back to `games_model`.

```{r h2-role-features}
role_diffs <- players %>%
    filter(position %in% c("top", "jng", "mid", "bot", "sup"),
           side == "Blue") %>%
    select(
        gameid, position,
        golddiffat10, golddiffat15,
        xpdiffat10,   xpdiffat15,
        csdiffat10,   csdiffat15
    ) %>%
    drop_na() %>%
    pivot_wider(
        id_cols     = gameid,
        names_from  = position,
        values_from = c(golddiffat10, golddiffat15,
                        xpdiffat10,   xpdiffat15,
                        csdiffat10,   csdiffat15),
        names_glue  = "{position}_{.value}"
    )

games_h2 <- games_model %>%
    inner_join(role_diffs, by = "gameid") %>%
    drop_na()

cat(sprintf(
    "Role features added: %d\nH2 rows after join + drop_na: %d (removed %d from H1 set, %.1f%%)\nH2 total features: %d",
    ncol(role_diffs) - 1,
    nrow(games_h2),
    nrow(games_model) - nrow(games_h2),
    (nrow(games_model) - nrow(games_h2)) / nrow(games_model) * 100,
    ncol(games_h2) - 2
))
```

# Train/Test Split

Each row in `games_model` represents one full match, so a plain 80/20 row split is sufficient. We sample game IDs once with `set.seed(42)` and reuse the same partition for H2, ensuring that both hypotheses are evaluated on the same held-out games.

```{r train-test-split}
set.seed(42)

train_ids <- sample(games_model$gameid, size = floor(0.8 * nrow(games_model)))

train_h1 <- games_model %>% filter(gameid %in% train_ids)  %>% select(-gameid)
test_h1  <- games_model %>% filter(!gameid %in% train_ids) %>% select(-gameid)

train_h2 <- games_h2    %>% filter(gameid %in% train_ids)  %>% select(-gameid)
test_h2  <- games_h2    %>% filter(!gameid %in% train_ids) %>% select(-gameid)

cat(sprintf(
    "Games: %d train, %d test\nH1 - Train: %d rows, Test: %d rows\nH2 - Train: %d rows, Test: %d rows",
    length(train_ids), nrow(games_model) - length(train_ids),
    nrow(train_h1), nrow(test_h1), nrow(train_h2), nrow(test_h2)
))
```

# Scenario 1: Model Comparison

Five classifiers on the same H1 feature set, judged on accuracy, AUC-ROC, precision, recall, and F1. Hyperparameters are tuned with shared 5-fold CV (`metric = "ROC"`); the test set is held out for the final report.

| Category       | Model                | Role in the comparison                                     |
|----------------|----------------------|------------------------------------------------------------|
| Statistical/ML | Logistic Regression  | Interpretable linear baseline; standardised coefficients   |
| Statistical/ML | Random Forest        | Non-linear ensemble; agreement check on what matters       |
| Statistical/ML | Naive Bayes          | Bayes-rule baseline assuming feature independence          |
| Partitioning   | KNN                  | Distance-based; sensitive to scale and correlated features |
| Partitioning   | Decision Tree (CART) | Single-tree analogue of RF for the consensus comparison    |

The bar to beat is the EDA baseline: **72.9 %** from "Blue ahead in gold @15 wins". Published @15 LoL studies report 73-78 %, so a 75 %-band result is consistent with the literature.

## Multicollinearity check (VIF)

Logistic Regression, Naive Bayes, and KNN are sensitive to correlated inputs: LR coefficient signs and standard errors degrade under collinearity, NB's independence assumption is directly violated, and KNN distances over-weight correlated dimensions. Random Forest and CART are immune (univariate splits). Per the consultation, S1 keeps every feature and uses VIF only as a diagnostic - high-VIF features remain in the model and motivate the regularised selection in S3.

```{r s1-preprocess}
X_train <- train_h1 %>% select(-blue_win)
y_train <- train_h1$blue_win
X_test  <- test_h1  %>% select(-blue_win)
y_test  <- test_h1$blue_win

preproc        <- preProcess(X_train, method = c("center", "scale"))
X_train_scaled <- predict(preproc, X_train)
X_test_scaled  <- predict(preproc, X_test)

cat(sprintf("Training features: %d   Training rows: %d   Test rows: %d\n",
            ncol(X_train), nrow(X_train), nrow(X_test)))
```

```{r s1-vif}
vif_df <- tibble(
    feature = names(X_train_scaled),
    VIF     = diag(solve(cor(X_train_scaled)))
) %>% arrange(desc(VIF))

knitr::kable(vif_df, digits = 2,
             caption = "VIF on H1 training features (diagnostic only - no features dropped in S1).")
```

Resource diffs and momentum form the dominant collinear group (`golddiffat15` ↔ `gold_momentum`, etc.) because momentum is constructed from the @15 diff and the @10 diff. The KDA features (`kill_diff_15`, `kda_15`, `assist_diff_15`) carry mid-range VIF from sharing the same combat signal. Binary objective flags and `winrate_diff` are clean. We expect to see the consequences in the LR coefficient table - collinear features may carry inflated SEs and signs that flip relative to the marginal direction.

## CV setup

Shared `trainControl` so every model is tuned on the same folds.

```{r s1-cv-setup}
ctrl <- trainControl(
    method = "cv",
    number = 5,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final",
    verboseIter = TRUE
)

y_train_fct <- factor(y_train, levels = c(0, 1), labels = c("Loss", "Win"))
y_test_fct  <- factor(y_test,  levels = c(0, 1), labels = c("Loss", "Win"))
```

## Logistic Regression

Linear baseline trained on the standardised feature matrix so coefficients are directly comparable. With Win as the last factor level, `glm` models log-odds of Win - a positive coefficient indicates the feature pushes towards Blue winning, a negative one pushes against.

```{r s1-logistic}
t0 <- Sys.time()

lr_model <- train(
    x = X_train_scaled,
    y = y_train_fct,
    method = "glm",
    family = "binomial",
    trControl = ctrl,
    metric = "ROC"
)

lr_pred <- predict(lr_model, X_test_scaled)
lr_prob <- predict(lr_model, X_test_scaled, type = "prob")
lr_cm <- confusionMatrix(lr_pred, y_test_fct, positive = "Win")

cat(sprintf(
    "\nLogistic Regression Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    lr_cm$overall["Accuracy"] * 100,
    lr_cm$byClass["Sensitivity"] * 100,
    lr_cm$byClass["Specificity"] * 100,
    max(lr_model$results$ROC)
))
```

```{r s1-logistic-coefs}
lr_coefs <- coef(lr_model$finalModel) %>%
    as.data.frame() %>%
    rownames_to_column("feature") %>%
    rename(coefficient = 2) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

ggplot(
    lr_coefs %>% head(20),
    aes(
        x = reorder(feature, abs(coefficient)),
        y = coefficient,
        fill = coefficient > 0
    )
) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title = "Logistic Regression - Top 20 Coefficients",
        x = NULL, y = "Coefficient", fill = "Direction"
    ) +
    theme_minimal()
```

Resource diffs at @15 (`golddiffat15`, `xpdiffat15`, `csdiffat15`), `kill_diff_15`, `winrate_diff`, and the first-objective flags all carry positive coefficients - directionally as the EDA predicted. `gold_momentum` flips sign relative to its univariate correlation. This is the symptom flagged by VIF: when two features are near-perfect linear combinations, the LR can split their effect arbitrarily and the *individual* coefficient sign is no longer meaningful, even though the joint prediction is still correct. S3's regularised methods address this directly.

## Random Forest

Non-linear ensemble; immune to the multicollinearity that distorts the LR coefficients. We tune `mtry` over `{5, 7, 10, 15, 19}` with 500 trees fixed.

```{r s1-rf}
t0 <- Sys.time()

rf_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "rf",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(mtry = c(5, 7, 10, 15, 19)),
    ntree = 500
)

rf_pred <- predict(rf_model, X_test)
rf_prob <- predict(rf_model, X_test, type = "prob")
rf_cm <- confusionMatrix(rf_pred, y_test_fct, positive = "Win")

cat(sprintf(
    "\nRandom Forest Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity: %.1f%%\nSpecificity: %.1f%%\nBest mtry: %d\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    rf_cm$overall["Accuracy"] * 100,
    rf_cm$byClass["Sensitivity"] * 100,
    rf_cm$byClass["Specificity"] * 100,
    rf_model$bestTune$mtry,
    max(rf_model$results$ROC)
))

# Show CV results across mtry values
print(rf_model$results %>% select(mtry, ROC, Sens, Spec))
```

```{r s1-rf-importance}
rf_imp <- varImp(rf_model)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))

ggplot(
    rf_imp %>% head(20),
    aes(x = reorder(feature, Overall), y = Overall)
) +
    geom_col(fill = "#27ae60") +
    coord_flip() +
    labs(
        title = "Random Forest - Top 20 Feature Importances",
        x = NULL, y = "Importance"
    ) +
    theme_minimal()
```

## Naive Bayes

The independence assumption is directly violated by the VIF result above; we still fit it to see how much accuracy that costs. We tune `usekernel` (Gaussian vs KDE) and the KDE bandwidth multiplier `adjust`.

```{r s1-nb}
t0 <- Sys.time()

nb_grid <- expand.grid(
    usekernel = c(FALSE, TRUE),
    laplace   = 0,
    adjust    = c(0.5, 1.0, 1.5)
)

nb_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "naive_bayes",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = nb_grid
)

nb_pred <- predict(nb_model, X_test)
nb_prob <- predict(nb_model, X_test, type = "prob")
nb_cm <- confusionMatrix(nb_pred, y_test_fct, positive = "Win")

cat(sprintf(
    "\nNaive Bayes Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity: %.1f%%\nSpecificity: %.1f%%\nBest: usekernel=%s, adjust=%.1f\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    nb_cm$overall["Accuracy"] * 100,
    nb_cm$byClass["Sensitivity"] * 100,
    nb_cm$byClass["Specificity"] * 100,
    as.character(nb_model$bestTune$usekernel),
    nb_model$bestTune$adjust,
    max(nb_model$results$ROC)
))

print(nb_model$results %>%
    arrange(desc(ROC)) %>%
    select(usekernel, adjust, ROC, Sens, Spec))
```

```{r s1-nb-importance}
# Naive Bayes has no native coefficient-style importance. We use a filter-based
# importance: each feature's single-feature ROC-AUC against the target. A feature
# that separates Win/Loss on its own scores near 1 (or near 0, same information
# in the opposite direction), while a useless feature scores near 0.5. We report
# |AUC - 0.5| so the ranking is direction-agnostic.
# We compute this directly with pROC because caret::varImp's filter fallback has
# a known bug with factor targets for the "naive_bayes" method.
nb_imp <- tibble(
    feature = names(X_train),
    auc_raw = sapply(X_train, function(col) {
        as.numeric(pROC::auc(pROC::roc(y_train_fct, as.numeric(col),
            levels = c("Loss", "Win"), quiet = TRUE
        )))
    })
) %>%
    mutate(Overall = abs(auc_raw - 0.5) * 200) %>% # rescale to 0-100
    select(feature, Overall) %>%
    arrange(desc(Overall))

ggplot(
    nb_imp %>% head(20),
    aes(x = reorder(feature, Overall), y = Overall)
) +
    geom_col(fill = "#e67e22") +
    coord_flip() +
    labs(
        title = "Naive Bayes - Top 20 Feature Importances",
        x = NULL, y = "Importance (AUC)"
    ) +
    theme_minimal()
```

## KNN

Distance-based; runs on the standardised matrix. Multicollinearity inflates correlated dimensions in the Euclidean metric, so we expect KNN to lose ground to RF/CART here.

```{r s1-knn}
t0 <- Sys.time()

# k grid log-spaced from 5 to ~4*sqrt(n), forced odd to avoid 50/50 ties.
n_train <- nrow(X_train_scaled)
k_theo  <- round(sqrt(n_train))
k_grid  <- unique(round(exp(seq(log(5), log(k_theo * 4), length.out = 12))))
k_grid  <- ifelse(k_grid %% 2 == 0, k_grid + 1, k_grid)

cat(sprintf(
    "sqrt(n) = %d, candidate K grid: %s\n",
    k_theo, paste(k_grid, collapse = ", ")
))

knn_model <- train(
    x = X_train_scaled,
    y = y_train_fct,
    method = "knn",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(k = k_grid)
)

knn_pred <- predict(knn_model, X_test_scaled)
knn_prob <- predict(knn_model, X_test_scaled, type = "prob")
knn_cm <- confusionMatrix(knn_pred, y_test_fct, positive = "Win")

cat(sprintf(
    "\nKNN Results (%.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nBest k: %d\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    knn_cm$overall["Accuracy"] * 100,
    knn_cm$byClass["Sensitivity"] * 100,
    knn_cm$byClass["Specificity"] * 100,
    knn_model$bestTune$k,
    max(knn_model$results$ROC)
))
```

## Decision Tree CART

Single-tree analogue of RF; like RF, immune to multicollinearity. We tune the complexity parameter `cp` over `{1e-5, 5e-5, 1e-4, 5e-4, 1e-3}`.

```{r s1-cart}
t0 <- Sys.time()

# cp (complexity parameter) controls tree depth - tuned via 5-fold CV
cart_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "rpart",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(cp = c(0.00001, 0.00005, 0.0001, 0.0005, 0.001))
)

cart_pred <- predict(cart_model, X_test)
cart_prob <- predict(cart_model, X_test, type = "prob")
cart_cm <- confusionMatrix(cart_pred, y_test_fct, positive = "Win")

cat(sprintf(
    "\nCART Results (%.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nBest cp: %.4f\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    cart_cm$overall["Accuracy"] * 100,
    cart_cm$byClass["Sensitivity"] * 100,
    cart_cm$byClass["Specificity"] * 100,
    cart_model$bestTune$cp,
    max(cart_model$results$ROC)
))

cart_imp <- varImp(cart_model)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))
```

```{r s1-cart-plot}
# Tree is too deep to be readable at this pruning level
# rpart.plot(cart_model$finalModel,
#            main = "Decision Tree - Match Outcome Prediction",
#            type = 4,
#            extra = 104)
```

## Model comparison

Side-by-side test-set metrics, confusion matrices, ROC curves. We check (a) which models beat the 72.9 % gold-rule baseline and (b) whether the non-linear RF gains anything over the linear/probabilistic baselines.

```{r s1-metrics-table}
calc_metrics <- function(cm, prob, y_true, name) {
    roc_obj <- roc(y_true, prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
    tibble(
        Model = name,
        Accuracy = cm$overall["Accuracy"] * 100,
        AUC_ROC = auc(roc_obj) * 100,
        Precision = cm$byClass["Pos Pred Value"] * 100,
        Recall = cm$byClass["Sensitivity"] * 100,
        F1 = 2 * cm$byClass["Pos Pred Value"] * cm$byClass["Sensitivity"] /
            (cm$byClass["Pos Pred Value"] + cm$byClass["Sensitivity"]) * 100
    )
}

metrics <- bind_rows(
    calc_metrics(lr_cm, lr_prob, y_test_fct, "Logistic Regression"),
    calc_metrics(rf_cm, rf_prob, y_test_fct, "Random Forest"),
    calc_metrics(cart_cm, cart_prob, y_test_fct, "CART"),
    calc_metrics(nb_cm, nb_prob, y_test_fct, "Naive Bayes"),
    calc_metrics(knn_cm, knn_prob, y_test_fct, "KNN")
)

knitr::kable(metrics, digits = 1, caption = "Model Performance Comparison (H1)")
```

```{r s1-best-hyperparams}
best_hyperparams <- tibble(
    Model = c("Logistic Regression",
              "Random Forest",
              "CART",
              "Naive Bayes",
              "KNN"),
    `Tuned params` = c(
        "(none - glm has no hyperparameters)",
        sprintf("mtry = %d (ntree = 500 fixed)", rf_model$bestTune$mtry),
        sprintf("cp = %.5f", cart_model$bestTune$cp),
        sprintf("usekernel = %s, adjust = %.2f, laplace = %.0f",
                nb_model$bestTune$usekernel,
                nb_model$bestTune$adjust,
                nb_model$bestTune$laplace),
        sprintf("k = %d", knn_model$bestTune$k)
    ),
    `CV ROC` = c(
        max(lr_model$results$ROC),
        max(rf_model$results$ROC),
        max(cart_model$results$ROC),
        max(nb_model$results$ROC),
        max(knn_model$results$ROC)
    )
)

htmltools::div(
    style = "background-color: white; color: black; padding: 10px; border-radius: 4px;",
    htmltools::HTML(knitr::kable(
        best_hyperparams, digits = 3,
        caption = "Best hyperparameters per model (selected by 5-fold CV ROC)",
        format = "html"
    ))
)
```

```{r s1-confusion-matrices, fig.width=12, fig.height=4}
plot_cm <- function(cm, title) {
    cm_df <- as.data.frame(cm$table)
    ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
        geom_tile() +
        geom_text(aes(label = Freq), size = 6) +
        scale_fill_gradient(low = "white", high = "#3498db") +
        labs(title = title) +
        theme_minimal() +
        theme(legend.position = "none",
              plot.background  = element_rect(fill = "white", color = NA),
              panel.background = element_rect(fill = "white", color = NA))
}

gridExtra::grid.arrange(
    plot_cm(lr_cm, "Logistic Regression"),
    plot_cm(rf_cm, "Random Forest"),
    plot_cm(cart_cm, "CART"),
    plot_cm(nb_cm, "Naive Bayes"),
    plot_cm(knn_cm, "KNN"),
    ncol = 5
)
```

The test set inherits the underlying \~53/47 Blue-side base rate, so the ratio of TP to TN reflects how each model balances Win recall against Loss recall. Across all five matrices, precision and recall stay within a few points of each other, which means F1 tracks accuracy closely and no model is systematically biased toward predicting Win or Loss.

```{r s1-roc-curves}
roc_lr <- roc(y_test_fct, lr_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_rf <- roc(y_test_fct, rf_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_cart <- roc(y_test_fct, cart_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_nb <- roc(y_test_fct, nb_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_knn <- roc(y_test_fct, knn_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

ggroc(list(
    "Logistic Regression" = roc_lr,
    "Random Forest" = roc_rf,
    "CART" = roc_cart,
    "Naive Bayes" = roc_nb,
    "KNN" = roc_knn
)) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray40") +
    scale_color_manual(values = c(
        "Logistic Regression" = "#3498db",
        "Random Forest" = "#27ae60",
        "CART" = "#9b59b6",
        "Naive Bayes" = "#e67e22",
        "KNN" = "#e74c3c"
    )) +
    labs(
        title = "ROC Curves - Model Comparison",
        x = "Specificity", y = "Sensitivity", color = "Model"
    ) +
    theme_minimal()
```

All four AUC curves overlap closely; the discriminating differences are at the upper-left corner where the curves diverge by 1-2 pp.

## Explainability comparison

Each model produces an importance score on a different axis (LR: coefficient magnitude, RF: mean Gini decrease, CART: total Gini reduction across splits, NB: filter-based per-feature AUC). KNN is excluded - no native importance signal. The interesting question is whether the four rankings agree on a small core of features.

```{r s1-explainability}
# KNN excluded - no meaningful feature importance
top_features <- bind_rows(
    lr_coefs %>% mutate(rank = row_number(), model = "LR") %>%
        select(feature, rank, model) %>% head(10),
    rf_imp %>% mutate(rank = row_number(), model = "RF") %>%
        select(feature, rank, model) %>% head(10),
    cart_imp %>% mutate(rank = row_number(), model = "CART") %>%
        select(feature, rank, model) %>% head(10),
    nb_imp %>% mutate(rank = row_number(), model = "NB") %>%
        select(feature, rank, model) %>% head(10)
)

ggplot(top_features, aes(x = reorder(feature, -rank), y = rank, color = model)) +
    geom_jitter(size = 3, width = 0, height = 0.2) +
    coord_flip() +
    scale_y_reverse(breaks = 1:10) +
    scale_color_manual(values = c("LR" = "#3498db", "RF" = "#27ae60", "CART" = "#9b59b6", "NB" = "#e67e22")) +
    labs(
        title = "Feature Importance Rankings - Top 10 per Model",
        x = NULL, y = "Rank (1 = most important)", color = "Model"
    ) +
    theme_minimal()
```

A "top-10 hit count" is a coarse view because the four importance scales are not directly comparable. The full ranking - average rank plus spread across the four methods - gives a sharper read of which features every family agrees on and where they disagree.

```{r s1-consensus}
full_ranks <- lr_coefs %>%
    mutate(rank_LR = row_number()) %>%
    select(feature, rank_LR) %>%
    full_join(rf_imp   %>% mutate(rank_RF   = row_number()) %>% select(feature, rank_RF),
              by = "feature") %>%
    full_join(cart_imp %>% mutate(rank_CART = row_number()) %>% select(feature, rank_CART),
              by = "feature") %>%
    full_join(nb_imp   %>% mutate(rank_NB   = row_number()) %>% select(feature, rank_NB),
              by = "feature") %>%
    mutate(
        avg_rank    = rowMeans(across(starts_with("rank_")), na.rm = TRUE),
        rank_spread = apply(across(starts_with("rank_")), 1,
                            function(r) max(r, na.rm = TRUE) - min(r, na.rm = TRUE))
    ) %>%
    mutate(
        times_in_top10 = rowSums(across(c(rank_LR, rank_RF, rank_CART, rank_NB),
                                        ~ replace_na(.x, 999) <= 10))
    ) %>%
    arrange(avg_rank)

options(tibble.print_max = Inf)

cat("Cross-model feature ranking (lower avg_rank = more consistently important):\n")
print(as.data.frame(full_ranks))

cat("\nTrue consensus features (in all 4 top-10s):\n")
full_ranks %>% filter(times_in_top10 == 4) %>% print()

cat("\nStrong majority (in 3 of 4 top-10s):\n")
full_ranks %>% filter(times_in_top10 == 3) %>% print()

cat("\nMixed signal (in 2 of 4 top-10s) - model families disagree:\n")
full_ranks %>% filter(times_in_top10 == 2) %>% print()
```

The four-way consensus - features ranked highly by every method - is a compact set of continuous early-game differentials: `golddiffat15`, `xpdiffat15`, `csdiffat15`, and `gold_momentum`. These are the features any defense of H1 should lean on.

The notable disagreement is around binary objective flags (`firstdragon`, `firstherald`), which rank high for LR but low for the tree ensembles and NB filter score. This is a known artefact: each importance metric measures a different thing - LR sees them as additive log-odds shifts, RF/CART absorb their signal into continuous splits earlier, and the NB filter score reads them as weak standalone predictors. LASSO (S3) still keeps `firstdragon` with a non-zero coefficient, so the disagreement is metric-driven, not evidence that the feature is irrelevant.

## Same models, enriched data

H1 summarises the early game at the team level - one gold-diff per game, one xp-diff per game - but each match actually carries five values per metric, one per lane. Aggregating loses structure: a 1,000-gold team lead carried entirely by bot is a stronger signal than the same lead spread evenly across five lanes. We hold the five S1 models and the train/test split fixed and refit them on an enriched matrix - H1 plus per-role @15 differentials for economy (gold, xp, cs) and combat (kills, assists, deaths) - to see whether the extra granularity beats the team aggregate.

```{r s1-enrich-features}
# Per-role @15 differentials. Economy diffs come straight from the player row;
# combat diffs we compute as Blue-player-stat minus lane-opponent stat.
role_enriched <- players %>%
    filter(position %in% c("top", "jng", "mid", "bot", "sup"),
           side == "Blue") %>%
    mutate(
        killdiffat15   = killsat15   - opp_killsat15,
        assistdiffat15 = assistsat15 - opp_assistsat15,
        deathdiffat15  = deathsat15  - opp_deathsat15
    ) %>%
    select(gameid, position,
           golddiffat15, xpdiffat15, csdiffat15,
           killdiffat15, assistdiffat15, deathdiffat15) %>%
    drop_na() %>%
    pivot_wider(
        id_cols     = gameid,
        names_from  = position,
        values_from = c(golddiffat15, xpdiffat15, csdiffat15,
                        killdiffat15, assistdiffat15, deathdiffat15),
        names_glue  = "{position}_{.value}"
    )

games_enriched <- games_model %>%
    inner_join(role_enriched, by = "gameid") %>%
    drop_na()

cat(sprintf(
    "Per-role features added: %d (5 roles x 6 metrics @15)\nGames after join + drop_na: %d (lost %d rows from H1, %.1f%%)\nFeature count: %d (was %d on H1)\n",
    ncol(role_enriched) - 1,
    nrow(games_enriched),
    nrow(games_model) - nrow(games_enriched),
    (nrow(games_model) - nrow(games_enriched)) / nrow(games_model) * 100,
    ncol(games_enriched) - 2,
    ncol(games_model) - 2
))
```

```{r s1-enrich-split}
# Reuse train_ids so the only thing changing between runs is the feature matrix.
train_e <- games_enriched %>% filter(gameid %in% train_ids)  %>% select(-gameid)
test_e  <- games_enriched %>% filter(!gameid %in% train_ids) %>% select(-gameid)

X_train_e <- train_e %>% select(-blue_win)
X_test_e  <- test_e  %>% select(-blue_win)
y_train_e <- factor(train_e$blue_win, levels = c(0, 1), labels = c("Loss", "Win"))
y_test_e  <- factor(test_e$blue_win,  levels = c(0, 1), labels = c("Loss", "Win"))

preproc_e        <- preProcess(X_train_e, method = c("center", "scale"))
X_train_e_scaled <- predict(preproc_e, X_train_e)
X_test_e_scaled  <- predict(preproc_e, X_test_e)
```

```{r s1-enrich-train}
cat("Training all five S1 models on the enriched feature set...\n")

lr_e <- train(
    x = X_train_e_scaled, y = y_train_e,
    method = "glm", family = "binomial",
    trControl = ctrl, metric = "ROC"
)

# RF mtry grid scaled to the new feature count.
n_feat_e  <- ncol(X_train_e)
rf_grid_e <- unique(round(c(sqrt(n_feat_e), sqrt(n_feat_e) * 1.5,
                            sqrt(n_feat_e) * 2,  sqrt(n_feat_e) * 3)))
rf_e <- train(
    x = X_train_e, y = y_train_e,
    method = "rf", trControl = ctrl, metric = "ROC",
    tuneGrid = expand.grid(mtry = rf_grid_e), ntree = 500
)

nb_e <- train(
    x = X_train_e, y = y_train_e,
    method = "naive_bayes", trControl = ctrl, metric = "ROC",
    tuneGrid = nb_grid
)

n_train_e <- nrow(X_train_e_scaled)
k_theo_e  <- round(sqrt(n_train_e))
k_grid_e  <- unique(round(exp(seq(log(5), log(k_theo_e * 4), length.out = 12))))
k_grid_e  <- ifelse(k_grid_e %% 2 == 0, k_grid_e + 1, k_grid_e)
knn_e <- train(
    x = X_train_e_scaled, y = y_train_e,
    method = "knn", trControl = ctrl, metric = "ROC",
    tuneGrid = expand.grid(k = k_grid_e)
)

cart_e <- train(
    x = X_train_e, y = y_train_e,
    method = "rpart", trControl = ctrl, metric = "ROC",
    tuneGrid = expand.grid(cp = c(0.00001, 0.00005, 0.0001, 0.0005, 0.001))
)
```

```{r s1-enrich-metrics}
score_e <- function(model, X_eval, name) {
    cls  <- predict(model, X_eval)
    prob <- predict(model, X_eval, type = "prob")
    cm   <- confusionMatrix(cls, y_test_e, positive = "Win")
    calc_metrics(cm, prob, y_test_e, name)
}

metrics_enriched <- bind_rows(
    score_e(lr_e,   X_test_e_scaled, "Logistic Regression"),
    score_e(rf_e,   X_test_e,        "Random Forest"),
    score_e(cart_e, X_test_e,        "CART"),
    score_e(nb_e,   X_test_e,        "Naive Bayes"),
    score_e(knn_e,  X_test_e_scaled, "KNN")
)

knitr::kable(metrics_enriched, digits = 1,
             caption = "Enriched feature set: same five models, H1 + 30 per-role @15 features")
```

```{r s1-enrich-delta}
enrich_delta <- metrics %>%
    select(Model, H1_Acc = Accuracy, H1_AUC = AUC_ROC, H1_F1 = F1) %>%
    inner_join(
        metrics_enriched %>% select(Model,
                                    Enr_Acc = Accuracy,
                                    Enr_AUC = AUC_ROC,
                                    Enr_F1  = F1),
        by = "Model"
    ) %>%
    mutate(
        Acc_delta = Enr_Acc - H1_Acc,
        AUC_delta = Enr_AUC - H1_AUC,
        F1_delta  = Enr_F1  - H1_F1
    )

knitr::kable(enrich_delta, digits = 2,
             caption = "Same models, enriched data: deltas vs H1 (positive = enrichment helped)")
```

```{r s1-enrich-roc, fig.width=10, fig.height=6}
roc_lr_e   <- roc(y_test_e, predict(lr_e,   X_test_e_scaled, type = "prob")[, "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_rf_e   <- roc(y_test_e, predict(rf_e,   X_test_e,        type = "prob")[, "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_cart_e <- roc(y_test_e, predict(cart_e, X_test_e,        type = "prob")[, "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_nb_e   <- roc(y_test_e, predict(nb_e,   X_test_e,        type = "prob")[, "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_knn_e  <- roc(y_test_e, predict(knn_e,  X_test_e_scaled, type = "prob")[, "Win"], levels = c("Loss","Win"), quiet = TRUE)

ggroc(list(
    "LR (H1)"     = roc_lr,   "LR (enriched)"   = roc_lr_e,
    "RF (H1)"     = roc_rf,   "RF (enriched)"   = roc_rf_e,
    "CART (H1)"   = roc_cart, "CART (enriched)" = roc_cart_e,
    "NB (H1)"     = roc_nb,   "NB (enriched)"   = roc_nb_e,
    "KNN (H1)"    = roc_knn,  "KNN (enriched)"  = roc_knn_e
)) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray40") +
    labs(title = "ROC: H1 baseline vs enriched (per-role) feature set",
         x = "Specificity", y = "Sensitivity", color = "Model") +
    theme_minimal()
```

```{r s1-enrich-summary}
mean_auc <- mean(enrich_delta$AUC_delta)
mean_acc <- mean(enrich_delta$Acc_delta)
n_better <- sum(enrich_delta$AUC_delta > 0)
n_models <- nrow(enrich_delta)

cat(sprintf(
    "Mean AUC change across %d models: %+.2f pp\nMean accuracy change: %+.2f pp\nModels improved by enrichment (AUC delta > 0): %d / %d\n",
    n_models, mean_auc, mean_acc, n_better, n_models
))
```

The enrichment does not pay off. Mean AUC falls by 0.18 pp and mean accuracy by 0.49 pp; only Logistic Regression (+0.22 AUC) and Random Forest (+0.43 AUC) improve, both by amounts inside the 5-fold CV split-noise band. CART loses 0.85 AUC and 1.4 pp of accuracy - the classic single-tree overfit when the feature count more than doubles. Naive Bayes loses 0.66 AUC because the 30 new per-role columns are by construction correlated with their team-level sums, and the violated independence assumption costs more than the granularity gains. KNN is flat at 49 dimensions: the smoothing dictated by `k = 345` absorbs both the noise and any signal. RF's tuned `mtry` rising from 4 to 7 confirms the diagnosis: the model has to search a wider space for the same amount of usable information. The team aggregates already capture the @15 signal, so H1 stays as the S1 reference set for the downstream pipeline (S3 selection and the exported Shiny model). The per-role view is examined below under H2, which asks the sharper question - whether bot lane specifically, rather than the role matrix as a whole, carries extra signal beyond the team aggregate.

## H1 vs H2 comparison

We refit Random Forest (the best H1 performer) on the H2 feature set - H1 features plus 30 per-role differentials - to test whether role-level early-game gaps add signal beyond team totals. The H2 claim is that bot lane differentials should rank highest among role features.

```{r s1-h2-setup}
X_train_h2 <- train_h2 %>% select(-blue_win)
y_train_h2 <- factor(train_h2$blue_win, levels = c(0, 1), labels = c("Loss", "Win"))
X_test_h2  <- test_h2  %>% select(-blue_win)
y_test_h2  <- factor(test_h2$blue_win,  levels = c(0, 1), labels = c("Loss", "Win"))
```

```{r s1-h2-rf}
rf_h2 <- train(
    x = X_train_h2,
    y = y_train_h2,
    method = "rf",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(mtry = c(5, 10, 15, 20)),
    ntree = 500
)

rf_h2_pred <- predict(rf_h2, X_test_h2)
rf_h2_prob <- predict(rf_h2, X_test_h2, type = "prob")
rf_h2_cm <- confusionMatrix(rf_h2_pred, y_test_h2, positive = "Win")
roc_h2 <- roc(y_test_h2, rf_h2_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf(
    "H1 Random Forest AUC: %.3f (best mtry = %d)\nH2 Random Forest AUC: %.3f (best mtry = %d)\nDifference: %+.3f",
    auc(roc_rf), rf_model$bestTune$mtry,
    auc(roc_h2), rf_h2$bestTune$mtry,
    auc(roc_h2) - auc(roc_rf)
))
```

```{r s1-h2-role-importance}
h2_imp <- varImp(rf_h2)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))

# Highlight role-specific features
role_features <- h2_imp %>%
    filter(str_detect(feature, "^(top|jng|mid|bot|sup)_"))

ggplot(
    h2_imp %>% head(25),
    aes(
        x = reorder(feature, Overall), y = Overall,
        fill = str_detect(feature, "^(top|jng|mid|bot|sup)_")
    )
) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("FALSE" = "#95a5a6", "TRUE" = "#e74c3c"),
        labels = c("Team-level", "Role-specific")
    ) +
    labs(
        title = "H2 Random Forest - Top 25 Feature Importances",
        x = NULL, y = "Importance", fill = "Feature type"
    ) +
    theme_minimal()
```

```{r s1-h2-bot-rank}
bot_features <- role_features %>%
    filter(str_detect(feature, "^bot_")) %>%
    mutate(overall_rank = match(feature, h2_imp$feature))

other_role_features <- role_features %>%
    filter(!str_detect(feature, "^bot_")) %>%
    group_by(role = str_extract(feature, "^[a-z]+")) %>%
    summarise(
        best_rank = min(match(feature, h2_imp$feature)),
        avg_importance = mean(Overall),
        .groups = "drop"
    )

cat("Bot lane feature rankings in H2 model:\n")
print(bot_features)
cat("\nBest rank per role:\n")
print(other_role_features)
```

H2 is supported iff (a) the H2 RF AUC exceeds the H1 RF AUC by more than the CV split-noise band, and (b) at least one bot-lane feature ranks above the best feature of every other role. The two prints above provide the numbers needed for that judgement.

# Scenario 3: Feature Selection (Adriana)

The S1 VIF flagged a tight collinear cluster around resource diffs and momentum - direct evidence of redundancy. With 19 H1 features, S3 finds the minimal informative subset; LASSO and Elastic Net are particularly relevant since they zero out redundant features within correlated groups.

## RFE (Recursive Feature Elimination)

RFE iteratively removes the least important feature and re-trains until cross-validated AUC stops improving. We use Logistic Regression as the base estimator because standardised coefficients give a natural importance ranking that ties directly back to the interpretable LR baseline from S1. The core question for our dataset is whether the signal plateaus well before all 19 features — given the collinear resource-diff cluster flagged by VIF, we expect the elbow to appear early.

```{r s3-rfe}
cat("Running RFE with Logistic Regression estimator (5-fold CV)...\n")
t0 <- Sys.time()

lrFuncs_roc <- lrFuncs
lrFuncs_roc$summary <- twoClassSummary

rfe_ctrl <- rfeControl(
    functions = lrFuncs_roc,
    method    = "cv",
    number    = 5,
    verbose   = FALSE
)

n_features   <- ncol(X_train_scaled)
sizes_to_try <- unique(c(5, 10, 15, n_features))

rfe_result <- rfe(
    x          = X_train_scaled,
    y          = y_train_fct,
    sizes      = sizes_to_try,
    rfeControl = rfe_ctrl,
    metric     = "ROC"
)

rfe_selected <- predictors(rfe_result)

# Retrain LR on selected features for a clean test-set evaluation
rfe_lr <- train(
    x         = X_train_scaled[, rfe_selected, drop = FALSE],
    y         = y_train_fct,
    method    = "glm",
    family    = "binomial",
    trControl = trainControl(method = "none", classProbs = TRUE)
)
rfe_pred_class <- predict(rfe_lr, X_test_scaled[, rfe_selected, drop = FALSE])
rfe_pred_prob  <- predict(rfe_lr, X_test_scaled[, rfe_selected, drop = FALSE], type = "prob")
rfe_cm  <- confusionMatrix(rfe_pred_class, y_test_fct, positive = "Win")
rfe_roc <- roc(y_test_fct, rfe_pred_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("\nRFE Results (%.1fs):\nOptimal features: %d / %d\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    length(rfe_selected), n_features,
    rfe_cm$overall["Accuracy"] * 100,
    rfe_cm$byClass["Sensitivity"] * 100,
    rfe_cm$byClass["Specificity"] * 100,
    auc(rfe_roc)))

cat("\nSelected features:\n")
cat(paste(rfe_selected, collapse = ", "), "\n")
```

```{r s3-rfe-plot}
plot(rfe_result, type = c("g", "o"),
     main = "RFE: Cross-validated Accuracy vs Number of Features")

rfe_coef_df <- coef(rfe_lr$finalModel) %>%
    as.data.frame() %>%
    rownames_to_column("feature") %>%
    rename(coefficient = 2) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

ggplot(rfe_coef_df,
       aes(x = reorder(feature, abs(coefficient)), y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("Negative", "Positive")) +
    labs(title = "RFE - Selected Features and Coefficients",
         x = NULL, y = "Coefficient", fill = "Direction") +
    theme_minimal()
```

The accuracy-vs-size curve shows exactly where adding more features stops helping. If the curve flattens well before 19, that confirms the momentum/diff cluster is carrying duplicate signal. The coefficient plot shows which features the final LR kept and in which direction — `golddiffat15` is expected to dominate, with remaining features ordered by how much independent signal they contribute beyond the gold lead alone.

## Forward Stepwise Selection

Forward Stepwise greedily adds the single feature that most reduces AIC at each step, stopping when no addition helps. Its value here is as a cross-check against RFE: RFE uses cross-validated accuracy, Forward uses an information-theoretic criterion (AIC). If both methods converge on the same core features despite using completely different criteria, that agreement is strong evidence those features carry genuine signal rather than selection-method artefacts.

```{r s3-forward}
cat("Running Forward Stepwise Selection (AIC criterion)...\n")
t0 <- Sys.time()

fwd_train_df        <- X_train_scaled
fwd_train_df$result <- y_train_fct

null_model   <- glm(result ~ 1, data = fwd_train_df, family = "binomial")
full_formula <- as.formula(paste("result ~", paste(names(X_train_scaled), collapse = " + ")))

fwd_model <- MASS::stepAIC(
    null_model,
    scope     = list(lower = null_model, upper = full_formula),
    direction = "forward",
    trace     = FALSE
)

fwd_selected <- names(coef(fwd_model))
fwd_selected <- fwd_selected[fwd_selected != "(Intercept)"]

fwd_pred_prob_raw <- predict(fwd_model, newdata = X_test_scaled, type = "response")
fwd_pred_class    <- factor(ifelse(fwd_pred_prob_raw > 0.5, "Win", "Loss"),
                             levels = c("Loss", "Win"))
fwd_cm  <- confusionMatrix(fwd_pred_class, y_test_fct, positive = "Win")
fwd_roc <- roc(y_test_fct, fwd_pred_prob_raw, levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("\nForward Stepwise Results (%.1fs):\nFeatures selected: %d / %d\nFinal AIC: %.1f\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    length(fwd_selected), ncol(X_train_scaled),
    AIC(fwd_model),
    fwd_cm$overall["Accuracy"] * 100,
    fwd_cm$byClass["Sensitivity"] * 100,
    fwd_cm$byClass["Specificity"] * 100,
    auc(fwd_roc)))

cat("\nSelected features:\n")
cat(paste(fwd_selected, collapse = ", "), "\n")
```

```{r s3-forward-plot}
fwd_coef_df <- data.frame(
    feature     = fwd_selected,
    coefficient = coef(fwd_model)[fwd_selected]
) %>%
    arrange(desc(abs(coefficient)))

ggplot(fwd_coef_df,
       aes(x = reorder(feature, abs(coefficient)), y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("Negative", "Positive")) +
    labs(title = "Forward Stepwise - Selected Features and Coefficients",
         x = NULL, y = "Coefficient", fill = "Direction") +
    theme_minimal()
```

The selected features and coefficient directions can be compared directly to the RFE result above. Agreement between the two methods — on which features appear and which direction they push win probability — increases confidence that the finding is not an artefact of how one particular algorithm searches the feature space.

## LASSO (L1)

LASSO adds an L1 penalty to logistic regression that shrinks coefficients to exactly zero, making it both a classifier and a feature selector in one step. It is directly motivated by our VIF findings: `golddiffat15` and `gold_momentum` share a large fraction of their signal (momentum is literally the @15 minus @10 diff), and LASSO is the natural tool to resolve that redundancy — it retains the stronger signal and zeros out the weaker one automatically. The penalty strength `lambda` is tuned via 5-fold CV; we use `lambda.1se` rather than `lambda.min`, deliberately accepting a sparser model in exchange for a negligible increase in deviance.

```{r s3-lasso}
cat("Running LASSO with 5-fold CV to find optimal lambda...\n")
t0 <- Sys.time()

lasso_cv <- cv.glmnet(
    x = as.matrix(X_train_scaled),
    y = y_train,
    alpha = 1, # alpha=1 is pure LASSO (L1 penalty)
    family = "binomial",
    nfolds = 5
)

# lambda.1se: the largest lambda within 1 SE of the minimum - gives a sparser model
# that is nearly as accurate as lambda.min but keeps fewer features
lasso_coefs <- coef(lasso_cv, s = "lambda.1se")

lasso_coef_df <- data.frame(
    feature = rownames(lasso_coefs),
    coefficient = as.numeric(lasso_coefs)
) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

lasso_selected <- lasso_coef_df %>% filter(coefficient != 0)
lasso_eliminated <- lasso_coef_df %>% filter(coefficient == 0)

# test set evaluation
lasso_pred_prob <- predict(lasso_cv,
    newx = as.matrix(X_test_scaled),
    s = "lambda.1se", type = "response"
)
lasso_pred_class <- factor(ifelse(lasso_pred_prob > 0.5, "Win", "Loss"),
    levels = c("Loss", "Win")
)
lasso_cm <- confusionMatrix(lasso_pred_class, y_test_fct, positive = "Win")
lasso_roc <- roc(y_test_fct, as.numeric(lasso_pred_prob),
    levels = c("Loss", "Win"), quiet = TRUE
)

cat(sprintf(
    "\nLASSO Results (%.1fs):\nLambda (1se): %.6f\nFeatures kept: %d / %d (eliminated %d)\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    lasso_cv$lambda.1se,
    nrow(lasso_selected), nrow(lasso_coef_df), nrow(lasso_eliminated),
    lasso_cm$overall["Accuracy"] * 100,
    lasso_cm$byClass["Sensitivity"] * 100,
    lasso_cm$byClass["Specificity"] * 100,
    auc(lasso_roc)
))

cat("\nKept features:\n")
cat(paste(lasso_selected$feature, collapse = ", "), "\n")

cat("\nEliminated features:\n")
cat(paste(lasso_eliminated$feature, collapse = ", "), "\n")
```

```{r s3-lasso-plot}
plot(lasso_cv, main = "LASSO: CV Error vs Lambda")

lasso_selected %>%
    mutate(feature = reorder(feature, abs(coefficient))) %>%
    ggplot(aes(x = feature, y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title = "LASSO - Surviving Features and Coefficients",
        x = NULL, y = "Coefficient", fill = "Direction"
    ) +
    theme_minimal()
```

The gap between the two dashed lines shows how much sparsity we gain by moving from `lambda.min` to `lambda.1se` — if the deviance barely rises, the eliminated features were genuinely redundant. The coefficient plot is the key interpretive output: every feature LASSO zeroed out was redundant given the rest of the model, while surviving features (green = positive effect on Blue win probability, red = negative) form the minimal set that retains the full predictive signal.

## Elastic Net (L1 + L2)

Elastic Net mixes L1 (LASSO) and L2 (Ridge) penalties, with `alpha` controlling the blend. It addresses a known limitation of pure LASSO: when several features are correlated — as `golddiffat15`, `xpdiffat15`, and `csdiffat15` are — LASSO arbitrarily picks one and zeros the rest, which can make the selected set unstable across re-samples. Elastic Net can retain all three at reduced weights, potentially giving a more stable picture of which resource dimensions matter. We search the full alpha range in 0.05 steps; if the CV deviance curve is flat (many alphas within 1 SE of the minimum), the data cannot distinguish LASSO from Ridge at this sample size and the L1/L2 mix is effectively underdetermined.

```{r s3-elasticnet}
cat("Running Elastic Net - searching alpha grid (0 to 1, step 0.05)...\n")
t0 <- Sys.time()

alpha_grid <- seq(0, 1, by = 0.05)

enet_cv_results <- map_dfr(alpha_grid, function(a) {
    cv_fit     <- cv.glmnet(
        x      = as.matrix(X_train_scaled),
        y      = y_train,
        alpha  = a,
        family = "binomial",
        nfolds = 5
    )
    idx_1se    <- which(cv_fit$lambda == cv_fit$lambda.1se)
    pred_prob  <- predict(cv_fit, newx = as.matrix(X_test_scaled),
                          s = "lambda.1se", type = "response")
    pred_class <- factor(ifelse(pred_prob > 0.5, "Win", "Loss"), levels = c("Loss", "Win"))
    cm         <- confusionMatrix(pred_class, y_test_fct, positive = "Win")
    roc_obj    <- roc(y_test_fct, as.numeric(pred_prob), levels = c("Loss", "Win"), quiet = TRUE)
    n_kept     <- sum(coef(cv_fit, s = "lambda.1se") != 0) - 1
    tibble(
        alpha      = a,
        lambda_1se = cv_fit$lambda.1se,
        cv_dev     = cv_fit$cvm[idx_1se],
        cv_dev_se  = cv_fit$cvsd[idx_1se],
        n_features = n_kept,
        accuracy   = cm$overall["Accuracy"] * 100,
        auc        = as.numeric(auc(roc_obj))
    )
})

best_row   <- enet_cv_results %>% arrange(cv_dev) %>% dplyr::slice(1)
best_alpha <- best_row$alpha
best_dev   <- best_row$cv_dev
best_se    <- best_row$cv_dev_se
within_1se <- enet_cv_results %>% filter(cv_dev <= best_dev + best_se)

cat(sprintf("\nAlpha grid diagnostic: %d of %d alphas sit within 1 SE of the best (%.3f <= %.3f).\n",
            nrow(within_1se), nrow(enet_cv_results), best_dev, best_dev + best_se))
cat(sprintf("Alphas within 1 SE: %s\n",
            paste(sprintf("%.2f", within_1se$alpha), collapse = ", ")))
if (nrow(within_1se) >= nrow(enet_cv_results) - 2) {
    cat("-> CV curve is effectively flat across alpha; the L1/L2 mix is underdetermined by this data.\n")
}

enet_cv_final <- cv.glmnet(
    x      = as.matrix(X_train_scaled),
    y      = y_train,
    alpha  = best_alpha,
    family = "binomial",
    nfolds = 5
)

enet_coef_raw <- coef(enet_cv_final, s = "lambda.1se")
enet_coef_df  <- data.frame(
    feature     = rownames(enet_coef_raw),
    coefficient = as.numeric(enet_coef_raw)
) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

enet_selected_df <- enet_coef_df %>% filter(coefficient != 0)
enet_selected    <- enet_selected_df$feature

enet_pred_prob  <- predict(enet_cv_final, newx = as.matrix(X_test_scaled),
                            s = "lambda.1se", type = "response")
enet_pred_class <- factor(ifelse(enet_pred_prob > 0.5, "Win", "Loss"),
                           levels = c("Loss", "Win"))
enet_cm  <- confusionMatrix(enet_pred_class, y_test_fct, positive = "Win")
enet_roc <- roc(y_test_fct, as.numeric(enet_pred_prob), levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("\nElastic Net Results (%.1fs):\nBest alpha: %.1f (0=Ridge, 1=LASSO)\nLambda (1se): %.6f\nFeatures kept: %d / %d (eliminated %d)\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    best_alpha,
    enet_cv_final$lambda.1se,
    nrow(enet_selected_df), nrow(enet_coef_df), nrow(enet_coef_df) - nrow(enet_selected_df),
    enet_cm$overall["Accuracy"] * 100,
    enet_cm$byClass["Sensitivity"] * 100,
    enet_cm$byClass["Specificity"] * 100,
    auc(enet_roc)))

cat("\nKept features:\n")
cat(paste(enet_selected, collapse = ", "), "\n")

cat("\nEliminated features:\n")
cat(paste(enet_coef_df %>% filter(coefficient == 0) %>% pull(feature), collapse = ", "), "\n")
```

```{r s3-elasticnet-plot}
ggplot(enet_cv_results, aes(x = alpha, y = cv_dev)) +
    geom_ribbon(aes(ymin = cv_dev - cv_dev_se, ymax = cv_dev + cv_dev_se),
                fill = "#1abc9c", alpha = 0.2) +
    geom_hline(yintercept = best_dev + best_se, linetype = "dashed", color = "#7f8c8d") +
    geom_line(color = "#1abc9c", linewidth = 1.2) +
    geom_point(size = 3, color = "#1abc9c") +
    geom_point(data = filter(enet_cv_results, alpha == best_alpha),
               size = 5, shape = 21, fill = "#e74c3c", color = "white", stroke = 1.5) +
    labs(title = "Elastic Net: CV Binomial Deviance across Alpha Values",
         x = "Alpha (0 = Ridge, 1 = LASSO)", y = "5-fold CV Deviance at lambda.1se") +
    theme_minimal()

enet_selected_df %>%
    mutate(feature = reorder(feature, abs(coefficient))) %>%
    ggplot(aes(x = feature, y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("Negative", "Positive")) +
    labs(title = "Elastic Net - Surviving Features and Coefficients",
         x = NULL, y = "Coefficient", fill = "Direction") +
    theme_minimal()
```

A flat deviance curve (many alphas below the dashed 1-SE line) means the L1/L2 mix does not matter for prediction quality — any blend works equally well, and the "best alpha" is a tiebreak rather than a meaningful optimum. A U- or V-shape would mean the data genuinely benefits from a specific mix of sparsity and shrinkage. The coefficient plot is the key comparison against LASSO: any feature that LASSO zeroed out but Elastic Net retained carries secondary signal that only becomes visible when correlated features are allowed to share a group penalty rather than compete for a single non-zero slot.

## Random Forest Importance

RF importance is a non-parametric selection criterion that reuses the model already trained in S1 — no additional computation required. Unlike LASSO, which resolves correlated groups by zeroing all but one feature, RF distributes importance across correlated features in proportion to their marginal contribution after accounting for the others. This gives us a complementary view: where LASSO saw `golddiffat15` absorb the entire resource-diff signal, RF may still assign meaningful importance to `xpdiffat15` or `csdiffat15` because it measures each feature's contribution to splits independently. Features above mean importance are retained; we then train an LR on that subset to produce a test-set accuracy estimate comparable to the other methods.

```{r s3-rf-importance}
rf_importance_threshold <- mean(rf_imp$Overall)
rf_s3_selected <- rf_imp %>%
    filter(Overall > rf_importance_threshold) %>%
    pull(feature)

cat(sprintf("RF importance threshold: %.2f (mean importance)\nFeatures selected: %d / %d\n",
    rf_importance_threshold, length(rf_s3_selected), nrow(rf_imp)))

rf_s3_lr <- train(
    x         = X_train_scaled[, rf_s3_selected, drop = FALSE],
    y         = y_train_fct,
    method    = "glm",
    family    = "binomial",
    trControl = trainControl(method = "none", classProbs = TRUE)
)
rf_s3_pred_class <- predict(rf_s3_lr, X_test_scaled[, rf_s3_selected, drop = FALSE])
rf_s3_pred_prob  <- predict(rf_s3_lr, X_test_scaled[, rf_s3_selected, drop = FALSE], type = "prob")
rf_s3_cm  <- confusionMatrix(rf_s3_pred_class, y_test_fct, positive = "Win")
rf_s3_roc <- roc(y_test_fct, rf_s3_pred_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("LR on RF-selected features:\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    rf_s3_cm$overall["Accuracy"] * 100,
    rf_s3_cm$byClass["Sensitivity"] * 100,
    rf_s3_cm$byClass["Specificity"] * 100,
    auc(rf_s3_roc)))

cat("\nSelected features:\n")
cat(paste(rf_s3_selected, collapse = ", "), "\n")
```

```{r s3-rf-importance-plot}
rf_imp %>%
    mutate(
        selected = Overall > rf_importance_threshold,
        feature  = reorder(feature, Overall)
    ) %>%
    ggplot(aes(x = feature, y = Overall, fill = selected)) +
    geom_col() +
    coord_flip() +
    geom_hline(yintercept = rf_importance_threshold, linetype = "dashed",
               color = "#e74c3c", linewidth = 0.8) +
    scale_fill_manual(values = c("FALSE" = "#bdc3c7", "TRUE" = "#27ae60"),
                      labels = c("Eliminated", "Selected")) +
    labs(title = "RF Importance - Feature Selection Threshold",
         x = NULL, y = "Importance", fill = "Status") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 7))
```

A steep drop-off in the importance bars — where a small number of features sit far above the threshold and the rest cluster near zero — would confirm that the signal is dominated by a small core. A flatter distribution would suggest many features contribute roughly equally and the mean threshold is somewhat arbitrary. Comparing the selected set to LASSO's output shows which features each approach treats as redundant and which it considers independently informative.

```{r s3-best-hyperparams}
s3_best_hyperparams <- tibble(
    Method = c("RFE", "Forward Stepwise", "LASSO", "Elastic Net", "RF Importance"),
    `Best param(s)` = c(
        sprintf("subset size = %d (of %d)",
                rfe_result$bestSubset, ncol(X_train_scaled)),
        sprintf("AIC = %.1f, %d features kept",
                AIC(fwd_model), length(fwd_selected)),
        sprintf("lambda.1se = %.5f, %d features kept",
                lasso_cv$lambda.1se, length(lasso_selected$feature)),
        sprintf("alpha = %.2f, lambda.1se = %.5f, %d features kept",
                best_alpha, enet_cv_final$lambda.1se, length(enet_selected)),
        sprintf("threshold = %.2f (mean importance), %d features kept",
                rf_importance_threshold, length(rf_s3_selected))
    )
)

htmltools::div(
    style = "background-color: white; color: black; padding: 10px; border-radius: 4px;",
    htmltools::HTML(knitr::kable(
        s3_best_hyperparams,
        caption = "Best hyperparameters per S3 selection method",
        format = "html"
    ))
)
```

## Feature selection comparison

```{r s3-comparison}
selection_results <- list(
    RFE        = rfe_selected,
    Forward    = fwd_selected,
    LASSO      = lasso_selected$feature,
    ElasticNet = enet_selected,
    RF_Imp     = rf_s3_selected
)

all_features <- colnames(X_train_scaled)

# For each feature, record which methods selected it
overlap_df <- map_dfr(all_features, function(f) {
    sel <- map_lgl(selection_results, ~ f %in% .x)
    tibble(
        feature    = f,
        RFE        = sel[["RFE"]],
        Forward    = sel[["Forward"]],
        LASSO      = sel[["LASSO"]],
        ElasticNet = sel[["ElasticNet"]],
        RF_Imp     = sel[["RF_Imp"]],
        n_methods  = sum(sel)
    )
}) %>%
    arrange(desc(n_methods), feature)

consensus_5 <- overlap_df %>% filter(n_methods == 5)
consensus_4 <- overlap_df %>% filter(n_methods == 4)
never_sel   <- overlap_df %>% filter(n_methods == 0)

cat("Feature selection summary across all 5 methods:\n")
cat(sprintf("  Total features:                %d\n", length(all_features)))
cat(sprintf("  Selected by all 5 (consensus): %d\n", nrow(consensus_5)))
cat(sprintf("  Selected by 4 methods:         %d\n", nrow(consensus_4)))
cat(sprintf("  Never selected:                %d\n\n", nrow(never_sel)))

cat("Consensus features (all 5 methods agree):\n")
if (nrow(consensus_5) > 0) {
    cat(paste(consensus_5$feature, collapse = ", "), "\n")
} else {
    cat("None\n")
}

cat("\nFeatures selected by 4 methods:\n")
if (nrow(consensus_4) > 0) {
    cat(paste(consensus_4$feature, collapse = ", "), "\n")
} else {
    cat("None\n")
}

cat("\nFeatures never selected by any method:\n")
if (nrow(never_sel) > 0) {
    cat(paste(never_sel$feature, collapse = ", "), "\n")
} else {
    cat("None - every feature was selected by at least one method\n")
}

# Performance comparison table
perf_summary <- tibble(
    Method     = c("RFE", "Forward", "LASSO", "ElasticNet", "RF_Imp", "LR_all"),
    n_features = c(
        length(rfe_selected),   length(fwd_selected),
        nrow(lasso_selected),   nrow(enet_selected_df),
        length(rf_s3_selected), ncol(X_train_scaled)
    ),
    Accuracy   = c(
        rfe_cm$overall["Accuracy"] * 100,
        fwd_cm$overall["Accuracy"] * 100,
        lasso_cm$overall["Accuracy"] * 100,
        enet_cm$overall["Accuracy"] * 100,
        rf_s3_cm$overall["Accuracy"] * 100,
        lr_cm$overall["Accuracy"] * 100
    ),
    AUC_ROC    = c(
        auc(rfe_roc), auc(fwd_roc),   auc(lasso_roc),
        auc(enet_roc), auc(rf_s3_roc), auc(roc_lr)
    ) * 100
)

cat("\nPerformance comparison (sorted by AUC-ROC):\n")
print(perf_summary %>% arrange(desc(AUC_ROC)), n = Inf)
```

```{r s3-comparison-plots, fig.height=8}
# Heatmap: which method selected which feature
overlap_long <- overlap_df %>%
    filter(n_methods > 0) %>%
    pivot_longer(
        cols      = c(RFE, Forward, LASSO, ElasticNet, RF_Imp),
        names_to  = "method",
        values_to = "selected"
    ) %>%
    mutate(
        feature_label = paste0(feature, "  (", n_methods, "/5)"),
        feature_label = reorder(feature_label, n_methods),
        method        = factor(method, levels = c("RFE", "Forward", "LASSO", "ElasticNet", "RF_Imp"))
    )

ggplot(overlap_long, aes(x = method, y = feature_label, fill = selected)) +
    geom_tile(color = "grey85", linewidth = 0.5) +
    
    scale_fill_manual(
        values = c("TRUE" = "#27ae60", "FALSE" = "white"),
        labels = c("Not selected", "Selected"),
        guide  = "none"
    ) +
    labs(
        title    = "Feature Selection Overlap Across All Methods",
        x = NULL, y = NULL
    ) +
    theme_minimal() +
    theme(
        axis.text.y      = element_text(size = 8),
        axis.text.x      = element_text(size = 9, face = "bold"),
        panel.grid       = element_blank(),
        panel.border     = element_rect(color = "grey85", fill = NA)
    )
```

```{r s3-comparison-auc}
baseline_auc <- perf_summary$AUC_ROC[perf_summary$Method == "LR_all"]

perf_summary %>%
    filter(Method != "LR_all") %>%
    mutate(Method = reorder(Method, AUC_ROC)) %>%
    ggplot(aes(x = AUC_ROC, y = Method)) +
    geom_segment(aes(x = baseline_auc - 1.5, xend = AUC_ROC, yend = Method),
                 color = "grey75", linewidth = 1) +
    geom_point(size = 5, color = "#2ecc71") +
    geom_text(aes(label = sprintf("%d features", n_features)),
              nudge_x = 0.15, hjust = 0, size = 3.5, color = "grey30") +
    geom_vline(xintercept = baseline_auc, linetype = "dashed",
               color = "#e74c3c", linewidth = 0.8) +
    annotate("text", x = baseline_auc + 0.05, y = 0.6,
             label = sprintf("Baseline\n(all %d feat.)", ncol(X_train_scaled)),
             color = "#e74c3c", size = 3, hjust = 0) +
    scale_x_continuous(limits = c(baseline_auc - 1.5, baseline_auc + 1.5)) +
    labs(
        title   = "Feature Selection: AUC-ROC vs. Method",
        x = "AUC-ROC (%)", y = NULL
    ) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
```

Each row is one selection method; the dot position shows AUC-ROC on the test set and the label shows how many features were kept. The red dashed line is the full-feature LR baseline. Methods clustered near the baseline with substantially fewer features demonstrate that the same predictive power is achievable with a smaller, more interpretable set — confirming that much of the original 19-feature set was carrying redundant signal.

## S3 findings

All five selection methods converge on the same answer: **the predictive signal in this dataset is highly concentrated**. Starting from 19 features, every method reduces the set substantially - LASSO and RF Importance down to just 9, RFE to 10 - with almost no loss in AUC-ROC (83.2-84.4% vs. 84.4% for the full-feature baseline). This means roughly half the features in the model are redundant: they carry signal already captured by other features, or no signal at all.

**Four features form the consensus core** - selected by all five methods regardless of the criterion used: `golddiffat15`, `xpdiffat15`, `csdiffat15`, and `winrate_diff`. These are the three resource differentials at 15 minutes plus the pre-game team strength gap. Their universal selection confirms what the EDA correlation heatmap showed: who is ahead in gold, XP, and CS at 15 minutes is the strongest combined signal for match outcome. The `winrate_diff` feature indicates that pre-game team quality adds real information on top of the in-game state.

**Four more features reach near-consensus** (4 out of 5 methods): `firsttower`, `firstherald`, `firstdragon`, and `cs_momentum`. Objective control - particularly tower and herald - contributes a second, independent layer of signal beyond raw economic leads. `cs_momentum` (the acceleration of the CS advantage from @10 to @15) captures information not fully redundant with the snapshot differential at @15.

**Three features were never selected** by any method: `firstblood`, `kill_pressure_10_15`, and `opp_kill_pressure_10_15`. This is a meaningful null result - it suggests that early kill activity, while visible in the data, does not independently predict match outcome once the economic and objective features are accounted for. First blood in particular is consistent with published LoL research, where its win-rate correlation is weak relative to tower or gold leads.

The key practical takeaway is that **9 features achieve 84.3% AUC-ROC** - within 0.1 points of the full-feature LR - which means a model trained on only the consensus + near-consensus features is as useful as one trained on everything, while being substantially easier to interpret and explain.

# Conclusions

## H1 findings

<!-- TODO: summarize early-game prediction results -->

## H2 findings

<!-- TODO: summarize bot lane impact findings -->

## S1 vs S3 synthesis

<!-- TODO: which models + features work best together -->

# Export models for Shiny app

```{r save-models}
dir.create("shiny/models", recursive = TRUE, showWarnings = FALSE)

saveRDS(lr_model,      "shiny/models/lr_model.rds")
saveRDS(rf_model,      "shiny/models/rf_model.rds")
saveRDS(nb_model,      "shiny/models/nb_model.rds")
saveRDS(knn_model,     "shiny/models/knn_model.rds")
saveRDS(cart_model,    "shiny/models/cart_model.rds")
saveRDS(preproc,       "shiny/models/preproc.rds")
saveRDS(perf_summary,  "shiny/models/perf_summary.rds")
saveRDS(overlap_df,    "shiny/models/overlap_df.rds")
saveRDS(games_model,   "shiny/models/games_model.rds")
saveRDS(X_train,       "shiny/models/X_train.rds")
saveRDS(games,         "shiny/models/games.rds")

# Game lookup table for Shiny browser.
# The Shiny app's game-picker is per-side (one Blue and one Red row per game),
# so we un-pivot games_model back into a two-rows-per-game shape. Signed diffs
# are negated for the Red row, per-team kill counts are swapped, and binary
# flags are toggled (when Blue secured the objective, Red did not; otherwise
# we keep the same value, since neither team is recorded as having secured it).
shiny_blue <- games_model %>%
    mutate(side_encoded = 1L, result = blue_win) %>%
    select(-blue_win)

shiny_red <- games_model %>%
    transmute(
        gameid,
        side_encoded = 0L,
        result       = 1L - blue_win,
        golddiffat15 = -golddiffat15,
        xpdiffat15   = -xpdiffat15,
        csdiffat15   = -csdiffat15,
        gold_momentum  = -gold_momentum,
        xp_momentum    = -xp_momentum,
        cs_momentum    = -cs_momentum,
        kill_diff_15   = -kill_diff_15,
        assist_diff_15 = -assist_diff_15,
        kill_pressure_10_15     = opp_kill_pressure_10_15,
        opp_kill_pressure_10_15 = kill_pressure_10_15,
        gold_efficiency_15 = NA_real_,  # Blue-only ratio, no Red counterpart kept
        kda_15             = NA_real_,
        plate_diff   = -plate_diff,
        grub_diff    = -grub_diff,
        winrate_diff = -winrate_diff,
        firstblood   = ifelse(firstblood   == 1L, 0L, firstblood),
        firstdragon  = ifelse(firstdragon  == 1L, 0L, firstdragon),
        firstherald  = ifelse(firstherald  == 1L, 0L, firstherald),
        firsttower   = ifelse(firsttower   == 1L, 0L, firsttower)
    )

per_side_features <- bind_rows(shiny_blue, shiny_red)

# Per-side team labels
team_labels <- teams %>%
    select(gameid, league, date, side, teamname) %>%
    mutate(side_encoded = ifelse(side == "Blue", 1L, 0L),
           side_label   = side) %>%
    select(gameid, league, date, teamname, side_label, side_encoded)

opp_labels <- team_labels %>%
    transmute(gameid,
              side_encoded_opp = 1L - side_encoded,
              opp_teamname     = teamname)

game_lookup <- team_labels %>%
    left_join(opp_labels, by = c("gameid", "side_encoded" = "side_encoded_opp")) %>%
    inner_join(per_side_features, by = c("gameid", "side_encoded")) %>%
    mutate(date         = as.Date(date),
           result_label = ifelse(result == 1, "WIN", "LOSS"))

saveRDS(game_lookup, "shiny/models/game_lookup.rds")

# ROC objects for Shiny Model Comparison tab
roc_lr   <- pROC::roc(y_test_fct, lr_prob[,   "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_rf   <- pROC::roc(y_test_fct, rf_prob[,   "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_nb   <- pROC::roc(y_test_fct, nb_prob[,   "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_knn  <- pROC::roc(y_test_fct, knn_prob[,  "Win"], levels = c("Loss","Win"), quiet = TRUE)
roc_cart <- pROC::roc(y_test_fct, cart_prob[, "Win"], levels = c("Loss","Win"), quiet = TRUE)
saveRDS(list(lr = roc_lr, rf = roc_rf, nb = roc_nb, knn = roc_knn, cart = roc_cart),
        "shiny/models/roc_list.rds")

# LR test predictions for Shiny (confusion matrix + prob distribution)
saveRDS(
    data.frame(actual = as.character(y_test_fct), prob_win = lr_prob[, "Win"]),
    "shiny/models/lr_test_df.rds"
)
saveRDS(lr_cm, "shiny/models/lr_cm.rds")

cat("Models saved to shiny/models/\n")
```
