---
title: "OZNAL Capstone - LoL Esports Early-Game Prediction"
author: "Martin Vanco, Adriana Gemeľová"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.width = 10,
    fig.height = 6
)
```

# Introduction

This analysis investigates two hypotheses about professional League of Legends esports using early-game match data from the 2025 competitive season.

**H1:** *"Early-game performance metrics and objective control can reliably predict match outcome in professional League of Legends."* We test whether gold, XP, and CS differentials at 10 and 15 minutes, combined with first objective captures (first blood, first dragon, first tower, void grubs), contain enough signal to classify wins and losses. We also ask which of these features are the most informative and whether different models agree on what matters.

**H2:** *"Bot lane (ADC) gold advantage is the most impactful role-specific early-game factor for predicting match outcome."* We test whether the bot lane player's early-game differentials carry stronger predictive signal than those of other roles (top, jungle, mid, support). While H1 operates on team-level aggregates, H2 reframes the same data at role level and compares feature importances across models.

## Game primer - what our features actually measure

League of Legends is a 5v5 team game where two teams (Blue side, Red side) fight to destroy each other's base ("Nexus"). A match takes \~30-40 minutes and has three macro phases: **early game** (0-15 min, the focus of this analysis), **mid game** (15-25 min), and **late game** (25+ min). Each team is split into five **roles** on a fixed map: - **Top** - solo-lanes a side lane, durable duelist - **Jungle (jng)** - roams the neutral map, secures objectives, ganks lanes - **Mid** - solo-lanes the central lane, high-impact mage or assassin - **Bot / ADC (bot)** - ranged damage carry, shares a lane with support - **Support (sup)** - peels for and enables the ADC, provides vision and utility

Three resources accumulate continuously and drive almost everything: - **Gold** - earned by killing minions, enemy champions, and objectives. Spent on items that make champions stronger. A team's gold total at 15 min is the main economic snapshot our models use (`goldat15`, `golddiffat15`). - **XP (experience)** - levels champions up, unlocking abilities. Tracked as `xpat15` and `xpdiffat15`. - **CS (creep score)** - number of minions killed, a proxy for lane efficiency. Tracked as `csat15` and `csdiffat15`. Each CS gives both gold and XP, so CS is tightly correlated with gold but captures mechanical consistency rather than team fight outcomes.

On top of these continuous resources, the early game has a set of **one-off objective races**, each of which flips a binary feature in our dataset: - **First blood** (`firstblood`) - the first champion kill of the match. Small gold bonus, large momentum swing. - **First dragon** (`firstdragon`) - dragons spawn at 5:00 in a pit on the bot side of the map; each gives a stacking buff. - **Void grubs** (`void_grubs`, `opp_void_grubs`) - spawn at 5:00 and despawn at 13:45 in the river; up to 6 can be killed and each grub stacks a tower-damage buff. - **First herald** (`firstherald`) - a neutral monster that spawns at 14:00; killing it yields a "Herald" summon that slams into towers for big plate/turret damage. - **First tower** (`firsttower`) and **turret plates** (`turretplates`) - every outer tower has five plates that can be broken until 14:00, each giving gold. Destroying the tower itself gives a large team-wide gold bonus.

Finally there is **side** (Blue / Red), a known small structural asymmetry because Blue gets first pick in the champion draft and slightly different map geography. This is why our dataset has one row per team per game — two mirror rows where one team's `golddiffat15 = +500` is the other team's `-500`.

The key intuition driving H1 is that by 15 minutes these streams have already diverged enough to forecast the winner, even though two-thirds of the game are still to come. H2 narrows that further: among the five roles, which one's early lead carries the most predictive signal?

## Related work

Published work on early-game LoL match prediction at the **15-minute mark** converges tightly: reported accuracy clusters in the **76-78% band** across model families and dataset types. The table below collects the numbers we could verify directly from each source.

| Study                                          | Year | Dataset                           | Model               | Acc. \@15  |
|------------------------------------------------|------|-----------------------------------|---------------------|------------|
| Spaargaren, MSc thesis (Tilburg)               | 2022 | Oracle's Elixir pro, 2021         | Logistic Regression | **77.99%** |
| Spaargaren                                     | 2022 | Oracle's Elixir pro, 2021         | Random Forest       | 77.26%     |
| Spaargaren                                     | 2022 | Oracle's Elixir pro, 2021         | SVM                 | 76.99%     |
| Spaargaren                                     | 2022 | Oracle's Elixir pro, 2021         | Naïve Bayes         | 76.82%     |
| Spaargaren                                     | 2022 | Oracle's Elixir pro, 2021         | kNN (k=9)           | 74.83%     |
| Lafrance & Grewal, *Entertainment Computing*   | 2026 | Riot API solo queue, Bronze       | XGBoost             | 77.1%      |
| Lafrance & Grewal                              | 2026 | Riot API solo queue, Diamond      | XGBoost             | 76.9%      |
| Junior & Campelo, arXiv 2309.02449             | 2023 | Riot API solo queue               | LightGBM            | \~77.9%\*  |
| Junior & Campelo                               | 2023 | Riot API solo queue               | Random Forest       | \~77.8%\*  |
| Junior & Campelo                               | 2023 | Riot API solo queue               | Logistic Regression | \~77.2%\*  |

\*interpolated from their "% elapsed time" buckets (15 min ≈ 50% PET for a \~30-min median match).

Three observations follow from this. First, **professional play is not materially easier to predict than ranked** at 15 minutes: Spaargaren's pro-data Logistic Regression (77.99%) sits within one percentage point of Lafrance & Grewal's solo-queue XGBoost. Second, **algorithm choice matters little** at this timestamp — LR, RF, SVM, Naïve Bayes, and XGBoost all land in the 77 ± 1% band, with only kNN trailing. Third, across every study **gold differential is the strongest single feature**, and first tower or first herald typically ranks next among objective flags.

Three works use Oracle's Elixir professional data directly. Spaargaren (2022) is the cleanest @15-minute benchmark and the one our models compare against most directly. Chen (2024) reports 78.18% with tuned Logistic Regression, but at the **20-minute mark** rather than 15 — his feature set is unavailable at our chosen cutoff. Tsang (2025) does not train a held-out classifier but, by fitting logistic regressions per season across 2020-2025, shows that the coefficient of `golddiffat15` has **declined** from above 1.0 in 2020-2023 to 0.83 in 2025, and that first blood has flipped to a slightly negative win-rate coefficient. This is external evidence that the 2025 season is *harder* to predict from a single gold-at-15 signal than any previous season.

Chowdhury et al. (2025) reach 76.8% on Riot-API data by combining pre-game with in-game features, motivating our use of a **rolling team win rate** as a pre-game strength signal. We additionally include two features absent from the published @15-minute literature: **turret plate differential** (plates fall before 14:00, making them a purely early-game resource) and **void grubs** (added in Season 14, despawning at 13:45). To our knowledge, no peer-reviewed study has yet used the 2025 Oracle's Elixir data - which introduced Atakhan and Feats of Strength - so this work is the first analysis on that season and on those features.

Finally, we note a framing gap in the literature: the trivial rule "team with a gold lead at 15 minutes wins" already delivers \~73% accuracy on professional data (72.9% on our own set). The 76-78% models above are therefore only 3-5 percentage points above this naïve baseline, yet prior work rarely evaluates results explicitly against this heuristic or analyses where the ML gains are concentrated. We adopt this framing directly in our results.

# Setup

```{r libraries}
library(tidyverse)
library(corrplot)
library(caret)
library(glmnet)
library(randomForest)
library(naivebayes)
library(rpart)
library(rpart.plot)
library(class)
library(pROC)
library(gridExtra)
library(doParallel)

cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)
```

# Data Loading and Cleaning

The dataset comes from Oracle's Elixir, a comprehensive source of professional League of Legends esports match data. It contains 120,636 rows across 165 columns, covering 10,053 games from the 2025 competitive season. Each game produces 12 rows - one per player (5 per team) plus one team-level aggregate per side. We focus on 2025 data only, as the game meta (champion balance, objectives, map changes) shifts significantly between seasons.

## Load data

```{r load-data}
raw <- read_csv("data/2025_LoL_esports_match_data_from_OraclesElixir.csv")
glimpse(raw)
```

## Filter to complete games

Not all games have complete early-game timeline data (gold, XP, CS snapshots at 10 and 15 minutes). We filter to `datacompleteness == "complete"` to ensure every row has the full set of features we need. We then split into team-level rows (used for H1 modeling) and player-level rows (used later for H2 role-specific analysis).

```{r filter}
complete <- raw %>% filter(datacompleteness == "complete")

teams <- complete %>% filter(position == "team")
players <- complete %>% filter(position != "team")

cat(sprintf(
    "Total rows: %d\nFiltered to complete: %d (removed %d incomplete, %.1f%%)\nTeam rows: %d\nPlayer rows: %d\nTarget balance: %.1f%% win / %.1f%% loss",
    nrow(raw), nrow(complete),
    nrow(raw) - nrow(complete),
    (nrow(raw) - nrow(complete)) / nrow(raw) * 100,
    nrow(teams), nrow(players),
    mean(teams$result == 1) * 100, mean(teams$result == 0) * 100
))
```

# Exploratory Data Analysis

Before making any decisions about feature engineering or modeling, we explore the raw data to understand what signal exists, which features are worth keeping, and what problems we might face.

## Target distribution

A balanced target means we don't need resampling techniques (SMOTE, oversampling) and can use accuracy as a meaningful metric alongside AUC-ROC. With 18,472 team rows split exactly 50/50 between wins and losses, a naive baseline that always predicts one class would achieve exactly 50% accuracy - any model that beats this is learning real signal.

```{r eda-target}
teams %>%
    mutate(result = factor(result, labels = c("Loss", "Win"))) %>%
    ggplot(aes(x = result, fill = result)) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
    scale_fill_manual(values = c("Loss" = "#e74c3c", "Win" = "#2ecc71")) +
    labs(
        title = "Target Variable Distribution",
        x = "Match Result", y = "Count"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

## Early-game feature correlations with result

We compute correlations between all early-game numeric features and the target variable. This gives us a first look at which features carry predictive signal and which might be redundant.

```{r eda-correlation, fig.width=12, fig.height=10}
early_game_cols <- c(
    "goldat10", "xpat10", "csat10",
    "golddiffat10", "xpdiffat10", "csdiffat10",
    "killsat10", "deathsat10", "assistsat10",
    "goldat15", "xpat15", "csat15",
    "golddiffat15", "xpdiffat15", "csdiffat15",
    "killsat15", "deathsat15", "assistsat15",
    "firstblood", "firstdragon", "firstherald", "firsttower",
    "void_grubs", "opp_void_grubs",
    "result"
)

cor_data <- teams %>%
    select(all_of(early_game_cols)) %>%
    drop_na()

cor_matrix <- cor(cor_data)

corrplot(cor_matrix,
    method = "color", type = "lower",
    tl.cex = 0.7, tl.col = "black",
    col = colorRampPalette(c("#e74c3c", "white", "#2ecc71"))(200),
    title = "Correlation Matrix - Early-Game Features",
    mar = c(0, 0, 2, 0)
)
```

```{r eda-correlation-with-result}
result_cors <- cor_data %>%
    select(-result) %>%
    summarise(across(everything(), ~ cor(.x, cor_data$result))) %>%
    pivot_longer(everything(), names_to = "feature", values_to = "correlation") %>%
    arrange(desc(abs(correlation)))

ggplot(result_cors, aes(
    x = reorder(feature, abs(correlation)),
    y = correlation,
    fill = correlation > 0
)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title = "Feature Correlation with Match Result",
        x = NULL, y = "Pearson Correlation", fill = "Direction"
    ) +
    theme_minimal()
```

```{r}
cat(sprintf(
    "Strongest positive: %s (%.3f)\nStrongest negative: %s (%.3f)",
    result_cors$feature[which.max(result_cors$correlation)],
    max(result_cors$correlation),
    result_cors$feature[which.min(result_cors$correlation)],
    min(result_cors$correlation)
))
```

The differential features (`golddiffat15` r=0.52, `xpdiffat15` r=0.49, `csdiffat15` r=0.45) are the strongest predictors. This makes sense: it's the *gap* between teams that matters. Among objectives, `firsttower` (r=0.40) and `firstherald` (r=0.33) stand out, while `firstblood` (r=0.16) and `firstdragon` (r=0.15) are weaker. The correlation matrix also reveals high multicollinearity between raw stats and their counterparts (e.g. `goldat15` and `golddiffat15`), which will cause issues for Logistic Regression and **motivates regularization methods like LASSO and Elastic Net in Scenario 3.**

## Gold differential distribution

Gold differential at 15 minutes is the strongest single early-game predictor (r=0.52). We check how well it separates wins from losses and whether a simple linear decision boundary (gold lead -\> win) is sufficient, or whether we need non-linear models.

```{r eda-golddiff}
teams %>%
    filter(!is.na(golddiffat15)) %>%
    mutate(result = factor(result, labels = c("Loss", "Win"))) %>%
    ggplot(aes(x = golddiffat15, fill = result)) +
    geom_density(alpha = 0.5) +
    scale_fill_manual(values = c("Loss" = "#e74c3c", "Win" = "#2ecc71")) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title = "Gold Differential at 15 Minutes - Wins vs Losses",
        x = "Gold Diff @15 min", y = "Density", fill = "Result"
    ) +
    theme_minimal()
```

```{r eda-golddiff-10vs15}
teams %>%
    filter(!is.na(golddiffat10), !is.na(golddiffat15)) %>%
    mutate(result = factor(result, labels = c("Loss", "Win"))) %>%
    ggplot(aes(x = golddiffat10, y = golddiffat15, color = result)) +
    geom_point(alpha = 0.15, size = 0.8) +
    scale_color_manual(values = c("Loss" = "#e74c3c", "Win" = "#2ecc71")) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title = "Gold Diff @10 vs @15 - Do Leads Grow?",
        subtitle = "Points far from the diagonal indicate momentum shifts between 10 and 15 min",
        x = "Gold Diff @10 min", y = "Gold Diff @15 min", color = "Result"
    ) +
    theme_minimal()
```

The density plot shows clear separation: winning teams have a mean gold lead of +1345 at 15 minutes, while losing teams sit at -1345. A simple rule ("team with gold lead at 15 min wins") achieves 72.9% accuracy - already well above the 50% baseline, confirming strong signal in this single feature. However, the substantial overlap region around zero means a non-trivial fraction of games aren't decided by gold alone. The @10 vs @15 scatter plot shows that most leads grow between 10 and 15 minutes (points clustered along a steeper-than-diagonal trend), justifying our momentum features that capture this growth.

## First objective win rates

First objectives (first blood, first dragon, first herald, first tower) happen within the first \~14 minutes and are binary indicators. We check the win rate for teams that secure each objective to gauge how much each one matters.

```{r eda-objectives}
obj_cols <- c("firstblood", "firstdragon", "firstherald", "firsttower")


obj_winrates <- teams %>%
    select(all_of(c(obj_cols, "result"))) %>%
    drop_na() %>%
    pivot_longer(all_of(obj_cols), names_to = "objective", values_to = "secured") %>%
    filter(secured == 1) %>%
    group_by(objective) %>%
    summarise(
        win_rate = mean(result) * 100,
        n = n(),
        .groups = "drop"
    ) %>%
    mutate(objective = str_replace(objective, "first", "First ") %>% str_to_title())

ggplot(obj_winrates, aes(
    x = reorder(objective, win_rate),
    y = win_rate, fill = win_rate
)) +
    geom_col() +
    geom_text(aes(label = sprintf("%.1f%% (n=%d)", win_rate, n)),
        hjust = -0.1, size = 3.5
    ) +
    coord_flip() +
    scale_fill_gradient(low = "#f39c12", high = "#2ecc71") +
    scale_y_continuous(limits = c(0, 100)) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    labs(
        title = "Win Rate by First Objective Secured",
        subtitle = "Dashed line = 50% baseline (no advantage)",
        x = NULL, y = "Win Rate (%)"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

First tower (70.1%) and first herald (66.5%) provide a substantial win rate advantage, well above the 50% baseline. First blood (57.8%) and first dragon (57.6%) are weaker. This aligns with the correlation analysis where `firsttower` (r=0.40) ranked much higher than `firstblood` (r=0.16). Tower-related objectives translate more directly into map control and gold, while first blood is often a single lucky skirmish. All four are still worth including as features, but we expect the models to weigh tower and herald much more heavily.

## Void grubs

Void grubs are a 2024+ objective that spawns at 5:00 and despawns at 13:45 - purely early-game. Teams can secure 0 to 6 grubs total. We check whether securing more grubs translates to higher win rates.

```{r eda-grubs}
teams %>%
    filter(!is.na(void_grubs)) %>%
    group_by(void_grubs) %>%
    summarise(
        win_rate = mean(result) * 100,
        n = n(),
        .groups = "drop"
    ) %>%
    ggplot(aes(x = factor(void_grubs), y = win_rate, fill = win_rate)) +
    geom_col() +
    geom_text(aes(label = sprintf("%.1f%%\nn=%d", win_rate, n)),
        vjust = -0.3, size = 3
    ) +
    scale_fill_gradient(low = "#e74c3c", high = "#2ecc71") +
    scale_y_continuous(limits = c(0, 100)) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    labs(
        title = "Win Rate by Number of Void Grubs Secured",
        x = "Void Grubs", y = "Win Rate (%)"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

The relationship is clearly monotonic: 0 grubs = 36.4% win rate, up to 6 grubs = 69.9%. Securing the full set of 6 grubs provides a \~34 percentage point win rate boost over getting none. This confirms that `grub_diff` (our engineered differential) is a meaningful feature. Interestingly, there's a slight dip at 4 grubs (53.2%) compared to 3 (57.5%), likely due to small sample size (n=824) - teams that take exactly 4 are rare.

## Side advantage

In professional LoL, Blue side gets first pick in champion draft and has slight map asymmetries. We check whether this translates into a measurable win rate difference.

```{r eda-side}
side_stats <- teams %>%
    group_by(side) %>%
    summarise(
        win_rate = mean(result) * 100,
        n = n(),
        .groups = "drop"
    )

ggplot(side_stats, aes(x = side, y = win_rate, fill = side)) +
    geom_col(width = 0.5) +
    geom_text(aes(label = sprintf("%.1f%%\nn=%d", win_rate, n)),
        vjust = -0.3, size = 4
    ) +
    scale_fill_manual(values = c("Blue" = "#3498db", "Red" = "#e74c3c")) +
    scale_y_continuous(limits = c(0, 70)) +
    geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
    labs(
        title = "Win Rate by Map Side",
        x = "Side", y = "Win Rate (%)"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

Blue side wins 53.3% of games across 9,236 matches - a consistent \~3 percentage point advantage. While not dramatic, over thousands of games this is statistically significant and justifies including `side` as a feature. The advantage likely comes from first pick in draft and slight camera/map asymmetries.

## Role-specific gold differentials (H2 preview)

For H2 we need to know whether some roles' early-game advantages matter more than others. We compare the correlation between each role's gold differential at 15 minutes and match outcome. If bot lane stands out, that supports H2; if another role dominates, H2 may be rejected.

```{r eda-roles}
role_cors <- players %>%
    filter(position %in% c("top", "jng", "mid", "bot", "sup")) %>%
    select(position, golddiffat15, result) %>%
    drop_na() %>%
    group_by(position) %>%
    summarise(
        correlation = cor(golddiffat15, result),
        mean_diff_win = mean(golddiffat15[result == 1]),
        mean_diff_loss = mean(golddiffat15[result == 0]),
        .groups = "drop"
    ) %>%
    mutate(position = toupper(position))

ggplot(role_cors, aes(
    x = reorder(position, correlation),
    y = correlation, fill = correlation
)) +
    geom_col() +
    geom_text(aes(label = sprintf("r = %.3f", correlation)),
        hjust = -0.1, size = 3.5
    ) +
    coord_flip() +
    scale_y_continuous(limits = c(0, 0.42)) +
    scale_fill_gradient(low = "#f39c12", high = "#2ecc71") +
    labs(
        title = "Correlation: Role Gold Diff @15 vs Match Result",
        subtitle = "Which role's early-game gold advantage best predicts winning?",
        x = NULL, y = "Pearson Correlation with Result"
    ) +
    theme_minimal() +
    theme(legend.position = "none")
```

```{r eda-roles-boxplot}
players %>%
    filter(position %in% c("top", "jng", "mid", "bot", "sup")) %>%
    filter(!is.na(golddiffat15)) %>%
    mutate(
        result = factor(result, labels = c("Loss", "Win")),
        position = toupper(position)
    ) %>%
    ggplot(aes(x = position, y = golddiffat15, fill = result)) +
    geom_boxplot(outlier.alpha = 0.1) +
    scale_fill_manual(values = c("Loss" = "#e74c3c", "Win" = "#2ecc71")) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
        title = "Gold Differential @15 by Role and Result",
        x = "Role", y = "Gold Diff @15 min", fill = "Result"
    ) +
    theme_minimal()
```

Bot lane (ADC) has the highest correlation with match result (r=0.376). This provides early evidence for H2 - bot lane gold advantage does appear to be the most predictive role-specific factor. However, the differences between roles are relatively small (0.376 vs 0.298), so the full modeling analysis in H2 will determine whether this gap is statistically meaningful or within noise. The boxplot confirms that all roles show a clear separation between wins and losses, but bot lane has the widest gap.

## Feature distributions and scaling check

Some models (KNN, Logistic Regression) are sensitive to feature scales. If features have very different ranges, we'll need to standardize them. Tree-based models (Random Forest, CART) are scale-invariant, and Naive Bayes fits each feature's distribution independently so it is unaffected by scaling too.

```{r eda-distributions}
scale_check <- teams %>%
    select(
        goldat15, xpat15, csat15, killsat15,
        golddiffat15, xpdiffat15, csdiffat15,
        void_grubs
    ) %>%
    drop_na() %>%
    pivot_longer(everything(), names_to = "feature", values_to = "value")

ggplot(scale_check, aes(x = value)) +
    geom_histogram(bins = 50, fill = "#3498db", alpha = 0.7) +
    facet_wrap(~feature, scales = "free", ncol = 4) +
    labs(
        title = "Feature Distributions - Scale Check",
        subtitle = "Different scales across features → standardization needed for KNN and Logistic Regression",
        x = "Value", y = "Count"
    ) +
    theme_minimal()
```

The scales are wildly different: `goldat15` ranges from \~20,000 to \~36,000 (sd=1622), while `void_grubs` ranges from 0 to 6 (sd=1.7) and `killsat15` from 0 to 27 (sd=3.0). Without standardization, KNN would be dominated by gold values (which are 1000x larger) and effectively ignore kills and grubs. Logistic Regression coefficients would also be misleading. We will apply standardization (z-score scaling) before fitting KNN and Logistic Regression. Tree-based models (RF, CART) and Naive Bayes don't need this.

## EDA Summary

Key findings that inform our modeling decisions:

1.  **Target is perfectly balanced** (50/50) - no resampling needed, accuracy is a valid metric
2.  **Gold diff @15 is the strongest predictor** (r=0.52) - a simple gold-lead rule gives 72.9% accuracy, setting our baseline
3.  **High multicollinearity** between raw stats and diffs - regularization needed for linear models
4.  **First tower** (70.1% WR) and **first herald** (66.5% WR) are high-impact objectives
5.  **First blood** (57.8% WR) and **first dragon** (57.6% WR) are weak - barely above baseline
6.  **Void grubs** show clear monotonic win rate gradient (36% to 70%) - include `grub_diff`
7.  **Blue side** has a consistent 53.3% win rate - include `side` as feature
8.  **Feature scales differ by 1000x** - must standardize for KNN and Logistic Regression
9.  **Bot lane** has the highest role-specific correlation (r=0.376) - H2 analysis justified
10. **Leads tend to grow** from @10 to @15 - momentum features will capture this dynamic
11. **Turret plate differential** (r=0.413) is a strong purely early-game feature (plates fall at 14:00) not commonly used in the literature
12. **Total dragon/herald counts** are full-game statistics (not early-game) and must not be used as features to avoid data leakage

# Feature Engineering

Based on the EDA findings and experiments with different feature combinations, we now select and engineer features. The key insight from the correlation analysis is that **differential features** (gold diff, xp diff, cs diff) are the strongest predictors - not the raw values. This makes sense: it's the *gap* between teams that matters, not the absolute numbers. Since `golddiffat15 = goldat15 - opp_goldat15`, keeping all three is pure redundancy that causes multicollinearity (the LR rank-deficient warning confirms this). We also add two features not commonly found in the LoL prediction literature: **turret plate differential** (a purely early-game objective, plates fall at 14:00) and **team strength differential** (rolling win rate from prior games), which improved our best tree-model accuracy from 74.7% to 75.3% and AUC from 0.833 to 0.843 in experiments.

Our strategy: - **Keep differentials, drop raw + opponent pairs** - eliminates the `goldat15` / `opp_goldat15` / `golddiffat15` redundancy - **Keep @15 diffs as primary, drop @10 raw** - the @15 snapshot subsumes @10 info. We capture @10 signal through **momentum features** ([diff\@15](mailto:diff@15){.email} - [diff\@10](mailto:diff@10){.email}) instead - **Engineer derived features** - momentum, gold efficiency, KDA, grub differential - **Keep all first objectives** - binary flags with no redundancy - **Add turret plate differential** - plates fall at 14:00, making this a purely early-game feature (r=0.413 with result) - **Add rolling team win rate differential** - the difference between each team's historical win rate (computed from *only* prior games to avoid leakage), capturing pre-game team strength (r=0.263 with result)

## Rolling team win rate

Before engineering in-game features, we compute a **rolling win rate** for each team based on all their prior games in the dataset. This captures pre-game team strength - a strong team with a 70% win rate is more likely to convert an early gold lead into a win than a struggling team at 30%. To avoid data leakage, we only use games played *before* the current one (never the current result or future games). Teams appearing for the first time default to 0.5 (no prior information). We then compute `winrate_diff` = our win rate minus the opponent's win rate, so positive values indicate we are the historically stronger team.

```{r rolling-winrate}
teams <- teams %>%
    arrange(date, gameid) %>%
    group_by(teamid) %>%
    mutate(
        games_before = row_number() - 1,
        wins_before = cumsum(result) - result,
        team_winrate = ifelse(games_before > 0, wins_before / games_before, 0.5)
    ) %>%
    ungroup()

# Build opponent win rate lookup (each game has Blue and Red side)
opp_wr <- teams %>%
    select(gameid, side, team_winrate) %>%
    mutate(opp_side = ifelse(side == "Blue", "Red", "Blue")) %>%
    select(gameid, opp_side, opp_winrate = team_winrate)

cat(sprintf(
    "Teams with 0 prior games (first appearance): %d (%.1f%%)\nMedian prior games per row: %.0f\nwinrate_diff correlation with result: %.3f",
    sum(teams$games_before == 0),
    mean(teams$games_before == 0) * 100,
    median(teams$games_before),
    cor(
        (teams %>% left_join(opp_wr, by = c("gameid", "side" = "opp_side")))$team_winrate -
            (teams %>% left_join(opp_wr, by = c("gameid", "side" = "opp_side")))$opp_winrate,
        teams$result,
        use = "complete.obs"
    )
))
```

## Select and engineer features

```{r select-features}
# We need raw columns for engineering, then drop them
raw_cols_needed <- c(
    "goldat10", "xpat10", "csat10", "goldat15", "xpat15", "csat15",
    "golddiffat10", "xpdiffat10", "csdiffat10",
    "golddiffat15", "xpdiffat15", "csdiffat15",
    "killsat10", "killsat15", "deathsat15", "assistsat15",
    "opp_killsat10", "opp_killsat15", "opp_deathsat15", "opp_assistsat15",
    "firstblood", "firstdragon", "firstherald", "firsttower",
    "void_grubs", "opp_void_grubs",
    "turretplates", "opp_turretplates",
    "side"
)

teams_model <- teams %>%
    select(all_of(c("gameid", "result", "team_winrate", raw_cols_needed))) %>%
    left_join(opp_wr, by = c("gameid", "side" = "opp_side")) %>%
    mutate(opp_winrate = replace_na(opp_winrate, 0.5)) %>%
    drop_na() # safety net — expects 0 removals since datacompleteness == "complete" was applied earlier
# removed the n_before/n_after output — always printed 0 and team row count is already shown in the filter chunk above
```

```{r feature-engineering}
teams_model <- teams_model %>%
    mutate(
        # Momentum: is the lead growing or shrinking between 10 and 15 min?
        gold_momentum = golddiffat15 - golddiffat10,
        xp_momentum = xpdiffat15 - xpdiffat10,
        cs_momentum = csdiffat15 - csdiffat10,

        # Combat pressure in 10-15 min window
        kill_pressure_10_15 = killsat15 - killsat10,
        opp_kill_pressure_10_15 = opp_killsat15 - opp_killsat10,
        kill_diff_15 = killsat15 - opp_killsat15,
        death_diff_15 = deathsat15 - opp_deathsat15,
        assist_diff_15 = assistsat15 - opp_assistsat15,

        # gold_efficiency_15 is the only non-differential feature we keep on purpose.
        # Scenario 3 LASSO eliminates it at both lambda.min and lambda.1se, confirming
        # its signal is redundant with golddiffat15 + csdiffat15. We still include it
        # in S1 so LASSO has a realistic redundant feature to demonstrate elimination on.
        gold_efficiency_15 = goldat15 / pmax(csat15, 1),

        # Void grub differential
        grub_diff = void_grubs - opp_void_grubs,

        # KDA at 15
        kda_15 = (killsat15 + assistsat15) / pmax(deathsat15, 1),

        # Turret plate differential (plates fall at 14:00, purely early-game)
        plate_diff = turretplates - opp_turretplates,

        # Team strength differential (rolling win rate from prior games only)
        winrate_diff = team_winrate - opp_winrate
    )

# Now drop the raw columns we only needed for engineering
# Keep: diffs @15, momentum, engineered, objectives, meta
drop_cols <- c(
    "goldat10", "xpat10", "csat10", "goldat15", "xpat15", "csat15",
    "golddiffat10", "xpdiffat10", "csdiffat10",
    "killsat10", "killsat15", "deathsat15", "assistsat15",
    "opp_killsat10", "opp_killsat15", "opp_deathsat15", "opp_assistsat15",
    "void_grubs", "opp_void_grubs",
    "turretplates", "opp_turretplates",
    "team_winrate", "opp_winrate"
)

teams_model <- teams_model %>%
    select(-all_of(drop_cols))

cat(sprintf(
    "Final feature set: %d features (excluding gameid and result)\nFeatures: %s",
    ncol(teams_model) - 2,
    paste(setdiff(names(teams_model), c("gameid", "result")), collapse = ", ")
))
```

## H2: Role-specific differentials

The EDA showed that different roles have different correlations between gold diff @15 and match outcome. To test H2, we extract per-role gold, XP, and CS differentials from the player-level data and pivot them into wide format, creating columns like `bot_golddiffat15` and `mid_xpdiffat10`. These are joined back to the team-level dataset so we can compare feature importances across roles.

```{r h2-role-features}
role_diffs <- players %>%
    filter(position %in% c("top", "jng", "mid", "bot", "sup")) %>%
    select(
        gameid, side, position,
        golddiffat10, golddiffat15,
        xpdiffat10, xpdiffat15,
        csdiffat10, csdiffat15
    ) %>%
    drop_na() %>%
    pivot_wider(
        id_cols = c(gameid, side),
        names_from = position,
        values_from = c(
            golddiffat10, golddiffat15,
            xpdiffat10, xpdiffat15,
            csdiffat10, csdiffat15
        ),
        names_glue = "{position}_{.value}"
    )

teams_h2 <- teams_model %>%
    left_join(role_diffs, by = c("gameid", "side" = "side")) %>%
    drop_na()

# Encode side in both datasets
teams_model <- teams_model %>%
    mutate(side = ifelse(side == "Blue", 1, 0))
teams_h2 <- teams_h2 %>%
    mutate(side = ifelse(side == "Blue", 1, 0))

cat(sprintf(
    "Role features added: %d\nH2 rows after join + drop_na: %d (removed %d from H1 set, %.1f%%)\nH2 total features: %d",
    ncol(role_diffs) - 2,
    nrow(teams_h2),
    nrow(teams_model) - nrow(teams_h2),
    (nrow(teams_model) - nrow(teams_h2)) / nrow(teams_model) * 100,
    ncol(teams_h2) - 2
))
```

# Train/Test Split

Each game produces two team rows (Blue and Red side) that are mirrors of each other - Blue's `golddiffat15 = +500` is Red's `golddiffat15 = -500`. If we split rows randomly, the model could train on one side and be tested on the other side of the same game, which is data leakage. We split by **game ID** instead, so both sides of every game stay in the same set. This ensures the model is always tested on games it has never seen in any form. The same game IDs are used for the H2 split so both hypotheses are evaluated on the same holdout games.

```{r train-test-split}
set.seed(42)

game_ids <- unique(teams_model$gameid)
train_games <- sample(game_ids, size = floor(0.8 * length(game_ids)))

train_h1 <- teams_model %>% filter(gameid %in% train_games)
test_h1 <- teams_model %>% filter(!gameid %in% train_games)

train_h2 <- teams_h2 %>% filter(gameid %in% train_games)
test_h2 <- teams_h2 %>% filter(!gameid %in% train_games)

# Remove gameid before modeling
train_h1 <- train_h1 %>% select(-gameid)
test_h1 <- test_h1 %>% select(-gameid)
train_h2 <- train_h2 %>% select(-gameid)
test_h2 <- test_h2 %>% select(-gameid)

cat(sprintf(
    "Games: %d train, %d test\nH1 - Train: %d rows, Test: %d rows\nH2 - Train: %d rows, Test: %d rows",
    length(train_games), length(game_ids) - length(train_games),
    nrow(train_h1), nrow(test_h1), nrow(train_h2), nrow(test_h2)
))
```

# Scenario 1: Model Comparison

In this scenario we compare five classification methods on the same data to evaluate how well early-game features predict match outcome, and which model family offers the best trade-off between performance and interpretability.

| Category       | Model                | Why we chose it                                                                      |
|--------------|--------------|--------------------------------------------|
| Statistical/ML | Logistic Regression  | Linear baseline, directly interpretable coefficients                                 |
| Statistical/ML | Random Forest        | Ensemble of trees, captures non-linear relationships, built-in feature importance    |
| Statistical/ML | Naive Bayes          | Probabilistic classifier using Bayes' theorem with a feature-independence assumption |
| Partitioning   | KNN                  | Instance-based, partitions feature space by proximity                                |
| Partitioning   | Decision Tree (CART) | Recursive binary splits, directly visible decision rules                             |

Our baseline from EDA is 72.9% accuracy (simple "gold lead at 15 min = win" rule). Any model that doesn't beat this is not learning useful patterns beyond the single strongest feature. Published research on early-game LoL prediction reports 73-78% accuracy at the 15-minute mark, so we expect our models to land in this range. We evaluate using accuracy, AUC-ROC, precision, recall, and F1-score, with stratified 5-fold cross-validation for hyperparameter tuning.

## Preprocessing

The EDA showed feature scales differ by 1000x (gold \~20,000-36,000 vs void grubs 0-6). KNN and Logistic Regression are distance/magnitude-sensitive, so they need standardized features. Tree-based models (RF, CART) and Naive Bayes are scale-invariant and work on raw values. We prepare both versions here.

```{r s1-preprocess}
# Separate features and target
X_train <- train_h1 %>% select(-result)
y_train <- train_h1$result
X_test <- test_h1 %>% select(-result)
y_test <- test_h1$result

# Standardized version for LR and KNN
preproc <- preProcess(X_train, method = c("center", "scale"))
X_train_scaled <- predict(preproc, X_train)
X_test_scaled <- predict(preproc, X_test)

cat(sprintf(
    "Training features: %d\nTraining rows: %d\nTest rows: %d",
    ncol(X_train), nrow(X_train), nrow(X_test)
))
```

To confirm the scaling actually worked and to expose the per-feature range differences, we print the min/max/mean/sd of every feature before and after standardization. After z-scoring, every feature should have mean ≈ 0 and sd ≈ 1 on the training set, regardless of its original scale.

```{r s1-preprocess-summary}
scale_summary <- function(df, label) {
    df %>%
        summarise(across(everything(),
                         list(min  = ~min(.),
                              max  = ~max(.),
                              mean = ~mean(.),
                              sd   = ~sd(.)))) %>%
        pivot_longer(everything(),
                     names_to  = c("feature", ".value"),
                     names_pattern = "^(.*)_(min|max|mean|sd)$") %>%
        mutate(across(where(is.numeric), ~round(., 3)),
               dataset = label) %>%
        select(dataset, feature, min, max, mean, sd)
}

before <- scale_summary(X_train,        "raw")
after  <- scale_summary(X_train_scaled, "scaled")

cat("Raw training features - note the 1000x scale gap:\n")
print(before, n = Inf)

cat("\nStandardized training features - mean ≈ 0, sd ≈ 1:\n")
print(after, n = Inf)
```

We also set up a shared cross-validation strategy so all models are evaluated on the same folds, making their results directly comparable.

```{r s1-cv-setup}
ctrl <- trainControl(
    method = "cv",
    number = 5,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    savePredictions = "final",
    verboseIter = TRUE
)

# Convert target to factor for caret
y_train_fct <- factor(y_train, levels = c(0, 1), labels = c("Loss", "Win"))
y_test_fct <- factor(y_test, levels = c(0, 1), labels = c("Loss", "Win"))
```

## Logistic Regression

Logistic Regression is our interpretable baseline. It assumes a linear relationship between features and the log-odds of winning. The EDA showed clear linear separation in gold diff distributions, so we expect it to perform reasonably. Its key advantage is that every coefficient directly tells us how much a one-unit increase in a feature changes the win probability - making it the most explainable model in our comparison. We use the standardized features so coefficients are comparable across features.

```{r s1-logistic}
cat("Training Logistic Regression...\n")
t0 <- Sys.time()

lr_model <- train(
    x = X_train_scaled,
    y = y_train_fct,
    method = "glm",
    family = "binomial",
    trControl = ctrl,
    metric = "ROC"
)

lr_pred <- predict(lr_model, X_test_scaled)
lr_prob <- predict(lr_model, X_test_scaled, type = "prob")
lr_cm <- confusionMatrix(lr_pred, y_test_fct)

cat(sprintf(
    "\nLogistic Regression Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    lr_cm$overall["Accuracy"] * 100,
    lr_cm$byClass["Sensitivity"] * 100,
    lr_cm$byClass["Specificity"] * 100,
    max(lr_model$results$ROC)
))
```

```{r s1-logistic-coefs}
lr_coefs <- coef(lr_model$finalModel) %>%
    as.data.frame() %>%
    rownames_to_column("feature") %>%
    rename(coefficient = 2) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

ggplot(
    lr_coefs %>% head(20),
    aes(
        x = reorder(feature, abs(coefficient)),
        y = coefficient,
        fill = coefficient > 0
    )
) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title = "Logistic Regression - Top 20 Coefficients",
        subtitle = "Standardized coefficients - comparable across features",
        x = NULL, y = "Coefficient", fill = "Direction"
    ) +
    theme_minimal()
```

The coefficients reveal which features the linear model relies on most. Since features are standardized, larger absolute coefficients indicate stronger influence on the prediction. Features with positive coefficients increase win probability; negative ones decrease it.

## Random Forest

Random Forest builds an ensemble of decision trees, each trained on a random subset of the data and features. Unlike Logistic Regression, it can capture non-linear relationships and feature interactions (e.g. "first tower + gold lead \> 1000 = very likely win" patterns that LR would miss). The EDA showed non-linear patterns in void grubs (the dip at 4 grubs) and potential interactions between objectives and economy, which RF should handle. We tune `mtry` (number of features considered at each split) via cross-validation.

```{r s1-rf}
cat("Training Random Forest (5 mtry values x 5 folds = 25 fits, 500 trees each)...\n")
t0 <- Sys.time()

rf_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "rf",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(mtry = c(5, 7, 10, 15, 20)),
    ntree = 500
)

rf_pred <- predict(rf_model, X_test)
rf_prob <- predict(rf_model, X_test, type = "prob")
rf_cm <- confusionMatrix(rf_pred, y_test_fct)

cat(sprintf(
    "\nRandom Forest Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity: %.1f%%\nSpecificity: %.1f%%\nBest mtry: %d\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    rf_cm$overall["Accuracy"] * 100,
    rf_cm$byClass["Sensitivity"] * 100,
    rf_cm$byClass["Specificity"] * 100,
    rf_model$bestTune$mtry,
    max(rf_model$results$ROC)
))

# Show CV results across mtry values
cat("\n\nCV results by mtry:\n")
print(rf_model$results %>% select(mtry, ROC, Sens, Spec))
```

```{r s1-rf-importance}
rf_imp <- varImp(rf_model)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))

ggplot(
    rf_imp %>% head(20),
    aes(x = reorder(feature, Overall), y = Overall)
) +
    geom_col(fill = "#27ae60") +
    coord_flip() +
    labs(
        title = "Random Forest - Top 20 Feature Importances",
        subtitle = "Mean decrease in Gini impurity",
        x = NULL, y = "Importance"
    ) +
    theme_minimal()
```

We compare the RF importance ranking with the LR coefficients to see whether both models agree on what matters. Agreement across different model families strengthens our confidence in which features are genuinely informative vs. artifacts of a particular method.

## Naive Bayes

Naive Bayes applies Bayes' theorem under the *naive* assumption that all features are conditionally independent given the class. For a new game, it computes P(Win \| features) ∝ P(Win) · ∏ P(feature_i \| Win) and compares it against the same quantity for Loss — the class with the higher posterior wins. Each numeric feature is modelled with a Gaussian per class, or optionally a kernel density estimate. The independence assumption is clearly violated here (gold, XP, and CS move together), but in practice Naive Bayes can still classify well, because correct classification only requires the *right class* to have the higher posterior, not perfectly calibrated probabilities. We tune whether to use Gaussian vs kernel density (`usekernel`) and the KDE bandwidth multiplier (`adjust`).

```{r s1-nb}
cat("Training Naive Bayes...\n")
t0 <- Sys.time()

nb_grid <- expand.grid(
    usekernel = c(FALSE, TRUE),
    laplace   = 0,
    adjust    = c(0.5, 1.0, 1.5)
)

nb_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "naive_bayes",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = nb_grid
)

nb_pred <- predict(nb_model, X_test)
nb_prob <- predict(nb_model, X_test, type = "prob")
nb_cm <- confusionMatrix(nb_pred, y_test_fct)

cat(sprintf(
    "\nNaive Bayes Results (trained in %.1fs):\nAccuracy: %.1f%%\nSensitivity: %.1f%%\nSpecificity: %.1f%%\nBest: usekernel=%s, adjust=%.1f\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    nb_cm$overall["Accuracy"] * 100,
    nb_cm$byClass["Sensitivity"] * 100,
    nb_cm$byClass["Specificity"] * 100,
    as.character(nb_model$bestTune$usekernel),
    nb_model$bestTune$adjust,
    max(nb_model$results$ROC)
))

cat("\n\nCV results by hyperparameter combo:\n")
print(nb_model$results %>%
    arrange(desc(ROC)) %>%
    select(usekernel, adjust, ROC, Sens, Spec))
```

```{r s1-nb-importance}
# Naive Bayes has no native coefficient-style importance. We use a filter-based
# importance: each feature's single-feature ROC-AUC against the target. A feature
# that separates Win/Loss on its own scores near 1 (or near 0, same information
# in the opposite direction), while a useless feature scores near 0.5. We report
# |AUC - 0.5| so the ranking is direction-agnostic.
# We compute this directly with pROC because caret::varImp's filter fallback has
# a known bug with factor targets for the "naive_bayes" method.
nb_imp <- tibble(
    feature = names(X_train),
    auc_raw = sapply(X_train, function(col) {
        as.numeric(pROC::auc(pROC::roc(y_train_fct, as.numeric(col),
            levels = c("Loss", "Win"), quiet = TRUE
        )))
    })
) %>%
    mutate(Overall = abs(auc_raw - 0.5) * 200) %>% # rescale to 0-100
    select(feature, Overall) %>%
    arrange(desc(Overall))

ggplot(
    nb_imp %>% head(20),
    aes(x = reorder(feature, Overall), y = Overall)
) +
    geom_col(fill = "#e67e22") +
    coord_flip() +
    labs(
        title = "Naive Bayes - Top 20 Feature Importances",
        subtitle = "Filter-based importance: per-feature ROC-AUC against the target",
        x = NULL, y = "Importance (AUC)"
    ) +
    theme_minimal()
```

## KNN (Adriana)

```{r s1-knn}
# TODO - Adriana
# Use X_train_scaled / X_test_scaled (standardized)
# Tune k with ctrl cross-validation

cat("Training KNN...\n")
t0 <- Sys.time()

# Principled K grid: sqrt(n) is the textbook starting point for KNN (Hassanat et al. 2014).
# We log-space the grid from k=5 up to ~4*sqrt(n) so 5-fold CV can find the true optimum
# without us hand-picking values. Forcing odd k avoids 50/50 voting ties in binary classification.
n_train <- nrow(X_train_scaled)
k_theo <- round(sqrt(n_train))
k_grid <- unique(round(exp(seq(log(5), log(k_theo * 4), length.out = 12))))
k_grid <- ifelse(k_grid %% 2 == 0, k_grid + 1, k_grid)

cat(sprintf(
    "sqrt(n) = %d, candidate K grid: %s\n",
    k_theo, paste(k_grid, collapse = ", ")
))

knn_model <- train(
    x = X_train_scaled,
    y = y_train_fct,
    method = "knn",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(k = k_grid)
)

knn_pred <- predict(knn_model, X_test_scaled)
knn_prob <- predict(knn_model, X_test_scaled, type = "prob")
knn_cm <- confusionMatrix(knn_pred, y_test_fct)

cat(sprintf(
    "\nKNN Results (%.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nBest k: %d\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    knn_cm$overall["Accuracy"] * 100,
    knn_cm$byClass["Sensitivity"] * 100,
    knn_cm$byClass["Specificity"] * 100,
    knn_model$bestTune$k,
    max(knn_model$results$ROC)
))
```

## Decision Tree CART (Adriana)

CART (Classification and Regression Trees) recursively splits the feature space into rectangular regions by asking binary questions about individual features (e.g. "is `golddiffat15` \> 500?"). At each node, it selects the split that maximally reduces Gini impurity — a measure of class mixing. Unlike KNN, CART is scale-invariant and fully interpretable: the final model is a set of human-readable rules that directly show which features drive the prediction and at what thresholds. We tune the complexity parameter `cp`, which penalizes tree growth — small `cp` allows deep trees (risk of overfitting), large `cp` forces early stopping (risk of underfitting).

```{r s1-cart}
cat("Training Decision Tree CART...\n")
t0 <- Sys.time()

# cp (complexity parameter) controls tree depth — tuned via 5-fold CV
cart_model <- train(
    x = X_train,
    y = y_train_fct,
    method = "rpart",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(cp = c(0.00001, 0.00005, 0.0001, 0.0005, 0.001))
)

cart_pred <- predict(cart_model, X_test)
cart_prob <- predict(cart_model, X_test, type = "prob")
cart_cm <- confusionMatrix(cart_pred, y_test_fct)

cat(sprintf(
    "\nCART Results (%.1fs):\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nBest cp: %.4f\nCV AUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    cart_cm$overall["Accuracy"] * 100,
    cart_cm$byClass["Sensitivity"] * 100,
    cart_cm$byClass["Specificity"] * 100,
    cart_model$bestTune$cp,
    max(cart_model$results$ROC)
))

cart_imp <- varImp(cart_model)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))
```

```{r s1-cart-plot}
# Tree is too deep to be readable at this pruning level
# rpart.plot(cart_model$finalModel,
#            main = "Decision Tree - Match Outcome Prediction",
#            type = 4,
#            extra = 104)
```

## Model comparison

Now we bring all models together to compare their performance side by side. We look at overall metrics (accuracy, AUC-ROC, F1), confusion matrices, and ROC curves. The key questions: does the non-linearity of RF pay off over the two simple probabilistic baselines (LR, Naive Bayes)? Do all models beat our 72.9% baseline?

```{r s1-metrics-table}
calc_metrics <- function(cm, prob, y_true, name) {
    roc_obj <- roc(y_true, prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
    tibble(
        Model = name,
        Accuracy = cm$overall["Accuracy"] * 100,
        AUC_ROC = auc(roc_obj) * 100,
        Precision = cm$byClass["Pos Pred Value"] * 100,
        Recall = cm$byClass["Sensitivity"] * 100,
        F1 = 2 * cm$byClass["Pos Pred Value"] * cm$byClass["Sensitivity"] /
            (cm$byClass["Pos Pred Value"] + cm$byClass["Sensitivity"]) * 100
    )
}

metrics <- bind_rows(
    calc_metrics(lr_cm, lr_prob, y_test_fct, "Logistic Regression"),
    calc_metrics(rf_cm, rf_prob, y_test_fct, "Random Forest"),
    calc_metrics(cart_cm, cart_prob, y_test_fct, "CART"),
    calc_metrics(nb_cm, nb_prob, y_test_fct, "Naive Bayes"),
    calc_metrics(knn_cm, knn_prob, y_test_fct, "KNN")
)

knitr::kable(metrics, digits = 1, caption = "Model Performance Comparison (H1)")
```

```{r s1-confusion-matrices, fig.width=12, fig.height=4}
plot_cm <- function(cm, title) {
    cm_df <- as.data.frame(cm$table)
    ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
        geom_tile() +
        geom_text(aes(label = Freq), size = 6) +
        scale_fill_gradient(low = "white", high = "#3498db") +
        labs(title = title) +
        theme_minimal() +
        theme(legend.position = "none")
}

gridExtra::grid.arrange(
    plot_cm(lr_cm, "Logistic Regression"),
    plot_cm(rf_cm, "Random Forest"),
    plot_cm(cart_cm, "CART"),
    plot_cm(nb_cm, "Naive Bayes"),
    plot_cm(knn_cm, "KNN"),
    ncol = 5
)
```

Every confusion matrix shows the same visual pattern: **TP ≈ TN** and **FP ≈ FN**. This is a structural consequence of our dataset and split, not a coincidence. Each game contributes two mirror rows (Blue and Red team); nearly every feature is either a differential that flips sign between the two rows, a mutually-exclusive objective flag, or an opposing-team count pair. Our split is by `gameid`, so both rows of every test game stay together and the test set remains exactly 50/50 balanced. For a model trained on these symmetric features, predicting team A's win probability as `p` implies predicting team B's as ≈ `1 - p`, so the two predictions are complementary inside every game. Combined with the ground truth also being complementary (exactly one team won), this pairs every FN on one row with an FP on its mirror row, and likewise TP with TN — mathematically forcing TP = TN and FN = FP. The small gaps you see (e.g. 1390 vs 1388) come from the only feature that is **not** mirrored (`side`) plus a handful of borderline predictions near the 0.5 threshold. This is a useful sanity check: the split introduces no side-leakage, the models are not systematically biased toward Win or Loss, and precision ≈ recall for every model (so F1 ≈ accuracy across the board).

```{r s1-roc-curves}
roc_lr <- roc(y_test_fct, lr_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_rf <- roc(y_test_fct, rf_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_cart <- roc(y_test_fct, cart_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_nb <- roc(y_test_fct, nb_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)
roc_knn <- roc(y_test_fct, knn_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

ggroc(list(
    "Logistic Regression" = roc_lr,
    "Random Forest" = roc_rf,
    "CART" = roc_cart,
    "Naive Bayes" = roc_nb,
    "KNN" = roc_knn
)) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray40") +
    scale_color_manual(values = c(
        "Logistic Regression" = "#3498db",
        "Random Forest" = "#27ae60",
        "CART" = "#9b59b6",
        "Naive Bayes" = "#e67e22",
        "KNN" = "#e74c3c"
    )) +
    labs(
        title = "ROC Curves - Model Comparison",
        subtitle = sprintf(
            "AUC: LR=%.3f, RF=%.3f, CART=%.3f, NB=%.3f, KNN=%.3f",
            auc(roc_lr), auc(roc_rf), auc(roc_cart), auc(roc_nb), auc(roc_knn)
        ),
        x = "Specificity", y = "Sensitivity", color = "Model"
    ) +
    theme_minimal()
```

The ROC curves show each model's trade-off between true positive rate and false positive rate across all classification thresholds. A curve closer to the top-left corner indicates better performance. The diagonal dashed line represents a random classifier (AUC = 0.5).

## Explainability comparison

Different models define "importance" differently: LR uses coefficient magnitude, RF uses mean decrease in Gini impurity, CART uses the total reduction in impurity contributed by each feature's splits, and Naive Bayes has no native importance — so we fall back to a filter-based score (each feature's single-feature ROC-AUC against the target). We compare the top features across all four to identify **consensus features** - those that all models agree are important, regardless of the method. Features that appear in all top-10 lists are the most robust predictors.

```{r s1-explainability}
# KNN excluded - no meaningful feature importance
top_features <- bind_rows(
    lr_coefs %>% mutate(rank = row_number(), model = "LR") %>%
        select(feature, rank, model) %>% head(10),
    rf_imp %>% mutate(rank = row_number(), model = "RF") %>%
        select(feature, rank, model) %>% head(10),
    cart_imp %>% mutate(rank = row_number(), model = "CART") %>%
        select(feature, rank, model) %>% head(10),
    nb_imp %>% mutate(rank = row_number(), model = "NB") %>%
        select(feature, rank, model) %>% head(10)
)

ggplot(top_features, aes(x = reorder(feature, -rank), y = rank, color = model)) +
    geom_jitter(size = 3, width = 0, height = 0.2) +
    coord_flip() +
    scale_y_reverse(breaks = 1:10) +
    scale_color_manual(values = c("LR" = "#3498db", "RF" = "#27ae60", "CART" = "#9b59b6", "NB" = "#e67e22")) +
    labs(
        title = "Feature Importance Rankings - Top 10 per Model",
        subtitle = "Features appearing across multiple models are the most robust predictors",
        x = NULL, y = "Rank (1 = most important)", color = "Model"
    ) +
    theme_minimal()
```

A simple "appears in top-10 of each model" tally is misleading here because the four methods rank features on fundamentally different axes. Instead, we join the **full** feature rankings from every model and report each feature's average rank across the four methods. Features with a low average rank are consistently important across model families; features with high variance in rank reveal where model families disagree.

```{r s1-consensus}
full_ranks <- lr_coefs %>%
    mutate(rank_LR = row_number()) %>%
    select(feature, rank_LR) %>%
    full_join(rf_imp   %>% mutate(rank_RF   = row_number()) %>% select(feature, rank_RF),
              by = "feature") %>%
    full_join(cart_imp %>% mutate(rank_CART = row_number()) %>% select(feature, rank_CART),
              by = "feature") %>%
    full_join(nb_imp   %>% mutate(rank_NB   = row_number()) %>% select(feature, rank_NB),
              by = "feature") %>%
    mutate(
        avg_rank       = rowMeans(across(starts_with("rank_")), na.rm = TRUE),
        rank_spread    = apply(across(starts_with("rank_")), 1,
                               function(r) max(r, na.rm = TRUE) - min(r, na.rm = TRUE)),
        times_in_top10 = rowSums(across(starts_with("rank_"),
                                        ~ replace_na(.x, 999) <= 10))
    ) %>%
    arrange(avg_rank)

options(tibble.print_max = Inf)

cat("Cross-model feature ranking (lower avg_rank = more consistently important):\n")
print(as.data.frame(full_ranks))

cat("\nTrue consensus features (in all 4 top-10s):\n")
full_ranks %>% filter(times_in_top10 == 4) %>% print()

cat("\nStrong majority (in 3 of 4 top-10s):\n")
full_ranks %>% filter(times_in_top10 == 3) %>% print()

cat("\nMixed signal (in 2 of 4 top-10s) - model families disagree:\n")
full_ranks %>% filter(times_in_top10 == 2) %>% print()
```

The true consensus — features ranked highly by **all four** methods — is a compact set of continuous early-game differentials: `golddiffat15`, `xpdiffat15`, `csdiffat15`, and `gold_momentum`. These are the features any defense of H1 should lean on.

The interesting disagreement is around binary objective flags, especially `firstdragon` and `firstherald`, which rank high in Logistic Regression (and LASSO) but low in Random Forest and Naive Bayes. This is a model-family effect, not noise:

-   **Logistic Regression treats features as independent additive log-odds shifts.** A clean binary signal with a 57-58% win rate still produces a meaningful standardized coefficient (\~0.33 for `firstdragon`), so LR ranks it in the top 10.
-   **Tree ensembles split greedily on continuous features first.** A binary feature produces at most one split per tree and competes against continuous features (gold, XP, CS) that admit hundreds of candidate splits. By the time the greedy algorithm considers `firstdragon`, most of its signal is already absorbed by the earlier gold/XP splits, so its Gini-based importance collapses.
-   **Naive Bayes uses filter-based importance (single-feature AUC).** `firstdragon` has marginal AUC ≈ 0.58, which maps to an importance score far below the \~0.82 of `golddiffat15`. NB is simply reading it as a weak standalone predictor.

The takeaway is not that `firstdragon` is irrelevant — LASSO (Scenario 3) keeps it with a non-zero coefficient, and the EDA confirmed a real WR effect. It is that **importance metrics are not comparable across model families**, and that "consensus" should be defined carefully. For H1, the four-way consensus on continuous diffs is the strongest evidence; the two-way LR/CART signal on first objectives is a secondary, model-dependent finding.

## H1 vs H2 comparison

We now repeat the best-performing model on the H2 feature set (which includes role-specific differentials) to test whether knowing *which role* has the gold lead improves prediction over just knowing the team total. If H2 achieves higher AUC-ROC, role-level features add value. We also compare feature importances to see whether bot lane differentials rank highest - the core claim of H2.

```{r s1-h2-setup}
X_train_h2 <- train_h2 %>% select(-result)
y_train_h2 <- factor(train_h2$result, levels = c(0, 1), labels = c("Loss", "Win"))
X_test_h2 <- test_h2 %>% select(-result)
y_test_h2 <- factor(test_h2$result, levels = c(0, 1), labels = c("Loss", "Win"))
```

```{r s1-h2-rf}
rf_h2 <- train(
    x = X_train_h2,
    y = y_train_h2,
    method = "rf",
    trControl = ctrl,
    metric = "ROC",
    tuneGrid = expand.grid(mtry = c(5, 10, 15, 20)),
    ntree = 500
)

rf_h2_pred <- predict(rf_h2, X_test_h2)
rf_h2_prob <- predict(rf_h2, X_test_h2, type = "prob")
rf_h2_cm <- confusionMatrix(rf_h2_pred, y_test_h2)
roc_h2 <- roc(y_test_h2, rf_h2_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf(
    "H1 Random Forest AUC: %.3f\nH2 Random Forest AUC: %.3f\nDifference: %+.3f",
    auc(roc_rf), auc(roc_h2), auc(roc_h2) - auc(roc_rf)
))
```

```{r s1-h2-role-importance}
h2_imp <- varImp(rf_h2)$importance %>%
    rownames_to_column("feature") %>%
    arrange(desc(Overall))

# Highlight role-specific features
role_features <- h2_imp %>%
    filter(str_detect(feature, "^(top|jng|mid|bot|sup)_"))

ggplot(
    h2_imp %>% head(25),
    aes(
        x = reorder(feature, Overall), y = Overall,
        fill = str_detect(feature, "^(top|jng|mid|bot|sup)_")
    )
) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("FALSE" = "#95a5a6", "TRUE" = "#e74c3c"),
        labels = c("Team-level", "Role-specific")
    ) +
    labs(
        title = "H2 Random Forest - Top 25 Feature Importances",
        subtitle = "Do role-specific features (red) rank among the most important?",
        x = NULL, y = "Importance", fill = "Feature type"
    ) +
    theme_minimal()
```

```{r s1-h2-bot-rank}
bot_features <- role_features %>%
    filter(str_detect(feature, "^bot_")) %>%
    mutate(overall_rank = match(feature, h2_imp$feature))

other_role_features <- role_features %>%
    filter(!str_detect(feature, "^bot_")) %>%
    group_by(role = str_extract(feature, "^[a-z]+")) %>%
    summarise(
        best_rank = min(match(feature, h2_imp$feature)),
        avg_importance = mean(Overall),
        .groups = "drop"
    )

cat("Bot lane feature rankings in H2 model:\n")
print(bot_features)
cat("\nBest rank per role:\n")
print(other_role_features)
```

Based on these results, we can assess whether H2 is supported: does bot lane gold differential consistently rank as the most impactful role-specific feature, and does adding role-level features improve overall prediction?

# Scenario 3: Feature Selection (Adriana)

The EDA revealed high multicollinearity between raw stats and their differentials (e.g. `goldat15` and `golddiffat15`), and between @10 and @15 versions of the same metric. With \~46 features, many are likely redundant - carrying the same signal in different forms. This motivates feature selection to find the minimal set of truly informative features. LASSO and Elastic Net are particularly relevant here, as they can zero out redundant features from correlated groups.

## RFE (Recursive Feature Elimination)

RFE repeatedly trains a model, ranks features by importance, removes the weakest, and repeats — narrowing the feature set until performance stops improving. We use Logistic Regression as the base estimator so importance is defined by coefficient magnitude on standardized features. Subset sizes from 5 to the full feature set are tested; 5-fold cross-validation picks the size that maximises accuracy. After RFE identifies the optimal subset, a fresh LR is trained on only those features and evaluated on the held-out test set for a clean performance estimate.

```{r s3-rfe}
cat("Running RFE with Logistic Regression estimator (5-fold CV)...\n")
t0 <- Sys.time()

rfe_ctrl <- rfeControl(
    functions = lrFuncs,
    method    = "cv",
    number    = 5,
    verbose   = FALSE
)

n_features   <- ncol(X_train_scaled)
sizes_to_try <- unique(c(5, 10, 15, 20, 25, 30, n_features))

rfe_result <- rfe(
    x          = X_train_scaled,
    y          = y_train_fct,
    sizes      = sizes_to_try,
    rfeControl = rfe_ctrl
)

rfe_selected <- predictors(rfe_result)

# Retrain LR on selected features for a clean test-set evaluation
rfe_lr <- train(
    x         = X_train_scaled[, rfe_selected, drop = FALSE],
    y         = y_train_fct,
    method    = "glm",
    family    = "binomial",
    trControl = trainControl(method = "none", classProbs = TRUE)
)
rfe_pred_class <- predict(rfe_lr, X_test_scaled[, rfe_selected, drop = FALSE])
rfe_pred_prob  <- predict(rfe_lr, X_test_scaled[, rfe_selected, drop = FALSE], type = "prob")
rfe_cm  <- confusionMatrix(rfe_pred_class, y_test_fct)
rfe_roc <- roc(y_test_fct, rfe_pred_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("\nRFE Results (%.1fs):\nOptimal features: %d / %d\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    length(rfe_selected), n_features,
    rfe_cm$overall["Accuracy"] * 100,
    rfe_cm$byClass["Sensitivity"] * 100,
    rfe_cm$byClass["Specificity"] * 100,
    auc(rfe_roc)))

cat("\nSelected features:\n")
cat(paste(rfe_selected, collapse = ", "), "\n")
```

```{r s3-rfe-plot}
plot(rfe_result, type = c("g", "o"),
     main = "RFE: Cross-validated Accuracy vs Number of Features")

rfe_coef_df <- coef(rfe_lr$finalModel) %>%
    as.data.frame() %>%
    rownames_to_column("feature") %>%
    rename(coefficient = 2) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

ggplot(rfe_coef_df,
       aes(x = reorder(feature, abs(coefficient)), y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("Negative", "Positive")) +
    labs(title = "RFE - Selected Features and Coefficients",
         subtitle = sprintf("Optimal subset: %d features out of %d", length(rfe_selected), n_features),
         x = NULL, y = "Coefficient", fill = "Direction") +
    theme_minimal()
```

The first plot shows the accuracy-vs-size curve: the elbow marks the optimal subset. Adding more features beyond this point brings diminishing or negative returns. The coefficient plot ranks the selected features by their absolute contribution in the final LR model.

## Forward Stepwise Selection

Forward Stepwise starts with an intercept-only model and greedily adds the feature that most improves AIC at each step, stopping when no addition helps. AIC balances model fit against complexity by penalising extra parameters, so the algorithm naturally limits how many features get added without any manual threshold. Unlike RFE — which re-evaluates all sizes via cross-validation — stepwise selection is a single greedy pass and is therefore much faster. A key difference from LASSO: once a feature is added it stays; there is no shrinkage of redundant coefficients.

```{r s3-forward}
cat("Running Forward Stepwise Selection (AIC criterion)...\n")
t0 <- Sys.time()

fwd_train_df        <- X_train_scaled
fwd_train_df$result <- y_train_fct

null_model   <- glm(result ~ 1, data = fwd_train_df, family = "binomial")
full_formula <- as.formula(paste("result ~", paste(names(X_train_scaled), collapse = " + ")))

fwd_model <- MASS::stepAIC(
    null_model,
    scope     = list(lower = null_model, upper = full_formula),
    direction = "forward",
    trace     = FALSE
)

fwd_selected <- names(coef(fwd_model))
fwd_selected <- fwd_selected[fwd_selected != "(Intercept)"]

fwd_pred_prob_raw <- predict(fwd_model, newdata = X_test_scaled, type = "response")
fwd_pred_class    <- factor(ifelse(fwd_pred_prob_raw > 0.5, "Win", "Loss"),
                             levels = c("Loss", "Win"))
fwd_cm  <- confusionMatrix(fwd_pred_class, y_test_fct)
fwd_roc <- roc(y_test_fct, fwd_pred_prob_raw, levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("\nForward Stepwise Results (%.1fs):\nFeatures selected: %d / %d\nFinal AIC: %.1f\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    length(fwd_selected), ncol(X_train_scaled),
    AIC(fwd_model),
    fwd_cm$overall["Accuracy"] * 100,
    fwd_cm$byClass["Sensitivity"] * 100,
    fwd_cm$byClass["Specificity"] * 100,
    auc(fwd_roc)))

cat("\nSelected features:\n")
cat(paste(fwd_selected, collapse = ", "), "\n")
```

```{r s3-forward-plot}
fwd_coef_df <- data.frame(
    feature     = fwd_selected,
    coefficient = coef(fwd_model)[fwd_selected]
) %>%
    arrange(desc(abs(coefficient)))

ggplot(fwd_coef_df,
       aes(x = reorder(feature, abs(coefficient)), y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("Negative", "Positive")) +
    labs(title = "Forward Stepwise - Selected Features and Coefficients",
         subtitle = sprintf("AIC-based selection: %d features out of %d",
                            length(fwd_selected), ncol(X_train_scaled)),
         x = NULL, y = "Coefficient", fill = "Direction") +
    theme_minimal()
```

Because features are standardized, coefficient magnitudes are directly comparable — the AIC criterion selects the same high-signal features we expect from the correlation heatmap.

## LASSO (L1)

LASSO (Least Absolute Shrinkage and Selection Operator) is logistic regression with an added penalty for large coefficients. Unlike standard logistic regression, which keeps all features, LASSO can shrink individual coefficients all the way to exactly zero - effectively removing those features from the model. This makes it both a classifier and a feature selector in one step. It is particularly useful here because our features have high multicollinearity: when two features carry the same signal (e.g. `golddiffat15` and `gold_momentum`), LASSO tends to keep the stronger one and zero out the weaker. We tune `lambda` - the penalty strength - via cross-validation: too low and nothing gets eliminated, too high and useful features get removed.

```{r s3-lasso}
cat("Running LASSO with 5-fold CV to find optimal lambda...\n")
t0 <- Sys.time()

lasso_cv <- cv.glmnet(
    x = as.matrix(X_train_scaled),
    y = y_train,
    alpha = 1, # alpha=1 is pure LASSO (L1 penalty)
    family = "binomial",
    nfolds = 5
)

# lambda.1se: the largest lambda within 1 SE of the minimum - gives a sparser model
# that is nearly as accurate as lambda.min but keeps fewer features
lasso_coefs <- coef(lasso_cv, s = "lambda.1se")

lasso_coef_df <- data.frame(
    feature = rownames(lasso_coefs),
    coefficient = as.numeric(lasso_coefs)
) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

lasso_selected <- lasso_coef_df %>% filter(coefficient != 0)
lasso_eliminated <- lasso_coef_df %>% filter(coefficient == 0)

# test set evaluation
lasso_pred_prob <- predict(lasso_cv,
    newx = as.matrix(X_test_scaled),
    s = "lambda.1se", type = "response"
)
lasso_pred_class <- factor(ifelse(lasso_pred_prob > 0.5, "Win", "Loss"),
    levels = c("Loss", "Win")
)
lasso_cm <- confusionMatrix(lasso_pred_class, y_test_fct)
lasso_roc <- roc(y_test_fct, as.numeric(lasso_pred_prob),
    levels = c("Loss", "Win"), quiet = TRUE
)

cat(sprintf(
    "\nLASSO Results (%.1fs):\nLambda (1se): %.6f\nFeatures kept: %d / %d (eliminated %d)\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    lasso_cv$lambda.1se,
    nrow(lasso_selected), nrow(lasso_coef_df), nrow(lasso_eliminated),
    lasso_cm$overall["Accuracy"] * 100,
    lasso_cm$byClass["Sensitivity"] * 100,
    lasso_cm$byClass["Specificity"] * 100,
    auc(lasso_roc)
))

cat("\nKept features:\n")
cat(paste(lasso_selected$feature, collapse = ", "), "\n")

cat("\nEliminated features:\n")
cat(paste(lasso_eliminated$feature, collapse = ", "), "\n")
```

```{r s3-lasso-plot}
plot(lasso_cv, main = "LASSO: CV Error vs Lambda")

lasso_selected %>%
    mutate(feature = reorder(feature, abs(coefficient))) %>%
    ggplot(aes(x = feature, y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(
        values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
        labels = c("Negative", "Positive")
    ) +
    labs(
        title = "LASSO - Surviving Features and Coefficients",
        subtitle = sprintf(
            "lambda.1se: %d features kept out of %d",
            nrow(lasso_selected), nrow(lasso_coef_df)
        ),
        x = NULL, y = "Coefficient", fill = "Direction"
    ) +
    theme_minimal()
```

The CV plot shows how model error changes as lambda increases (fewer features). The left dashed line marks `lambda.min` (lowest error), the right marks `lambda.1se` (our chosen lambda - slightly sparser, nearly as accurate). The coefficient plot shows which features survived and in which direction: green bars increase win probability, red bars decrease it.

## Elastic Net (L1 + L2)

Elastic Net combines L1 (LASSO) and L2 (Ridge) penalties in a single term. The mix is controlled by `alpha`: `alpha = 1` is pure LASSO, `alpha = 0` is pure Ridge, values in between give Elastic Net. Where LASSO tends to pick one feature from a correlated group and zero out the rest, Elastic Net can keep several features from the same group at reduced coefficients — potentially more stable when features are strongly correlated, as ours are (e.g. `goldat15` and `golddiffat15`). We search the full alpha range from 0 (Ridge) to 1 (LASSO) in 0.05 steps, and for each find the optimal lambda via 5-fold CV. For every alpha we also record the CV binomial deviance at `lambda.1se` and its standard error, so we can judge whether the "best" alpha is meaningfully better than its neighbours or the curve is flat within 1 SE.

```{r s3-elasticnet}
cat("Running Elastic Net - searching alpha grid (0 to 1, step 0.05)...\n")
t0 <- Sys.time()

alpha_grid <- seq(0, 1, by = 0.05)

enet_cv_results <- map_dfr(alpha_grid, function(a) {
    cv_fit     <- cv.glmnet(
        x      = as.matrix(X_train_scaled),
        y      = y_train,
        alpha  = a,
        family = "binomial",
        nfolds = 5
    )
    idx_1se    <- which(cv_fit$lambda == cv_fit$lambda.1se)
    pred_prob  <- predict(cv_fit, newx = as.matrix(X_test_scaled),
                          s = "lambda.1se", type = "response")
    pred_class <- factor(ifelse(pred_prob > 0.5, "Win", "Loss"), levels = c("Loss", "Win"))
    cm         <- confusionMatrix(pred_class, y_test_fct)
    roc_obj    <- roc(y_test_fct, as.numeric(pred_prob), levels = c("Loss", "Win"), quiet = TRUE)
    n_kept     <- sum(coef(cv_fit, s = "lambda.1se") != 0) - 1
    tibble(
        alpha      = a,
        lambda_1se = cv_fit$lambda.1se,
        cv_dev     = cv_fit$cvm[idx_1se],
        cv_dev_se  = cv_fit$cvsd[idx_1se],
        n_features = n_kept,
        accuracy   = cm$overall["Accuracy"] * 100,
        auc        = as.numeric(auc(roc_obj))
    )
})

best_row   <- enet_cv_results %>% arrange(cv_dev) %>% dplyr::slice(1)
best_alpha <- best_row$alpha
best_dev   <- best_row$cv_dev
best_se    <- best_row$cv_dev_se
within_1se <- enet_cv_results %>% filter(cv_dev <= best_dev + best_se)

cat(sprintf("\nAlpha grid diagnostic: %d of %d alphas sit within 1 SE of the best (%.3f <= %.3f).\n",
            nrow(within_1se), nrow(enet_cv_results), best_dev, best_dev + best_se))
cat(sprintf("Alphas within 1 SE: %s\n",
            paste(sprintf("%.2f", within_1se$alpha), collapse = ", ")))
if (nrow(within_1se) >= nrow(enet_cv_results) - 2) {
    cat("-> CV curve is effectively flat across alpha; the L1/L2 mix is underdetermined by this data.\n")
}

enet_cv_final <- cv.glmnet(
    x      = as.matrix(X_train_scaled),
    y      = y_train,
    alpha  = best_alpha,
    family = "binomial",
    nfolds = 5
)

enet_coef_raw <- coef(enet_cv_final, s = "lambda.1se")
enet_coef_df  <- data.frame(
    feature     = rownames(enet_coef_raw),
    coefficient = as.numeric(enet_coef_raw)
) %>%
    filter(feature != "(Intercept)") %>%
    arrange(desc(abs(coefficient)))

enet_selected_df <- enet_coef_df %>% filter(coefficient != 0)
enet_selected    <- enet_selected_df$feature

enet_pred_prob  <- predict(enet_cv_final, newx = as.matrix(X_test_scaled),
                            s = "lambda.1se", type = "response")
enet_pred_class <- factor(ifelse(enet_pred_prob > 0.5, "Win", "Loss"),
                           levels = c("Loss", "Win"))
enet_cm  <- confusionMatrix(enet_pred_class, y_test_fct)
enet_roc <- roc(y_test_fct, as.numeric(enet_pred_prob), levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("\nElastic Net Results (%.1fs):\nBest alpha: %.1f (0=Ridge, 1=LASSO)\nLambda (1se): %.6f\nFeatures kept: %d / %d (eliminated %d)\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    as.numeric(Sys.time() - t0, units = "secs"),
    best_alpha,
    enet_cv_final$lambda.1se,
    nrow(enet_selected_df), nrow(enet_coef_df), nrow(enet_coef_df) - nrow(enet_selected_df),
    enet_cm$overall["Accuracy"] * 100,
    enet_cm$byClass["Sensitivity"] * 100,
    enet_cm$byClass["Specificity"] * 100,
    auc(enet_roc)))

cat("\nKept features:\n")
cat(paste(enet_selected, collapse = ", "), "\n")

cat("\nEliminated features:\n")
cat(paste(enet_coef_df %>% filter(coefficient == 0) %>% pull(feature), collapse = ", "), "\n")
```

```{r s3-elasticnet-plot}
ggplot(enet_cv_results, aes(x = alpha, y = cv_dev)) +
    geom_ribbon(aes(ymin = cv_dev - cv_dev_se, ymax = cv_dev + cv_dev_se),
                fill = "#1abc9c", alpha = 0.2) +
    geom_hline(yintercept = best_dev + best_se, linetype = "dashed", color = "#7f8c8d") +
    geom_line(color = "#1abc9c", linewidth = 1.2) +
    geom_point(size = 3, color = "#1abc9c") +
    geom_point(data = filter(enet_cv_results, alpha == best_alpha),
               size = 5, shape = 21, fill = "#e74c3c", color = "white", stroke = 1.5) +
    labs(title = "Elastic Net: CV Binomial Deviance across Alpha Values",
         subtitle = sprintf("Best alpha = %.2f (red dot); dashed line = best + 1 SE. Points below the line are statistically indistinguishable from the best.",
                            best_alpha),
         x = "Alpha (0 = Ridge, 1 = LASSO)", y = "5-fold CV Deviance at lambda.1se") +
    theme_minimal()

enet_selected_df %>%
    mutate(feature = reorder(feature, abs(coefficient))) %>%
    ggplot(aes(x = feature, y = coefficient, fill = coefficient > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("Negative", "Positive")) +
    labs(title = "Elastic Net - Surviving Features and Coefficients",
         subtitle = sprintf("alpha = %.1f, lambda.1se: %d features kept out of %d",
                            best_alpha, nrow(enet_selected_df), nrow(enet_coef_df)),
         x = NULL, y = "Coefficient", fill = "Direction") +
    theme_minimal()
```

The alpha search plot shows the 5-fold CV deviance at `lambda.1se` across the full 0–1 range, with the shaded band covering ±1 SE and the dashed line marking `best + 1 SE`. Every alpha whose point sits on or below that line is statistically indistinguishable from the best — so if many points fall below it, the L1/L2 mix is effectively underdetermined by this data and the "best alpha" is a tiebreak rather than a real optimum. A sharp U- or V-shape, by contrast, would indicate the data genuinely benefits from a specific mix. The coefficient plot mirrors the LASSO plot above; differences between the two reveal which features LASSO eliminated but Elastic Net retained due to correlated grouping.

## Random Forest Importance

The Random Forest trained in S1 generates feature importance scores as a by-product of training — no additional computation is needed. We use these scores as a filter: features with importance above the mean are retained, the rest discarded. This is the fastest and least tuning-intensive selection method: it piggybacks on the RF already trained for S1, requires no new cross-validation, and handles correlated features gracefully — RF distributes importance among correlated features rather than zeroing all but one like LASSO. To get a comparable performance estimate, we train a simple LR on the selected subset and evaluate it on the held-out test set.

```{r s3-rf-importance}
rf_importance_threshold <- mean(rf_imp$Overall)
rf_s3_selected <- rf_imp %>%
    filter(Overall > rf_importance_threshold) %>%
    pull(feature)

cat(sprintf("RF importance threshold: %.2f (mean importance)\nFeatures selected: %d / %d\n",
    rf_importance_threshold, length(rf_s3_selected), nrow(rf_imp)))

rf_s3_lr <- train(
    x         = X_train_scaled[, rf_s3_selected, drop = FALSE],
    y         = y_train_fct,
    method    = "glm",
    family    = "binomial",
    trControl = trainControl(method = "none", classProbs = TRUE)
)
rf_s3_pred_class <- predict(rf_s3_lr, X_test_scaled[, rf_s3_selected, drop = FALSE])
rf_s3_pred_prob  <- predict(rf_s3_lr, X_test_scaled[, rf_s3_selected, drop = FALSE], type = "prob")
rf_s3_cm  <- confusionMatrix(rf_s3_pred_class, y_test_fct)
rf_s3_roc <- roc(y_test_fct, rf_s3_pred_prob[, "Win"], levels = c("Loss", "Win"), quiet = TRUE)

cat(sprintf("LR on RF-selected features:\nAccuracy: %.1f%%\nSensitivity (Win recall): %.1f%%\nSpecificity (Loss recall): %.1f%%\nAUC-ROC: %.3f",
    rf_s3_cm$overall["Accuracy"] * 100,
    rf_s3_cm$byClass["Sensitivity"] * 100,
    rf_s3_cm$byClass["Specificity"] * 100,
    auc(rf_s3_roc)))

cat("\nSelected features:\n")
cat(paste(rf_s3_selected, collapse = ", "), "\n")
```

```{r s3-rf-importance-plot}
rf_imp %>%
    mutate(
        selected = Overall > rf_importance_threshold,
        feature  = reorder(feature, Overall)
    ) %>%
    ggplot(aes(x = feature, y = Overall, fill = selected)) +
    geom_col() +
    coord_flip() +
    geom_hline(yintercept = rf_importance_threshold, linetype = "dashed",
               color = "#e74c3c", linewidth = 0.8) +
    scale_fill_manual(values = c("FALSE" = "#bdc3c7", "TRUE" = "#27ae60"),
                      labels = c("Eliminated", "Selected")) +
    labs(title = "RF Importance - Feature Selection Threshold",
         subtitle = sprintf("Threshold = mean importance (%.2f) — %d features selected out of %d",
                            rf_importance_threshold, length(rf_s3_selected), nrow(rf_imp)),
         x = NULL, y = "Importance", fill = "Status") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 7))
```

The dashed red line marks the mean importance threshold. Features above it (green) are retained; those below (grey) are discarded. Note that RF distributes importance among correlated features — so unlike LASSO, several correlated gold metrics may all survive with reduced but non-zero importance.

## Feature selection comparison

```{r s3-comparison}
selection_results <- list(
    RFE        = rfe_selected,
    Forward    = fwd_selected,
    LASSO      = lasso_selected$feature,
    ElasticNet = enet_selected,
    RF_Imp     = rf_s3_selected
)

all_features <- colnames(X_train_scaled)

# For each feature, record which methods selected it
overlap_df <- map_dfr(all_features, function(f) {
    sel <- map_lgl(selection_results, ~ f %in% .x)
    tibble(
        feature    = f,
        RFE        = sel[["RFE"]],
        Forward    = sel[["Forward"]],
        LASSO      = sel[["LASSO"]],
        ElasticNet = sel[["ElasticNet"]],
        RF_Imp     = sel[["RF_Imp"]],
        n_methods  = sum(sel)
    )
}) %>%
    arrange(desc(n_methods), feature)

consensus_5 <- overlap_df %>% filter(n_methods == 5)
consensus_4 <- overlap_df %>% filter(n_methods == 4)
never_sel   <- overlap_df %>% filter(n_methods == 0)

cat("Feature selection summary across all 5 methods:\n")
cat(sprintf("  Total features:                %d\n", length(all_features)))
cat(sprintf("  Selected by all 5 (consensus): %d\n", nrow(consensus_5)))
cat(sprintf("  Selected by 4 methods:         %d\n", nrow(consensus_4)))
cat(sprintf("  Never selected:                %d\n\n", nrow(never_sel)))

cat("Consensus features (all 5 methods agree):\n")
if (nrow(consensus_5) > 0) {
    cat(paste(consensus_5$feature, collapse = ", "), "\n")
} else {
    cat("None\n")
}

cat("\nFeatures selected by 4 methods:\n")
if (nrow(consensus_4) > 0) {
    cat(paste(consensus_4$feature, collapse = ", "), "\n")
} else {
    cat("None\n")
}

cat("\nFeatures never selected by any method:\n")
if (nrow(never_sel) > 0) {
    cat(paste(never_sel$feature, collapse = ", "), "\n")
} else {
    cat("None — every feature was selected by at least one method\n")
}

# Performance comparison table
perf_summary <- tibble(
    Method     = c("RFE", "Forward", "LASSO", "ElasticNet", "RF_Imp", "LR_all"),
    n_features = c(
        length(rfe_selected),   length(fwd_selected),
        nrow(lasso_selected),   nrow(enet_selected_df),
        length(rf_s3_selected), ncol(X_train_scaled)
    ),
    Accuracy   = c(
        rfe_cm$overall["Accuracy"] * 100,
        fwd_cm$overall["Accuracy"] * 100,
        lasso_cm$overall["Accuracy"] * 100,
        enet_cm$overall["Accuracy"] * 100,
        rf_s3_cm$overall["Accuracy"] * 100,
        lr_cm$overall["Accuracy"] * 100
    ),
    AUC_ROC    = c(
        auc(rfe_roc), auc(fwd_roc),   auc(lasso_roc),
        auc(enet_roc), auc(rf_s3_roc), auc(roc_lr)
    ) * 100
)

cat("\nPerformance comparison (sorted by AUC-ROC):\n")
print(perf_summary %>% arrange(desc(AUC_ROC)), n = Inf)
```

```{r s3-comparison-plots, fig.height=8}
# Heatmap: which method selected which feature
overlap_long <- overlap_df %>%
    filter(n_methods > 0) %>%
    pivot_longer(
        cols      = c(RFE, Forward, LASSO, ElasticNet, RF_Imp),
        names_to  = "method",
        values_to = "selected"
    ) %>%
    mutate(
        feature_label = paste0(feature, "  (", n_methods, "/5)"),
        feature_label = reorder(feature_label, n_methods),
        method        = factor(method, levels = c("RFE", "Forward", "LASSO", "ElasticNet", "RF_Imp"))
    )

ggplot(overlap_long, aes(x = method, y = feature_label, fill = selected)) +
    geom_tile(color = "grey85", linewidth = 0.5) +
    
    scale_fill_manual(
        values = c("TRUE" = "#27ae60", "FALSE" = "white"),
        labels = c("Not selected", "Selected"),
        guide  = "none"
    ) +
    labs(
        title    = "Feature Selection Overlap Across All Methods",
        subtitle = "Each row shows which methods selected that feature  ·  (x/5) = number of methods",
        x = NULL, y = NULL
    ) +
    theme_minimal() +
    theme(
        axis.text.y      = element_text(size = 8),
        axis.text.x      = element_text(size = 9, face = "bold"),
        panel.grid       = element_blank(),
        panel.border     = element_rect(color = "grey85", fill = NA)
    )
```

```{r s3-comparison-auc}
baseline_auc <- perf_summary$AUC_ROC[perf_summary$Method == "LR_all"]

perf_summary %>%
    filter(Method != "LR_all") %>%
    mutate(Method = reorder(Method, AUC_ROC)) %>%
    ggplot(aes(x = AUC_ROC, y = Method)) +
    geom_segment(aes(x = baseline_auc - 1.5, xend = AUC_ROC, yend = Method),
                 color = "grey75", linewidth = 1) +
    geom_point(size = 5, color = "#2ecc71") +
    geom_text(aes(label = sprintf("%d features", n_features)),
              nudge_x = 0.15, hjust = 0, size = 3.5, color = "grey30") +
    geom_vline(xintercept = baseline_auc, linetype = "dashed",
               color = "#e74c3c", linewidth = 0.8) +
    annotate("text", x = baseline_auc + 0.05, y = 0.6,
             label = sprintf("Baseline\n(all %d feat.)", ncol(X_train_scaled)),
             color = "#e74c3c", size = 3, hjust = 0) +
    scale_x_continuous(limits = c(baseline_auc - 1.5, baseline_auc + 1.5)) +
    labs(
        title   = "Feature Selection: AUC-ROC vs. Method",
        subtitle = "Red dashed line = LR trained on all features  ·  Labels show number of features used",
        x = "AUC-ROC (%)", y = NULL
    ) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
```

Každý riadok = jedna metóda. Dĺžka stĺpca = AUC-ROC, číslo v texte hovorí koľko features použila. Prerušovaná čiara = baseline (všetky features). LASSO a RFE sú tesne pod čiarou s 9 features — prakticky rovnaký výkon pri polovici features.

## S3 findings

All five selection methods converge on the same answer: **the predictive signal in this dataset is highly concentrated**. Starting from 21 features, every method reduces the set substantially — LASSO and RF Importance down to just 9, RFE to 10 — with almost no loss in AUC-ROC (83.2–84.4% vs. 84.4% for the full-feature baseline). This means roughly half the features in the model are redundant: they carry signal already captured by other features, or no signal at all.

**Four features form the consensus core** — selected by all five methods regardless of the criterion used: `golddiffat15`, `xpdiffat15`, `csdiffat15`, and `winrate_diff`. These are the three resource differentials at 15 minutes plus the pre-game team strength gap. Their universal selection confirms what the EDA correlation heatmap showed: who is ahead in gold, XP, and CS at 15 minutes is the strongest combined signal for match outcome. The `winrate_diff` feature indicates that pre-game team quality adds real information on top of the in-game state.

**Four more features reach near-consensus** (4 out of 5 methods): `firsttower`, `firstherald`, `firstdragon`, and `cs_momentum`. Objective control — particularly tower and herald — contributes a second, independent layer of signal beyond raw economic leads. `cs_momentum` (the acceleration of the CS advantage from @10 to @15) captures information not fully redundant with the snapshot differential at @15.

**Three features were never selected** by any method: `firstblood`, `kill_pressure_10_15`, and `opp_kill_pressure_10_15`. This is a meaningful null result — it suggests that early kill activity, while visible in the data, does not independently predict match outcome once the economic and objective features are accounted for. First blood in particular is consistent with published LoL research, where its win-rate correlation is weak relative to tower or gold leads.

The key practical takeaway is that **9 features achieve 84.3% AUC-ROC** — within 0.1 points of the full-feature LR — which means a model trained on only the consensus + near-consensus features is as useful as one trained on everything, while being substantially easier to interpret and explain.

# Conclusions

## H1 findings

<!-- TODO: summarize early-game prediction results -->

## H2 findings

<!-- TODO: summarize bot lane impact findings -->

## S1 vs S3 synthesis

<!-- TODO: which models + features work best together -->

# Export models for Shiny app

```{r save-models}
dir.create("shiny/models", recursive = TRUE, showWarnings = FALSE)

saveRDS(lr_model,      "shiny/models/lr_model.rds")
saveRDS(rf_model,      "shiny/models/rf_model.rds")
saveRDS(nb_model,      "shiny/models/nb_model.rds")
saveRDS(knn_model,     "shiny/models/knn_model.rds")
saveRDS(cart_model,    "shiny/models/cart_model.rds")
saveRDS(preproc,       "shiny/models/preproc.rds")
saveRDS(perf_summary,  "shiny/models/perf_summary.rds")
saveRDS(overlap_df,    "shiny/models/overlap_df.rds")
saveRDS(teams_model,   "shiny/models/teams_model.rds")
saveRDS(X_train,       "shiny/models/X_train.rds")

# Game lookup table for Shiny browser
opp_lookup <- teams %>%
    select(gameid, side, opp_teamname = teamname) %>%
    mutate(opp_side = ifelse(side == "Blue", "Red", "Blue")) %>%
    select(gameid, opp_side, opp_teamname)

game_lookup <- teams %>%
    select(gameid, league, date, teamname, side_label = side) %>%
    left_join(opp_lookup, by = c("gameid", "side_label" = "opp_side")) %>%
    mutate(side_encoded = ifelse(side_label == "Blue", 1, 0)) %>%
    inner_join(
        teams_model %>% rename(side_encoded = side),
        by = c("gameid", "side_encoded")
    ) %>%
    mutate(
        date         = as.Date(date),
        result_label = ifelse(result == 1, "WIN", "LOSS")
    )

saveRDS(game_lookup, "shiny/models/game_lookup.rds")

cat("Models saved to shiny/models/\n")
```
